# Gradient Boosting in Practice

Imagine you are trying to predict house prices and your first simple model is wrong in a patterned way: it overshoots cheap houses and undershoots expensive ones. A natural fix is to fit a second model to those leftover errors, add it to the first, look at what is still wrong, fit a third model to that, and keep going. Each new model is a small correction aimed at the part of the problem the running total has not yet captured. That, in one sentence, is gradient boosting: build a strong predictor by stacking many weak ones, where each weak learner is trained on the mistakes of the ensemble so far.

This idea turns out to be one of the most reliable workhorses in all of machine learning. On the kind of tabular data that fills spreadsheets and databases (rows of mixed numeric and categorical columns), gradient-boosted trees routinely win against neural networks and almost everything else. They are the default first thing many practitioners reach for.

The [Boosting] chapter introduced the foundations: forward stagewise additive modeling, AdaBoost, and the basic gradient boosting machine. This chapter picks up where that left off and focuses on the modern gradient-boosted-tree systems that dominate tabular prediction in practice: XGBoost, LightGBM, and CatBoost. By the end you will be able to read these systems in two complementary ways. First, as an optimization story: each one performs functional gradient descent against a regularized objective, and the formulas they use for leaf values and splits fall directly out of that view. Second, as a set of practical knobs you actually turn, with runnable training and validation loss curves that show overfitting happening and early stopping catching it.

> **Note:** This chapter assumes you have seen the [Boosting] chapter. If the phrase "fit a tree to the residuals" is not yet familiar, read that chapter first; here we move quickly from intuition to the regularized objective that the production systems optimize.

## Boosting as Functional Gradient Descent

The opening picture (fit each new model to the leftover errors) is intuitive, but to design and tune real systems we need a precise version of it. The precise version reframes boosting as gradient descent, not over a few numbers in a parameter vector, but over the predictions themselves. This section builds that reframing one step at a time.

A boosted model is an additive expansion of base learners,

$$
F_M(x) = \sum_{m=1}^M \nu \, f_m(x),
$$

where each $f_m$ is a small regression tree and $\nu \in (0,1]$ is the learning rate. We build this expansion one term at a time. Given the current model $F_{m-1}$, we add a new base learner $f_m$ chosen to reduce the total loss

$$
\mathcal{L}(F) = \sum_{i=1}^n l\big(y_i, F(x_i)\big).
$$

This is forward stagewise additive modeling: previously added terms stay fixed, and we only solve for the next one.^[Refitting all the earlier terms at every step would be the "fully corrective" version. Keeping them frozen is cheaper and, with shrinkage, works better in practice because it limits how aggressively the model adapts to the training data.]

> **Key idea:** Treat the predictions $F(x_1), \dots, F(x_n)$ as the quantities we are optimizing, one per training point. The loss $\mathcal{L}$ is then an ordinary function of those $n$ numbers, and we can do gradient descent on it just as we would on any other parameters.

Steepest descent on $\mathcal{L}$ in this function space moves each prediction in the direction of the negative gradient. Define the pseudo-residual^[Called a "pseudo"-residual because for a general loss it is not literally the observed-minus-predicted difference; it is the negative gradient of the loss, which only coincides with the ordinary residual for squared error, as we show next.] at observation $i$ and step $m$ as

$$
r_{im} = - \left[ \frac{\partial l(y_i, F(x_i))}{\partial F(x_i)} \right]_{F = F_{m-1}}.
$$

We then fit the base learner $f_m$ to the pseudo-residuals $\{(x_i, r_{im})\}$ by least squares. Here is why a tree is the right tool: pure gradient descent would tell us how to nudge each of the $n$ predictions individually, but that gives us no rule for predicting at a new $x$ we have never seen. Fitting a tree to $r_{im}$ forces the descent direction to be a function of $x$, so it generalizes. The new tree points the whole model downhill in loss while still being something we can evaluate anywhere. After fitting, the leaf values can be re-optimized for the actual loss (a line search per leaf), and the tree is added with shrinkage $\nu$.

> **Intuition:** Gradient descent says "move in this direction." A tree is how we make "this direction" into a reusable function of the features instead of a one-off adjustment to the training points.

### Squared error gives ordinary residuals

For squared error loss $l(y, F) = \tfrac12 (y - F)^2$, the gradient with respect to $F$ is

$$
\frac{\partial l}{\partial F} = -(y - F), \qquad r_{im} = y_i - F_{m-1}(x_i).
$$

The pseudo-residual is exactly the ordinary residual, which recovers the "fit a tree to the residuals" algorithm from the [Boosting] chapter. The optimal leaf value for a terminal region is the mean of the residuals in that region. So the intuitive picture from the start of the chapter is not an analogy; for squared error it is literally what the gradient view prescribes.

### Logistic loss gives probability residuals

The same machinery handles classification once we pick the right loss. The only change is the formula for the gradient; everything else (fit a tree to it, re-optimize leaves, shrink, repeat) is identical.

For binary classification with $y_i \in \{0,1\}$, let $F(x)$ be the log-odds (a real number) and $p = \sigma(F) = 1/(1 + e^{-F})$.^[Working in log-odds space keeps $F$ unconstrained on the whole real line, which is what lets us add trees freely. The sigmoid $\sigma$ maps the running total back to a probability in $(0,1)$ only when we want to read off $p$.] The logistic (binomial deviance) loss is

$$
l(y, F) = -\big[\, y \log p + (1-y) \log(1-p) \,\big] = \log\!\big(1 + e^{F}\big) - y F.
$$

Differentiating with respect to $F$ and using $\sigma'(F) = \sigma(F)(1-\sigma(F))$,

$$
\frac{\partial l}{\partial F} = \sigma(F) - y = p - y, \qquad r_{im} = y_i - p_{i,m-1}.
$$

The pseudo-residual is the difference between the observed label and the current predicted probability, a satisfying mirror of the regression case where it was observed minus predicted value. The second derivative (Hessian), which the next section uses to take a smarter step than plain gradient descent, is

$$
\frac{\partial^2 l}{\partial F^2} = p (1 - p).
$$

> **Note:** The Hessian $p(1-p)$ is largest when $p \approx 0.5$ (the model is most uncertain) and shrinks toward zero as predictions become confident. XGBoost uses exactly this curvature to decide how big a step each leaf should take, which is the subject of the next section.

## The Tuning Knobs and Their Roles

Before the production-system formulas, it helps to know what you will actually be turning. Every gradient-boosting library exposes essentially the same handful of parameters, and they all do one job: control where the model sits on the bias-variance spectrum. Too little flexibility and the model underfits; too much and it memorizes the training set. The knobs below are the levers for that trade-off.

- Learning rate (shrinkage) $\nu$: scales each tree's contribution. Smaller $\nu$ slows learning, which usually improves generalization but requires more trees. Think of it as taking many small, careful steps instead of a few large, greedy ones. $\nu$ and the number of trees trade off against each other directly.
- Number of trees (rounds) $M$: more rounds keep reducing training loss, but validation loss eventually rises. This is the main overfitting control, and it is what early stopping monitors.
- Tree depth / interaction order: a tree of depth $d$ can model interactions among up to $d$ features. Depth 1 (stumps) gives a purely additive model; deeper trees capture higher-order interactions at the cost of variance.
- Subsampling rows (stochastic gradient boosting): fitting each tree on a random fraction of the rows decorrelates the trees and acts as a regularizer, often improving accuracy and speed.
- Subsampling columns: sampling a fraction of features per tree or per split further decorrelates trees, in the same spirit as random forests.

> **Intuition:** The learning rate and the number of trees are two ends of the same lever. Halve the learning rate and you will roughly need to double the trees to fit the same amount. This is why we almost never tune them independently: we fix a small learning rate and let early stopping pick the count.

The pattern that holds across most tabular problems is worth memorizing: a small learning rate, moderate depth (often 3 to 8), some row and column subsampling, and the number of trees chosen by early stopping on a validation set. The rest of this chapter explains why these defaults work and shows them in code.

## XGBoost: a Regularized, Second-Order Objective

So far we have used only the first derivative of the loss (the gradient). XGBoost (Chen and Guestrin 2016) makes two changes that account for much of its success. First, it writes down an explicit penalty for tree complexity and optimizes loss plus penalty together, rather than relying only on tree depth to control overfitting. Second, it uses the second derivative (the Hessian) as well as the first, which lets it take a Newton step instead of a plain gradient step and gives a clean, closed-form rule for both leaf values and split decisions. The payoff is that the algorithm's core formulas are not heuristics; they are the exact solution to a well-defined optimization problem.

XGBoost optimizes a penalized objective that adds a complexity term for each tree $f_k$,

$$
\mathcal{L} = \sum_{i=1}^n l(y_i, \hat y_i) + \sum_k \Omega(f_k), \qquad \Omega(f) = \gamma T + \tfrac12 \lambda \lVert w \rVert^2,
$$

where $T$ is the number of leaves, $w \in \mathbb{R}^T$ is the vector of leaf weights, $\gamma$ penalizes adding leaves, and $\lambda$ is an L2 penalty on the leaf weights. Read $\Omega$ as a price list: every leaf you add costs $\gamma$, and large leaf values are taxed by $\lambda$. Splitting only pays off when it buys enough loss reduction to cover that price.

### Newton step on the objective

We now derive the rules XGBoost uses. The strategy is to approximate the loss around the current prediction by a quadratic, because a quadratic is easy to minimize exactly. This is the same move Newton's method makes in ordinary optimization, applied here per training point.

At boosting round $t$ we add a tree $f_t$, so the prediction becomes $\hat y_i^{(t)} = \hat y_i^{(t-1)} + f_t(x_i)$. Take a second-order Taylor expansion of the loss around the current prediction, using the gradient and Hessian

$$
g_i = \frac{\partial l(y_i, \hat y_i^{(t-1)})}{\partial \hat y_i^{(t-1)}}, \qquad h_i = \frac{\partial^2 l(y_i, \hat y_i^{(t-1)})}{\partial (\hat y_i^{(t-1)})^2}.
$$

Dropping the constant term, the objective for round $t$ is approximately

$$
\tilde{\mathcal{L}}^{(t)} = \sum_{i=1}^n \Big[ g_i f_t(x_i) + \tfrac12 h_i f_t(x_i)^2 \Big] + \gamma T + \tfrac12 \lambda \sum_{j=1}^T w_j^2.
$$

A tree assigns every $x_i$ to exactly one leaf. Let $I_j = \{ i : x_i \text{ falls in leaf } j \}$ be the set of observations in leaf $j$, where $f_t(x_i) = w_j$. Group the sum by leaf,

$$
\tilde{\mathcal{L}}^{(t)} = \sum_{j=1}^T \Big[ \big(\textstyle\sum_{i \in I_j} g_i\big) w_j + \tfrac12 \big(\textstyle\sum_{i \in I_j} h_i + \lambda\big) w_j^2 \Big] + \gamma T.
$$

Grouping by leaf is the step that makes everything tractable: because every point in a leaf gets the same weight $w_j$, the whole objective splits into one independent little problem per leaf. We solve those next.

### Optimal leaf weight and the structure score

For a fixed tree structure, each leaf weight is decoupled and the objective is quadratic in $w_j$. Let $G_j = \sum_{i \in I_j} g_i$ and $H_j = \sum_{i \in I_j} h_i$. Setting the derivative to zero,

$$
\frac{\partial}{\partial w_j}\Big[ G_j w_j + \tfrac12 (H_j + \lambda) w_j^2 \Big] = G_j + (H_j + \lambda) w_j = 0,
$$

gives the optimal weight

$$
w_j^* = - \frac{G_j}{H_j + \lambda} = - \frac{\sum_{i \in I_j} g_i}{\sum_{i \in I_j} h_i + \lambda}.
$$

Substituting back, the best achievable objective for that structure (the structure score) is

$$
\tilde{\mathcal{L}}^{(t)} = -\frac12 \sum_{j=1}^T \frac{G_j^2}{H_j + \lambda} + \gamma T.
$$

Lower is better, so $\sum_j G_j^2 / (H_j + \lambda)$ measures how good a tree structure is.

> **Key idea:** The structure score turns "how good is this tree shape?" into a single number we can compute without ever fitting the leaf values by trial and error. Every decision XGBoost makes about growing the tree reduces to comparing this score before and after a candidate change.

### Split gain

The structure score evaluates a finished tree, but we build trees greedily, one split at a time. So we need the change in the score from a single split, which is the quantity the algorithm maximizes when choosing where to cut.

We grow the tree greedily. Consider splitting one leaf into a left child $L$ and a right child $R$. The reduction in the objective from making the split is the gain

$$
\text{Gain} = \frac12 \left[ \frac{G_L^2}{H_L + \lambda} + \frac{G_R^2}{H_R + \lambda} - \frac{(G_L + G_R)^2}{H_L + H_R + \lambda} \right] - \gamma.
$$

The first two terms are the structure scores of the children, the third is the score of the parent before splitting, and $\gamma$ is the cost of adding a leaf. A split is only kept when its gain is positive, which is a built-in form of pruning: the $\gamma$ term and the $\lambda$ shrinkage in the denominators both discourage splits that rely on few observations.^[This is why $\gamma$ is sometimes called `min_split_loss`: it is the minimum gain a split must earn to be worth keeping. Notice the family resemblance between split gain here and the impurity-reduction criterion of an ordinary CART tree; XGBoost has simply replaced impurity with a loss-and-penalty-aware score.]

It is worth seeing that this general machinery collapses to the familiar cases. For squared error, $g_i = \hat y_i - y_i$ and $h_i = 1$, so $w_j^* = -G_j / (|I_j| + \lambda)$, a shrunken mean of the residuals: the residual-averaging rule from before, now with a regularization term in the denominator. For logistic loss, $g_i = p_i - y_i$ and $h_i = p_i(1-p_i)$, which is exactly the gradient and Hessian we derived in the classification subsection. The same two-line objective handles regression and classification by plugging in a different $g$ and $h$.

## LightGBM, CatBoost, and the Speed Tricks

The objective and split formulas above describe what XGBoost computes. The remaining differences between the major libraries are mostly about how fast they compute it on large data, and how gracefully they handle categorical features. Understanding these tricks helps you choose a library and read its parameter list, even though the underlying boosting story is the same.

XGBoost's original exact greedy split search scans every candidate split point on every feature, which is accurate but slow when there are many rows. LightGBM (Ke et al. 2017) trades a little precision for large speedups using four ideas:

- Histogram-based splits: continuous features are bucketed into a fixed number of bins (for example 255). Split candidates are evaluated on the histogram of binned values rather than on raw sorted values, so split finding cost scales with the number of bins, not the number of distinct values. The per-row gradient and Hessian sums are accumulated per bin.^[XGBoost later added its own histogram method (`tree_method = "hist"`), so this is no longer unique to LightGBM. The other three tricks below are more distinctive.]
- Leaf-wise (best-first) growth: instead of growing all nodes at a given depth (level-wise), LightGBM repeatedly splits the leaf with the largest gain. This reaches lower loss for a given number of leaves but can overfit, so the number of leaves and a maximum depth are the main controls.
- Gradient-based One-Side Sampling (GOSS): observations with large gradients are far from being fit well, so GOSS keeps all large-gradient rows and randomly subsamples the small-gradient rows, reweighting the kept small-gradient rows to keep the gradient sums roughly unbiased. This shrinks the training set while preserving the important signal.
- Exclusive Feature Bundling (EFB): in sparse data many features are rarely nonzero at the same time. EFB bundles such mutually exclusive features into a single feature, reducing the effective feature count with little information loss.

> **Warning:** LightGBM's leaf-wise growth is powerful but easy to overfit with. A tree controlled only by `num_leaves` can grow much deeper along one branch than a level-wise tree of the same leaf count. If you see training loss far below validation loss, cap `max_depth` or lower `num_leaves` before touching anything else.

CatBoost rounds out the trio with a different concern: bias. Ordinary gradient boosting uses the same rows to both fit the model and compute the residuals it will fit next, which subtly leaks information and biases the predictions, an effect its authors call prediction shift. CatBoost addresses this with ordered boosting, which computes the residual for each observation using a model trained only on earlier observations in a random permutation. It also has strong native handling of categorical features through ordered target statistics, so categorical columns can be passed in directly rather than one-hot encoded by hand.

> **When to use this:** Reach for LightGBM when you have many rows and want speed; reach for CatBoost when categorical features dominate and you want sensible defaults with minimal preprocessing; reach for XGBoost as a mature, widely supported baseline. In practice all three are competitive, and the choice often comes down to ecosystem and familiarity.

## Simulation: Loss Curves, Learning Rates, and Importance

Theory predicts that training loss falls forever while validation loss eventually turns around; this section makes that prediction visible. We use the `Boston` housing data from `ISLR2` as a regression example, where the goal is to predict the median home value `medv` from neighborhood characteristics. We hold out a validation set, train XGBoost while recording training and validation RMSE at every round, and plot the curves to see overfitting happen and to locate where early stopping would cut off. We then compare learning rates and read off feature importance.

The setup splits the data 70/30 and wraps each part in an `xgb.DMatrix`, XGBoost's internal data structure:

```{r gb-setup, message=FALSE, warning=FALSE}
library(xgboost)
library(ISLR2)
library(ggplot2)

set.seed(1)
data(Boston, package = "ISLR2")

n <- nrow(Boston)
train_idx <- sample(seq_len(n), size = floor(0.7 * n))
x_all <- as.matrix(Boston[, setdiff(names(Boston), "medv")])
y_all <- Boston$medv

dtrain <- xgb.DMatrix(data = x_all[train_idx, ], label = y_all[train_idx])
dvalid <- xgb.DMatrix(data = x_all[-train_idx, ], label = y_all[-train_idx])
```

We train a deliberately under-regularized model (large learning rate, moderate depth, many rounds, no subsampling) so the validation curve turns upward and the overfitting is plain to see. The `watchlist` tells XGBoost to evaluate RMSE on both sets after every round and store the history in `evaluation_log`.

```{r gb-overfit, message=FALSE, warning=FALSE}
watchlist <- list(train = dtrain, valid = dvalid)

set.seed(1)
fit_overfit <- xgb.train(
  params = list(
    objective = "reg:squarederror",
    eta = 0.3,
    max_depth = 6
  ),
  data = dtrain,
  nrounds = 500,
  watchlist = watchlist,
  verbose = 0
)

log_df <- as.data.frame(fit_overfit$evaluation_log)
head(log_df)
```

```{r gradient-boosting-losscurve, message=FALSE, warning=FALSE, fig.cap="Training and validation RMSE versus number of boosting rounds. Training error keeps falling while validation error flattens then rises, the signature of overfitting. The dashed line marks the round with minimum validation RMSE (the early-stopping point)."}
best_round <- which.min(log_df$valid_rmse)

ggplot(log_df, aes(x = iter)) +
  geom_line(aes(y = train_rmse, color = "Training")) +
  geom_line(aes(y = valid_rmse, color = "Validation")) +
  geom_vline(xintercept = best_round, linetype = "dashed") +
  labs(x = "Boosting round", y = "RMSE", color = NULL,
       title = "XGBoost training vs validation RMSE") +
  theme_minimal()

best_round
min(log_df$valid_rmse)
```

Figure \@ref(fig:gradient-boosting-losscurve) is the whole lesson in one picture. The training curve marches steadily downward, because adding trees can always fit the training data a little better. The validation curve drops at first, because early trees capture real structure, but then bottoms out and creeps back up, because later trees start fitting noise that does not generalize. The dashed line marks the round where validation RMSE is lowest; everything to the right of it is overfitting we would rather avoid.

> **Tip:** Reading a loss curve is the single most useful diagnostic in boosting. A validation curve that is still falling means you stopped too early (train longer or lower the learning rate). One that turned up long ago means you trained too long. A flat curve far above the training curve means the model is too flexible (reduce depth or add subsampling).

Eyeballing the dashed line is fine for a plot, but we do not want to retrain hundreds of rounds just to throw most away. Early stopping automates picking that round: training continues until the validation metric fails to improve for a fixed number of rounds (`early_stopping_rounds`), then stops and keeps the best iteration. Here we also switch to sensible regularization (smaller learning rate, shallower trees, row and column subsampling) so the model generalizes better.

```{r gb-earlystop, message=FALSE, warning=FALSE}
set.seed(1)
fit_es <- xgb.train(
  params = list(
    objective = "reg:squarederror",
    eta = 0.1,
    max_depth = 4,
    subsample = 0.8,
    colsample_bytree = 0.8
  ),
  data = dtrain,
  nrounds = 2000,
  watchlist = watchlist,
  early_stopping_rounds = 25,
  verbose = 0
)

fit_es$best_iteration
fit_es$best_score
```

The reported `best_iteration` is the round XGBoost will use for prediction, and `best_score` is the validation RMSE there. Note that this regularized model stops after far fewer effective rounds and reaches a similar or better error than the over-trained one above, with much less risk of overfitting.

### Comparing learning rates

To see the learning-rate-versus-rounds trade-off concretely, we sweep a few learning rates, each with its own early stopping, and record how many rounds each one needs and the best validation RMSE it reaches. The expectation from theory is that smaller learning rates need more rounds but tend to reach a comparable or better minimum.

```{r gb-lr-compare, message=FALSE, warning=FALSE}
etas <- c(0.01, 0.05, 0.1, 0.3)

lr_results <- do.call(rbind, lapply(etas, function(e) {
  set.seed(1)
  fit <- xgb.train(
    params = list(
      objective = "reg:squarederror",
      eta = e,
      max_depth = 4,
      subsample = 0.8,
      colsample_bytree = 0.8
    ),
    data = dtrain,
    nrounds = 3000,
    watchlist = watchlist,
    early_stopping_rounds = 30,
    verbose = 0
  )
  data.frame(
    learning_rate = e,
    best_round = fit$best_iteration,
    valid_rmse = round(fit$best_score, 3)
  )
}))

lr_results
```

The `best_round` column confirms the pattern: as the learning rate shrinks, the number of rounds needed climbs sharply, while the validation RMSE stays in a narrow band. This is exactly why the standard recipe is to fix a small learning rate and let early stopping spend whatever rounds are required, rather than tuning the round count by hand.

### Feature importance

A fitted booster also tells us which features mattered. XGBoost reports three importance measures: gain (the average split-gain contribution of a feature, summed over the splits that use it), cover (the average number of observations affected by its splits), and frequency (how often it is used). Gain is the usual ranking because it reflects how much each feature actually reduced the loss.

```{r gradient-boosting-importance, message=FALSE, warning=FALSE}
imp <- xgb.importance(model = fit_es)
knitr::kable(
  imp,
  digits = 4,
  caption = "XGBoost feature importance (gain, cover, frequency) for the Boston model."
)
```

The ranking in Table \@ref(tab:gradient-boosting-importance) is intuitive for housing data: `rm` (average number of rooms) and `lstat` (percent lower-status population) dominate, which matches both economic intuition and the linear-model coefficients you would get on the same data.

> **Warning:** Importance scores describe what the model used, not causal effect, and they can be unstable when features are correlated (the model may "credit" one of two redundant features almost arbitrarily). Treat them as a rough guide for the model at hand, not as a measure of how the world works. For more careful attribution, the [Interpretable Machine Learning] chapter covers SHAP values and partial dependence.

### LightGBM on the same data

To show that the workflow carries over, we fit LightGBM on the same train/validation split. The API differs in surface details (`lgb.Dataset` instead of `xgb.DMatrix`, `feature_fraction`/`bagging_fraction` instead of `colsample_bytree`/`subsample`), but the logic (early stopping on a validation set, then read importance) is identical. Its key controls are `num_leaves`, which governs the leaf-wise growth discussed above, and `learning_rate`.

```{r gradient-boosting-lgb-importance, message=FALSE, warning=FALSE}
library(lightgbm)

ltrain <- lgb.Dataset(data = x_all[train_idx, ], label = y_all[train_idx])
lvalid <- lgb.Dataset.create.valid(ltrain, data = x_all[-train_idx, ],
                                   label = y_all[-train_idx])

set.seed(1)
fit_lgb <- lgb.train(
  params = list(
    objective = "regression",
    metric = "rmse",
    learning_rate = 0.05,
    num_leaves = 31,
    feature_fraction = 0.8,
    bagging_fraction = 0.8,
    bagging_freq = 1,
    verbose = -1
  ),
  data = ltrain,
  nrounds = 2000,
  valids = list(valid = lvalid),
  early_stopping_rounds = 30,
  verbose = -1
)

fit_lgb$best_iter
fit_lgb$best_score

lgb_imp <- lgb.importance(fit_lgb)
knitr::kable(
  lgb_imp,
  digits = 4,
  caption = "LightGBM feature importance for the Boston model."
)
```

Reassuringly, the LightGBM ranking in Table \@ref(tab:gradient-boosting-lgb-importance) lands on the same top features (`rm` and `lstat`) and a similar validation error, despite its different growth strategy and split search. When two well-tuned boosting libraries agree on both the ranking and the error, that is a good sign the signal is real and not an artifact of one implementation.

## Practical Notes

Everything above comes together into a short, repeatable workflow. The following order is what works in practice on most tabular problems:

1. Fix a small learning rate (0.05 or 0.1), set the number of rounds high, and let early stopping choose the actual count.
2. Tune tree shape next: `max_depth` or `num_leaves`, and the minimum child weight or minimum data per leaf.
3. Tune the subsampling fractions: `subsample`/`bagging_fraction` for rows and `colsample_bytree`/`feature_fraction` for columns.
4. Add L1/L2 penalties (`alpha`, `lambda`) and the split-cost `gamma` last, only if the model still overfits.
5. Optionally, after the shape is settled, drop the learning rate further and retrain with early stopping for a small accuracy gain at the cost of more rounds.

> **Warning:** Early stopping needs a validation set (or cross-validation via `xgb.cv` and `lgb.cv`) and a metric to monitor, and that set has been used for model selection. Always report your final performance on a separate test set the model never saw during stopping, or you will report an optimistic number.

A recurring practical question is how each library handles categorical features and missing values, since this is where the most preprocessing time is spent. The three systems differ:

- XGBoost takes a numeric matrix, so categoricals must be one-hot or target encoded beforehand. It handles missing values natively: at each split a default direction is learned, and missing values go that way, so you can pass `NA` without imputation.
- LightGBM handles missing values natively and can treat integer-coded columns as categorical when told which columns they are (`categorical_feature`), using an optimized partition of categories rather than one-hot expansion.
- CatBoost is built around categorical features and applies ordered target statistics internally, which avoids manual encoding and reduces leakage.

> **Tip:** The native missing-value handling in these libraries is a real convenience, but it makes "missing" itself a signal the model can use. If your data is missing for reasons related to the target (informative missingness), that can help; if it is missing by accident, double-check that the learned default directions are not picking up a quirk of your collection process.

To summarize the chapter: gradient boosting is functional gradient descent that builds an additive model of trees, each fit to the negative gradient of the loss. XGBoost sharpens this with a regularized, second-order objective whose leaf weights and split gains have closed forms. LightGBM and CatBoost keep that core and add speed and categorical-handling tricks. In practice you fix a small learning rate, let early stopping pick the number of trees, and tune tree shape and subsampling around that, watching the validation loss curve the entire time.

## Further Reading

- Friedman, J. H. (2001). Greedy function approximation: a gradient boosting machine. Annals of Statistics. The functional gradient descent view used throughout this chapter.
- Chen, T. and Guestrin, C. (2016). XGBoost: A scalable tree boosting system. KDD. The regularized objective, second-order split gain, and systems design.
- Ke, G. et al. (2017). LightGBM: A highly efficient gradient boosting decision tree. NeurIPS. Histogram splits, GOSS, and EFB.
