# Bayesian Additive Regression Trees {#sec-bart}

```{r}
#| include: false
source("_common.R")
```

Imagine you want to predict the price of a house from its features (number of rooms, neighborhood income, distance to work, and so on). A single regression tree (@sec-regression-trees) is easy to read but usually too crude. A forest of many trees (@sec-random-forest) predicts well but gives you a single point estimate with no honest sense of how uncertain that estimate is. Bayesian Additive Regression Trees, or BART [@sparapani2021], try to give you the best of both: the flexibility of a sum of many small trees, plus a full probability distribution over the prediction so you can attach uncertainty to every number you report.

This chapter introduces BART from intuition first, then states the algorithm precisely, and finally walks through worked examples in R using the `BART` package. By the end you should understand why BART is often called a "good out-of-the-box" method, how its three main settings (number of trees, number of iterations, and burn-in) work, and how to read its output, including convergence checks, variable importance, and partial dependence plots.


::: {.callout-tip title="When to use this"}
BART is a strong default when you want flexible, nonlinear regression or classification with little tuning, and you also want uncertainty intervals around predictions. It pairs naturally with the tree-based and ensemble ideas you have already seen.
:::


BART sits in the same family as the methods in the chapters on bagging (@sec-bagging), random forests (@sec-random-forest), and boosting (@sec-boosting), and it borrows one key idea from each. As in bagging and random forests, every tree is built using randomness rather than a single deterministic greedy search. As in boosting, every tree tries to capture the part of the signal that the rest of the model has not yet explained.^[The difference is that boosting fits each new tree to the residuals once, in a forward, greedy pass, while BART repeatedly *revisits and revises* every tree using a Markov chain Monte Carlo sampler. This revisiting is what produces a posterior distribution instead of a single fitted model.]

## The model: a sum of trees

The core idea is simple to state. Instead of one big tree, BART uses a *sum* of many small trees, and each tree contributes a small piece of the final prediction.

Let $K$ be the number of regression trees, and let $B$ be the number of MCMC iterations for which the BART algorithm will be run. Let $\hat{f}_k^b(x)$ represent the prediction at $x$ for the $k$-th regression tree used in the $b$-th iteration. The prediction from a single iteration is the sum of all $K$ trees in that iteration:

$$
\hat{f}^b(x) = \sum_{k=1}^K \hat{f}_k^b(x), b = 1, \dots, B
$$


::: {.callout-important title="Key idea"}
No single tree has to be good on its own. Each tree is deliberately kept small (a "weak learner"), and the prediction comes from adding many of them together. Keeping each tree small is the main guard against overfitting.
:::


### Starting point

We need a place to begin the chain. In the first iteration, all trees are initialized to have a single root node with $\hat{f}_k^1(x) = (1/nK)\sum_{i=1}^n y_i$, which is the mean of the response values divided by the total number of trees. Splitting the overall mean equally across the $K$ trees means that when we add the trees back up, we recover exactly the mean response:

$$
\hat{f}^1(x) = \sum_{k=1}^K \hat{f}_k^1 (x) = \frac{1}{n} \sum_{i=1}^n y_i \text{ (mean response)}
$$


::: {.callout-tip title="Intuition"}
BART starts from the simplest possible prediction, "guess the average for everyone," and then lets the trees gradually carve out structure from there.
:::


### Updating one tree at a time using partial residuals

In each later iteration, BART updates each of the $K$ trees, one at a time. The trick that makes this work is the *partial residual*: when we are about to update tree $k$, we temporarily remove its contribution and ask how far off the *rest* of the model is. Whatever the other trees fail to explain is exactly what tree $k$ should try to capture.

Concretely, in the $b$-th iteration, to update the $k$-th tree, we subtract from each response value the predictions from all but the $k$-th tree to obtain a partial residual:

$$
r_i = y_i - \sum_{k'<k} \hat{f}_{k'}^b (x_i) - \sum_{k'>k} f_{k'}^{b-1}(x_i), i = 1, \dots, n
$$

Notice the bookkeeping in the two sums: trees with index $k' < k$ have already been updated in the current iteration $b$, so we use their fresh values, while trees with $k' > k$ have not been touched yet this round, so we use their values from the previous iteration $b-1$.

Here is the step that makes BART Bayesian rather than just another boosting scheme. Instead of fitting a brand new tree to this partial residual, BART randomly chooses a perturbation to the tree from the previous iteration ($\hat{f}_k^{b-1}$) out of a set of possible perturbations (this is where MCMC comes in), favoring ones that improve the fit to the partial residual.^[Mechanically, this is a Metropolis-Hastings step: a candidate change to the tree is proposed at random, and it is accepted or rejected with a probability that depends on how much it improves the fit and on the prior. Over many iterations the trees are effectively samples from a posterior distribution.] There are two components to this perturbation, which we apply together:

-   change the structure of the tree by adding or pruning branches, and
-   change the prediction value stored in each terminal (leaf) node of the tree.


::: {.callout-note}
Because we *perturb* the existing tree rather than regrowing it from scratch, no single update can move the model very far. This deliberate timidity is what keeps BART from chasing noise.
:::


### Collecting the output

After running the chain, the output of BART is not one model but a whole collection of prediction models, one per iteration:

$$
\hat{f}^b(x) = \sum_{k =1}^K \hat{f}_k^b (x) , b = 1, \dots, B
$$

Early iterations are unreliable because the chain has not yet settled down, so we discard the first $L$ iterations as a burn-in period and average over the rest:

$$
\hat{f}^b(x) = \frac{1}{B - L} \sum_{b = L +1}^B \hat{f}_k^b (x)
$$

The payoff of the Bayesian framing appears here: because each retained iteration is a posterior sample, we are not limited to the average. We can use the spread of the posterior samples to get distributional estimates of the predictions, that is, credible intervals rather than just point predictions.


::: {.callout-tip}
When someone asks "how confident are you in this prediction?", BART has a direct answer: look at the variability of $\hat{f}^b(x)$ across the retained iterations.
:::


## The BART algorithm

Putting the pieces together, the full procedure is summarized below, following [@james2013] Alg 8.3. The outer loop walks through MCMC iterations, the middle loop walks through the $K$ trees, and the inner loop computes the partial residual for each observation before perturbing the current tree.

1.  Let $\hat{f}_1^1(x) = f^1_2(x) = \dots = \hat{f}_K^1(x) = \frac{1}{nK} \sum_{i=1}^n y_i$

2.  Compute $\hat{f}^1 (x) = \sum_{k=1}^K \hat{f}_k^1 (x) = \frac{1}{n}\sum_{i=1}^n y_i$

3.  For $b = 2, \dots, B$

    1.  For $k = 1, \dots, K$

        1.  For $i = 1, \dots , n$, compute the current partial residual ($r_i$)

        2.  Fit a new tree $\hat{f}_k^b(x)$ to $r_i$ by randomly perturbing the $k$-th tree from the previous iteration, $\hat{f}_k^{b-1}(x)$ (favor perturbations that improve the fit)

    2.  Compute $\hat{f}^b (x) = \sum_{k=1}^K \hat{f}^b_k (x)$

4.  Compute the mean after $L$ burn-in samples $\hat{f}(x) = \frac{1}{B - L} \sum_{b = L +1}^B \hat{f}^b (x)$

A few remarks make the algorithm easier to reason about. Step 3.1.2 prevents overfitting because it limits how hard we fit the data in each iteration: a single perturbation can only nudge a tree a little. The individual trees are kept quite small on purpose; limiting tree size is another guard against overfitting, since very large trees would memorize the training data. And, as noted above, each time we randomly perturb a tree to fit the residuals we are effectively drawing a new tree from a posterior distribution.

When we use BART, we have to specify three quantities:

-   the number of trees $K$,
-   the number of iterations $B$, and
-   the number of burn-in iterations $L$.

In practice these are easy to set. We typically choose large values such as $B = 1000$ and $K = 200$, and a moderate value such as $L = 100$. Because these defaults work well across many problems, BART is a good out-of-the-box tool that needs minimal tuning.


::: {.callout-warning}
The defaults are robust, but they are not free. With $K = 200$ trees and $B = 1000$ iterations, BART does a lot of computation, and runtime grows with the number of observations and predictors. If a run is slow, the thinning trick shown later in this chapter can help.
:::


## Worked example: predicting Boston housing prices {#bart-boston-james}

We start with the example from [@james2013, Ch 8.3], which predicts the median home value `medv` in the `Boston` data from all other predictors. The workflow is the familiar one: split into training and test sets, fit the model on the training data, then measure error on the held-out test set. The function `gbart` is the BART package's general-purpose fitter for a continuous response.

```{r, message=FALSE, warning=FALSE}
library(BART)
library(ISLR2)
library(tree)

set.seed(1)

train <- sample(1:nrow(ISLR2::Boston), nrow(ISLR2::Boston)/2)
# tree.boston <- tree(medv ~ ., ISLR2::Boston, subset = train)

x <- ISLR2::Boston[,1:12]
y <- ISLR2::Boston[,"medv"]
xtrain <- x[train,]
ytrain <- y[train]
xtest <- x[-train,]
ytest <- y[-train]

bartfit <- gbart(xtrain, ytrain, x.test = xtest)

# test error
yhat.bart <- bartfit$yhat.test.mean
mean((ytest - yhat.bart)^2)

# how many times each variable appeared in the colelction of trees
ord <- order(bartfit$varcount.mean, decreasing = T)
bartfit$varcount.mean[ord]
```

Two pieces of output are worth reading carefully. The first, the mean squared error on the test set, is the headline accuracy number: smaller is better, and on this split BART is competitive with or better than boosting and random forests. The second, `varcount.mean`, counts how many times each variable appeared in the collection of trees, averaged over iterations. Variables that show up more often are the ones BART relied on more heavily, so this gives a quick, model-based ranking of variable importance.


::: {.callout-note}
Variable appearance counts are a useful first look at importance, but they are not the whole story. A variable can be used many times in shallow splits without changing predictions much. The partial dependence plot later in this chapter shows the *direction and shape* of a variable's effect, which counts alone cannot.
:::


## Worked example: the package authors' walkthrough

The next sequence follows the example from the package's own authors [@sparapani2021]. To keep things visual, it restricts attention to two predictors so we can plot everything. The response and predictors are:

-   `y = medv`, the median value of owner-occupied homes,
-   `x1 = rm`, the average number of rooms, and
-   `x2 = lstat`, the percent of the population that is lower status.

Before fitting anything, it pays to look at the raw relationships. @fig-BART-scatter-raw shows each predictor against the response and the two predictors against each other.

```{r fig-BART-scatter-raw, fig.cap="Scatterplots of the raw Boston data: median home value against average number of rooms (x1=rm), median home value against percent lower status (x2=lstat), and the two predictors against each other."}
library(MASS)
x <- MASS::Boston[,c(6,13)]
y <- MASS::Boston$medv
# head(cbind(x,y))
par(mfrow = c(2,2))
plot(x[, 1], y, xlab = "x1=rm", ylab = "y=mdev")
plot(x[, 2], y, xlab = "x2=lstat", ylab = "y=mdev")
plot(x[, 1], x[, 2], xlab = "x1=rm", ylab = "x2=lstat")
par(mfrow = c(1, 1))


```

The plots already hint at what any good model should recover: home value rises with the number of rooms and falls as the lower-status percentage increases, and the two predictors are themselves negatively related. We now fit BART and see whether it captures these patterns.

Here we use `wbart`, the function for a continuous (weighted) response, and we ask for `nd = 200` posterior draws after a burn-in of `burn = 50` draws.^[`nskip` is the burn-in count $L$ and `ndpost` is the number of retained posterior draws. They correspond directly to the $L$ and $B - L$ in the algorithm above.]

```{r}
library("BART")
set.seed(99)
nd <- 200 # number of posterior draws
burn <- 50
# use wbart for continuous varaibles
post <- wbart(x, y, nskip = burn, ndpost = nd)
```

### Checking convergence

Because BART is an MCMC method, we should never trust the output until we have checked that the chain actually converged. A standard diagnostic is to plot the sampled error standard deviation `sigma` across iterations: if the chain has settled, the trace should stop drifting and fluctuate around a stable level. @fig-BART-sigma-trace shows this trace, where the red line marks the end of the burn-in period.

```{r fig-BART-sigma-trace, fig.cap="Trace of the sampled error standard deviation sigma across MCMC iterations. The vertical red line marks the end of the burn-in period; the trace flattens and fluctuates around a stable level, indicating good convergence."}
plot(post$sigma, type = "l")
abline(v = burn, lwd = 2, col = "red")
# good convergence
```

The trace flattens out quickly and stays level after burn-in, which indicates good convergence. If instead you saw a slow drift or a trend that never stabilized, the remedy would be to increase the burn-in (and possibly the total number of draws) and refit.


::: {.callout-warning}
Skipping this convergence check is the most common BART mistake. Point predictions can look reasonable even when the chain has not mixed, but the uncertainty intervals you report will be wrong.
:::


### Comparing predictions with linear regression

To put BART's fitted values in context, we compare them against ordinary linear regression on the same two predictors. We line up the observed `y`, the BART fitted means, and the linear model fitted values, then look at their correlations and the pairwise scatterplots in @fig-BART-fit-pairs.

```{r fig-BART-fit-pairs, fig.cap="Pairwise scatterplots of the observed response, BART fitted means, and linear-regression fitted values. The BART fitted values track the observed response more closely than the linear fit."}
lmf <- lm(y~., data.frame(x, y))
fitmat <- cbind(y, post$yhat.train.mean, lmf$fitted.values)
colnames(fitmat) <- c("y", "BART", "Linear")
cor(fitmat)
pairs(fitmat)
```

The correlation matrix and the scatterplots tell the story: BART's fitted values track the observed response more closely than the linear fit, which is what we expect when the true relationship is nonlinear (the rooms and lower-status effects both bend rather than running in straight lines).

### From point predictions to uncertainty

One of BART's selling points is that every prediction comes with a posterior distribution, not just a single number. @fig-BART-posterior-boxplot shows, for each training observation (sorted by its fitted mean), the spread of predictions across the retained posterior draws. Wider boxes mean more uncertainty about that observation's prediction.

```{r fig-BART-posterior-boxplot, fig.cap="Boxplots of the posterior predictive draws for each training observation, sorted by fitted mean. Wider boxes indicate greater predictive uncertainty for that observation."}
i <- order(post$yhat.train.mean)
boxplot(post$yhat.train[, i])

n <- length(y)
set.seed(14)
i <- sample(1:n, floor(0.75 * n))
x.train <- x[i, ]; y.train = y[i]
x.test <- x[-i, ]; y.test = y[-i]
cat("training sample size = ", length(y.train), "\n")
cat("testing sample size = ", length(y.test), "\n")

set.seed(99)
post1 <- wbart(x.train, y.train, x.test)
length(post1$yhat.test.mean)
```

In the same chunk we also create a proper 75/25 train/test split and refit, producing test-set predictions in `post1`. The length check confirms we get one predicted mean per test observation.

### Thinning to save time and memory

Long MCMC runs can produce huge output objects, and consecutive draws are correlated anyway, so we often *thin* the chain: run many iterations but keep only a subset. Thinning trades a little statistical efficiency for large savings in memory and storage. Below we run a long chain (1000 burn-in, 10000 draws) but keep only 200 tree draws, then predict on the test set and compare against the earlier, unthinned predictions in @fig-BART-thin-pairs.^[The `nkeep*` arguments control which parts of the output are stored. Setting `nkeeptrain`, `nkeeptest`, and `nkeeptestmean` to 0 tells `wbart` not to keep those potentially large matrices, while `nkeeptreedraws = 200` keeps 200 tree ensembles so we can still predict later with `predict`.]

```{r fig-BART-thin-pairs, fig.cap="Pairwise scatterplots of the observed test response, the full unthinned predictions (yhat), and the thinned predictions (yhatThin). The thinned predictions line up closely with the full predictions."}
set.seed(4)
post3 <-
    wbart(
        x.train,
        y.train,
        nskip = 1000,
        ndpost = 10000,
        nkeeptrain = 0,
        nkeeptest = 0,
        nkeeptestmean = 0,
        nkeeptreedraws = 200
    )
yhatthin <- predict(post3, x.test)

fmat <- cbind(y.test, post1$yhat.test.mean, apply(yhatthin, 2, mean))
colnames(fmat) <- c("y", "yhat", "yhatThin")
pairs(fmat)

```

The scatterplots show that the thinned predictions (`yhatThin`) line up closely with the full predictions (`yhat`), confirming that thinning here costs us almost nothing in accuracy while keeping the stored model small.

### Variable importance via partial dependence

Finally, we look at variable importance in a way that shows not just *whether* a variable matters but *how* it shapes the prediction. Friedman's partial dependence function answers the question: if we sweep one predictor across its range while averaging over the distribution of all the other predictors, how does the predicted response move?

The procedure builds a grid of values for `lstat` (the 13th column), replaces that column with each grid value while holding the other columns at their observed combinations, predicts, and averages. In @fig-BART-partial-dependence, the solid line is the average partial-dependence curve and the dashed lines are pointwise 2.5% and 97.5% posterior quantiles, giving an uncertainty band around the effect.

```{r fig-BART-partial-dependence, fig.cap="Partial dependence of predicted median home value on the lower-status percentage (lstat). The solid line is the average partial-dependence curve and the dashed lines are pointwise 2.5% and 97.5% posterior quantiles."}
x.train <- as.matrix(MASS::Boston[i,-14])
set.seed(12)
post4 <- wbart(x.train, y.train)
H <- length(y.train)
L <- 41
x <- seq(min(x.train[, 13]), max(x.train[, 13]), length.out = L)
x.test <- cbind(x.train[,-13], x[1])
for (j in 2:L)
    x.test <- rbind(x.test, cbind(x.train[,-13], x[j]))
pred <- predict(post4, x.test)
partial <- matrix(nrow = 1000, ncol = L)
for (j in 1:L) {
    h <- (j - 1) * H + 1:H
    partial[, j] <- apply(pred[, h], 1, mean)
}
plot(
    x,
    apply(partial, 2, mean),
    type = "l",
    ylim = c(10, 50),
    xlab = "lstat",
    ylab = "mdev"
)

lines(x, apply(partial, 2, quantile, probs = 0.025), lty = 2)
lines(x, apply(partial, 2, quantile, probs = 0.975), lty = 2)
```

The curve slopes clearly downward: as the lower-status percentage increases, predicted median home value falls, and the decline is steeper at low values of `lstat` than at high ones (a nonlinear effect a linear model would miss). The dashed band reminds us that this estimated effect is itself uncertain, which is precisely the kind of honest summary BART was built to provide.


::: {.callout-tip}
Partial dependence plots are model-agnostic. You met the same idea in the chapter on interpretable machine learning (@sec-interpretable-machine-learning), and you can apply it to random forests, boosting, or BART alike. What is special here is that BART hands you a credible band for free, straight from its posterior draws.
:::


## Summary

BART models a response as a sum of many small trees and fits them with an MCMC sampler that repeatedly perturbs each tree toward the partial residual left by the others. Keeping trees small and updates gentle controls overfitting, while the Bayesian machinery delivers full posterior distributions, so every prediction, variable-importance summary, and partial-dependence curve comes with honest uncertainty. With sensible defaults (many trees, a long chain, a moderate burn-in), it works well with little tuning, which is why it is a reliable first choice for flexible regression and classification.

For examples that extend BART to categorical responses, see [@sparapani2021].
