# Introduction {#intro}

Resources

-   [INTRODUCTION TO MACHINE LEARNING](http://alex.smola.org/drafts/thebook.pdf)
-   [An introduction to Statistical Learning](https://web.stanford.edu/~hastie/ISLRv2_website.pdf)
-   [The Elements of Statistical Learning](https://web.stanford.edu/~hastie/ElemStatLearn/printings/ESLII_print12.pdf)

Machine learning is also known as statistical learning

-   two types of learning:

    -   supervised learning: predicting or estimating output based on inputs

    -   unsupervised learning: understanding the relationships among inputs without output

+-------------+----------------------------------------------------------+-------------------------------------------------------------+
|             | Unsupervised                                             | Supervised                                                  |
+-------------+----------------------------------------------------------+-------------------------------------------------------------+
| Continuous  | -   Clustering & Dimension Reduction (SVD, PCA, K-means) | -   Regression (linear, polynomial)                         |
|             |                                                          |                                                             |
|             |                                                          | -   Tree-based methods                                      |
+-------------+----------------------------------------------------------+-------------------------------------------------------------+
| Categorical | -   Association Analysis (Apriori, FP-growth)            | -   Classification (KNN, Trees, Logistic, Naive Bayes, SVM) |
|             |                                                          |                                                             |
|             | -   Hidden Markov                                        |                                                             |
+-------------+----------------------------------------------------------+-------------------------------------------------------------+

3 types of data

-   Structured Data (tables form)

-   Semi-structured Data (e.g., XML, JSON, HTML)

-   Unstructured Data

    -   Texts
    -   Images: Videos, photos,
    -   Networks: Social Networks

-   ML is concerned with the the development, the analysis, and the application of algorithm that allow computers to learn

-   Learning:

    -   A computer learns if it improves its performance at some tasks with experience (i.e., data)
    -   Extracting a model of system from the sole observation (or the simulation)
    -   A model = some relationships between the variables used to describe the system.

-   Goals:

    -   Prediction
    -   Inference

-   Learning if useful when:

    -   Human expertise does not exist (navigating on Mars)
    -   Humans are unable to explain their expertise (speech recognition)
    -   Solution changes in time (routing on a computer network)
    -   Solution needs to be adapted to particular cases (user biometrics)

-   Applications in different domains:

    -   Retail: Market basket analysis, customer relationship management (CRM)
    -   Finance: Credit scoring, fraud detection.
    -   Manufacturing: Optimization, troubleshooting
    -   Medicine: medical diagnosis
    -   Telecommunications: Quality of service optimization, routing
    -   Bioinformatics: Motifs, alignment WEb mining, search engines.

-   **Steps**

    -   Data generation: data types, sample size
    -   Preprocessing: normalization, missing values, feature selection/extraction
    -   Machine Learning: Hypothesis, choice of learning paradigm/algorithm.
    -   Hypothesis validation: cross-validation, model deployment.

-   Types of machine learning:

    -   Supervised: manually label variables
    -   Unsupervised:

-   Training and testing

    -   Training is the process of making the machine learn

    -   No free lunch:

        -   Training set and testing set come from the same distribution
        -   Need to make some assumptions.

-   Factors affecting the performance:

    -   Types of training provided
    -   The form and extend of prior knowledge
    -   Type of feedback provided
    -   The learning algorithms used

-   Two factors:

    -   Modeling
    -   Optimization

-   Algorithms:

    -   ML depends on algorithms.
    -   The algorithms control the search to find and build the knowledge structures
    -   The learning algorithms should extract useful information from training examples.

-   Types of learning (algorithms):

    -   Supervised learning ($\{ X_n \in R^d, y_n \in R\}^N_{n=1}$)

        -   Prediction
        -   Classification (discrete labels), regression (real values)

    -   Unsupervised learning ($\{ X_n \in R^d\}^N_{n=1}$)

        -   Clustering
        -   Probability distribution estimation
        -   Finding association (in features)
        -   Dimension reduction

    -   Semi-supervised learning

    -   Reinforcement learning

        -   Decision making (robot, chess machine)

<br>

## General Framework

-   $Y$ = output

-   $X = (X_1,...,X_p)$ = p inputs

Under supervised learning, additive error model:

$$
Y = f(x) + \epsilon
$$

where

-   $f(.)$ = unknown fixed function

-   $\epsilon$ = random error ($E(\epsilon) = 0; \epsilon \perp X$)

For prediction (blackbox)

$$
\hat{Y} = \hat{f}(X)
$$

where

-   Reducible Error: $\hat{f}(.)$ can be improved (in principle)

-   Irreducible Error: $\epsilon$ because it's not a function of the inputs, which might contain unknown variables or uncontrolled variation

$$
\begin{aligned}
E(Y - \hat{Y})^2 &= E[f(X) + \epsilon - \hat{f}(X)]^2 \\
&= [f(X) - \hat{f}(X)]^2 + var(\epsilon) \\
&= \text{reducible} + \text{irreducible}
\end{aligned}
$$

For inference, $f(X) = E(Y|X)$, but the distribution of Y\|X is unknown; hence, must be estimated.

We try to find $\hat{f}$ such that $Y \approx \hat{f}(X) \forall (X,Y)$

Ways to estimate $f$:

-   Parametric

    1.  Make an assumption about the function form of $f(X)$

    2.  Try to fit data to that specified form (i.e., estimate the parameters) using OLS in regression or MLE in general.

-   Nonparametric

Tradeoff between prediction accuracy and model interpretability

+--------------------+-----------------+
| Restrictive Model  | Flexible Model  |
+====================+=================+
| More inference     | More prediction |
|                    |                 |
| More interpretable |                 |
+--------------------+-----------------+

## Assessing Model Accuracy - Model fit

### Continuous output: Mean squared error (MSE)

$$
MSE = \frac{1}{n} \sum_{i=1}^n (y_i - \hat{f}(x_i))^2
$$

Generalization Error is a measure of how accurately an algorithm or model is able to predict outcome values for test set.

Tradeoff between flexibility of the method and training MSE , but there is an U-shaped relationship between flexibility and test MSE. This is a manifestation of bias-variance tradeoff

$$
E(MSE) = E(y_0 - \hat{f}(x_0))^2\\
= var(\hat{f}(x_0)) + [bias(\hat{f}(x_0))]^2 + var(\epsilon)
$$

where $0$ subscript means training data set

-   $E(MSE) \ge var(\epsilon)$ because $\epsilon$ is irreducible model error and serves as the lower bound on the quality of potential model fit

-   More flexible means more variance, but less bias.

-   Less flexible (parametric) means less variance, but more bias.

### Categorical output: training error rate

**Training error rate** is the proportion of mistakes in classification from applying $\hat{f}$ to the training set:

$$
\frac{1}{n} \sum_{i=1}^n I(y_i \neq \hat{y}_i)
$$

and **test error rate** is

$$
\frac{1}{n} \sum_{i=1}^n I(y_0 \neq \hat{y}_0)
$$

**Bayes classifier** is a simple classifier where it assigns each observation to the most likely class, given its predictor values (i.e., conditional probability that $Y = j | x_0$ )

In other words, assigns $x_i$ to class j where $P(Y=j | X=x_0)$ is largest

Bayes classifier should produce the lowest possible test error rate, known as Bayes error rate:

$$
1 - E_x(max_j [P(Y = j |X)])
$$

-   where it is greater than 0 if the classes overlap

-   analogous to the irreducible error

In reality, we don't know $P(Y|X)$, hence, we try to approximate this conditional probability.

For example, KNN

$$
P(Y =j |X = x_0 ) \approx \frac{1}{K}\sum_{i \in N_0} I(y_i = j)
$$

### Curse of dimensionality

Example: with a p-dimensional hypercube of unit volume, to capture a fraction (r) of the observations in a subcude, the length of a side of the sub-cube averages to $e_p(r) = r^{1/p}$

$$
e_1(.01) = .01 ; e_1(.1) = .1\\
e_{10}(.01) = .63; e_{10} (.1) = .8
$$

To cover 1% of the observation in 10 dimensional cub, we have to cover 60% of the entire sample space.

Hence, there is no "local" neighborhood in high-dimensions.

### Reminder

Linear Regression

$$
Y = f(X) + \epsilon
$$

where $f(X) = \beta_0 + \sum_{j=1}^p X_j \beta_j$ and $X_J$ can be:

-   quantitative inputs

-   transformations of quantitative inputs

-   polynomial terms

-   interactions

-   Discretize (i.e., dummy) variable of quantitative inputs.

We assume additive independent errors.

| Relationship between n and p | Result                                                      |
|------------------------------|-------------------------------------------------------------|
| $n >>p$                      | low bias, low predictive variance, fit well with test data  |
| $n \approx p$                | substantial variability, overfit, fit poorly with test data |
| $n <p$                       | no unique solution, variance is infinite                    |

And we fit to (i.e., find $\beta$) to minimize the residual sum of squares (RSS):

$$
RSS(\beta) = \sum_{i=1}^n(y_i - f(x_i))^2 = \sum_{i=1}^n (y_i -\beta_0 - \sum_{j=1}^p x_{ij} \beta_j)^2 \\
= (\mathbf{y} - \mathbf{X\beta})^T (\mathbf{y} - \mathbf{X\beta}) \equiv  ||\mathbf{y} - \mathbf{X\beta}||^2
$$

which is the square of the $L^2$ norm

differentiating with respect to $\beta$

$$
\mathbf{X}^T  (\mathbf{y} - \mathbf{X\beta}) = 0 
$$

then

$$
\hat{\beta} = \mathbf{X}^T\mathbf{X}^{-1}\mathbf{X}^T \mathbf{y}
$$

and

$$
var(\hat{\beta}) = (\mathbf{X}^T \mathbf{X})^{-1} \sigma^2 \\
 = (\mathbf{X}^T \mathbf{X})^{-1} var(\epsilon)
$$

**Prediction**

$$
\hat{\mathbf{y}} = \mathbf{X} \hat{\beta} = \mathbf{X}(\mathbf{X}^T \mathbf{X})^{-1}\mathbf{X}^T\mathbf{y}
$$

where

$$
\mathbf{H} \equiv \mathbf{X}(\mathbf{X}^T \mathbf{X})^{-1}\mathbf{X}^T
$$

known as the hat matrix, which projects orthogonally the $\mathbf{y}$ vector onto the subspace of $\mathbf{X}$

From the Gauss-Markov theorem, OLS estimates are unbiased and have the smallest variance among all linear unbiased estimates.

For an etimator $\tilde{\beta}$

$$
MSE(\tilde{\beta}) = E(\tilde{\beta} - \beta)^2 = var(\tilde{\beta} + [E(\tilde{\beta}) - \beta]^2 \\
 = variance + bias ^2 
$$

Since OLS is an unbiased estimator $\tilde{\beta} = \hat{\beta}$

We could get smaller MSE if we allow the estimator to be biased. Hence, OLS might not be the best variance-bias trade-off.

Inference

$$
\hat{\sigma}^2 = \frac{1}{n - p - 1} \sum_{i=1}^n (y_i - \hat{y}_i)^2
$$

assuming that $\epsilon \sim N(0, \sigma^2)$ then

$$
\hat{\beta} \sim N(\beta, (\mathbf{X}^T \mathbf{X})^{-1}\sigma^2)
$$

$$
(n-p-1) \hat{\sigma}^2 \sim \sigma^2 \chi^2_{n-p-1}
$$

If p is large (i.e., a lot of covariates), you will likely find at least one variable to be significant by chance, thus, you could find a significant partial t-statistics. The F-test does not suffer from this issue because it accounts for the number of predictors.

Issues with linear regression

-   Nonlinearity of the response -predictor relationship

-   correlated errors

-   Non-constant error variance

-   high-leverage points

-   Collinearity

Rule of thumb:

-   parametric methods will tend to outperform non-parametric methods when the number of observations per predictor is small.

<br>

### Model Assessment

The "generalization" performance of a learning method is how well it predicts independent test data, which is an objective way to choose between models (in terms of **model selection** and **model assessment**)

Suppose we have

-   Y = output variable

-   X = vector of inputs

-   $\hat{f}(X)$ = prediction model estimated form a training set $T$

Consider errors between Y and $\hat{f}(X)$ through some loss function. For example, **squared-error loss**

$$
L(Y, \hat{f}(X)) = (Y- \hat{f}(X))^2
$$

or **absolute error loss**

$$
L(Y, \hat{f}(X)) = |Y- \hat{f}(X)|
$$

#### Types of Error

**Test error (or generalization error)** is the prediction error over an independent test sample, and is defined as

$$
Err_T = E_{X^\circ, Y^\circ}[L(Y^\circ, \hat{f}(X^\circ))|T]
$$

where X and Y are assumed to be drawn randomly from their joint population distribution and training set $T$ is fixed

Thus, the test error refers to the error over an independent sample when $f(.)$ is estimated with this specific training set

<br>

**Expected prediction error (expected test error)**

$$
Err = E_T E_{X^\circ, Y^\circ}[L(Y^\circ, \hat{f}(X^\circ))|T] = E_T [Err_T]
$$

which averages over possible training sets

**Training error** is the average loss over the specific training sample, given by

$$
\bar{err} = \frac{1}{n} \sum_{i=1}^n L(y_i, \hat{f}(x_i))
$$

Even though the interest is in the test error, $Err_T$ , which is almost impossible without a separate test data set. because within a single data set, we attempt to approximate such a validation step by adjusting the training error (e.g., AIC model selection) or by efficient re-use (cross-validation/bootstrap) of the training sample, but then we can only get good estimates of $Err$ (expected prediction error), not $Err_T$

The bias-variance tradeoff when considering loss function is related to the performance of the training error as a measure of test error

#### Model Error/ Model Selection

Training error is more optimistic (less than) the true test error

To read more on quantifying this optimism [@hastie2001a] Ch. 7.4 where

the amount that $\bar{err}$ underestimates the true test error depends on how strongly $y_i$ affects its own prediction.

In other words, the larger $cov(\hat{y}_i, y_i)$, the more optimistic the training error is.

The greater the effective number of parameters (less degrees of freedom) (more flexible model), the larger this co variance

[@hastie2001a] Ch. 7.5 showed that most of model selection criterion (e.g., AIC, BIC $C_p$) try to estimate prediction error by estimating the optimism and adjusting the training error accordingly to obtain an unbiased estimate of prediction error.

Alternatively, cross-validation and bootstrap methods provide direct estimates of the expected prediction (test) error, $Err$

Note:

-   A model with more effective number of parameters is more flexible

Let

$$
\mathbf{y} = (y_1, \dots, y_n)'
$$

then a linear fitting method can be written as $\mathbf{\hat{y}} = \mathbf{Sy}$

where $\mathbf{S} = n \times n$ matrix depending on the input vectors $x_i$, but not on $y_i$

For example, in linear regression $\mathbf{S = X(X'X)^{-1}X'}$

Then the effective number of parameters (df) is

$$
df(\hat{\mathbf{y}}) = df (\mathbf{S}) = trace(\mathbf{S})
$$

which is the sum of the diagonal elements fo $\mathbf{S}$

If $\hat{\mathbf{y}}$ arises from an additive-error model, $Y = f(X) + \epsilon$, with $var(\epsilon) = \sigma^2_\epsilon$, then

$$
df(\hat{\mathbf{y}}) = \frac{\sum_{i=1}^ncov(\hat{y}_i, y_i)}{\sigma^2_\epsilon}
$$

<br>

#### Cross-validation

If you have a large data set, you could randomly divide it into a training set (to fit the model) and a validation (or holdout, or test) set (to evaluate the test error rate relative to your loss function of choice such as MSE). But it has 2 issues:

-   It can be highly variable

-   With fewer observations to fit the model, the statistical models perform worse, and validation set error rate may overestimate the test error rate

```{r, message = FALSE}
library(tidyverse)
library(ISLR2)
train <- sample(392, 120) # select random subset of 120 out of 392
Auto <- Auto %>% drop_na()
lm.fit <- lm(mpg ~ horsepower , data = Auto , subset = train)

attach(Auto)
mean((mpg - predict(lm.fit, Auto))[-train] ^ 2) # MSE in the validation set (not the training set)

# Quadratic regression
lm.fit2 <- lm(mpg ~ poly(horsepower, 2), data = Auto, subset = train)
mean((mpg - predict(lm.fit2, Auto))[-train] ^ 2) # MSE

# Cubic regression
lm.fit3 <- lm(mpg ~ poly(horsepower, 3), data = Auto, subset = train)
mean((mpg - predict(lm.fit3, Auto))[-train] ^ 2) # MSE
```

<br>

##### Leave-One-Out CV (LOOCV)

A special case of K-fold CV when K = n

$$
CV_{(n)} = \frac{1}{n} \sum_{i=1}^n MSE_i
$$

where $MSE_i = (y_i - \hat{f}^{-i}(x_i))^2$ (alternative loss function could also be used)

-   LOOCV has low bias as an estimate of the expected test error rate, but very high variance

-   Individual MSE's in LOOCV are highly correlated because they pretty much have the same data set expect one observation.

-   LOOCV is expensive to implement

```{r, message=FALSE}
Auto <- Auto %>% drop_na()
library(tidyverse)
glm.fit <- glm (mpg ~ horsepower , data = Auto) # allows to be passed to cv.glm()
coef(glm.fit)

# exactly the same as glm without binomial link
lm.fit <- lm(mpg ~ horsepower , data = Auto)
coef(lm.fit)

library(boot)
glm.fit <- glm (mpg ~ horsepower , data = Auto)
cv.err <- cv.glm (Auto , glm.fit)
cv.err$delta # cross-validation estimate for the test error
# the first delta is the standard CV estimate, the second delta is a bias-corrected version


# automate to different polynomial
cv.error <- rep (0, 10)
for (i in 1:10) {
    glm.fit <- glm(mpg ~ poly(horsepower , i), data = Auto)
    cv.error[i] <- cv.glm(Auto , glm.fit)$delta[1] # cross-validation estimate for the test error
}
cv.error # a drop from lienar to quadratic fit
```

<br>

##### K-Fold Cross-Validation

We randomly split the data set into K equal size parts (or folds). Each fold is held out and the model is trained on the remaining K -1 folds and then the loss function of choice is evaluated on the fold that was held out

Consider $k = 1, \dots, K$ folds

let $\hat{f}^{-k}(x)$ denote the fitted function computed with the k-th fold removed. We compute the loss based on the samples in the k-th fold.

For example, with squared-error loss, we compute $MSE_k$ for the observations in the k-th fold, $k = 1,\dots,K$. The K-fold cross-validation estimate is

$$
CV_{(K)} = \frac{1}{K} \sum_{k=1}^K MSE_k
$$

Empirically, good choices for K are 5 or 10.

```{r}
cv.error.10 <- rep(0, 10)
for (i in 1:10) {
    glm.fit <- glm(mpg ~ poly(horsepower, i), data = Auto)
    cv.error.10[i] <- cv.glm(Auto, glm.fit, K = 10)$delta[1] # K-fold cross validation
}
cv.error.10
```

<br>

##### Generalized Cross-Validation

Recall PRESS statistics

$$
\frac{1}{n} \sum_{i=1}^n [y_i - \hat{f}^{-i}(x_i)]^2 = \frac{1}{n} \sum_{i=1}^n [\frac{y_i - \hat{f}(x_i)}{1- S_{ii}}]^2
$$

where $S_{ii}$ is the i-th diagonal element of $\mathbf{S}$ (or the i-th element of the "hat matrix" diagonal in linear regression). Hence, we only have to fit the model once.

Generalized cross-validation (GCV)is a similar approximation when the trace of $\mathbf{S}$ can be computed more easily than the individual elements $S_{ii}$

$$
GCV(\hat{f})= \frac{1}{n} \sum_{i=1}^n [\frac{y_i - \hat{f}(x_i)}{1- trace(\mathbf{S})/n}]^2
$$

<br>

##### CV and Classification

Common loss function is 0-1 loss where the error rate is defined as

$$
L(G, \hat{G}(X)) = I(G \neq \hat{G}(X))
$$

where $\hat{G}(X)$ is the highest estimated conditional probability given the inputs.

Alternatively, we could also use deviances that factor in the estimated conditional probability directly. If $p_k(X) = P(G = k|X)$ and $\hat{G}(X)$ = the k that gives the maximum $\hat{p}_k(X)$

$$
L(G, \hat{p}(X)) = -2 \sum_{k=1}^K I(G = k) \log \hat{p}_k (X) \equiv -2 \log \hat{p}_G (X)
$$

which is -2 times the log-likelihood (i.e., the deviance).

Note:

-   CV must be applied to the entire sequence of modeling steps (i.e., samples must be removed before any selection or filtering steps are applied so long as these steps involve the outputs

    -   An exception: if there was an unsupervised step in the process associated with the inputs. Since it's unrelated to the outputs, we don't need to do CV on this component.

#### Bootstrapping

-   To quantify the uncertainty associated with a given estimator

-   Get new data sets by sampling from the observed data set many times with replacement

-   Sample $B$ different sets and calculate the quantity of interest about which you want to do inference $\hat{\alpha}^{*1},\dots, \hat{\alpha}^{*B}$

-   Then, you look at the distribution of samples and apply Monte Carlo procedures to do inference

-   For example, you want to get the standard error

$$
SE_B (\hat{\alpha}) = \sqrt{\frac{1}{B-1} \sum_{r=1}^B (\hat{\alpha}^{*r} - \frac{1}{B} \sum_{r'=1}^B \hat{\alpha}^{*r'})^2}
$$

-   In principle, you could evaluate model error, but it often overfits because the training and test samples contain some of the same observations, which can be improved by "leave-one-out" bootstrap procedures, but there are still large training set bias problems with the bootstrap.

-   [@hastie2001a] give an estimator that helps reduce this bias.

-   In general, cross-validation is more effective for evaluating model performance, and bootstrapping is more effective when evaluating the properties of estimators (for inference).

-   Rule of thumb for "big data" prediction problem, we can combine cross-validation and hold-out validation:

    -   Use K-fold CV on the training data to choose the parameters that control flexibility of the model

    -   Use the entire training data to fit the model selected by CV

    -   Use a large hold-out sample to evaluate the test error from this fitted model (compare to the CV error).

```{r}
summary(Portfolio)
alpha.fn <- function(data, index) {
    X <- data$X[index]
    Y <- data$Y[index]
    (var(Y) - cov(X, Y)) / (var(X) + var(Y) - 2 * cov(X, Y)) # statistic of interest
}

alpha.fn(Portfolio, 1:100)
alpha.fn(Portfolio, sample(100, 100, replace = T)) # randomly select 100 obs from 1:100 with replacement


# Another faster way
boot(Portfolio, alpha.fn, R = 100) # R = number of replicates (samples)
```

Another example: SE estimate for linear model

```{r}
boot.fn <- function(data,index){
    coef(lm(mpg ~ horsepower, data = data, subset = index))
}
boot.fn(Auto, 1:392)
boot.fn(Auto, sample(392,392,replace =T))

boot(Auto, boot.fn,1000)

# there is a difference between the 2 SE estimates
summary(lm(mpg ~ horsepower, data = Auto))$coef




# Quadratic example
# boot.fn <- function(data,index){
#     coef(
#         lm(mpg ~ horsepower + I(horsepower^2),
#            data = data, subset = index)
#     )
# }
# boot(Auto, boot.fn, 1000)
# summary(
#     lm(mpg ~ horsepower + I(horsepower^2), data = Auto)$coef
# )

```

The difference in SE estimates from bootstrapping and from a linear model comes from:

1.  LM estimates $\sigma^2$ suing the RSS, which relies heavily on the linear model being correct. In other words, for example, if we have non-linearity in our data, the residuals ($e$) will be inflated, and leads to $\hat{\sigma}^2$ being inflated as well.
2.  LM assumes $x_i$ fixed and variability comes from $var(\epsilon_i)$

<br>

## Improving Prediction Accuracy and Interpretability

If we have

1.  large number of predictors
2.  highly dependent predictors
3.  low interpretability

then, we can either

1.  [Subset Selection]

2.  [Shrinkage Methods] (also known as "*regularization*"): force (shrink) p predictors' parameters towards zero to reduce variance. It can also be used for variable selection

3.  [Dimension Reduction]: Project the p predictors in an M-dimensional subpsace where $M <p$ (i.e., compute M different linear combination or projection of the p predictors and use them as new predictors). (e.g., principal component analysis)

### Subset Selection

-   Also known as [Variable Selection](https://bookdown.org/mike/data_analysis/ordinary-least-squares.html#variable-selection) (for more detail click the hyperlink)

#### Best Subset Selection

-   Fit all $p$ one by one, then all $p(p-1)/2$ two predicotrs models, etc. to have $2^p$ possible models. Then you find the best one using

    -   $C_p$

    -   AIC

    -   BIC

    -   Adjusted $R^2$

    -   cross-validation error measure

-   A possible algorithm is the leaps and bounds algorithm by @furnival1974 (up to 40 p)

-   Usually, we do best-subset first then apply cross-validation

```{r}
library(tidyverse)
library(ISLR2)
head(Hitters) # check dataset 

# check missing values 
sum(is.na(Hitters$Salary)) # 59 missing data poitns

# eliminate missing values 
Hitters <- Hitters %>% 
    drop_na(Salary)

# Best subset seleciton (by RSS)
library(leaps)
regfit.full <- regsubsets(Salary ~ ., Hitters)
# summary(regfit.full)

```

```{r}

```

<br>

#### Stepwise Selection

-   With very large numbers of p, stepwise selection can be less computational intensive

-   Forward stepwise selection

    -   can be used when $p >n$, but only models up to order $n-1$

-   Backward stepwise selection

    -   can't be applied when $p>n$

-   Hybrid stepwise

-   When applying cross-validation, to account fro the variability, we apply the **one-standard-error rule**

    -   Evaluates cross-validation criteria (MSE) for each model size

    -   Select the smallest model where the estimated test error is within one SE of the lowest point on the curve (i.e.more parsimonious model)

### Shrinkage Methods

-   Constrain the Least Squares solution in a way that regularizes (shrinks) the coefficients towards 0.

-   This leads to biased estimates, but reduce variance

#### Ridge regression

-   can also deal with multicolinnearity

-   Adding a constant to the diagonal of the $\mathbf{X}^T \mathbf{X}$ matrix to get well-behaved (low variance) regression estimates even when $\mathbf{X}^T \mathbf{X}$ is nearly singular.

In LS regression, we minimize the residual sum of squares

$$
RSS = \sum_{i=1}^n(y_i - \beta_0 - \sum_{j=1}^p \beta_j x_{ij})^2
$$

whereas in ridge regression, we estimate $\hat{\beta}^R$ that minimize

$$
\sum_{i=1}^n (y_i- \beta - \sum_{j=1}^p \beta_j x_{ij})^2 + \lambda \sum_{j=1}^p \beta_j^2
$$

where $\lambda$ = tuning parameter

$\lambda \sum_{j=1}^p \beta_j^2$ is an $l_2$-norm shrinkage penalty that is small if the $\beta$ are near 0.

Minimizing this objective function subject to this constraint forces $\hat{\beta}^R$ to be closer to 0 than the LS estimates.

-   When $\lambda = 0$, we get the LS estimates

-   When $\lambda \to \infty$ , the coefficients go to 0.

Because in the $l_2$-norm does not contain the intercept (i.e, we don't shrink the intercept), we have to center all the covariates by removing the mean of each column before we do the regression ($\hat{\beta}_0= \bar{y}$).

$$
\hat{\beta}^R = (\mathbf{X}^T \mathbf{X} + \lambda \mathbf{I})^{-1}\mathbf{X}^T \mathbf{y}
$$

To estimate $\lambda$, we use cross-validation

-   Ridge estimator is not scale-invariant, so coefficients change substantially depending on the scale of the covariates. Hence, before performing ridge regression, we have to scaled and centered covariates (i.e., after removing the mean, divided each column of $\mathbf{X}$ by its standard deviation.

-   $\lambda$ allows for a continuous range of bias-variance trade-offs.

    -   As $\lambda$ increases, the variance decreases but the bias increases. Hence, it works well when LS estimates have high variance.

-   Possible even when $p>n$

-   Under Bayesian framework, if $y_i \sim N(\beta_0 + x^T_i\beta, \sigma^2)$ is the likelihood and the parameters have the independent prior distributions $\beta_j \sim N(0, \tau^2)$ with known $\tau^2, \sigma^2$, then $\lambda = \sigma^2 / \tau^2$ (the ratio of regression to prior variances), and the ridge estimate is the posterior mode (and mean, since the posterior is Gaussian)

-   The ridge estimates are related to principal component regression. When principal components associated with the predictors, ridge regression shrinks the low variance principal component the most.

<br>

#### Lasso regression

-   Even though ridge regression shrinks the parameters to 0, unless $\lambda= \infty$, it cant set any parameter exactly equal 0. Hence, @tibshirani1996 proposed lasso to handle this problem

The lasso coefficients, $\hat{\beta}^L$ are those that minimize the objective function

$$
\sum_{i=1}^n (y_i - \beta_0- \sum_{j=1}^p \beta_j x_{ij})^2 + \lambda\sum_{j=1}^p|\beta_j|
$$

The only difference between ridge and lasso is the penalty term:

-   Lasso considers $l_1$ norm penalty (given by $||\beta||_1 = \sum_j |\beta_j|$)

-   Ridge considers $l_2$ norm penalty (given by $||\beta||_2 = \sqrt{\sum_j \beta_j^2}$) similar to Euclidean distance

When tuning parameter ($\lambda$) is large enough, some coefficients are 0 (are dropped from the model)

There is no closed form for the lasso estimates $\hat{\beta}^L$ , the choice of $\lambda$ can be derived from cross-validation

Note:

-   Lasso outperforms ridge when there are redundant predictors, but ridge outperforms lasso when all included predictors are needed

-   Lasso shrinks each parameter by approximately the same amount, while ridge shrinks proportionally

-   Lasso equivalent Bayesian interpretation is that: if the prior distribution on the parameters $\beta_j$ are independent double exponential (Laplace) distribution with mean 0 and scale parameter $\lambda$, the posterior mode for $\beta$ is the lasso solution (but no the posterior mean); the Laplace prior puts more "mass" at 0, giving more a priori belief that the coefficients are 0.

#### Comparison

$$
\displaystyle{\min_{\beta}} \sum_{i=1}^n (y_i - \beta_0- \sum_{j=1}^p \beta_j x_{ij})^2 \text{ subject to }
\begin{cases}
\sum_{j=1}^p |\beta_j| \le s \text{ (lasso)} \\
\sum_{j=1}^p \beta_j^2 \le s \text{ (ridge)} \\
\sum_{j=1}^p I(\beta_j \neq 0) \le s \text{ (best subset)}
\end{cases}
$$

All seek to minimize the RSS subject to constraints on the parameters.

In the best subset method, no more than s coefficients can be nonzero, and can't be computed for large p

#### Cross-validation Fitting

Steps:

1.  Choose a "grid" of $\lambda$ values
2.  Compute the cross-validation error for each vlaue of $\lambda$
3.  Select the $\lambda$ for which the cross-validation error is smallest (one-standard error rule)
4.  Refit the model using all available observations for the selected tuning parameter.

<br>

#### Generalization

We could generalize the objective function to

$$
\sum_{i=1}^n (y_i - \beta_0- \sum_{j=1}^p \beta_j x_{ij})^2 + \lambda \sum_{j=1}^p |\beta_j|^q
$$

for $q \ge 0$, and $q =0,1,2$ correspond to subset selection lasso, and ridge respectively

We could estimate q from the data, but it isn't worth.

Elastic net penalty (= combination of ridge and lasso) will be an alternative

<br>

The penalty component in the objective function is the priori distribution on the $\beta$ under Bayesian framework (e.g., a normal prior = ridge regression and a Laplace prior = lasso). Hence, we can use **stochastic search variable selection (SSVS)** to consider model selection @george1993

SSVS is a hierarchical mixture of normal distribution

$$
\beta_j |\gamma_j \sim \gamma_j N(0, c_j \tau^2_j) + (1- \gamma_j)N(0, \tau^2_j)
$$

$$
\gamma_j \sim^{idd}Bern(\pi_j)
$$

where $\gamma_j=1$ means the j-th variable is in the model with probability $\pi_j$

We would want $\tau_j$ to be small so that when $\gamma_j =0$, it's reasonable to estimate $\beta_j$ close to 0.

We would want $c_j$ to be large so that we are more likely to get a non-zero $\beta)j$ when $\gamma_j = 1$ (which is analogous to ridge regression)

An alternative to SSVS produce is **spike and slab** which replaces the second mixture distribution with a delta function at 0 (analogous to lasso)

### Dimension Reduction

Let $Z_1, \dots, Z_M$ represent $M<p$ linear combination of the original $p$ predictors:

$$
Z_m = \sum_{j=1}^p \phi_{jm} X_j, m = 1,\dots, M
$$

for constants $\phi_{1m}, \dots, \phi_{pm}$

Consider the least squared linear regression model

$$
y_i = \theta_0 + \sum_{m=1}^M \theta_m z_{im} + \epsilon_i, i = 1, \dots,n
$$

Hence, we will try to find an optimal way to choose $\phi_{jm}$ so that we will have better results than the least squared solution on the original inputs

Special note

$$
\begin{aligned}
\sum_{m=1}^M \theta_m z_{im} &= \sum_{m=1}^M \theta_m \sum_{j=1}^p \phi_{jm} x_{ij} \\
&= \sum_{j=1}^p  \sum_{m=1}^M \theta_m \phi_{jm} x_{ij} \\
&= \sum_{j=1}^p \beta_j x_{ij} \\
\end{aligned}
$$

Hence,

$$
\beta_j = \sum_{m=1}^M \theta_m \phi_{jm}
$$

This means the reduced dimension regression is a special case of OLS except that the parameters $\beta_j$ are constrained.

Hence, when $M = p$ and $Z_m$ are all linearly independent (no constraints), the reduced dimension regression is equivalent to the OLS . The only benefit is that $Z_m$'s are orthogonal to each other

Similarly to the [Shrinkage Methods], the constrained solution is biased, but typically leads to lower variance in the fitted coefficients - especially when $M << p$

#### Principal Component Regression

Steps:

1.  Find the linear combination of the inputs that accounted for the largest fraction of variance of the inputs
2.  Find another linear combination of the inputs that accounted for the next highest amount of variation (subject to being orthogonal to the first)
3.  Continue the process

Mathematically, principal components could also be obtained using **singular value decomposition** (SVD) fo the centered data matrix

Let $\mathbf{X}$ ($n \times p$) be the centered data matrix, then

$$
\mathbf{X} = \mathbf{UDV}^T
$$

where $\mathbf{U}$ and $\mathbf{V}$ are $n \times p$ and $p \times p$ orthogonal matrices

$\mathbf{D}$ (**singular values**) is $p \times p$ diagonal matrix with diagonal entries $d_1 \ge d_2 \ge \dots \ge d_p \ge 0$

Then,

$$
\mathbf{X}^T \mathbf{X} = \mathbf{VD}^2 \mathbf{V}^T
$$

which is the eigen-decomposition of $\mathbf{X}^T \mathbf{X}$ and of the sample covariance matrix $\mathbf{S} = \frac{\mathbf{X}^T \mathbf{X}}{n}$ up to the factor n

And the columns of $\mathbf{V}$ are **eigenvectors** and correspond to the **principal component weights**

Hence, the new variables are

$$
\mathbf{z}_1 = \mathbf{Xv}_1 , \dots, \mathbf{z}_M = \mathbf{Xv}_M
$$

where $\mathbf{v}_m$ is the m-th column of $\mathbf{V}$

These $\mathbf{z}$'s are the principal components.

The elements of $\mathbf{v}_m$ are $\phi$'s in the dimension reduction summary

$$
var(\mathbf{z}_m) = \frac{d^2_m}{n}
$$

Thus, large (small) singular values correspond to directions in the column space of $\mathbf{X}$ that have large (small) variance

Then principle component regression basically construct the first $M$ principal components (i.e., $Z_1, \dots, Z_M$) and uses these components as predictors int the linear regression model fit by least squares

Assumption:

-   Directions in which $X_1, \dots , X_p$ have the most variation are the directions that are most associated with $Y$ (not always the case, but typically effective). In other words, these X's are mostly associated with Y (best predict), but we cannot be certain that X's variation causes Y's variation.

Notes:

-   While capturing most of the variance from inputs, since $M <<p$ we don't run into the problem of overfitting

-   Another benefit is that the new predictors are orthogonal ($\mathbf{z}^T_j \mathbf{z}_k =0, j \neq k$), which in the LS estimation we typically run into the problem of collinearity

-   Even though PCR uses a low-dimensional set of predictors, no model selection was performed because we only translate all $p$ variables into a linear combination that makes the new PC inputs. Hence, it's closely related to [Ridge regression] than [Lasso regression] or [Best Subset Selection]. Seeing Singular Value Decomposition

    -   $\mathbf{X}\hat{\beta}^R = \sum_{j=1}^p \mathbf{u}_j \frac{d_j^2}{d_j^2+ \lambda}\mathbf{u}_j^T \mathbf{y}$

    -   where $\mathbf{u}_j$ corresponds to the j-th column of $\mathbf{U}$ from the SVD

    -   Since LS estimate gives $\mathbf{X}\hat{\beta} = \mathbf{UU}^T \mathbf{y}$, ridge regression goes to the LS solution when $\lambda = 0$

    -   The singular vectors that have smaller variation (i.e., smaller $d_j^2$) are shrunk more than those with larger variation

-   Rule of thumb: standardize the predictors before performing the Singular Value Decomposition to generate the principle components.

    -   If we don't standardize, the predictors with the largest scale typically account for the most variance, regardless of their importance relative to the output.

#### Partial Least Squares

-   Find linear combinations of the data (or directions in the $\mathbf{X}$ matrix) that maximize **variance** and are **orthogonal**

-   Find the weight just like [Principle Component Regression], but PCR in a unsupervised way because the choice of these directions is not dependent on $Y$. Hence, might not provide good predictions of $Y$

-   PLS is a supervised way to do dimension reduction that finds new features $Z_1, \dots, Z_M$ that are linear combinations of the original inputs, but it selects these in a away to account for the covariability withe the response Y

Steps to calculate the PLS directions ($Z_m = \sum_{j=1}^p\phi_{jm}X_j$)

1.  Standardize the $p$ predictors

2.  For $Z_1$, set each $\phi_{j1}$ = the coefficient from the simple linear regression of Y on $X_j$ (this is proportional to the correlation between Y and $X_j$)

3.  Calculate $Z_2$, by calculating the residuals of a regression of each variable (separately) on $Z_1$ (call these $e_j$), which corresponds to the info remaining that has not been accounted for by $Z_1$

4.  Set $\phi_{j2}$ = the coefficient from the simple linear regression of $Y$ on each $e_j$ to get $Z_2$

5.  Repeat $M$ to get $Z_1, \dots, Z_2$

**Comparison to PCR:** @hastie2001a (chapter 3.5)

The m-th PC direction coefficients $\mathbf{v}_m$ are the $\alpha$ that solve

$$
\max_\alpha[var(\mathbf{X\alpha})]
$$

subject to $||\alpha||=1, \alpha^T \mathbf{Sv}_l = 0, l = 1,\dots, m-1$ (the last condition ensures that each $\mathbf{z}_m = \mathbf{Xv}_m$ is uncorrelated with all of the previous $\mathbf{z}_l$)

where $\mathbf{S}$ is the sample covariance matrix associated with $\mathbf{X}$

The m-th PLS direction $\hat{\phi}_m$ have the $\alpha$ that solves:

$$
\max_\alpha  [corr^2(\mathbf{y}, \mathbf{X \alpha}) var(\mathbf{X \alpha})]
$$

subject to $||\alpha|| =1, \alpha^T \mathbf{S}\hat{\phi}_l =0,l = 1, \dots, m-1$

Notes:

-   The extra component in PLS helps make the weight more relatable to Y (but sometimes it's more efficient, but sometimes it isn't)

-   Both X and Y should be standardized before performing PLS

-   The choice of $M$ is made based on [Cross-validation]

-   Although PLS tries to maximize both the correlation to the output and the variance in the inputs, the variance tends to dominate and PLS behaves very similarly to ridge regression and PCR

-   The use of the output to supervise the selection of the reduced dimension can reduce bias, but increase variance (bias-variance trade-off), and reduce the added value of the approach relative to PCR

-   PLS is closely related to canonical correlation analysis (CCA) in multivariate statistic, which seeks to find a reduced set of linear combinations of inputs and linear combination of outputs that maximize the correlation (e.lg., finding the coefficient $\alpha, \beta$ s.t

$$
\max_{\alpha, \beta} [corr(\mathbf{Y \beta}, \mathbf{X \alpha})]
$$

where a set of $M$ with orthogonality constraints

-   In high-dimensional space ( $p \approx n; p >n$ ), models are easily overfitted (too flexible). Hence, they might appear better than they really are. Hence, to deal with high dimensional data, we use less flexible models (i.e., those that are regularized or constrained), such as [Best Subset Selection], [Ridge regression], [Lasso regression], [Principal Component Regression], [Partial Least Squares]. But we still need to consider their issues

    -   Large $p$ always has collinearity; [Best Subset Selection] or [Lasso regression] can't guarantee the chosen model is a scientific one

    -   p-values are not considered under high-dimensional regressions

    -   Tuning parameter selection is utmost important for good predictive performance

    -   The test error tends to increase as the dimensionality of the problem increases ([Curse of dimensionality])
