# k-Nearest Neighbors

Imagine you move to a new town and want to guess whether a house will sell quickly. A reasonable first instinct is to look at the handful of most similar houses nearby (same size, same neighborhood, same age) and see what happened to them. If most of those sold fast, you predict yours will too. That everyday reasoning is exactly the k-Nearest Neighbors algorithm: predict by looking at the closest known examples and copying what they did.

This chapter builds that intuition into a precise method. You will learn how kNN turns "look at similar cases" into a formula for both classification and regression, why the single tuning parameter $k$ controls the tradeoff between fitting noise and oversmoothing, why the method breaks down when the number of features is large (the curse of dimensionality), and the practical steps (standardizing features, choosing $k$ by cross-validation, speeding up search) that make it work in practice. We close with a simulation that shows the bias-variance tradeoff and the decision boundary directly.

kNN fits near the start of the predictive-modeling story. It is the most direct embodiment of the principle that nearby points should have similar responses, and the same principle reappears, in more refined form, in the kernel and local-regression methods covered earlier. Because it is so transparent, kNN is a useful baseline: if an elaborate model cannot beat nearest neighbors on your problem, that is worth knowing.

## Intuition

k-Nearest Neighbors (kNN) is one of the simplest predictive methods, and it is a useful reference point for understanding more elaborate models. The idea is direct: to predict the response at a query point $x$, look at the $k$ training points closest to $x$ and let them vote (for classification) or average their responses (for regression).

> **Intuition:** A prediction is just a poll of the neighbors. For classification the neighbors vote on a class; for regression they report a number and you take the average. Everything else in this chapter is about who counts as a neighbor and how many to ask.

kNN sits in a particular corner of the prediction landscape:

- It is nonparametric. The method does not assume a fixed functional form (no linear predictor, no fixed number of coefficients). The complexity of the fitted function grows with the data.
- It is instance-based (also called memory-based). Rather than summarizing the data into a small set of parameters, kNN keeps the training set and consults it at prediction time.
- It does no training in the usual sense. There is no optimization step that fits parameters. All of the work happens when a prediction is requested, so the cost shifts from fit time to query time. This is sometimes called lazy learning.^[Lazy learning is the opposite of eager learning. An eager method (like linear regression) does its hard work up front, distilling the data into coefficients, and then predicts cheaply. A lazy method postpones all the work to query time and keeps the full training set around. Neither is better in general; the right choice depends on whether you predict often or rarely and how much data you can afford to store.]

Because it makes few assumptions, kNN can approximate very irregular decision boundaries. The price is that it needs enough data to populate the neighborhood of every query point, and it is sensitive to the scale of the features and to the number of features. Those tradeoffs are the subject of the rest of this chapter.

> **When to use this:** kNN is a strong choice when you have a moderate number of features, plenty of training examples relative to the number of features, and a decision boundary that may be irregular. It is a poor choice when features are high-dimensional, when predictions must be very fast on a large training set, or when many features are irrelevant (they add noise to the distance).

## The kNN Estimator

### Setup

We have training data $(x_1, y_1), \dots, (x_n, y_n)$ with $x_i \in \mathbb{R}^p$, meaning each example has $p$ features and a known response.^[Here $n$ is the number of training examples and $p$ is the number of features, so $x_i$ is the vector of features for the $i$th example and $y_i$ is its response. We use this notation throughout the book.] For a query point $x$, let $N_k(x)$ denote the index set of the $k$ training points nearest to $x$ under a chosen distance $d(\cdot, \cdot)$. Formally, $N_k(x)$ contains the indices of the $k$ smallest values among $d(x, x_1), \dots, d(x, x_n)$.

With the neighborhood $N_k(x)$ defined, the prediction rule depends only on what we do with the responses inside it. Classification and regression differ just in that final step.

### Classification

For a classification problem with classes $c \in \{1, \dots, C\}$, kNN estimates the class posterior probability by the fraction of neighbors in each class:

$$
\hat p(y = c \mid x) = \frac{1}{k} \sum_{i \in N_k(x)} \mathbb{1}(y_i = c).
$$

The predicted class is the majority vote among the neighbors,

$$
\hat y(x) = \arg\max_{c} \ \hat p(y = c \mid x) = \arg\max_{c} \ \sum_{i \in N_k(x)} \mathbb{1}(y_i = c).
$$

In words, the estimated probability that $x$ belongs to class $c$ is simply the share of its $k$ neighbors that carry label $c$, and the prediction is whichever class wins that local poll.

This is a plug-in approximation to the Bayes classifier, which assigns $x$ to $\arg\max_c p(y = c \mid x)$.^[The Bayes classifier is the best any method could do if the true conditional probabilities were known exactly. Its error rate, the Bayes error, is the irreducible floor set by the overlap between classes. kNN tries to estimate those conditional probabilities from local frequencies.] As $n \to \infty$ and $k \to \infty$ with $k/n \to 0$, the kNN classifier is consistent, and its error rate converges to the Bayes rate under mild conditions. The classic finite-$k$ result of Cover and Hart (1967) shows that the asymptotic error of the $1$-NN rule is at most twice the Bayes error.

> **Key idea:** Even the crudest version of kNN (a single nearest neighbor) is never more than twice as bad as the theoretical optimum in the large-sample limit. That is a remarkable guarantee for such a simple rule, and it explains why kNN remains a respected baseline.

### Regression

For regression, kNN predicts the average response over the neighborhood:

$$
\hat f(x) = \frac{1}{k} \sum_{i \in N_k(x)} y_i.
$$

This is a locally constant estimate: within any neighborhood, the fit is a flat line at the local average. It can be viewed as a kernel-weighted average with a uniform weight over the $k$ nearest points and zero weight elsewhere, which connects kNN to the kernel smoothing methods covered earlier in the book. The neighborhood is defined by a fixed count $k$ rather than a fixed bandwidth in distance, so the effective window adapts to local data density: it shrinks where data are dense and widens where data are sparse.

> **Note:** Kernel smoothing fixes a window width (a bandwidth $h$) and averages whatever falls inside. kNN fixes a count $k$ and lets the window width float. The two are mirror images of the same locally-weighted-averaging idea, and which one behaves better depends on how evenly your data are spread.

### Distance metrics

So far "nearest" has been left undefined. We now make it precise, because the choice of distance is what determines who the neighbors are. For $x, z \in \mathbb{R}^p$, common choices are:

- Euclidean ($L_2$): $d(x, z) = \left( \sum_{j=1}^p (x_j - z_j)^2 \right)^{1/2}$.
- Manhattan ($L_1$): $d(x, z) = \sum_{j=1}^p |x_j - z_j|$.
- Minkowski ($L_q$): $d(x, z) = \left( \sum_{j=1}^p |x_j - z_j|^q \right)^{1/q}$, which gives Manhattan for $q = 1$ and Euclidean for $q = 2$.
- Mahalanobis: $d(x, z) = \left( (x - z)^\top \Sigma^{-1} (x - z) \right)^{1/2}$, where $\Sigma$ is a covariance matrix (often estimated from the data). This metric accounts for correlation among features and for differences in their spread, so it is scale-invariant in a way the others are not.

Euclidean distance is the default and works well once features are on comparable scales. Manhattan distance can be more robust when features are noisy or when large coordinate differences should not be squared. Mahalanobis distance is the right tool when features are correlated or measured in very different units, since it folds standardization and decorrelation into the metric itself.

### Distance weighting

Plain kNN gives every neighbor an equal vote. A refinement weights neighbors by closeness, so that a point right next to $x$ counts more than one at the edge of the neighborhood. With weights $w_i \ge 0$, the weighted estimates are

$$
\hat p(y = c \mid x) = \frac{\sum_{i \in N_k(x)} w_i \, \mathbb{1}(y_i = c)}{\sum_{i \in N_k(x)} w_i},
\qquad
\hat f(x) = \frac{\sum_{i \in N_k(x)} w_i \, y_i}{\sum_{i \in N_k(x)} w_i}.
$$

A common weight is the inverse distance $w_i = 1 / d(x, x_i)$ (with a small offset to avoid division by zero), or a kernel of the distance such as $w_i = K\!\left( d(x, x_i) / h \right)$. Distance weighting reduces the sensitivity to the exact value of $k$, because far-off neighbors contribute little.

## Bias, Variance, and the Choice of k

The parameter $k$ controls the smoothness of the fit and is the main knob in kNN. Understanding how it works is the heart of using the method well, so we take it slowly: first the intuition for small and large $k$, then an exact derivation for regression, and then the special difficulty that high dimensions create.

### Small versus large k

With $k = 1$, the prediction at any training point is that point's own label, so the training error is zero, and the decision boundary is highly irregular, tracking individual points. This is low bias and high variance: the fit closely follows the data, but a small change in the training set can change the prediction substantially.

With large $k$, the prediction averages over a wide region. The boundary smooths out and becomes more stable across training sets, but it can miss real local structure. This is high bias and low variance. In the limit $k = n$, every query receives the same prediction (the overall majority class or the global mean), which is the most biased and least variable estimate possible.

> **Warning:** A training error of zero (which $k = 1$ always achieves) is not a sign of a good model. It means the model has memorized the training set, including its noise. The quantity to watch is error on held-out data, not training error.

### A regression derivation

The bias-variance behavior can be made precise for regression. Assume $y_i = f(x_i) + \varepsilon_i$ with $\mathbb{E}[\varepsilon_i] = 0$ and $\operatorname{Var}(\varepsilon_i) = \sigma^2$, errors independent. Treat the neighborhood points as fixed and let $\hat f(x) = \frac{1}{k} \sum_{i \in N_k(x)} y_i$. Then

$$
\mathbb{E}[\hat f(x)] = \frac{1}{k} \sum_{i \in N_k(x)} f(x_i),
\qquad
\operatorname{Var}(\hat f(x)) = \frac{\sigma^2}{k}.
$$

The expected squared error at $x$ decomposes as

$$
\mathbb{E}\left[ (y - \hat f(x))^2 \right]
= \sigma^2
+ \underbrace{\left( f(x) - \frac{1}{k} \sum_{i \in N_k(x)} f(x_i) \right)^2}_{\text{bias}^2}
+ \underbrace{\frac{\sigma^2}{k}}_{\text{variance}}.
$$

The variance term $\sigma^2 / k$ falls as $k$ grows: averaging more points reduces the noise. The squared bias term tends to grow with $k$, because larger neighborhoods reach farther from $x$, where $f(x_i)$ can differ from $f(x)$. The sum of these two terms is the familiar U-shaped curve in $k$, and the best $k$ balances them. In practice $k$ is chosen by cross-validation.^[Cross-validation splits the training data into folds, fits with each candidate $k$ on all-but-one fold, and measures error on the held-out fold, averaging over folds. The $k$ with the lowest average held-out error is selected. This estimates generalization error without touching the test set.] The effective number of parameters of a kNN fit is often described as roughly $n / k$, which makes the role of $k$ as a complexity control explicit.

> **Tip:** Read $n/k$ as the model's complexity. Small $k$ means many effective parameters (a flexible, wiggly fit); large $k$ means few (a rigid, smooth fit). This is the same dial as the degrees of freedom in a spline or the depth of a tree, just expressed through the neighborhood size.

### The curse of dimensionality

kNN relies on neighbors being genuinely close to the query point. In high dimensions this becomes hard to satisfy, for two related reasons.

Neighborhoods are not local. Suppose the $n$ training points are spread uniformly in the unit cube $[0,1]^p$. To capture a fraction $r$ of the data in a cubical neighborhood, the side length of that cube must be about

$$
e_p(r) = r^{1/p}.
$$

For example, to capture $r = 1\%$ of the data: in $p = 1$ dimension the neighborhood spans $0.01$ of each axis, but in $p = 10$ dimensions it spans $0.01^{1/10} \approx 0.63$ of each axis, and in $p = 100$ it spans $0.01^{1/100} \approx 0.95$. The "local" neighborhood is no longer local at all; it reaches across most of the range of every feature, so the estimate is no longer based on nearby points.

Distances concentrate. A second effect is that pairwise distances become nearly equal as $p$ grows. Consider points whose coordinates are independent. The squared Euclidean distance between two such points is a sum of $p$ independent per-coordinate contributions, so by a law-of-large-numbers argument its mean grows like $p$ while its standard deviation grows only like $\sqrt{p}$. The ratio of spread to typical distance therefore shrinks like $1/\sqrt{p}$,

$$
\frac{\text{sd of distance}}{\text{mean of distance}} \to 0 \quad \text{as } p \to \infty.
$$

When the nearest and farthest neighbors of a query are almost the same distance away, the ranking that kNN depends on carries little information. This is why kNN often performs poorly in high dimensions unless the data lie on a low-dimensional structure or the features are reduced first.

> **Key idea:** In high dimensions, "near" and "far" stop meaning much. The neighborhood needed to gather enough points spans almost the whole feature range, and all points sit at roughly the same distance. kNN's core assumption (nearby points are similar) quietly fails, so reduce dimensionality or drop irrelevant features before reaching for kNN.

Having seen how kNN behaves in theory, we turn to the choices that decide whether it works in practice.

## Practical Considerations

### Standardization

Because distance mixes all features together, a feature measured on a large numeric scale dominates the distance and crowds out the others. A variable in dollars (range in the thousands) will swamp a variable that is a proportion (range $0$ to $1$). Standardizing each feature to zero mean and unit variance (or scaling to a common range) puts the features on comparable footing before computing distances. This step is essential for kNN, not optional. The Mahalanobis distance addresses the same issue internally by dividing out the covariance.

> **Warning:** Compute the standardization (the means and standard deviations) on the training set only, then apply those same values to the test set. Standardizing the full dataset before splitting leaks information from the test set into training and inflates your performance estimate. The simulation below follows the correct order.

### Ties

Two kinds of ties arise. A tie in distance (two training points equally far from the query, with only one slot left in the neighborhood) is usually broken at random or by index order. A tie in the vote (equal class counts) can be broken at random, by choosing the class of the single nearest neighbor, or by using an odd $k$ in two-class problems to avoid the tie entirely. The `class::knn` function in R breaks voting ties at random, so results can vary run to run unless a seed is set.

### Computational cost

A naive query compares the point to all $n$ training points in $O(np)$ time, then selects the $k$ smallest distances. For $m$ queries this is $O(mnp)$, which is acceptable for small data but expensive at scale. Memory is also a concern, since the whole training set must be retained.

### kd-trees and ball-trees

Spatial data structures speed up neighbor search by avoiding a full scan. A kd-tree recursively partitions the space with axis-aligned splits; a query descends the tree and uses the partition geometry to prune branches that cannot contain a nearer neighbor. Average query cost can approach $O(\log n)$ in low dimensions. A ball-tree partitions the data into nested hyperspheres and prunes using the triangle inequality; it tends to hold up better than a kd-tree as $p$ grows. Both degrade toward the cost of brute force in high dimensions, which is another face of the curse of dimensionality. When exact search is too slow, approximate nearest-neighbor methods trade a small loss in accuracy for large speedups.

## Simulation

The ideas above are easier to trust once you see them. We illustrate the bias-variance tradeoff and the decision boundary on simulated two-class data in two dimensions. The two classes are drawn from Gaussian clusters with overlapping support, so the Bayes error is nonzero and no classifier can drive the test error to zero. Working in two dimensions lets us plot the decision boundary directly and keeps the curse of dimensionality at bay, so the bias-variance story comes through cleanly.

We first generate a training set and a larger test set, then standardize both using statistics computed from the training set alone, as the warning above recommends.

```{r knn-data}
set.seed(1)

# Two classes, each a mixture of Gaussian blobs in 2-D.
gen_data <- function(n_per_class) {
  # class 0 centers
  c0 <- rbind(c(-1, -1), c(1.5, -1.5))
  # class 1 centers
  c1 <- rbind(c(1, 1.2), c(-1.5, 1.5))
  draw <- function(centers, n, label) {
    idx <- sample(seq_len(nrow(centers)), n, replace = TRUE)
    pts <- centers[idx, , drop = FALSE] +
      matrix(rnorm(2 * n, sd = 0.9), ncol = 2)
    data.frame(x1 = pts[, 1], x2 = pts[, 2], y = factor(label))
  }
  rbind(
    draw(c0, n_per_class, 0),
    draw(c1, n_per_class, 1)
  )
}

train <- gen_data(150)
test  <- gen_data(500)

# Standardize features using the training mean and sd.
mu <- colMeans(train[, c("x1", "x2")])
sdv <- apply(train[, c("x1", "x2")], 2, sd)

scale_xy <- function(df) {
  out <- df
  out$x1 <- (df$x1 - mu["x1"]) / sdv["x1"]
  out$x2 <- (df$x2 - mu["x2"]) / sdv["x2"]
  out
}

train_s <- scale_xy(train)
test_s  <- scale_xy(test)
```

### Test error versus k

We fit kNN for a grid of $k$ and record the test error. Table \@ref(tab:knn-error-table) lists the test error for each candidate $k$, and the resulting curve is the U-shape predicted by the bias-variance decomposition: very small $k$ overfits (high variance), very large $k$ underfits (high bias).

```{r knn-error-table}
library(class)

ks <- c(1, 3, 5, 7, 9, 15, 25, 45, 75, 125, 201)

train_X <- train_s[, c("x1", "x2")]
test_X  <- test_s[, c("x1", "x2")]

set.seed(2)
err <- sapply(ks, function(k) {
  pred <- knn(train = train_X, test = test_X, cl = train_s$y, k = k)
  mean(pred != test_s$y)
})

err_tab <- data.frame(k = ks, test_error = round(err, 4))
knitr::kable(
  err_tab,
  caption = "kNN test error as a function of k. The error is high for very small k (high variance), reaches a minimum at a moderate k, and rises again for large k (high bias)."
)
```

Figure \@ref(fig:knn-error-plot) plots the same numbers, making the U-shape and the location of the best $k$ easy to read off.

```{r knn-error-plot, fig.width=6, fig.height=4, fig.cap="Test error rate of the kNN classifier as a function of the number of neighbors k. The dashed vertical line marks the k with the lowest test error; the curve is high at small k (high variance) and rises again at large k (high bias)."}
library(ggplot2)

best_k <- err_tab$k[which.min(err_tab$test_error)]

ggplot(err_tab, aes(x = k, y = test_error)) +
  geom_line() +
  geom_point() +
  geom_vline(xintercept = best_k, linetype = "dashed") +
  labs(
    x = "k (number of neighbors)",
    y = "Test error rate",
    title = "Bias-variance tradeoff in kNN"
  )
```

### Decision boundaries

To see the effect of $k$ directly, we color a fine grid of the feature space by the predicted class for a small $k$ and a large $k$. Figure \@ref(fig:knn-boundary) shows the two boundaries side by side: the small-$k$ boundary is jagged and follows individual points, while the large-$k$ boundary is smooth and stable.

```{r knn-boundary, fig.width=8, fig.height=4, fig.cap="kNN decision boundaries for a small k (left) and the cross-validated-best k (right) on the simulated two-class data. Shaded regions show the predicted class over a fine grid; points are the standardized training data. The small-k boundary is jagged (high variance) and the larger-k boundary is smooth (lower variance)."}
# Grid over the (standardized) feature space.
g1 <- seq(min(test_X$x1), max(test_X$x1), length.out = 200)
g2 <- seq(min(test_X$x2), max(test_X$x2), length.out = 200)
grid <- expand.grid(x1 = g1, x2 = g2)

boundary_df <- function(k) {
  set.seed(3)
  pred <- knn(train = train_X, test = grid, cl = train_s$y, k = k)
  data.frame(grid, pred = pred, k = paste0("k = ", k))
}

bd <- rbind(boundary_df(1), boundary_df(best_k))
pts <- data.frame(
  x1 = rep(train_s$x1, 2),
  x2 = rep(train_s$x2, 2),
  y  = rep(train_s$y, 2),
  k  = rep(c("k = 1", paste0("k = ", best_k)), each = nrow(train_s))
)

ggplot(bd, aes(x = x1, y = x2)) +
  geom_raster(aes(fill = pred), alpha = 0.35) +
  geom_point(data = pts, aes(color = y), size = 0.7) +
  facet_wrap(~ k) +
  labs(
    x = "x1 (standardized)",
    y = "x2 (standardized)",
    fill = "Predicted",
    color = "Class",
    title = "kNN decision boundary: small k versus large k"
  ) +
  coord_equal()
```

The left panel ($k = 1$) carves out small islands around individual training points, a signature of high variance. The right panel (the cross-validated-best $k$ from the error table) gives a smoother boundary that generalizes better to the test set. Notice that these two pictures are the same phenomenon as the two ends of the U-shaped error curve: the jagged boundary is what overfitting looks like geometrically, and the smooth one is the balance that cross-validation found.

## Practical Notes

The chapter's advice condenses into a short checklist. Keep these in mind whenever you apply kNN:

- Always standardize the features before computing distances, unless you use a scale-invariant metric like Mahalanobis.
- Choose $k$ by cross-validation rather than by a rule of thumb. The optimal $k$ depends on the noise level and the sample size.
- Use an odd $k$ for two-class problems to reduce voting ties.
- Reduce dimensionality when $p$ is large. Feature selection or a projection (for example PCA) can restore the locality that kNN needs.
- Mind the prediction cost. For large training sets, use a kd-tree or ball-tree (available through the `FNN` package in R), or an approximate nearest-neighbor library.
- kNN handles multiclass naturally through the majority vote, with no change to the algorithm.
- Class imbalance distorts the vote: a rare class is easily outvoted. Distance weighting, resampling, or per-class thresholds can help.

## Further Reading

- Cover, T. and Hart, P. (1967). Nearest neighbor pattern classification. Foundational analysis of the $1$-NN rule and its asymptotic error bound.
- Fix, E. and Hodges, J. L. (1951). Discriminatory analysis, nonparametric discrimination. The original technical report introducing nonparametric nearest-neighbor classification.
- Hastie, T., Tibshirani, R., and Friedman, J. (2009). The Elements of Statistical Learning, 2nd edition. See the treatment of kNN, the bias-variance decomposition, and the curse of dimensionality.
- James, G., Witten, D., Hastie, T., and Tibshirani, R. (2013). An Introduction to Statistical Learning. Accessible introduction to kNN classification and the choice of $k$.
