# Recurrent Neural Networks: LSTMs and GRUs {#sec-rnn}

```{r}
#| include: false
source("_common.R")
```

Most of the models in this book treat each observation as a fixed bundle of
features with no inherent order. A row is a row, and shuffling the rows changes
nothing. Sequences break that assumption. A sentence is not a bag of words, a
patient's chart is not a bag of lab values, and a stock price today depends on
where it was yesterday. The order carries the meaning, and a model that ignores
order throws away exactly the signal we care about.

Recurrent neural networks (RNNs) are the classical neural answer to this. The
idea is almost embarrassingly simple: read the sequence one step at a time, and
carry a running summary of everything seen so far. That summary, the *hidden
state*, is updated at each step by the same small network, so the model has a kind
of memory that it reuses across positions. A feedforward network (@sec-neural-networks)
sees the whole input at once; an RNN sees it as a story unfolding in time.

The simplicity hides a serious difficulty. Carrying memory across many steps means
multiplying many derivatives together when we train, and products of many numbers
either collapse to zero or explode to infinity. This is the vanishing and
exploding gradient problem, and it is the reason vanilla RNNs struggle to learn
dependencies that span more than a handful of steps. The long short-term memory
(LSTM) cell and the gated recurrent unit (GRU) were engineered specifically to keep
a gradient alive across long spans, and they did so well enough to power machine
translation, speech recognition, and time series forecasting for the better part of
a decade before attention and Transformers (@sec-transformers) took over.

This chapter builds the recurrence from intuition to mechanism. We define the
vanilla RNN precisely, derive backpropagation through time and the exact condition
under which its gradients vanish or explode, then show how the gating in LSTMs and
GRUs creates an additive path that lets gradients survive. We close with
sequence-to-sequence modelling, guidance on when to reach for a Transformer instead,
and a runnable base-R demonstration of an RNN forward pass together with a numerical
illustration of gradient vanishing.

::: {.callout-important title="Key idea"}
An RNN applies the *same* update at every time step and threads a hidden state
through the sequence. Sharing the update is what lets the model generalize across
positions; threading the state is what gives it memory. Both come at the cost of a
long multiplicative chain in the gradient.
:::

## The Recurrence

Let the input be a sequence of vectors $x_1, x_2, \dots, x_T$, with each
$x_t \in \mathbb{R}^{d}$. An RNN maintains a hidden state $h_t \in \mathbb{R}^{m}$
that summarizes the prefix $x_1, \dots, x_t$. The defining equation is the
recurrence

$$
h_t = f(h_{t-1}, x_t),
$$ {#eq-rnn-recurrence}

with $h_0$ a fixed or learned initial state (often the zero vector). The same
function $f$ is applied at every step, which is the parameter sharing that
distinguishes an RNN from simply stacking $T$ separate layers.

The classic "Elman" choice for $f$ is an affine map followed by a squashing
nonlinearity,

$$
h_t = \tanh\!\big(W_{h} h_{t-1} + W_{x} x_t + b_h\big),
$$ {#eq-rnn-elman}

where $W_h \in \mathbb{R}^{m \times m}$, $W_x \in \mathbb{R}^{m \times d}$, and
$b_h \in \mathbb{R}^{m}$ are shared across all time steps. Each step produces an
output through a read-out layer applied to the hidden state,

$$
\hat{y}_t = g\!\big(W_{y} h_t + b_y\big),
$$ {#eq-rnn-output}

where $g$ is the identity for regression or a softmax for classification over a
vocabulary. Depending on the task we use the output at every step (for example,
tagging each word), only the final output $\hat{y}_T$ (for example, classifying a
whole review), or a separate decoder (for sequence-to-sequence, below).

::: {.callout-note}
Unrolling the recurrence over $T$ steps gives a feedforward network that is $T$
layers deep but with every layer tied to the same weights. This "unrolled" view is
what makes ordinary backpropagation applicable, and it is also where the depth, and
hence the gradient trouble, comes from.
:::

The number of parameters does not grow with $T$: it is $m(m + d) + m$ for the
recurrent layer plus the read-out, regardless of how long the sequence is. The
computational cost of a forward pass is $O(T m^2)$ for the hidden updates, and
because each $h_t$ needs $h_{t-1}$, the steps must run in sequence and cannot be
parallelized along the time axis. That serial dependency is one of the practical
reasons Transformers eventually displaced RNNs.

## Backpropagation Through Time

Training minimizes a loss summed over time,

$$
\mathcal{L} = \sum_{t=1}^{T} \mathcal{L}_t(\hat{y}_t, y_t),
$$ {#eq-rnn-loss}

with respect to the shared parameters $\theta = \{W_h, W_x, b_h, W_y, b_y\}$.
Because the same weights appear at every step, the gradient with respect to, say,
$W_h$ collects contributions from all steps. Backpropagation through time (BPTT) is
ordinary backpropagation applied to the unrolled graph, with one twist: gradients
that flow into the shared weights are accumulated across all the time steps at which
those weights were used.

### Deriving the gradient

Write the pre-activation at step $t$ as $a_t = W_h h_{t-1} + W_x x_t + b_h$, so that
$h_t = \tanh(a_t)$. We want $\partial \mathcal{L} / \partial W_h$. By the chain rule,
summing over the steps where $W_h$ acts,

$$
\frac{\partial \mathcal{L}}{\partial W_h}
= \sum_{t=1}^{T} \frac{\partial \mathcal{L}_t}{\partial h_t}\,
  \frac{\partial h_t}{\partial W_h}\Bigg|_{\text{total}},
$$ {#eq-rnn-grad-sum}

where the "total" derivative of $h_t$ with respect to $W_h$ must account for the
fact that $h_t$ depends on $W_h$ both directly and through every earlier $h_k$.
Unrolling that dependence, the loss at a single step $t$ sends gradient back to
every earlier step $k \le t$ along the chain of hidden states. The contribution of
step $t$ to the gradient of the hidden state at step $k$ is

$$
\frac{\partial \mathcal{L}_t}{\partial h_k}
= \frac{\partial \mathcal{L}_t}{\partial h_t}\,
  \frac{\partial h_t}{\partial h_k},
\qquad k \le t,
$$ {#eq-rnn-deltat}

and the crucial factor is the Jacobian product that carries the signal from step
$t$ back to step $k$,

$$
\frac{\partial h_t}{\partial h_k}
= \prod_{i=k+1}^{t} \frac{\partial h_i}{\partial h_{i-1}}.
$$ {#eq-rnn-jacprod}

Each factor in this product is the per-step Jacobian. Differentiating
@eq-rnn-elman,

$$
\frac{\partial h_i}{\partial h_{i-1}}
= D_i \, W_h,
\qquad
D_i = \operatorname{diag}\!\big(1 - \tanh^2(a_i)\big),
$$ {#eq-rnn-jacstep}

where $D_i$ is the diagonal matrix of $\tanh'$ values, each entry in $(0, 1]$.
Substituting @eq-rnn-jacstep into @eq-rnn-jacprod shows that carrying gradient
across $t - k$ steps means multiplying $t - k$ copies of $D_i W_h$ together. The
behaviour of that long product is the whole story.

### The vanishing and exploding gradient

Take norms of the Jacobian product in @eq-rnn-jacprod. By submultiplicativity,

$$
\left\| \frac{\partial h_t}{\partial h_k} \right\|
\le \prod_{i=k+1}^{t} \| D_i \, W_h \|
\le \prod_{i=k+1}^{t} \| D_i \| \, \| W_h \|.
$$ {#eq-rnn-normbound}

Since every diagonal entry of $D_i$ lies in $(0, 1]$ for the $\tanh$ activation,
$\|D_i\| \le 1$ in the spectral norm, so the entire product is bounded by
$\|W_h\|^{\,t-k}$. Let $\gamma = \|W_h\|$ (equivalently its largest singular value,
and closely related to its largest eigenvalue $\lambda_{\max}$). Then

$$
\left\| \frac{\partial h_t}{\partial h_k} \right\| \lesssim \gamma^{\,t-k}.
$$ {#eq-rnn-gammapow}

The consequences are immediate and sharp:

- If $\gamma < 1$, the gradient norm decays geometrically as the gap $t - k$ grows.
  A signal from 50 steps back is multiplied by $\gamma^{50}$, which for
  $\gamma = 0.9$ is about $0.005$. The early steps receive essentially no learning
  signal. This is the **vanishing gradient** problem.
- If $\gamma > 1$, the same product grows geometrically and the gradient
  **explodes**, producing wild updates and numerical overflow.

A more precise statement uses the eigendecomposition. If $W_h = V \Lambda V^{-1}$
with eigenvalues $\lambda_j$, then the repeated product is dominated by
$\lambda_{\max}^{\,t-k}$. The borderline case $\lambda_{\max} = 1$ is the only one
that neither vanishes nor explodes, and it is measure-zero in practice, so a generic
deep recurrence sits on one side of the knife edge or the other.

::: {.callout-warning}
The diagonal factor $D_i$ makes things worse, not better, for vanishing.
Because $\tanh' = 1 - \tanh^2 \le 1$ and is strictly below $1$ whenever the unit is
even slightly saturated, the realized decay is usually faster than the $\gamma^{t-k}$
bound suggests. Saturated $\tanh$ units kill the gradient on their own.
:::

Exploding gradients have a cheap practical fix: **gradient clipping**, where the
gradient vector is rescaled to a maximum norm before each update,

$$
\tilde{g} = g \cdot \frac{\min(\tau, \|g\|)}{\|g\|},
$$ {#eq-rnn-clip}

for a threshold $\tau$. This caps the explosion without changing the gradient's
direction. Vanishing gradients are the harder problem, because there is no
information left to rescale once it has decayed. Solving vanishing required changing
the architecture itself, which is what gating does.

## The LSTM Cell

The LSTM (Hochreiter and Schmidhuber, 1997) introduces a second state, the *cell
state* $c_t$, that runs along the sequence with a nearly linear update. Where the
vanilla hidden state is transformed by a full nonlinear map at every step, the cell
state is mostly *added to*, and addition has a gradient of one. Three learned gates,
each a sigmoid-squashed linear function of the input and previous hidden state,
decide what to erase, what to write, and what to read out.

Writing $\sigma$ for the logistic sigmoid and $\odot$ for elementwise product, the
LSTM computes, at each step,

$$
\begin{aligned}
f_t &= \sigma\!\big(W_f [h_{t-1}, x_t] + b_f\big), \\
i_t &= \sigma\!\big(W_i [h_{t-1}, x_t] + b_i\big), \\
o_t &= \sigma\!\big(W_o [h_{t-1}, x_t] + b_o\big), \\
\tilde{c}_t &= \tanh\!\big(W_c [h_{t-1}, x_t] + b_c\big),
\end{aligned}
$$ {#eq-rnn-lstm-gates}

where $[h_{t-1}, x_t]$ denotes concatenation. The forget gate $f_t$, input gate
$i_t$, and output gate $o_t$ all take values in $(0,1)$ per coordinate. The state
updates are then

$$
c_t = f_t \odot c_{t-1} + i_t \odot \tilde{c}_t,
\qquad
h_t = o_t \odot \tanh(c_t).
$$ {#eq-rnn-lstm-state}

The cell update in @eq-rnn-lstm-state is the heart of the design. The previous cell
$c_{t-1}$ is scaled by the forget gate and then a new candidate is added in. When
the forget gate is near one and the input gate near zero, the cell simply copies
itself forward: $c_t \approx c_{t-1}$. That is a near-identity path along the time
axis, sometimes called the *constant error carousel*.

### Why gating preserves gradients

Repeat the Jacobian argument from @eq-rnn-jacprod, but now along the cell state.
The relevant per-step derivative is $\partial c_t / \partial c_{t-1}$. From the cell
update, treating the gates as approximately constant over the path (they depend on
$h_{t-1}$, but the dominant term is explicit),

$$
\frac{\partial c_t}{\partial c_{t-1}} \approx \operatorname{diag}(f_t).
$$ {#eq-rnn-lstm-jac}

The product that carries gradient from step $t$ back to step $k$ is therefore
approximately

$$
\frac{\partial c_t}{\partial c_k} \approx \prod_{i=k+1}^{t} \operatorname{diag}(f_i)
= \operatorname{diag}\!\left( \prod_{i=k+1}^{t} f_i \right).
$$ {#eq-rnn-lstm-jacprod}

Compare this with the vanilla product $\prod_i D_i W_h$ in @eq-rnn-jacprod. Two
things have changed. First, there is no repeated multiplication by a weight matrix
$W_h$, so the eigenvalues of $W_h$ no longer drive the decay. Second, the
multiplicative factors are the forget gates $f_i \in (0,1)$, which the network
*learns* to set. If a unit needs to remember something for a long time, gradient
descent can push the relevant forget gates toward one, making
$\prod_i f_i \approx 1$ and keeping the gradient alive across the whole span.

The vanilla RNN had no such control: its decay rate $\gamma = \|W_h\|$ is a single
number shared by all paths and all time scales. The LSTM makes the per-step memory
coefficient a learned, input-dependent gate, so different units can remember over
different horizons. This is precisely why initializing the forget-gate bias $b_f$ to
a positive value (so gates start near one) is a standard and effective trick: it
starts the network in a "remember by default" regime.

::: {.callout-tip}
The intuition that "addition has gradient one" is the same one behind residual
connections in very deep feedforward networks and behind the skip path in
Transformers. Whenever you see an additive update $z_t = z_{t-1} + (\text{something})$,
expect a clean gradient highway through it.
:::

## The GRU

The gated recurrent unit (Cho et al., 2014) is a streamlined gating scheme. It drops
the separate cell state and uses only two gates, an update gate $z_t$ and a reset
gate $r_t$:

$$
\begin{aligned}
z_t &= \sigma\!\big(W_z [h_{t-1}, x_t] + b_z\big), \\
r_t &= \sigma\!\big(W_r [h_{t-1}, x_t] + b_r\big), \\
\tilde{h}_t &= \tanh\!\big(W_h [\, r_t \odot h_{t-1}, \, x_t \,] + b_h\big),
\end{aligned}
$$ {#eq-rnn-gru-gates}

with the hidden state formed as a convex blend of the old state and the candidate,

$$
h_t = (1 - z_t) \odot h_{t-1} + z_t \odot \tilde{h}_t.
$$ {#eq-rnn-gru-state}

The update gate $z_t$ plays the combined role of the LSTM's forget and input gates:
when $z_t \approx 0$ the state is copied forward unchanged, giving the same
near-identity path and the same gradient-preserving product
$\prod_i (1 - z_i)$ as in @eq-rnn-lstm-jacprod. The reset gate $r_t$ controls how
much of the past enters the candidate, letting the unit "forget" the prior state
when computing a fresh proposal.

A GRU has fewer parameters than an LSTM (three weight blocks rather than four) and
is faster to train. Empirically the two perform comparably on most tasks; LSTMs
sometimes have an edge on very long dependencies because the separate cell state
gives them an extra degree of freedom, while GRUs are often preferred when data or
compute is limited. There is no universal winner, and the choice is usually settled
by validation performance.

```{r}
#| label: tbl-rnn-compare
#| tbl-cap: "Comparison of the three recurrent cells discussed in this chapter."
#| echo: false
cmp <- data.frame(
  Property   = c("States carried", "Number of gates", "Long-memory path",
                 "Gradient factor", "Relative parameters"),
  `Vanilla RNN` = c("h_t", "0", "none", "D_i W_h (uncontrolled)", "1x"),
  LSTM       = c("h_t and c_t", "3 (f, i, o)", "additive cell state",
                 "diag(f_i) (learned)", "~4x"),
  GRU        = c("h_t", "2 (z, r)", "convex blend",
                 "diag(1 - z_i) (learned)", "~3x"),
  check.names = FALSE
)
knitr::kable(cmp)
```

## Sequence to Sequence and When to Prefer Transformers

Many tasks map one sequence to another of a different length: translating a
sentence, summarizing a paragraph, transcribing speech. The classic RNN solution is
the *encoder-decoder* architecture. An encoder RNN reads the input
$x_1, \dots, x_T$ and compresses it into its final hidden state $h_T$, the
*context vector*. A decoder RNN is then initialized from that context and generates
the output $y_1, \dots, y_{T'}$ one token at a time, feeding each predicted token
back in as the next input,

$$
s_t = f_{\text{dec}}(s_{t-1}, y_{t-1}), \qquad s_0 = h_T.
$$ {#eq-rnn-seq2seq}

The weakness is the bottleneck: the entire input must be squeezed through the single
fixed-size vector $h_T$, and for long inputs the early content is lost (the same
vanishing problem, now framed as forgetting). The historical fix was *attention*,
which lets the decoder look back at all of the encoder's hidden states rather than
just the last one. Attention proved so effective that it was eventually used
*without* recurrence at all, giving the Transformer (@sec-transformers).

When should you still reach for an RNN? The honest answer in modern practice is:
rarely for large-scale language, where Transformers dominate, but RNNs remain a
sensible default in a few situations.

- **Streaming or online inference.** An RNN processes one step at a time with
  constant memory, which suits real-time signals (sensors, audio) where you cannot
  wait for the whole sequence.
- **Short sequences and small data.** With limited data, the smaller parameter count
  of a GRU can generalize better than a data-hungry Transformer.
- **Strict latency or memory budgets.** Self-attention costs $O(T^2)$ in time and
  memory; an RNN is $O(T)$. For very long sequences on constrained hardware, that
  difference matters.

Prefer a Transformer when you have long-range dependencies, enough data to feed it,
and hardware that rewards parallelism, which is to say most modern problems at
scale. The connection to probabilistic sequence models such as those in
@sec-structured-prediction is worth keeping in mind: an RNN with a softmax read-out
is a learned autoregressive factorization of $p(y_1, \dots, y_T)$, just with a
neural parameterization of each conditional.

::: {.callout-note}
Bidirectional RNNs run one recurrence forward and another backward, then
concatenate the two hidden states, so each position sees both past and future
context. They are useful for tagging and classification where the whole sequence is
available at once, but not for streaming generation, where the future is unknown.
:::

## A Worked Demonstration in R

We now make two things concrete in base R: the forward pass of a vanilla RNN, and
the numerical reality of vanishing gradients across many steps. No deep learning
library is needed; the recurrence is a short loop over matrix multiplications.

The first chunk runs a vanilla RNN forward over a random sequence and then measures
how the magnitude of the backpropagated gradient changes as we increase the number
of steps. We do this for two recurrent weight matrices, one scaled to have spectral
radius below one (vanishing) and one above one (exploding), to show both regimes
predicted by @eq-rnn-gammapow.

```{r}
#| label: fig-rnn-vanishing
#| fig-cap: "Norm of the backpropagated gradient of the final hidden state with respect to the first, in the linearized recurrence, as a function of the number of steps. Each recurrent matrix is a scaled orthogonal matrix, so its spectral norm equals its gain (and equals its spectral radius). With gain below one the gradient vanishes geometrically; above one it grows geometrically. The dashed lines are the proven bound gamma^(t-k) from @eq-rnn-gammapow with gamma the spectral norm."
set.seed(1)

m <- 8                          # hidden dimension
T_max <- 60                     # maximum number of steps

# A recurrent matrix with a prescribed gain: scale a random orthogonal matrix,
# so its spectral norm and spectral radius both equal `gain` exactly.
make_W <- function(m, gain) {
  Q <- qr.Q(qr(matrix(rnorm(m * m), m, m)))
  gain * Q
}

# Linearized backprop through time: with tanh' = 1 the Jacobian dh_T/dh_0 is the
# matrix product W^steps (eq-rnn-jacprod); we report its spectral norm. The
# linearization isolates the gain of W from the forward-pass saturation that, in
# a real tanh RNN, keeps every tanh' <= 1 and can therefore damp explosion.
grad_norm_through_time <- function(W, steps) {
  J <- diag(nrow(W))
  for (t in seq_len(steps)) J <- W %*% J
  norm(J, type = "2")
}

W_vanish  <- make_W(m, 0.7)
W_explode <- make_W(m, 1.3)

steps_grid <- seq(2, T_max, by = 2)
g_vanish  <- sapply(steps_grid, function(s) grad_norm_through_time(W_vanish,  s))
g_explode <- sapply(steps_grid, function(s) grad_norm_through_time(W_explode, s))

df <- rbind(
  data.frame(steps = steps_grid, gnorm = g_vanish,  regime = "gain 0.7"),
  data.frame(steps = steps_grid, gnorm = g_explode, regime = "gain 1.3")
)
bound <- rbind(
  data.frame(steps = steps_grid, gnorm = 0.7^steps_grid, regime = "gain 0.7"),
  data.frame(steps = steps_grid, gnorm = 1.3^steps_grid, regime = "gain 1.3")
)

library(ggplot2)
ggplot(df, aes(steps, gnorm, colour = regime)) +
  geom_line() +
  geom_point(size = 1.2) +
  geom_line(data = bound, aes(steps, gnorm, colour = regime),
            linetype = "dashed", linewidth = 0.5) +
  scale_y_log10() +
  labs(x = "number of recurrent steps (t - k)",
       y = "gradient norm  (log scale)",
       colour = NULL,
       title = "Vanishing and exploding gradients in a vanilla RNN") +
  scale_colour_viridis_d(end = 0.7)
```

The measured gradient norms (solid) sit exactly on the $\gamma^{t-k}$ reference
(dashed): below one they collapse toward machine zero, and above one they grow by
many orders of magnitude over sixty steps. This is @eq-rnn-gammapow made visible.
With the bounded tanh nonlinearity each factor $\operatorname{diag}(1 - \tanh^2)$
has spectral norm at most one, so in a real forward pass saturation can damp the
explosion (exploding gradients are most acute with unbounded activations), while
vanishing remains the dominant tanh failure mode that gating was invented to cure.

The second chunk contrasts the per-step gradient factor of a vanilla RNN with that
of an LSTM-style cell, showing how a forget gate near one keeps the cumulative
product from decaying.

```{r}
#| label: fig-rnn-gating
#| fig-cap: "Cumulative gradient factor across steps. The vanilla recurrence multiplies by the spectral radius each step and decays; the gated cell multiplies by a learned forget gate near one and keeps the gradient alive."
set.seed(7)
steps <- 1:60

# Vanilla: product of spectral radius (0.7) per step.
vanilla_factor <- 0.7^steps

# Gated: product of forget gates, each drawn close to one.
forget_gates <- plogis(rnorm(60, mean = 3, sd = 0.3))   # sigmoid of large bias
gated_factor <- cumprod(forget_gates)

dfg <- rbind(
  data.frame(steps = steps, factor = vanilla_factor, cell = "vanilla RNN (radius 0.7)"),
  data.frame(steps = steps, factor = gated_factor,   cell = "gated cell (forget ~ 0.95)")
)

ggplot(dfg, aes(steps, factor, colour = cell)) +
  geom_line() +
  scale_y_log10() +
  labs(x = "number of steps back in time",
       y = "cumulative gradient factor (log scale)",
       colour = NULL,
       title = "Gating preserves the gradient across long spans") +
  scale_colour_viridis_d(end = 0.7)
```

The gated curve stays within an order of magnitude of one across all sixty steps,
while the vanilla curve has long since vanished. A learnable per-step coefficient,
tuned toward one when long memory is needed, is the entire mechanism behind LSTMs
and GRUs.

### Training a real model

Fitting an LSTM or GRU on real data uses a deep learning backend. The following
sketch with `keras` trains a small LSTM for sequence classification; it is shown but
not run, since it requires a TensorFlow installation.

```{r}
#| eval: false
library(keras)

# x_train: integer-encoded sequences padded to length `maxlen`
# y_train: binary labels
model <- keras_model_sequential() |>
  layer_embedding(input_dim = vocab_size, output_dim = 32,
                  input_length = maxlen) |>
  layer_lstm(units = 64, dropout = 0.2, recurrent_dropout = 0.2) |>
  layer_dense(units = 1, activation = "sigmoid")

model |> compile(
  optimizer = "adam",
  loss      = "binary_crossentropy",
  metrics   = "accuracy"
)

# Gradient clipping (eq-rnn-clip) guards against exploding gradients;
# in keras pass clipnorm to the optimizer, e.g. optimizer_adam(clipnorm = 1.0).
history <- model |> fit(
  x_train, y_train,
  epochs = 10, batch_size = 64,
  validation_split = 0.2
)
```

The `torch` package offers an equivalent path with `nn_lstm()` or `nn_gru()` and a
manual training loop, which exposes the gradient clipping step explicitly via
`nn_utils_clip_grad_norm_()`. Either backend handles the BPTT bookkeeping
automatically; the concepts that matter for getting it to train, the forget-gate
bias initialization, gradient clipping, and choosing a sequence length that the
gating can actually bridge, are all direct consequences of the analysis in this
chapter.

## Connections and Further Reading

RNNs sit at the intersection of several threads in this book. As neural sequence
models they extend the feedforward networks of @sec-neural-networks with weight
sharing across time. Their probabilistic reading as autoregressive models of a
sequence links them to @sec-structured-prediction. The attention mechanism that
fixed the encoder-decoder bottleneck grew into the Transformers of
@sec-transformers, which now dominate the tasks RNNs once owned. And the practice of
learning compressed representations of inputs, central to the encoder half of a
sequence-to-sequence model, connects to the representation learning ideas in
@sec-autoencoders.

::: {.callout-important title="Takeaways"}
The recurrence $h_t = f(h_{t-1}, x_t)$ gives a neural model memory, but BPTT turns
that memory into a product of Jacobians whose norm scales like $\gamma^{t-k}$, so
gradients vanish when $\gamma < 1$ and explode when $\gamma > 1$. Clipping tames the
explosion; gating cures the vanishing by replacing the uncontrolled weight product
with a learned, near-identity additive path. LSTMs and GRUs implement that path with
forget and update gates. For long-range, large-scale sequence modelling, prefer
Transformers; for streaming, small data, or tight resource budgets, a GRU or LSTM is
still a strong choice.
:::
