# Attention and Transformers

```{r}
#| include: false
suppressPackageStartupMessages(library(tidyverse))
```

Imagine reading a long sentence and trying to remember the very first word by the
time you reach the last. If you could only hold one running summary in your head,
updated word by word, the early details would fade. That is essentially how a
recurrent neural network reads a sequence: one step at a time, carrying a single
hidden state forward and updating it as each new token^[A *token* is one unit of
the input sequence, for example a word or a sub-word piece in text, or a time step
in a time series.] arrives. This design made recurrent networks the default tool
for language and time series for many years, but it has structural limits.

Attention takes a different stance. Instead of squeezing the past into one running
summary, it lets every position in a sequence look directly at every other
position and decide, on the fly, which ones matter. The Transformer architecture
builds an entire model out of this idea, and it is the backbone of nearly every
modern large language model.

This chapter builds attention up from intuition. We start by spelling out why
recurrence runs into trouble, then derive *scaled dot-product attention* (the core
operation) and explain each piece, including the scaling factor that many
treatments leave unexplained. From there we assemble multi-head attention,
positional encoding, and the full Transformer block, distinguishing the encoder
from the decoder. To make the equations concrete, we implement attention from
scratch in base R on a five-word toy sentence and inspect the attention weights
directly. We close with the efficient variants that tame attention's cost on long
inputs and the link to BERT, covered in Chapter \@ref(bert).

> **Key idea:** Attention replaces "remember everything in one state" with "look
> up whatever is relevant, whenever you need it." Every position can reach every
> other position in a single step.

## Why Move Beyond Recurrence

A recurrent network reads a sequence $x_1, x_2, \dots, x_n$ and produces hidden
states $h_t = f(h_{t-1}, x_t)$. Three problems follow from this recurrence.

First, the computation is inherently sequential. To compute $h_t$ you need
$h_{t-1}$, so the work cannot be spread across the sequence in parallel. Long
sequences mean long chains of dependent operations, which makes training slow and
underuses modern hardware that is built for parallel matrix math.

Second, long-range dependencies are hard to capture. Information from $x_1$ that
matters for a prediction at position $n$ has to survive being repeatedly
transformed through every intermediate state. The path length between two
positions grows linearly with their distance, so distant tokens influence each
other only weakly.

Third, training suffers from vanishing and exploding gradients. Backpropagation
through time multiplies many Jacobian factors together. If their norms are below
one the gradient shrinks toward zero over long spans; if above one it blows up.
Gated units such as LSTMs and GRUs soften this problem but do not eliminate it.

Attention attacks all three at once. It connects any two positions with a path of
length one, so distant tokens interact in a single step. The core computation is a
set of matrix multiplications that run in parallel across positions. And because
there is no long multiplicative chain along the sequence axis, the gradient flow
between distant positions is far healthier.

> **Note:** Attention trades one problem for another. By comparing every position
> with every other position, it does work that grows with the *square* of the
> sequence length. That quadratic cost is the price of the short paths, and we
> return to it at the end of the chapter.

With the motivation in place, we can now define the operation itself.

## Scaled Dot-Product Attention

Attention answers a retrieval question, and a useful mental model is a soft
dictionary lookup. In an ordinary dictionary you match a search term exactly
against one key and get back its value. Attention softens that: it matches a search
term against *all* keys by degree, then returns a blend of the values weighted by
how well each key matched.

The three roles carry standard names. For each position we form a *query* that
describes what that position is looking for. Every position also exposes a *key*
that describes what it contains, and a *value* that holds the content to be
returned. A position compares its query against all keys, turns the comparisons
into weights, and reads out a weighted blend of the values.

> **Intuition:** Query = "what am I looking for?", key = "what do I offer?",
> value = "what do I hand over if chosen?". A pronoun's query might match the key
> of the noun it refers to, and pull in that noun's value.

Stack the queries, keys, and values for all $n$ positions into matrices
$Q \in \mathbb{R}^{n \times d_k}$, $K \in \mathbb{R}^{n \times d_k}$, and
$V \in \mathbb{R}^{n \times d_v}$. The similarity between query $i$ and key $j$ is
the dot product $q_i^\top k_j$. Collecting all pairs gives the score matrix
$QK^\top \in \mathbb{R}^{n \times n}$, whose entry $(i, j)$ is how strongly
position $i$ attends to position $j$ before normalization.

We turn each row of scores into a probability distribution with a row-wise
softmax, then use those weights to average the value vectors. The full operation
is

$$
\text{Attention}(Q, K, V) = \text{softmax}\!\left(\frac{QK^\top}{\sqrt{d_k}}\right) V .
$$

The softmax acts along each row, so row $i$ of the resulting weight matrix sums to
one and tells us how position $i$ distributes its attention over all positions.^[The
softmax of a row $s$ has entries $\text{softmax}(s)_j = e^{s_j} / \sum_m e^{s_m}$,
which are non-negative and sum to one, so each row is a genuine probability
distribution.] Multiplying by $V$ replaces each position with a convex combination
of value vectors, that is, a weighted average where the weights are non-negative
and sum to one.

### Why Divide by the Square Root of the Key Dimension

The scaling factor $1/\sqrt{d_k}$ is not cosmetic. Suppose the entries of a query
$q$ and a key $k$ are independent with mean zero and variance one. The dot product
is $q^\top k = \sum_{m=1}^{d_k} q_m k_m$. Each term has mean zero and variance one,
and the terms are independent, so the sum has mean zero and variance $d_k$. Its
standard deviation grows like $\sqrt{d_k}$.

When $d_k$ is large the raw scores spread over a wide range. Feeding large values
into a softmax pushes it toward a one-hot distribution where almost all weight
lands on a single position. In that regime the softmax gradient is tiny, which
stalls learning. Dividing by $\sqrt{d_k}$ rescales the scores back to roughly unit
variance, keeping the softmax in a responsive range.

> **Why it matters:** Without the scaling, simply making the model wider (larger
> $d_k$) would silently saturate the softmax and freeze learning. The
> $1/\sqrt{d_k}$ factor decouples the temperature of the attention from the width
> of the representation.

### Computational Cost

The score matrix $QK^\top$ has $n^2$ entries, and each entry is a dot product of
length $d_k$, giving $O(n^2 d_k)$ work to form the scores. Applying the weights to
$V$ costs $O(n^2 d_v)$. Writing $d$ for the representation width, the cost is
$O(n^2 d)$ in time and $O(n^2)$ in memory for the attention matrix. The quadratic
dependence on sequence length $n$ is the main scaling concern, and it motivates
the efficient variants discussed later.

## Multi-Head Attention

A single attention computation forms one set of weights and one blend of values.
That forces every relationship in the sequence through a single similarity
pattern, which is limiting: a sentence has many kinds of relationships at once
(grammatical agreement, reference, topic) and one pattern cannot capture all of
them. Multi-head attention runs several attention computations in parallel, each
with its own learned projections, so different heads can specialize. One head might
track local word order while another links a pronoun to its referent.

Given input representations $X \in \mathbb{R}^{n \times d}$, head $i$ uses learned
projection matrices $W_i^Q, W_i^K \in \mathbb{R}^{d \times d_k}$ and
$W_i^V \in \mathbb{R}^{d \times d_v}$ to build its own queries, keys, and values,

$$
\text{head}_i = \text{Attention}(X W_i^Q,\; X W_i^K,\; X W_i^V) .
$$

The $h$ head outputs are concatenated along the feature axis and mixed by an output
projection $W^O \in \mathbb{R}^{h d_v \times d}$,

$$
\text{MultiHead}(X) = \text{Concat}(\text{head}_1, \dots, \text{head}_h)\, W^O .
$$

A common choice is $d_k = d_v = d / h$, so the total work of multi-head attention
is about the same as a single full-width head, but spread across specialized
subspaces.

> **Tip:** Multi-head attention is not "more compute for more power." With
> $d_k = d / h$ the heads split the existing width among themselves, so you get
> several specialized views at roughly the cost of one wide view.

### Self-Attention and Cross-Attention

In *self-attention* the queries, keys, and values all come from the same sequence.
Every position attends to every position in its own sequence, which is how a model
builds context-aware representations of its input.

In *cross-attention* the queries come from one sequence while the keys and values
come from another. A translation decoder, for example, forms queries from the
target sentence it is generating and reads keys and values from the encoded source
sentence. Cross-attention is how one sequence conditions on another.

## Positional Encoding

There is a catch hiding in everything so far: attention has no built-in notion of
order. If you permute the rows of $Q$, $K$, and $V$ together, the set of outputs is
permuted the same way and otherwise unchanged. The operation is *permutation
equivariant*,^[Permutation equivariant means that reordering the inputs reorders
the outputs in the same way without changing their values. Contrast with
permutation *invariant*, where reordering leaves the output completely unchanged.]
so on its own it treats a sentence as a bag of tokens. To attention, "dog bites
man" and "man bites dog" look identical, which is clearly wrong for language. Word
order has to be injected explicitly.

The original Transformer adds a fixed *positional encoding* to the input
embeddings. For position $\text{pos}$ and embedding dimension index $i$, the
encoding uses sines and cosines at geometrically spaced frequencies,

$$
PE_{(\text{pos},\, 2i)} = \sin\!\left(\frac{\text{pos}}{10000^{\,2i/d}}\right),
\qquad
PE_{(\text{pos},\, 2i+1)} = \cos\!\left(\frac{\text{pos}}{10000^{\,2i/d}}\right).
$$

Each dimension is a sinusoid with its own wavelength, ranging from short to very
long across the $d$ dimensions. This gives every position a distinct signature, and
because shifting position by a fixed offset corresponds to a linear rotation of
these sinusoids, the model can express relative position with linear operations.
Adding $PE$ to the token embeddings lets attention see where each token sits without
changing the attention mechanism itself. Learned positional embeddings are a common
alternative that replaces the fixed formula with a trainable table.

## The Transformer Block

The Transformer, introduced in "Attention Is All You Need" (Vaswani et al., 2017),
stacks identical blocks built from a small set of pieces.

Each block has two sublayers: a multi-head attention sublayer and a position-wise
feed-forward sublayer. Around each sublayer sits a *residual connection* followed by
*layer normalization*. If a sublayer computes a function $\text{Sublayer}(\cdot)$,
the block produces

$$
\text{LayerNorm}\big(x + \text{Sublayer}(x)\big).
$$

The residual connection^[A *residual connection* adds the sublayer's input back to
its output, so the sublayer only has to learn a correction to the identity rather
than the whole mapping. This is the same trick that made very deep image networks
trainable.] lets the input pass through unchanged when the sublayer has little to
add, which keeps gradients flowing through deep stacks. Layer normalization
standardizes each position's feature vector to stabilize training.

> **Note:** The two ingredients play complementary roles. Residual connections
> keep a clean gradient highway through the depth of the network; layer
> normalization keeps the scale of activations stable so that highway does not
> drift. Together they are what let Transformers stack dozens of blocks.

The *position-wise feed-forward* sublayer applies the same two-layer network to each
position independently,

$$
\text{FFN}(x) = \max(0,\, x W_1 + b_1)\, W_2 + b_2 ,
$$

with an inner width several times larger than $d$. It adds nonlinear capacity on top
of the mixing that attention provides.

### Encoder and Decoder

The encoder is a stack of these blocks using self-attention, where every position
attends freely in both directions. It maps an input sequence to a sequence of
context-aware representations.

The decoder adds a cross-attention sublayer that reads the encoder output, and it
uses *causal masking* in its self-attention. Causal masking sets the scores for
future positions to $-\infty$ before the softmax, so position $i$ can attend only to
positions $1, \dots, i$.^[Why $-\infty$ and not zero? The mask is applied to the
scores *before* the softmax. Exponentiating $-\infty$ gives exactly zero weight,
while a raw score of zero would still receive non-trivial weight after the softmax.]
This keeps the model from peeking at tokens it has not yet generated, which is what
allows the decoder to generate one token at a time at inference while still training
in parallel.

> **Key idea:** Encoder self-attention is bidirectional (every token sees the whole
> sequence), decoder self-attention is causal (every token sees only the past).
> That single difference is what separates an understanding model like BERT from a
> generative model like GPT.

## A From-Scratch Implementation in Base R

The code below implements scaled dot-product attention and a small multi-head
version using only base R matrix algebra on a toy sequence of five tokens. No deep
learning framework is involved. The point is to make the equations concrete and to
inspect the attention weights directly.

> **When to use this:** This implementation is for learning, not production. Real
> models use optimized framework layers (shown at the end of the chapter). Read it
> to see exactly how the formula turns into matrix operations, then reach for a
> framework when you build something real.

We begin with the two building blocks: a numerically safe row-wise softmax, and the
attention function itself. The softmax subtracts each row's maximum before
exponentiating, a standard trick that prevents overflow without changing the
result.^[Subtracting a constant $c$ from every score in a row leaves the softmax
unchanged, because the factor $e^{-c}$ cancels between numerator and denominator.
Choosing $c$ to be the row maximum keeps every exponent at most zero, so $e^{\cdot}$
never overflows.]

```{r transformers-softmax}
# Row-wise softmax with the standard max-subtraction trick for numerical safety.
row_softmax <- function(M) {
    M <- M - apply(M, 1, max)        # subtract row max before exponentiating
    E <- exp(M)
    E / rowSums(E)
}

# Scaled dot-product attention.
# Q: n_q x d_k, K: n_k x d_k, V: n_k x d_v
# Returns the output matrix and the attention weight matrix.
scaled_dot_product_attention <- function(Q, K, V, mask = NULL) {
    d_k <- ncol(K)
    scores <- (Q %*% t(K)) / sqrt(d_k)   # n_q x n_k raw similarities, scaled
    if (!is.null(mask)) {
        scores[mask] <- -Inf             # masked positions get -Inf before softmax
    }
    weights <- row_softmax(scores)       # each row sums to 1
    list(output = weights %*% V, weights = weights)
}
```

We build a tiny sequence. Five positions, each with a four-dimensional embedding,
plus a sinusoidal positional encoding so order is represented.

```{r transformers-toy-data}
set.seed(2017)

n <- 5    # sequence length
d <- 4    # embedding dimension

tokens <- c("the", "cat", "sat", "on", "mat")

# Random token embeddings standing in for a learned embedding table.
X_tok <- matrix(rnorm(n * d), nrow = n, ncol = d)

# Sinusoidal positional encoding, following the formula in the text.
positional_encoding <- function(n, d) {
    PE <- matrix(0, nrow = n, ncol = d)
    pos <- 0:(n - 1)
    for (i in 0:(d / 2 - 1)) {
        denom <- 10000^(2 * i / d)
        PE[, 2 * i + 1] <- sin(pos / denom)
        PE[, 2 * i + 2] <- cos(pos / denom)
    }
    PE
}

PE <- positional_encoding(n, d)
X  <- X_tok + PE          # inputs carry both content and position
rownames(X) <- tokens
round(X, 3)
```

Single-head self-attention. The queries, keys, and values are linear projections of
the same input $X$, which is what makes this self-attention.

```{r transformers-single-head}
d_k <- 4

W_Q <- matrix(rnorm(d * d_k), d, d_k)
W_K <- matrix(rnorm(d * d_k), d, d_k)
W_V <- matrix(rnorm(d * d_k), d, d_k)

Q <- X %*% W_Q
K <- X %*% W_K
V <- X %*% W_V

attn <- scaled_dot_product_attention(Q, K, V)

A <- attn$weights
rownames(A) <- tokens
colnames(A) <- tokens
round(A, 3)
```

Each row of `A` is the attention distribution for one query token over all key
tokens. Reading across a row tells you where that token "looks." With random
projections these patterns are arbitrary, a trained model would learn meaningful
ones, but the mechanics are identical. By construction every row is a probability
distribution, so the rows sum to one. We check this explicitly rather than trusting
it, since a single coding slip in the softmax would break the guarantee.

```{r transformers-rowsums}
rowsum_check <- rowSums(A)
print(round(rowsum_check, 8))
stopifnot(all(abs(rowsum_check - 1) < 1e-8))
cat("All attention rows sum to 1.\n")
```

The output vectors are the attention-weighted blends of the value vectors, one row
per token. @tbl-transformers-output-table lists these output vectors for
the toy sequence.

```{r tbl-transformers-output-table}
out <- attn$output
rownames(out) <- tokens
colnames(out) <- paste0("dim", seq_len(ncol(out)))

knitr::kable(
    round(out, 3),
    caption = "Self-attention output vectors, one row per input token."
)
```

A heatmap makes the attention pattern easy to read. Figure
@fig-transformers-heatmap shows the attention weight matrix, where brighter
cells mean a query token (row) places more weight on a key token (column).

```{r fig-transformers-heatmap, fig.cap="Attention weight matrix from single-head self-attention on the toy sequence. Each row is a probability distribution over the key tokens.", fig.width=6, fig.height=5}
library(ggplot2)

A_df <- expand.grid(query = tokens, key = tokens)
A_df$weight <- as.vector(A)   # column-major matches expand.grid ordering

# Keep token order on the axes instead of alphabetical sorting.
A_df$query <- factor(A_df$query, levels = rev(tokens))
A_df$key   <- factor(A_df$key,   levels = tokens)

ggplot(A_df, aes(x = key, y = query, fill = weight)) +
    geom_tile(color = "white", linewidth = 0.5) +
    geom_text(aes(label = sprintf("%.2f", weight)), size = 3) +
    scale_fill_viridis_c(limits = c(0, 1)) +
    labs(
        x = "Key token (attended to)",
        y = "Query token (attending)",
        fill = "Weight",
        title = "Scaled dot-product attention weights"
    ) +
    coord_fixed()
```

Now a small multi-head version. We run two heads with their own projections, each of
width $d_k = 2$, concatenate the head outputs, and mix them with an output
projection. This mirrors the multi-head formula from earlier.

```{r transformers-multi-head}
multi_head_attention <- function(X, h, d_k, mask = NULL) {
    d <- ncol(X)
    heads <- vector("list", h)
    weights_list <- vector("list", h)
    for (i in seq_len(h)) {
        WQ <- matrix(rnorm(d * d_k), d, d_k)
        WK <- matrix(rnorm(d * d_k), d, d_k)
        WV <- matrix(rnorm(d * d_k), d, d_k)
        a  <- scaled_dot_product_attention(X %*% WQ, X %*% WK, X %*% WV, mask)
        heads[[i]] <- a$output
        weights_list[[i]] <- a$weights
    }
    concat <- do.call(cbind, heads)          # n x (h * d_k)
    W_O <- matrix(rnorm((h * d_k) * d), h * d_k, d)
    list(output = concat %*% W_O, weights = weights_list)
}

set.seed(7)
mha <- multi_head_attention(X, h = 2, d_k = 2)

cat("Head 1 attention rows sum to 1:",
    all(abs(rowSums(mha$weights[[1]]) - 1) < 1e-8), "\n")
cat("Head 2 attention rows sum to 1:",
    all(abs(rowSums(mha$weights[[2]]) - 1) < 1e-8), "\n")

mh_out <- mha$output
rownames(mh_out) <- tokens
colnames(mh_out) <- paste0("dim", seq_len(ncol(mh_out)))
round(mh_out, 3)
```

We can also see causal masking in action. The mask blocks every position from
attending to later positions, which produces a lower-triangular weight matrix.

```{r transformers-causal-mask}
# TRUE where a query (row) should NOT see a key (column): strictly future keys.
causal_mask <- upper.tri(matrix(0, n, n))

masked <- scaled_dot_product_attention(Q, K, V, mask = causal_mask)
Am <- masked$weights
rownames(Am) <- tokens
colnames(Am) <- tokens
round(Am, 3)
```

The upper triangle is exactly zero, so each token attends only to itself and the
tokens before it, and each row still sums to one. The first token can attend only
to itself, which is why its row puts all weight on `the`. This lower-triangular
pattern is precisely the constraint a generative decoder needs.

To recap the implementation: a row-wise softmax plus three matrix products is the
entire core of attention, multi-head attention is the same function run several
times with different projections and stitched together, and a single boolean mask
turns bidirectional attention into causal attention. Everything else in a
Transformer (residual connections, normalization, the feed-forward sublayer) wraps
around this core.

## Efficient and Variant Attention

We flagged the quadratic cost at the start of the chapter; here is how the field
addresses it. The $O(n^2)$ cost of full attention becomes the binding constraint
for long sequences: doubling the input quadruples the attention work and memory.
Two broad families of variants reduce it.

*Sparse attention* restricts each query to a subset of keys instead of all of them.
Patterns include local windows, strided or dilated connections, and a few global
tokens that every position can reach. Limiting the number of key comparisons per
query brings the cost below quadratic while preserving the ability to reach distant
context through the global tokens.

*Linear attention* rewrites the computation to avoid forming the full $n \times n$
matrix. By replacing the softmax with a feature map $\phi(\cdot)$ and reordering the
products as $\phi(Q)\,(\phi(K)^\top V)$, the cost becomes linear in $n$. The
approximation trades some expressiveness for scalability on long inputs.

> **Warning:** Sub-quadratic attention is an approximation, not a free lunch. Each
> variant assumes the important interactions follow a particular structure (local
> windows, a low-rank pattern, a chosen feature map). When that assumption fits the
> data it scales beautifully; when it does not, accuracy drops. Match the variant
> to the structure you actually expect in your sequences.

## Connection to BERT

The encoder stack described here is exactly what BERT is built from. BERT uses the
bidirectional self-attention encoder with no causal mask, so every token attends to
its full left and right context, and it is pretrained on large text corpora before
being fine-tuned for downstream prediction tasks. The Chapter \@ref(bert) on BERT
develops that pretraining and fine-tuning story in detail and shows how to use the
resulting contextual embeddings as features for prediction.

When you reach for a deep learning framework, you will not write the matrix algebra
by hand. Attention and Transformer blocks are available as ready-made layers, and
your job is to wire them together. The snippet below sketches what a single
Transformer block looks like in Keras, and you should be able to recognize every
piece from this chapter: a multi-head attention layer, a residual add, layer
normalization, the feed-forward sublayer, and a second residual-plus-norm. It is
illustrative only and is not run here, since it requires the Keras/TensorFlow
backend, which this book does not build against.

```{r transformers-keras-illustration, eval=FALSE}
# Illustrative only. Requires TensorFlow/Keras, which is not run in this book.
library(keras)

inputs <- layer_input(shape = c(seq_len, d_model))

attn_out <- layer_multi_head_attention(
    num_heads = 8,
    key_dim   = d_model %/% 8
)(inputs, inputs)                              # self-attention: query = value

x <- layer_add(list(inputs, attn_out)) %>%     # residual connection
    layer_layer_normalization()

ffn <- x %>%
    layer_dense(units = 4 * d_model, activation = "relu") %>%
    layer_dense(units = d_model)

block_out <- layer_add(list(x, ffn)) %>%       # second residual connection
    layer_layer_normalization()

model <- keras_model(inputs, block_out)
```

## Further Reading

Vaswani, A., Shazeer, N., Parmar, N., Uszkoreit, J., Jones, L., Gomez, A. N.,
Kaiser, L., and Polosukhin, I. (2017). "Attention Is All You Need." *Advances in
Neural Information Processing Systems*. This paper introduced the Transformer and
scaled dot-product attention.

Bahdanau, D., Cho, K., and Bengio, Y. (2015). "Neural Machine Translation by Jointly
Learning to Align and Translate." *International Conference on Learning
Representations*. This earlier work introduced the attention mechanism in the
context of recurrent encoder-decoder translation.
