# BERT {#sec-bert}

```{r}
#| include: false
suppressPackageStartupMessages(library(tidyverse))
```

Most of the methods in this book expect their inputs as numbers: a row of
predictors, a design matrix, a feature vector. Text does not arrive in that form.
A product review, a clinical note, or a news headline is just a string of
characters, and before any classifier or regressor can use it we need a way to
turn that string into numbers that capture what it *means*. This chapter is about
one of the most effective tools for doing exactly that.

BERT (Bidirectional Encoder Representations from Transformers) is a family of
pretrained language models that produce *contextual* representations of text.
Introduced by Google in 2018, BERT became one of the foundational tools for
modern natural language processing (NLP)^[Natural language processing is the
branch of machine learning concerned with getting computers to work with human
language: classifying it, extracting information from it, translating it, or
generating it.] and remains a workhorse for prediction tasks where the input is
text: document classification, sentiment analysis, named-entity recognition, and
question answering.

For a prediction-focused course, the key idea is this: BERT lets you turn raw
text into high-quality numeric features (embeddings) that you can feed into a
downstream classifier or regressor, *or* fine-tune the whole network end-to-end
for your specific task. Either way, you inherit knowledge learned from a massive
text corpus without having to train a language model yourself.

> **Key idea:** BERT is a reusable "text-to-numbers" engine. Someone else paid
> the enormous cost of teaching it language; you reuse the result, either as
> off-the-shelf features or as a starting point you nudge toward your own task.

By the end of this chapter you will understand what makes BERT's representations
"contextual," the core mechanics of the Transformer encoder it is built from, how
it is trained before you ever see it, and how to put it to work from both R and
Python. We begin with the idea that motivated the whole approach: the difference
between a fixed word vector and one that pays attention to context.

## From Static to Contextual Embeddings

Earlier word-embedding methods such as word2vec and GloVe map each word
type to a single fixed vector (@sec-embeddings-vector-search). The word "bank" gets *one* vector regardless of
whether the sentence is about a *river bank* or a *savings bank*. These are
static embeddings: useful, but blind to context.

BERT instead produces contextual embeddings. The vector it assigns to a token
depends on the entire surrounding sentence, so "bank" in "river bank" and "bank"
in "deposit money in the bank" receive different representations. This sensitivity
to context is what makes BERT dramatically more powerful than static embeddings on
tasks that hinge on meaning in context.

The word *bidirectional* in the name is central. Earlier contextual models (e.g.,
left-to-right language models) read text in one direction. BERT conditions each
token's representation on words to both its left and its right
simultaneously, giving a fuller picture of context.

> **Intuition:** To guess the missing word in "I deposited my check at the
> \_\_\_," reading only left-to-right ("I deposited my check at the") already
> points strongly toward "bank." But for "the \_\_\_ overflowed after three days
> of rain," you need the words *after* the blank to tell that this is a river,
> not a vault. Seeing both sides at once is what lets BERT resolve meaning that a
> one-directional reader would miss.

Now that we know *what* BERT produces, let us look at the machinery that produces
it.

## The Transformer Encoder

BERT is built entirely from the encoder half of the Transformer architecture
(@sec-transformers). You do not need the full mathematical machinery to use BERT,
but the following intuitions are worth carrying.

### Self-Attention

The core operation is self-attention. For each token, the model computes how
much it should "attend to" every other token in the sequence, then forms a
weighted combination of their representations. Concretely, each token is projected
into a query ($Q$), a key ($K$), and a value ($V$) vector, and
attention weights are computed as a scaled dot product:

$$
\text{Attention}(Q, K, V) = \text{softmax}\!\left(\frac{QK^\top}{\sqrt{d_k}}\right) V
$$

where $d_k$ is the dimension of the key vectors and the $\sqrt{d_k}$ term keeps the
dot products from growing too large. The softmax turns the similarities between a
query and all keys into a probability distribution; the output is the
correspondingly weighted average of the values. The practical effect: a token can
pull in information from any other token in the sentence, no matter how far away,
in a single step.

> **Intuition:** A useful mental model is a small lookup or search. Each token
> issues a query ("which other words should I be paying attention to?"), every
> token advertises a key ("here is what I am about"), and the dot product between
> query and key measures how well they match. The softmax converts those match
> scores into weights that sum to one, and the token then collects a weighted mix
> of the values. The word "it" can thereby reach back and grab information from
> the noun it refers to, even if that noun sat ten words earlier.

### Multi-Head Attention and Positional Encodings

A single round of attention is powerful but limited: it produces one pattern of
"who attends to whom." Two refinements address that limitation and a structural
blind spot of attention.

- Multi-head attention. Rather than computing attention once, the model runs
  several attention "heads" in parallel, each with its own learned projections.
  Different heads can specialize, e.g., one tracking syntactic dependencies,
  another tracking coreference. Their outputs are concatenated and recombined.
- Positional encodings. Self-attention is permutation-invariant: on its own it
  has no notion of word order. BERT adds (learned) positional embeddings to the
  token embeddings so the model knows which token is first, second, and so on.

Stacking many such attention-plus-feed-forward layers (12 layers in BERT-base, 24
in BERT-large) yields a deep network whose hidden states are the contextual
representations we use downstream.

> **Note:** You do not have to memorize these mechanics to use BERT effectively.
> What matters for practice is the consequence: every token's output vector is a
> context-aware summary that has already had the chance to look at every other
> token in the input.

## Pretraining Objectives

An architecture alone learns nothing; the network only becomes useful after it is
trained. BERT is pretrained on a large unlabeled corpus (originally English
Wikipedia and the BooksCorpus) using two self-supervised objectives
(@sec-self-supervised-learning). "Self-supervised" means the labels come from the
text itself, so no manual annotation is required.

> **Why this matters:** Labeled text is expensive, but raw text is essentially
> free and almost unlimited. Self-supervision lets BERT learn from billions of
> words without anyone hand-labeling a single one, which is precisely why the
> pretrained model carries so much general linguistic knowledge.

### Masked Language Modeling (MLM)

A random subset (roughly 15%) of input tokens is replaced with a special `[MASK]`
token, and the model is trained to predict the original tokens from context. Because
it must use both left and right context to fill in the blanks, MLM is what gives
BERT its deep bidirectional understanding. This is analogous to a cloze
("fill-in-the-blank") test.

### Next Sentence Prediction (NSP)

BERT is also given pairs of sentences and trained to predict whether the second
sentence actually followed the first in the original text or was a random
distractor. NSP was intended to help with tasks that involve relationships between
sentences (e.g., question answering, natural-language inference). Later work (notably
RoBERTa) found NSP contributes little, and many successors drop it. It is still
worth knowing as part of the original recipe.

## Tokenization and Special Tokens

BERT does not operate on whole words. It uses WordPiece tokenization, which
splits text into subword units drawn from a fixed vocabulary. Rare or unseen words
are broken into known pieces (e.g., "playing" $\rightarrow$ "play" + "##ing"). This
keeps the vocabulary manageable while avoiding out-of-vocabulary tokens entirely.

Beyond the ordinary subword tokens, two special tokens do bookkeeping that you
will see referenced constantly in code and tutorials:

- `[CLS]` is prepended to every input. Its final-layer hidden state is used as
  an aggregate, sentence-level representation for classification tasks.
- `[SEP]` separates segments (e.g., the two sentences in a pair, or a question
  and a passage).

> **Tip:** When you see code grab "index 0" of the hidden states (as in the
> Python example below), that is the `[CLS]` vector being used as a single
> fixed-length summary of the whole input. It is the most common way to reduce a
> variable-length sentence to one feature vector.

## Pretrain then Fine-Tune

The defining paradigm is pretrain $\rightarrow$ fine-tune:

1. Pretrain once, expensively, on a huge corpus (done by the model's authors;
   you download the result).
2. Fine-tune cheaply on your labeled task by adding a small task-specific head
   (often a single linear layer on top of the `[CLS]` representation) and training
   the whole network for a few epochs on your data.

This transfer-learning recipe (@sec-transfer-multitask-learning) is why BERT is so
practical: the heavy lifting of learning language is amortized across everyone who
reuses the pretrained weights.
Fine-tuning typically needs only modest amounts of labeled data and compute.^[The
original pretraining ran for days on many specialized accelerators. Fine-tuning a
task-specific head, by contrast, is often a matter of minutes to a few hours on a
single GPU, and sometimes feasible on a CPU for small datasets.]

> **When to use this:** Reach for the pretrain-then-fine-tune recipe when you
> have a text-prediction task and only a modest labeled dataset (say, a few
> hundred to a few tens of thousands of examples). If you have so little labeled
> data that even fine-tuning overfits, fall back to using BERT purely as a frozen
> feature extractor and train a simple model on top.

## Downstream Tasks and Why BERT Transfers Well

Once the network is fine-tuned, the same backbone supports several families of
prediction tasks; the difference is mainly which hidden states the task head
reads and what it predicts from them.

- Sequence classification (sentiment, topic, spam): use the `[CLS]` vector with
  a classification head.
- Token classification / Named-Entity Recognition (NER): attach a classifier to
  each token's representation to label spans (person, location, organization).
- Question answering (e.g., SQuAD-style extractive QA): predict the start and end
  positions of the answer span within a passage.

BERT transfers well because pretraining forces it to encode broad, reusable
linguistic structure (syntax, semantics, some world knowledge) into its hidden
states. A downstream task then mostly needs to learn how to *read off* the relevant
parts of that representation rather than learning language from scratch.

## Practical Use in R and Python

With the concepts in place, here is how you actually call BERT. In Python, the
standard interface is Hugging Face
`transformers`^[<https://huggingface.co/docs/transformers>], which provides
pretrained weights and tokenizers for BERT and its many variants. In R, the
`text`^[<https://www.r-text.org/>] package wraps Hugging Face models (via
reticulate, which bridges R and Python) so you can obtain embeddings or
predictions without leaving R.

> **Warning:** The chunks below are shown with `eval = FALSE` because they
> require a Python backend (transformers and PyTorch) and download large model
> weights on first use. To run them yourself, install the backend once with
> `text::textrpp_install()` in R, or `pip install transformers torch` for the
> Python version, and expect the first call to fetch several hundred megabytes.

A minimal R sketch using the `text` package to obtain contextual embeddings:

```{r, eval = FALSE}
# install.packages("text")
library(text)

# One-time setup: installs the Python backend (transformers, torch) in a
# managed conda environment via reticulate.
textrpp_install()
textrpp_initialize()

sentences <- c(
  "The river bank was flooded after the storm.",
  "She deposited her paycheck at the bank."
)

# Contextual embeddings from a pretrained BERT model.
embeddings <- textEmbed(
  texts = sentences,
  model = "bert-base-uncased"
)

# `embeddings$texts` holds per-sentence numeric features you can feed
# into any downstream model (e.g., glmnet, ranger, xgboost).
str(embeddings$texts)
```

The equivalent in Python, for comparison:

```{python, eval = FALSE}
from transformers import AutoTokenizer, AutoModel
import torch

tokenizer = AutoTokenizer.from_pretrained("bert-base-uncased")
model = AutoModel.from_pretrained("bert-base-uncased")

inputs = tokenizer("She deposited her paycheck at the bank.",
                   return_tensors="pt")
with torch.no_grad():
    outputs = model(**inputs)

# Last hidden state: (batch, tokens, hidden_size).
# The [CLS] vector (index 0) is a common sentence-level feature.
cls_embedding = outputs.last_hidden_state[:, 0, :]
```

For an end-to-end classification task, you would instead load
`AutoModelForSequenceClassification`, attach your labels, and fine-tune with the
Hugging Face `Trainer` API or PyTorch directly.

## Variants

The original BERT was a starting point, not a final word. Researchers quickly
found that the same architecture could be trained better, made smaller, or
specialized to a domain, and the result is a large family of models you will
encounter in practice. The most common ones are these.

- RoBERTa removes NSP, trains longer on more data with larger batches, and
  generally outperforms the original BERT.
- DistilBERT is a smaller, faster model distilled from BERT, retaining most of
  the accuracy at a fraction of the size, useful when latency or memory is tight.
- Domain-specific BERTs retrain or continue pretraining on specialized corpora,
  e.g., BioBERT (biomedical), SciBERT (scientific text), FinBERT
  (finance), and ClinicalBERT (clinical notes).

For most prediction problems, a sensible default is to start with a small
pretrained model (e.g., DistilBERT) for fast iteration, then scale up to a larger
or domain-specific model only if accuracy demands it.

## Summary

BERT turns text into context-aware numeric features by stacking Transformer
encoder layers whose self-attention lets every token attend to every other token.
It learns this skill once, through self-supervised pretraining (masked language
modeling, and originally next-sentence prediction) on a large unlabeled corpus.
You then reuse it in one of two ways: as a frozen feature extractor whose
`[CLS]` vector or token embeddings feed an ordinary downstream model, or by
fine-tuning the whole network with a small task head when you have labeled data.
The payoff, and the reason BERT earned a place alongside the other predictive
methods in this book, is transfer: the expensive part of understanding language
is done for you, and your job is reduced to teaching the model how to read off the
answer to your particular question.

## Further Reading

- Official implementation and weights: <https://github.com/google-research/bert>
- A gentle tutorial: <https://www.freecodecamp.org/news/google-bert-nlp-machine-learning-tutorial/>
- A visual guide to using BERT for the first time:
  <https://jalammar.github.io/a-visual-guide-to-using-bert-for-the-first-time/>
- BERT explained, theory and tutorial:
  <https://towardsml.com/2019/09/17/bert-explained-a-complete-guide-with-theory-and-tutorial/>
