# BERT {#sec-bert}

```{r}
#| include: false
source("_common.R")
```

Most of the methods in this book expect their inputs as numbers: a row of
predictors, a design matrix, a feature vector. Text does not arrive in that form.
A product review, a clinical note, or a news headline is just a string of
characters, and before any classifier or regressor can use it we need a way to
turn that string into numbers that capture what it *means*. This chapter is about
one of the most effective tools for doing exactly that.

BERT (Bidirectional Encoder Representations from Transformers) is a family of
pretrained language models that produce *contextual* representations of text.
Introduced by Google in 2018, BERT became one of the foundational tools for
modern natural language processing (NLP)^[Natural language processing is the
branch of machine learning concerned with getting computers to work with human
language: classifying it, extracting information from it, translating it, or
generating it.] and remains a workhorse for prediction tasks where the input is
text: document classification, sentiment analysis, named-entity recognition, and
question answering.

For a prediction-focused course, the key idea is this: BERT lets you turn raw
text into high-quality numeric features (embeddings) that you can feed into a
downstream classifier or regressor, *or* fine-tune the whole network end-to-end
for your specific task. Either way, you inherit knowledge learned from a massive
text corpus without having to train a language model yourself.


::: {.callout-important title="Key idea"}
BERT is a reusable "text-to-numbers" engine. Someone else paid
the enormous cost of teaching it language; you reuse the result, either as
off-the-shelf features or as a starting point you nudge toward your own task.
:::


By the end of this chapter you will understand what makes BERT's representations
"contextual," the core mechanics of the Transformer encoder it is built from, how
it is trained before you ever see it, and how to put it to work from both R and
Python. We begin with the idea that motivated the whole approach: the difference
between a fixed word vector and one that pays attention to context.

## From Static to Contextual Embeddings

Earlier word-embedding methods such as word2vec and GloVe map each word
type to a single fixed vector (@sec-embeddings-vector-search). The word "bank" gets *one* vector regardless of
whether the sentence is about a *river bank* or a *savings bank*. These are
static embeddings: useful, but blind to context.

BERT instead produces contextual embeddings. The vector it assigns to a token
depends on the entire surrounding sentence, so "bank" in "river bank" and "bank"
in "deposit money in the bank" receive different representations. This sensitivity
to context is what makes BERT dramatically more powerful than static embeddings on
tasks that hinge on meaning in context.

The word *bidirectional* in the name is central. Earlier contextual models (e.g.,
left-to-right language models) read text in one direction. BERT conditions each
token's representation on words to both its left and its right
simultaneously, giving a fuller picture of context.


::: {.callout-tip title="Intuition"}
To guess the missing word in "I deposited my check at the
\_\_\_," reading only left-to-right ("I deposited my check at the") already
points strongly toward "bank." But for "the \_\_\_ overflowed after three days
of rain," you need the words *after* the blank to tell that this is a river,
not a vault. Seeing both sides at once is what lets BERT resolve meaning that a
one-directional reader would miss.
:::


Now that we know *what* BERT produces, let us look at the machinery that produces
it.

## The Transformer Encoder

BERT is built entirely from the encoder half of the Transformer architecture
(@sec-transformers). You do not need the full mathematical machinery to use BERT,
but the following intuitions are worth carrying.

### Self-Attention

The core operation is self-attention. For each token, the model computes how
much it should "attend to" every other token in the sequence, then forms a
weighted combination of their representations. Concretely, each token is projected
into a query ($Q$), a key ($K$), and a value ($V$) vector, and
attention weights are computed as a scaled dot product:

$$
\text{Attention}(Q, K, V) = \text{softmax}\!\left(\frac{QK^\top}{\sqrt{d_k}}\right) V
$$

where $d_k$ is the dimension of the key vectors and the $\sqrt{d_k}$ term keeps the
dot products from growing too large. The softmax turns the similarities between a
query and all keys into a probability distribution; the output is the
correspondingly weighted average of the values. The practical effect: a token can
pull in information from any other token in the sentence, no matter how far away,
in a single step.


::: {.callout-tip title="Intuition"}
A useful mental model is a small lookup or search. Each token
issues a query ("which other words should I be paying attention to?"), every
token advertises a key ("here is what I am about"), and the dot product between
query and key measures how well they match. The softmax converts those match
scores into weights that sum to one, and the token then collects a weighted mix
of the values. The word "it" can thereby reach back and grab information from
the noun it refers to, even if that noun sat ten words earlier.
:::


#### Formulation

Fix notation. Let the input sequence have $n$ tokens, each embedded as a row
vector in $\mathbb{R}^{d_{\text{model}}}$, and collect them as the rows of
$X \in \mathbb{R}^{n \times d_{\text{model}}}$. A single attention head is
defined by three learned projection matrices
$W_Q \in \mathbb{R}^{d_{\text{model}} \times d_k}$,
$W_K \in \mathbb{R}^{d_{\text{model}} \times d_k}$, and
$W_V \in \mathbb{R}^{d_{\text{model}} \times d_v}$, which produce the query, key,
and value matrices

$$
Q = X W_Q, \qquad K = X W_K, \qquad V = X W_V .
$$ {#eq-BERT-qkv}

The head output is then

$$
\text{Attention}(Q,K,V)
= \underbrace{\operatorname{softmax}\!\left(\frac{QK^\top}{\sqrt{d_k}}\right)}_{A \,\in\, \mathbb{R}^{n\times n}} V ,
$$ {#eq-BERT-attn}

where the softmax is applied independently to each row, so each row of the
attention matrix $A$ is a probability vector: $A_{ij} \ge 0$ and
$\sum_{j=1}^{n} A_{ij} = 1$. Writing this row by row makes the operation
transparent. With $q_i = x_i W_Q$ the query for token $i$ and $k_j = x_j W_K$ the
key for token $j$, the weight token $i$ places on token $j$ is

$$
A_{ij} = \frac{\exp\!\big(q_i^\top k_j / \sqrt{d_k}\big)}
              {\sum_{\ell=1}^{n} \exp\!\big(q_i^\top k_\ell / \sqrt{d_k}\big)},
\qquad
\text{output}_i = \sum_{j=1}^{n} A_{ij}\, v_j .
$$ {#eq-BERT-attn-row}

So token $i$'s new representation is a convex combination of all value vectors,
with the weights determined by query-key compatibility. Because every $A_{ij}$ is
nonzero, information can flow between any pair of positions in a single layer;
this is the precise sense in which attention has a "receptive field" spanning the
whole sequence, in contrast to the local receptive field of a convolution or the
sequentially decaying memory of a recurrent network.

#### Why divide by $\sqrt{d_k}$

The scaling factor is not cosmetic; it controls the variance of the logits and
keeps the softmax in its well-conditioned regime. Suppose the entries of $q_i$
and $k_j$ are independent, mean-zero, unit-variance random variables (a
reasonable model at initialization). The unscaled logit is
$q_i^\top k_j = \sum_{m=1}^{d_k} q_{im} k_{jm}$. Each summand has mean
$\mathbb{E}[q_{im} k_{jm}] = 0$ and variance
$\operatorname{Var}(q_{im}k_{jm}) = \mathbb{E}[q_{im}^2]\,\mathbb{E}[k_{jm}^2] = 1$
by independence, so

$$
\mathbb{E}\big[q_i^\top k_j\big] = 0,
\qquad
\operatorname{Var}\big(q_i^\top k_j\big) = \sum_{m=1}^{d_k} \operatorname{Var}(q_{im}k_{jm}) = d_k .
$$ {#eq-BERT-logit-var}

The logits therefore have standard deviation $\sqrt{d_k}$, which grows with the
head dimension. Dividing by $\sqrt{d_k}$ rescales them back to unit variance.
This matters because the softmax saturates: if one logit is much larger than the
rest, the output approaches a one-hot vector and the Jacobian of the softmax
collapses toward zero. Concretely, for $p = \operatorname{softmax}(z)$,

$$
\frac{\partial p_i}{\partial z_j} = p_i(\delta_{ij} - p_j),
$$ {#eq-BERT-softmax-jac}

which vanishes whenever any $p_i \to 0$ or $p_i \to 1$. Letting the logits scale
as $\sqrt{d_k}$ would push the softmax into these flat regions and starve the
upstream projections of gradient. The $1/\sqrt{d_k}$ factor is exactly the
normalization that holds the logit variance at $O(1)$ regardless of $d_k$.

We can confirm @eq-BERT-logit-var, that the variance of the unscaled dot
product equals $d_k$ and the scaled version has unit variance, with a quick
simulation in base R.

```{r}
set.seed(1)
dot_var <- function(d_k, reps = 20000) {
  vals <- replicate(reps, {
    q <- rnorm(d_k); k <- rnorm(d_k)   # unit-variance, independent
    sum(q * k)                          # unscaled logit q^T k
  })
  c(unscaled = var(vals), scaled = var(vals / sqrt(d_k)))
}
sapply(c(8, 64, 512), dot_var)         # columns: d_k = 8, 64, 512
```

The first row tracks $d_k$ almost exactly, while the second row stays near one,
which is the whole point of the scaling.


### Multi-Head Attention and Positional Encodings

A single round of attention is powerful but limited: it produces one pattern of
"who attends to whom." Two refinements address that limitation and a structural
blind spot of attention.

- Multi-head attention. Rather than computing attention once, the model runs
  several attention "heads" in parallel, each with its own learned projections.
  Different heads can specialize, e.g., one tracking syntactic dependencies,
  another tracking coreference. Their outputs are concatenated and recombined.
- Positional encodings. Self-attention is permutation-invariant: on its own it
  has no notion of word order. BERT adds (learned) positional embeddings to the
  token embeddings so the model knows which token is first, second, and so on.

Stacking many such attention-plus-feed-forward layers (12 layers in BERT-base, 24
in BERT-large) yields a deep network whose hidden states are the contextual
representations we use downstream.

#### Multi-head attention, formally

With $h$ heads, each head $r = 1,\dots,h$ has its own projections
$W_Q^{(r)}, W_K^{(r)}, W_V^{(r)}$ and computes
$\text{head}_r = \text{Attention}(XW_Q^{(r)}, XW_K^{(r)}, XW_V^{(r)})
\in \mathbb{R}^{n \times d_v}$. The heads are concatenated along the feature axis
and mixed by an output projection
$W_O \in \mathbb{R}^{h d_v \times d_{\text{model}}}$,

$$
\text{MultiHead}(X) = \big[\text{head}_1 \;\Vert\; \cdots \;\Vert\; \text{head}_h\big]\, W_O .
$$ {#eq-BERT-multihead}

In BERT the per-head dimensions are tied to the model width by
$d_k = d_v = d_{\text{model}}/h$ (for BERT-base, $d_{\text{model}}=768$, $h=12$,
so $d_k = 64$). This keeps the total computation and parameter count of
multi-head attention essentially equal to that of a single full-width head while
allowing $h$ distinct attention patterns. The reason a single head is
insufficient is structural: each head produces a single $n \times n$ stochastic
matrix $A$, and a convex average through one such matrix can emphasize only one
relational pattern at a time. Multiple heads let the layer attend to several
relations (syntactic head, coreferent antecedent, adjacent token) in parallel
before $W_O$ recombines them.

#### The full encoder block

Each BERT layer wraps multi-head attention and a position-wise feed-forward
network in residual connections with layer normalization. Writing the
sublayer computation as

$$
\begin{aligned}
Z &= \text{LayerNorm}\big(X + \text{MultiHead}(X)\big), \\
\text{FFN}(z) &= \text{GELU}(z W_1 + b_1)\,W_2 + b_2, \\
\text{Output} &= \text{LayerNorm}\big(Z + \text{FFN}(Z)\big),
\end{aligned}
$$ {#eq-BERT-block}

where the feed-forward network is applied to each position independently with
$W_1 \in \mathbb{R}^{d_{\text{model}} \times d_{\text{ff}}}$ and
$W_2 \in \mathbb{R}^{d_{\text{ff}} \times d_{\text{model}}}$ (BERT-base uses
$d_{\text{ff}} = 4 d_{\text{model}} = 3072$). The residual connections
$X + (\cdot)$ are what make a stack of 12 to 24 such blocks trainable: the
identity path gives gradients an unobstructed route back to early layers, so the
gradient of the loss with respect to an early hidden state always retains an
additive term of magnitude one, mitigating vanishing gradients.

#### Computational complexity

Self-attention's cost is dominated by forming $QK^\top$ and applying $A$ to $V$,
each $O(n^2 d)$ in time, plus the $O(n^2)$ memory needed to store the attention
matrix. The feed-forward sublayer costs $O(n\, d\, d_{\text{ff}})$. Per layer the
total is

$$
O\big(n^2 d + n\, d\, d_{\text{ff}}\big).
$$ {#eq-BERT-complexity}

The quadratic dependence on sequence length $n$ is the defining scaling property
(and limitation) of the Transformer: it is the reason BERT caps inputs at 512
tokens and the motivation for the many "efficient attention" variants
(Longformer, BigBird, Performer) that replace the dense $n \times n$ matrix with
sparse or low-rank approximations.


::: {.callout-note}
You do not have to memorize these mechanics to use BERT effectively.
What matters for practice is the consequence: every token's output vector is a
context-aware summary that has already had the chance to look at every other
token in the input.
:::


## Pretraining Objectives

An architecture alone learns nothing; the network only becomes useful after it is
trained. BERT is pretrained on a large unlabeled corpus (originally English
Wikipedia and the BooksCorpus) using two self-supervised objectives
(@sec-self-supervised-learning). "Self-supervised" means the labels come from the
text itself, so no manual annotation is required.


::: {.callout-important title="Why this matters"}
Labeled text is expensive, but raw text is essentially
free and almost unlimited. Self-supervision lets BERT learn from billions of
words without anyone hand-labeling a single one, which is precisely why the
pretrained model carries so much general linguistic knowledge.
:::


### Masked Language Modeling (MLM)

A random subset (roughly 15%) of input tokens is replaced with a special `[MASK]`
token, and the model is trained to predict the original tokens from context. Because
it must use both left and right context to fill in the blanks, MLM is what gives
BERT its deep bidirectional understanding. This is analogous to a cloze
("fill-in-the-blank") test.

#### The MLM objective, precisely

Let $x = (x_1,\dots,x_n)$ be the token sequence and let
$\mathcal{M} \subset \{1,\dots,n\}$ be the randomly chosen set of masked
positions ($|\mathcal{M}| \approx 0.15\, n$). Write $x_{\mathcal{M}}$ for the
masked tokens and $x_{\setminus \mathcal{M}}$ for the corrupted context the model
sees. The encoder maps the (corrupted) input to final hidden states
$H = (h_1,\dots,h_n)$ with $h_i \in \mathbb{R}^{d_{\text{model}}}$, and a
prediction head turns each masked position's hidden state into a distribution
over the vocabulary $\mathcal{V}$ by a softmax,

$$
p_\theta\big(x_i \mid x_{\setminus \mathcal{M}}\big)
= \operatorname{softmax}\!\big(W_{\text{emb}}\, g(h_i) + b\big),
\qquad i \in \mathcal{M},
$$ {#eq-BERT-mlm-head}

where $g$ is a small transform (a dense layer plus GELU plus LayerNorm) and
$W_{\text{emb}} \in \mathbb{R}^{|\mathcal{V}| \times d_{\text{model}}}$ is tied to
the input embedding matrix. Training minimizes the average negative
log-likelihood over masked positions, which is precisely the cross-entropy
between the one-hot truth and the predicted distribution,

$$
\mathcal{L}_{\text{MLM}}(\theta)
= -\,\mathbb{E}_{x,\,\mathcal{M}}
\left[\frac{1}{|\mathcal{M}|}\sum_{i \in \mathcal{M}}
\log p_\theta\big(x_i \mid x_{\setminus \mathcal{M}}\big)\right].
$$ {#eq-BERT-mlm-loss}

This is the key formal distinction from a left-to-right language model. A standard
autoregressive model factorizes
$p(x) = \prod_{i} p(x_i \mid x_{<i})$ and conditions only on the left context. MLM
instead estimates the conditionals $p(x_i \mid x_{\setminus \mathcal{M}})$ given
context on both sides, which is what licenses the term "bidirectional." The price
is that MLM does not define a proper joint distribution over $x$ (the conditionals
need not be compatible with any single joint), so BERT is a representation learner
and a denoiser rather than a generative model you can sample from.

::: {.callout-note title="The 80/10/10 masking rule"}
A subtlety: the `[MASK]` token appears during pretraining but never at
fine-tuning time, which would create a train/test mismatch. BERT's remedy is that
of the 15% selected positions, 80% are replaced with `[MASK]`, 10% are replaced
with a random vocabulary token, and 10% are left unchanged. The loss in
@eq-BERT-mlm-loss is still computed at all selected positions. Leaving some
tokens unchanged forces the model to build a useful representation for every
position (it cannot tell which unchanged tokens it will be scored on), and the
random replacements make the model robust rather than reliant on the literal
`[MASK]` symbol.
:::

### Next Sentence Prediction (NSP)

BERT is also given pairs of sentences and trained to predict whether the second
sentence actually followed the first in the original text or was a random
distractor. NSP was intended to help with tasks that involve relationships between
sentences (e.g., question answering, natural-language inference). Later work (notably
RoBERTa) found NSP contributes little, and many successors drop it. It is still
worth knowing as part of the original recipe.

Formally, NSP is a binary classification on the `[CLS]` hidden state $h_{\texttt{[CLS]}}$,
with $y = 1$ if the second segment is the true continuation and $y = 0$ if it was
sampled at random,

$$
p_\theta(y \mid x) = \sigma\!\big(w^\top h_{\texttt{[CLS]}} + b\big),
\qquad
\mathcal{L}_{\text{NSP}}(\theta)
= -\,\mathbb{E}\big[y \log p_\theta + (1-y)\log(1-p_\theta)\big],
$$ {#eq-BERT-nsp-loss}

and BERT is pretrained on the sum
$\mathcal{L} = \mathcal{L}_{\text{MLM}} + \mathcal{L}_{\text{NSP}}$. The reason
NSP turned out to be weak is that distinguishing a random sentence from the true
next sentence is dominated by topic mismatch, a signal already abundant in MLM, so
NSP adds little beyond what masked prediction teaches. RoBERTa drops it; ALBERT
replaces it with sentence-order prediction (predicting whether two consecutive
segments have been swapped), a harder task that removes the topic shortcut and is
more useful.

## Tokenization and Special Tokens

BERT does not operate on whole words. It uses WordPiece tokenization, which
splits text into subword units drawn from a fixed vocabulary. Rare or unseen words
are broken into known pieces (e.g., "playing" $\rightarrow$ "play" + "##ing"). This
keeps the vocabulary manageable while avoiding out-of-vocabulary tokens entirely.

Beyond the ordinary subword tokens, two special tokens do bookkeeping that you
will see referenced constantly in code and tutorials:

- `[CLS]` is prepended to every input. Its final-layer hidden state is used as
  an aggregate, sentence-level representation for classification tasks.
- `[SEP]` separates segments (e.g., the two sentences in a pair, or a question
  and a passage).


::: {.callout-tip}
When you see code grab "index 0" of the hidden states (as in the
Python example below), that is the `[CLS]` vector being used as a single
fixed-length summary of the whole input. It is the most common way to reduce a
variable-length sentence to one feature vector.
:::


## Pretrain then Fine-Tune

The defining paradigm is pretrain $\rightarrow$ fine-tune:

1. Pretrain once, expensively, on a huge corpus (done by the model's authors;
   you download the result).
2. Fine-tune cheaply on your labeled task by adding a small task-specific head
   (often a single linear layer on top of the `[CLS]` representation) and training
   the whole network for a few epochs on your data.

#### The fine-tuning objective

For a sequence-classification task with $C$ classes and labeled data
$\{(x^{(t)}, y^{(t)})\}_{t=1}^{N}$, fine-tuning adds a single linear head on the
`[CLS]` representation and minimizes the regularized cross-entropy

$$
p_\phi(y \mid x) = \operatorname{softmax}\!\big(W\, h_{\texttt{[CLS]}}(x;\theta) + b\big),
\qquad
\min_{\theta,\,W,\,b}\;
-\frac{1}{N}\sum_{t=1}^{N}\log p_\phi\big(y^{(t)} \mid x^{(t)}\big),
$$ {#eq-BERT-finetune}

where $h_{\texttt{[CLS]}}(x;\theta)$ now depends on the full encoder parameters
$\theta$, which are themselves updated (not frozen). The crucial point is the
starting condition: rather than initializing $\theta$ at random, fine-tuning
initializes it at the pretrained weights $\theta_{\text{pre}}$ and runs only a few
epochs of gradient descent. The optimization therefore stays in the basin around
$\theta_{\text{pre}}$, which already encodes general linguistic structure, and the
task head $W$ only has to learn a linear read-out of that structure.

#### Why transfer works: a bias-variance view

The transfer benefit has a clean statistical reading. Training a Transformer from
scratch on a few thousand labeled examples is a high-variance estimation problem:
the hypothesis class (hundreds of millions of parameters) is enormous relative to
the sample, so the estimator overfits badly. Pretraining acts as an extremely
informative, data-driven prior. Initializing at $\theta_{\text{pre}}$ and taking a
small number of steps constrains the effective hypothesis class to a neighborhood
of weights that already produce linguistically sensible representations,
collapsing the variance at the cost of a small bias (the pretrained features may
not be perfectly aligned with the target task). On modest labeled datasets this
trade is overwhelmingly favorable, which is exactly the regime where fine-tuned
BERT dominates models trained from scratch. As labeled data grows, the variance
penalty of training from scratch shrinks and the advantage of transfer narrows,
the familiar shape of a learning curve crossover.

This transfer-learning recipe (@sec-transfer-multitask-learning) is why BERT is so
practical: the heavy lifting of learning language is amortized across everyone who
reuses the pretrained weights.
Fine-tuning typically needs only modest amounts of labeled data and compute.^[The
original pretraining ran for days on many specialized accelerators. Fine-tuning a
task-specific head, by contrast, is often a matter of minutes to a few hours on a
single GPU, and sometimes feasible on a CPU for small datasets.]


::: {.callout-tip title="When to use this"}
Reach for the pretrain-then-fine-tune recipe when you
have a text-prediction task and only a modest labeled dataset (say, a few
hundred to a few tens of thousands of examples). If you have so little labeled
data that even fine-tuning overfits, fall back to using BERT purely as a frozen
feature extractor and train a simple model on top.
:::


## Downstream Tasks and Why BERT Transfers Well

Once the network is fine-tuned, the same backbone supports several families of
prediction tasks; the difference is mainly which hidden states the task head
reads and what it predicts from them.

- Sequence classification (sentiment, topic, spam): use the `[CLS]` vector with
  a classification head.
- Token classification / Named-Entity Recognition (NER): attach a classifier to
  each token's representation to label spans (person, location, organization).
- Question answering (e.g., SQuAD-style extractive QA): predict the start and end
  positions of the answer span within a passage.

For extractive question answering the head structure is worth making explicit,
since it is less obvious than the classification case. Given the token hidden
states $h_1,\dots,h_n$ of the (question, passage) pair, two learned vectors
$s, e \in \mathbb{R}^{d_{\text{model}}}$ define independent softmax distributions
over positions for the answer's start and end,

$$
P_{\text{start}}(i) = \frac{\exp(s^\top h_i)}{\sum_{j} \exp(s^\top h_j)},
\qquad
P_{\text{end}}(i) = \frac{\exp(e^\top h_i)}{\sum_{j} \exp(e^\top h_j)},
$$ {#eq-BERT-qa}

trained by summing the cross-entropies of the true start and end indices. At
inference the predicted span $(\hat\imath, \hat\jmath)$ maximizes
$s^\top h_i + e^\top h_j$ subject to $i \le j$, computable in $O(n)$ after sorting.

BERT transfers well because pretraining forces it to encode broad, reusable
linguistic structure (syntax, semantics, some world knowledge) into its hidden
states. A downstream task then mostly needs to learn how to *read off* the relevant
parts of that representation rather than learning language from scratch.

#### Failure modes

BERT is not a universal solution, and knowing where it breaks is part of using it
responsibly.

- Sequence length. The $O(n^2)$ attention cost and the learned positional
  embeddings cap the input at 512 tokens. Longer documents must be truncated or
  chunked, which loses long-range dependencies; for genuinely long inputs use a
  long-context variant rather than forcing BERT.
- Fine-tuning instability. On small datasets, fine-tuning all parameters can be
  unstable across random seeds, with a non-trivial fraction of runs failing to
  beat the majority-class baseline. The usual culprits are too high a learning
  rate and too few warmup steps; remedies are lower learning rates, more warmup,
  and averaging or selecting over several seeds.
- Domain shift. Pretraining on Wikipedia and books leaves BERT weak on text
  whose vocabulary and style differ sharply (clinical notes, legal contracts,
  social-media slang), where WordPiece shatters domain terms into many subwords.
  Domain-adaptive continued pretraining or a domain-specific variant fixes this.
- Distribution and spurious cues. Like any empirical-risk minimizer, a fine-tuned
  BERT will exploit dataset artifacts (annotation shortcuts, lexical overlap
  heuristics) and can fail under distribution shift even when in-distribution
  accuracy is high.

## Practical Use in R and Python

With the concepts in place, here is how you actually call BERT. In Python, the
standard interface is Hugging Face
`transformers`^[<https://huggingface.co/docs/transformers>], which provides
pretrained weights and tokenizers for BERT and its many variants. In R, the
`text`^[<https://www.r-text.org/>] package wraps Hugging Face models (via
reticulate, which bridges R and Python) so you can obtain embeddings or
predictions without leaving R.


::: {.callout-warning}
The chunks below are shown with `eval = FALSE` because they
require a Python backend (transformers and PyTorch) and download large model
weights on first use. To run them yourself, install the backend once with
`text::textrpp_install()` in R, or `pip install transformers torch` for the
Python version, and expect the first call to fetch several hundred megabytes.
:::


A minimal R sketch using the `text` package to obtain contextual embeddings:

```{r, eval = FALSE}
# install.packages("text")
library(text)

# One-time setup: installs the Python backend (transformers, torch) in a
# managed conda environment via reticulate.
textrpp_install()
textrpp_initialize()

sentences <- c(
  "The river bank was flooded after the storm.",
  "She deposited her paycheck at the bank."
)

# Contextual embeddings from a pretrained BERT model.
embeddings <- textEmbed(
  texts = sentences,
  model = "bert-base-uncased"
)

# `embeddings$texts` holds per-sentence numeric features you can feed
# into any downstream model (e.g., glmnet, ranger, xgboost).
str(embeddings$texts)
```

The equivalent in Python, for comparison:

```{python, eval = FALSE}
from transformers import AutoTokenizer, AutoModel
import torch

tokenizer = AutoTokenizer.from_pretrained("bert-base-uncased")
model = AutoModel.from_pretrained("bert-base-uncased")

inputs = tokenizer("She deposited her paycheck at the bank.",
                   return_tensors="pt")
with torch.no_grad():
    outputs = model(**inputs)

# Last hidden state: (batch, tokens, hidden_size).
# The [CLS] vector (index 0) is a common sentence-level feature.
cls_embedding = outputs.last_hidden_state[:, 0, :]
```

For an end-to-end classification task, you would instead load
`AutoModelForSequenceClassification`, attach your labels, and fine-tune with the
Hugging Face `Trainer` API or PyTorch directly.

### Choosing Hyperparameters and Diagnostics

The original BERT paper recommends a deliberately narrow search space for
fine-tuning, and it remains a good default.

- Learning rate. Use a small value, typically in $\{2,3,5\}\times 10^{-5}$ with the
  AdamW optimizer. Fine-tuning is sensitive here: rates above $\sim 10^{-4}$
  frequently destabilize training because they push $\theta$ out of the pretrained
  basin in a single step. Pair the rate with a linear warmup over the first
  6 to 10 percent of steps followed by linear decay; the warmup is what prevents
  the early, high-variance gradients from corrupting the pretrained weights.
- Epochs. Two to four epochs are usually enough. More tends to overfit because the
  model starts from an already-competent initialization, so the marginal labeled
  signal is small.
- Batch size. 16 or 32 are standard; larger batches need a proportionally larger
  learning rate.
- Frozen versus fine-tuned. If labeled data is very scarce or compute is tight,
  freeze $\theta$ and train only the head (equivalently, use BERT as the feature
  extractor of @eq-BERT-finetune with $\theta$ held at $\theta_{\text{pre}}$).
  This trades a little accuracy for stability and speed.

A practical note on pooling. The `[CLS]` vector is the conventional sentence
summary, but for the frozen-feature use case mean-pooling the token hidden states
(averaging $h_1,\dots,h_n$ over non-padding positions) is often a stronger
sentence representation than `[CLS]`, because the raw pretrained `[CLS]` vector was
optimized for NSP rather than for general similarity. When fine-tuning end to end
the distinction largely washes out, since the head and encoder co-adapt.

For diagnostics, watch the gap between training and validation loss across the few
epochs (overfitting appears quickly at these dataset sizes), and because
fine-tuning can be seed-sensitive on small data, run several seeds and report the
median rather than the best.

## Variants

The original BERT was a starting point, not a final word. Researchers quickly
found that the same architecture could be trained better, made smaller, or
specialized to a domain, and the result is a large family of models you will
encounter in practice. The most common ones are these.

- RoBERTa removes NSP, trains longer on more data with larger batches, and
  generally outperforms the original BERT.
- DistilBERT is a smaller, faster model distilled from BERT, retaining most of
  the accuracy at a fraction of the size, useful when latency or memory is tight.
- Domain-specific BERTs retrain or continue pretraining on specialized corpora,
  e.g., BioBERT (biomedical), SciBERT (scientific text), FinBERT
  (finance), and ClinicalBERT (clinical notes).

For most prediction problems, a sensible default is to start with a small
pretrained model (e.g., DistilBERT) for fast iteration, then scale up to a larger
or domain-specific model only if accuracy demands it.

## Summary

BERT turns text into context-aware numeric features by stacking Transformer
encoder layers whose self-attention lets every token attend to every other token.
It learns this skill once, through self-supervised pretraining (masked language
modeling, and originally next-sentence prediction) on a large unlabeled corpus.
You then reuse it in one of two ways: as a frozen feature extractor whose
`[CLS]` vector or token embeddings feed an ordinary downstream model, or by
fine-tuning the whole network with a small task head when you have labeled data.
The payoff, and the reason BERT earned a place alongside the other predictive
methods in this book, is transfer: the expensive part of understanding language
is done for you, and your job is reduced to teaching the model how to read off the
answer to your particular question.

## Further Reading

- Official implementation and weights: <https://github.com/google-research/bert>
- A gentle tutorial: <https://www.freecodecamp.org/news/google-bert-nlp-machine-learning-tutorial/>
- A visual guide to using BERT for the first time:
  <https://jalammar.github.io/a-visual-guide-to-using-bert-for-the-first-time/>
- BERT explained, theory and tutorial:
  <https://towardsml.com/2019/09/17/bert-explained-a-complete-guide-with-theory-and-tutorial/>
