# Large Language Models and Foundation Models {#sec-llms}

```{r}
#| include: false
source("_common.R")
```

A language model assigns probabilities to sequences of text. For a
prediction-focused course, the value is direct: once you can score how likely a
piece of text is, you can predict the next token, fill in blanks, classify
documents by comparing likelihoods, or generate text by sampling. **Large
language models** (LLMs) are the same idea scaled up by orders of magnitude in
data, parameters, and compute, trained once and then reused across many tasks.
That reuse pattern is what people mean by a foundation model.

This chapter builds the concept from the ground up. We start with what a language
model is mathematically, trace the path from simple counting models to
Transformer decoders, describe how these models are pretrained and then adapted,
and end with a runnable base R demonstration that builds a small n-gram language
model, generates text, and measures perplexity.

The reassuring news for anyone coming from the rest of this book is that nothing
here is conceptually new at its core. A language model is a conditional
probability estimator, the same object you have fit all along, just pointed at
text and scaled up dramatically. By the end you will be able to say precisely
what an LLM computes, why "next-token prediction" is enough to learn a great deal
about language, how perplexity measures a model's quality, and how a pretrained
model gets adapted to a specific task. The small n-gram model at the end makes
all of this concrete: it is a complete language model you can read line by line,
and it behaves, in miniature, exactly like the giants.


::: {.callout-tip title="Intuition"}
If you can estimate "given what came before, what comes next?"
then prediction, classification, fill-in-the-blank, and generation are all the
same machine used in different ways. Everything in this chapter is a variation
on that single estimate.
:::


## What a Language Model Is

Let a piece of text be a sequence of tokens $x_1, x_2, \dots, x_T$. A language
model defines a probability distribution over such sequences. Using the chain
rule of probability, any joint distribution factorizes exactly into a product of
conditional distributions:

$$
p(x_1, x_2, \dots, x_T) = \prod_{t=1}^{T} p(x_t \mid x_1, \dots, x_{t-1})
= \prod_{t=1}^{T} p(x_t \mid x_{<t}),
$$

where $x_{<t}$ denotes all tokens before position $t$. This factorization is not
an approximation. It is the definition of a joint distribution written one token
at a time. A model that estimates each factor $p(x_t \mid x_{<t})$ is called an
autoregressive language model, because it predicts the next token from the
tokens it has already seen.^[A token is the unit a model reads and predicts.
For now you can think of it as roughly a word; the section on tokenization below
explains why real models use subword pieces instead.]


::: {.callout-important title="Key idea"}
The chain rule turns the impossible-looking task of modeling whole
documents into a sequence of ordinary one-step prediction problems. Estimate
$p(x_t \mid x_{<t})$ well at every position and you have, by construction, a
model of the entire sequence.
:::


### Next-Token Prediction

Training an autoregressive model is a supervised learning problem in disguise.
The input is the context $x_{<t}$, the target is the next token $x_t$, and the
labels come for free from the text itself. No human annotation is needed, which
is what makes it possible to train on enormous corpora. The model outputs a
probability distribution over the vocabulary $V$ at each position, and we want
that distribution to place high probability on the token that actually came next.


::: {.callout-note}
This is why LLMs train on text scraped from the open web, books, and
code without paying for labels. The "label" for each position is simply the
word that already follows it. Self-supervision of this kind is what makes
training on trillions of tokens possible.
:::


### Cross-Entropy Loss

The standard training objective is the cross-entropy loss, which is the
average negative log-likelihood of the observed tokens under the model. For a
single sequence,

$$
\mathcal{L} = -\frac{1}{T} \sum_{t=1}^{T} \log p(x_t \mid x_{<t}).
$$

Minimizing this loss is the same as maximizing the likelihood the model assigns
to real text. The logarithm is taken in natural base here, though base 2 is also
common (the two differ only by a constant factor of $\log 2$).^[Negative
log-likelihood is the same loss you have seen for logistic regression and softmax
classifiers elsewhere in the book. A language model is, at each position, a
classifier over the vocabulary, so it inherits exactly that objective.]

#### The model class and the softmax head {.unnumbered}

To be precise about what is being optimized, fix the parametric form. A
Transformer decoder with parameters $\theta$ maps a context $x_{<t}$ to a
real-valued vector of logits $z_t = f_\theta(x_{<t}) \in \mathbb{R}^{|V|}$, one
entry per vocabulary item. The conditional distribution is the softmax of those
logits,

$$
p_\theta(x_t = v \mid x_{<t}) = \frac{\exp(z_{t,v})}{\sum_{w \in V} \exp(z_{t,w})},
\qquad v \in V.
$$ {#eq-llms-softmax}

The full objective over a corpus of $N$ sequences $\{x^{(i)}\}$ is the negative
conditional log-likelihood,

$$
\hat{\theta} = \arg\min_\theta \; -\sum_{i=1}^{N} \sum_{t=1}^{T_i}
\log p_\theta\!\left(x^{(i)}_t \mid x^{(i)}_{<t}\right),
$$

which is exactly the maximum-likelihood estimator under the autoregressive
factorization. The only assumptions are (i) the chain-rule factorization, which
is exact, and (ii) the parametric family $f_\theta$, which is the modeling
choice. Everything that distinguishes an n-gram from a Transformer lives in
$f_\theta$; the loss is identical.

#### Gradient of the cross-entropy through the softmax {.unnumbered}

The reason this objective is convenient is that its gradient with respect to the
logits is strikingly simple. Writing the per-position loss as
$\ell_t = -\log p_\theta(x_t \mid x_{<t})$ and letting $y_t \in \{0,1\}^{|V|}$ be
the one-hot indicator of the true next token, the derivative with respect to the
logit of class $v$ is

$$
\frac{\partial \ell_t}{\partial z_{t,v}}
= p_\theta(x_t = v \mid x_{<t}) - y_{t,v}.
$$ {#eq-llms-softmax-grad}

To see this, note $\ell_t = -z_{t,x_t} + \log \sum_w \exp(z_{t,w})$. The first
term contributes $-y_{t,v}$. The second contributes
$\exp(z_{t,v}) / \sum_w \exp(z_{t,w}) = p_\theta(x_t = v \mid x_{<t})$ by direct
differentiation of the log-sum-exp. The gradient is therefore the predicted
distribution minus the target, the same "prediction error" form that drives
logistic and softmax regression. Backpropagation simply carries this residual
back into $\theta$ through $f_\theta$.

The following base R check confirms @eq-llms-softmax-grad by comparing the
analytic gradient $p - y$ against a finite-difference approximation of the
cross-entropy loss at random logits.

```{r softmax-grad-check}
set.seed(1)
softmax <- function(z) exp(z - max(z)) / sum(exp(z - max(z)))
z <- rnorm(6)            # random logits over a 6-token vocabulary
true_class <- 3          # index of the observed next token
loss <- function(z) -log(softmax(z)[true_class])

# Analytic gradient: predicted distribution minus the one-hot target.
y <- numeric(6); y[true_class] <- 1
analytic <- softmax(z) - y

# Finite-difference gradient.
eps <- 1e-6
numeric_grad <- sapply(seq_along(z), function(j) {
  zp <- z; zp[j] <- zp[j] + eps
  zm <- z; zm[j] <- zm[j] - eps
  (loss(zp) - loss(zm)) / (2 * eps)
})

max(abs(analytic - numeric_grad))   # ~1e-9, confirming the derivation
```

### Perplexity

The headline evaluation metric for a language model is perplexity (PPL),
defined as the exponential of the average negative log-likelihood:

$$
\mathrm{PPL} = \exp\!\left( -\frac{1}{T} \sum_{t=1}^{T} \log p(x_t \mid x_{<t}) \right)
= \exp(\mathcal{L}).
$$

Perplexity has a clean interpretation. If a model has perplexity $k$ on some
text, it is on average as uncertain as if it had to choose uniformly among $k$
equally likely tokens at each step. A perplexity of 1 means perfect prediction;
a perplexity equal to the vocabulary size means the model learned nothing beyond
a uniform guess. Lower is better. Because perplexity is a deterministic function
of cross-entropy, reducing the loss and reducing perplexity are the same goal.

#### Perplexity, cross-entropy, and KL divergence {.unnumbered}

The "effective number of guesses" reading is not a metaphor; it is the
information-theoretic content of perplexity. Suppose the test tokens are drawn
from a true conditional distribution $p^\star$ and the model predicts $p_\theta$.
As $T \to \infty$, the sample average in @eq-llms-softmax converges by the law
of large numbers to a cross-entropy expectation,

$$
\frac{1}{T}\sum_{t=1}^{T} -\log p_\theta(x_t \mid x_{<t})
\;\xrightarrow{\;a.s.\;}\;
H(p^\star, p_\theta)
= \underbrace{H(p^\star)}_{\text{irreducible}}
+ \underbrace{D_{\mathrm{KL}}\!\left(p^\star \,\|\, p_\theta\right)}_{\ge 0},
$$ {#eq-llms-ce-decomp}

using $H(p^\star, p_\theta) = H(p^\star) + D_{\mathrm{KL}}(p^\star \| p_\theta)$.
Exponentiating, $\mathrm{PPL} = \exp(H(p^\star, p_\theta))$. Two consequences
follow. First, perplexity is bounded below by $\exp(H(p^\star))$, the entropy of
language itself; no model, however large, can drive perplexity below the
intrinsic uncertainty of the next token. Second, the only term a model can reduce
is the KL gap $D_{\mathrm{KL}}(p^\star \| p_\theta) \ge 0$, which vanishes if and
only if $p_\theta = p^\star$ almost everywhere. Minimizing cross-entropy is
therefore exactly forward-KL projection of the model onto the true distribution,
and scaling laws (below) are empirically the rate at which that KL gap shrinks.

The uniform-guess bound is the special case $p_\theta(\cdot \mid x_{<t}) =
1/|V|$, which gives $\mathcal{L} = \log |V|$ and $\mathrm{PPL} = |V|$, confirming
the claim that a vocabulary-size perplexity signals a model that learned nothing.


::: {.callout-tip title="Intuition"}
Read perplexity as an "effective number of guesses." A perplexity
of 5 means the model is about as confused as someone flipping fairly between 5
options at each word. Cutting perplexity in half means the model has genuinely
narrowed down its choices.
:::



::: {.callout-warning}
Perplexity is only comparable across models that share the same
tokenization and vocabulary. A model that splits text into more, smaller tokens
faces an easier per-token prediction and can post a lower perplexity without
being a better model. Never compare perplexity numbers across different
tokenizers.
:::


## From N-grams to Transformer Decoders

The definitions above tell us what to estimate; they say nothing about how. The
hard part of $p(x_t \mid x_{<t})$ is that the context $x_{<t}$ grows without
bound: by the thousandth word, the model must condition on the previous nine
hundred ninety-nine. No method can store a separate probability for every
possible history, so every model family makes a compromise in how it summarizes
that context. The history of language modeling is largely the history of better
and better summaries, and the three families below mark the main steps.

### N-gram Models

The oldest approach truncates the context to the previous $n-1$ tokens, an
$n$-th order Markov assumption:

$$
p(x_t \mid x_{<t}) \approx p(x_t \mid x_{t-n+1}, \dots, x_{t-1}).
$$

A bigram model ($n = 2$) conditions on one previous token, a trigram
($n = 3$) on two. The conditional probabilities are estimated by counting how
often each continuation follows each context in a training corpus and
normalizing.

#### Why counting is the maximum-likelihood estimate {.unnumbered}

The "count and normalize" recipe is not an ad hoc heuristic; it is the MLE for a
multinomial model. Fix a context $c = (x_{t-n+1}, \dots, x_{t-1})$ and let
$\theta_{w \mid c} = p(w \mid c)$ be the parameters of the conditional
distribution, subject to $\sum_{w} \theta_{w \mid c} = 1$. If the context $c$
occurs with the various continuations $w$ counted as $N(c, w)$ times, the
log-likelihood contributed by that context is $\sum_w N(c,w) \log \theta_{w\mid
c}$. Maximizing under the simplex constraint with a Lagrange multiplier
$\lambda$,

$$
\frac{\partial}{\partial \theta_{w\mid c}}
\left[ \sum_w N(c,w)\log\theta_{w\mid c}
+ \lambda\Big(1 - \sum_w \theta_{w\mid c}\Big) \right]
= \frac{N(c,w)}{\theta_{w\mid c}} - \lambda = 0,
$$

so $\theta_{w\mid c} = N(c,w)/\lambda$, and the constraint forces $\lambda =
\sum_w N(c,w) = N(c)$. The MLE is therefore the relative frequency

$$
\hat{\theta}^{\,\mathrm{MLE}}_{w\mid c} = \frac{N(c,w)}{N(c)},
$$ {#eq-llms-ngram-mle}

which is exactly the unsmoothed counting estimator.

#### Add-one smoothing as MAP estimation {.unnumbered}

Add-one smoothing is the Bayesian counterpart of this MLE. Place a symmetric
Dirichlet prior $\theta_{\cdot \mid c} \sim \mathrm{Dir}(\alpha, \dots, \alpha)$
over the $|V|$ continuations. The Dirichlet is conjugate to the multinomial, so
the posterior is $\mathrm{Dir}\big(\alpha + N(c, w)\big)_w$, and the posterior
mean (equivalently the MAP estimate up to the usual mode-versus-mean offset) is

$$
\hat{\theta}^{\,\mathrm{smooth}}_{w\mid c}
= \frac{N(c,w) + \alpha}{N(c) + \alpha |V|}.
$$ {#eq-llms-ngram-smooth}

Setting $\alpha = 1$ recovers add-one (Laplace) smoothing exactly, the formula
implemented as `(count_full + 1) / (count_ctx + vocab_size)` in the demo below.
The prior acts as $\alpha$ pseudo-counts spread uniformly over the vocabulary: it
pulls every conditional toward the uniform distribution, guarantees strictly
positive probability for unseen continuations (hence finite perplexity), and its
influence fades as $N(c)$ grows. Smaller $\alpha$ (for example $\alpha = 0.01$,
Lidstone smoothing) smooths less and is usually preferable when contexts are
moderately well observed; add-one is known to over-smooth large vocabularies
because it moves a mass proportional to $|V|$ into the denominator.

N-gram models are fast and transparent, but they suffer from data
sparsity (most long contexts are never seen) and they cannot capture
dependencies beyond the fixed window. Smoothing techniques such as add-one
(Laplace) smoothing are used to avoid assigning zero probability to unseen
continuations.^[Without smoothing, a single unseen continuation makes the
sequence probability zero and the perplexity infinite. Add-one smoothing pretends
every continuation was seen one extra time, which guarantees nonzero
probabilities everywhere.] The runnable demo at the end of the chapter builds
exactly this kind of model, so the summary here will become concrete shortly.

### Neural Language Models

Instead of counting discrete contexts, a neural language model maps each
token to a dense vector (an embedding) and uses a neural network to predict the
next token. Early versions used feed-forward networks over a fixed window;
recurrent networks (RNNs, LSTMs) then allowed the context to be summarized into a
hidden state that is carried forward step by step, in principle removing the
fixed-window limit. Recurrent models struggle with very long dependencies and are
hard to parallelize across positions, because each step must wait for the
previous one to finish.


::: {.callout-important title="Key idea"}
Neural models replace counting with learning. Similar contexts
map to similar vectors, so the model can generalize to histories it never saw,
the exact weakness that sinks n-grams.
:::


### Transformer Decoders

Modern LLMs are Transformer decoders. Rather than a recurrent hidden state,
they use self-attention so that each position can read directly from every
earlier position in a single step. To preserve the autoregressive property, the
attention is causally masked: position $t$ may attend to positions
$1, \dots, t$ but not to anything in the future. Stacking many masked
self-attention and feed-forward layers gives a model that predicts
$p(x_t \mid x_{<t})$ with a rich, learned summary of the entire prior context.
The attention mechanism, multi-head attention, and positional information are
covered in the attention and Transformers chapter (@sec-transformers) and in the
chapter on BERT (@sec-bert), which uses the encoder half of the same architecture. The decoder used by LLMs is
that architecture with causal masking and a next-token objective.

#### Causal self-attention, made precise {.unnumbered}

Concretely, let $X \in \mathbb{R}^{T \times d}$ stack the $d$-dimensional
representations of the $T$ context positions. A single attention head projects
$X$ into queries, keys, and values, $Q = X W_Q$, $K = X W_K$, $V = X W_V$ with
$W_Q, W_K \in \mathbb{R}^{d \times d_k}$ and $W_V \in \mathbb{R}^{d \times d_v}$,
and computes

$$
\mathrm{Attn}(Q, K, V) = \mathrm{softmax}\!\left(
\frac{Q K^\top}{\sqrt{d_k}} + M \right) V,
$$ {#eq-llms-attention}

where the softmax is applied row-wise. The causal mask $M$ has $M_{tj} = 0$ for
$j \le t$ and $M_{tj} = -\infty$ for $j > t$, which drives the softmax weight on
future positions to zero and so enforces $p(x_t \mid x_{<t})$ depending only on
the past. The $\sqrt{d_k}$ scaling keeps the dot products $q_t^\top k_j$ at unit
variance: if the entries of $q_t$ and $k_j$ are independent with variance $1$,
then $\mathrm{Var}(q_t^\top k_j) = d_k$, and dividing by $\sqrt{d_k}$ rescales to
variance $1$, preventing the softmax from saturating into near one-hot vectors
with vanishing gradients. Multi-head attention runs $h$ such heads in parallel on
$d_k = d/h$-dimensional subspaces and concatenates the outputs, letting different
heads specialize on different relations.

The computational signature of this mechanism is the one that governs context
length. Forming $Q K^\top$ costs $O(T^2 d_k)$ time and $O(T^2)$ memory for the
attention matrix, quadratic in sequence length $T$. This quadratic term is why
context windows are expensive to extend and why a large body of work pursues
linear-attention and sparse-attention approximations.


::: {.callout-tip title="Intuition"}
Self-attention lets every word "look back" at every earlier word
in one step and decide which ones matter, instead of squeezing all of history
through a single running hidden state. Because the lookups happen in parallel,
these models train efficiently on the huge corpora that scaling demands.
:::


### Tokenization

Models do not operate on raw characters or whole words. They operate on
tokens produced by a subword tokenizer. Two common schemes are **Byte Pair
Encoding** (BPE), which starts from characters and greedily merges the most
frequent adjacent pairs into a fixed-size vocabulary, and WordPiece, which
makes similar merges guided by likelihood. Subword tokenization keeps the
vocabulary small while still representing rare and novel words by composing them
from known pieces. Every probability and perplexity figure is defined relative to
this token vocabulary, which is why comparisons require a shared tokenizer.

## Pretraining and the Foundation-Model Paradigm

The defining recipe is pretrain then adapt. A model is first pretrained once,
at great expense, on a broad corpus using the next-token objective. The result is
a general-purpose foundation model. It is then adapted, cheaply, to many
downstream tasks. The cost of learning language is paid once and amortized across
every later use.


::: {.callout-tip title="When to use this"}
The foundation-model pattern pays off whenever you have a
task with limited labeled data but plenty of related unlabeled text. Instead of
training from scratch, you start from a model that already understands language
and nudge it toward your task. This is the text analogue of the transfer
learning idea (@sec-transfer-multitask-learning) used with pretrained image
networks.
:::


### Scaling Laws

A central empirical finding is that the loss of a Transformer language model falls
smoothly and predictably as you increase three quantities: the number of model
parameters, the size of the training dataset, and the amount of compute. Kaplan
et al. (2020) reported that test loss follows approximate power-law relationships
in each of these factors over many orders of magnitude. This means you can
forecast how much better a larger model will be before training it, which turns
model design into something closer to an engineering calculation.^[A power law
means the loss falls by a roughly constant fraction each time you multiply a
resource by a constant factor. Plotted on log-log axes it looks like a straight
line, which is what makes extrapolation feasible.]

#### The parametric form of the scaling law {.unnumbered}

The Chinchilla analysis makes the power laws explicit. Writing $N$ for the number
of parameters and $D$ for the number of training tokens, the test loss is fit by

$$
L(N, D) = E + \frac{A}{N^{\alpha}} + \frac{B}{D^{\beta}},
$$ {#eq-llms-scaling}

with empirical exponents near $\alpha \approx \beta \approx 0.34$ and $E$ the
irreducible loss (the entropy floor $H(p^\star)$ of @eq-llms-ce-decomp). The
two additive terms are the finite-model and finite-data penalties; each falls as
a straight line on log-log axes, which is what "power law" means. The
compute-optimal allocation follows by constrained optimization. Training compute
is approximately $C \approx 6 N D$ (the factor $6$ counts the forward and
backward floating-point operations per parameter per token). Minimizing
@eq-llms-scaling subject to $6ND = C$ with a Lagrange multiplier gives the
stationarity condition $\alpha A / N^{\alpha} = \beta B / D^{\beta}$, whose
solution scales as $N^\star \propto C^{\,\beta/(\alpha+\beta)}$ and $D^\star
\propto C^{\,\alpha/(\alpha+\beta)}$. With $\alpha \approx \beta$ both exponents
are near $1/2$, so the optimum grows $N$ and $D$ in near-equal proportion. This
is the formal content of the Chinchilla rule: doubling the compute budget should
roughly double both model size and data, not pour it all into parameters.

Hoffmann et al. (2022), the Chinchilla study, refined this picture for a
fixed compute budget. The earlier work tended to make models very large relative
to the data they were trained on. Chinchilla showed that, for a given amount of
compute, parameters and training tokens should be scaled roughly in proportion,
so many models of that era were undertrained: smaller models trained on more data
matched or beat larger models trained on less. The practical lesson is that data
and model size must be balanced, not just made big.


::: {.callout-tip}
If you ever have a fixed compute budget for training a model from
scratch, the Chinchilla finding is the rule to remember: spend it on a smaller
model fed more data rather than a giant model that never sees enough text.
:::


### Emergent In-Context Learning

A striking property of large models is in-context learning. Without any
gradient updates, a sufficiently large pretrained model can perform a new task
when the task is described or demonstrated inside its input. You provide a few
input-output examples in the prompt (few-shot), and the model continues the
pattern. This behavior tends to appear or sharpen as scale increases and is not
explicitly trained for; it falls out of next-token prediction on diverse text.
It is the mechanism behind much of how LLMs are used in practice today.

## Adaptation Methods

Once you have a pretrained foundation model, several routes turn it into a useful
predictor for your problem. They differ in how many parameters they change and
how much labeled data and compute they require, and they form a ladder from
heaviest to lightest: full fine-tuning changes everything, parameter-efficient
methods change a sliver, and prompting changes nothing at all. The right rung
depends on how much labeled data, compute, and per-task isolation you need.

### Full Fine-Tuning

The most direct method updates all of the model's parameters by continuing
training on your labeled task data. It can reach high accuracy but is expensive in
memory and compute, and it produces a full separate copy of the model for each
task.

### Parameter-Efficient Fine-Tuning

Parameter-efficient fine-tuning (PEFT) freezes most of the pretrained weights
and trains only a small number of new ones. Adapters insert small trainable
layers between existing layers. LoRA (Low-Rank Adaptation) freezes the
original weight matrices and learns a low-rank update for each, so the number of
trainable parameters drops by orders of magnitude while accuracy stays close to
full fine-tuning. PEFT makes it feasible to maintain many task-specific
adaptations of one shared base model.

#### The LoRA reparameterization {.unnumbered}

Make the LoRA construction precise. Let $W_0 \in \mathbb{R}^{d \times k}$ be a
frozen pretrained weight matrix. Full fine-tuning would learn an updated
$W_0 + \Delta W$ with $\Delta W$ a dense $d \times k$ matrix, that is $dk$ free
parameters. LoRA constrains the update to low rank by writing

$$
W = W_0 + \Delta W = W_0 + \frac{\alpha}{r}\, B A,
\qquad B \in \mathbb{R}^{d \times r}, \; A \in \mathbb{R}^{r \times k},
$$ {#eq-llms-lora}

with rank $r \ll \min(d, k)$. Only $A$ and $B$ are trained, so the parameter
count drops from $dk$ to $r(d + k)$, a reduction by a factor of roughly
$dk / (r(d+k))$, which for typical $d = k = 4096$ and $r = 8$ is over two hundred
fold. The scalar $\alpha/r$ decouples the update magnitude from the rank so that
tuning $\alpha$ does not require retuning the learning rate as $r$ changes. At
initialization $A$ is random and $B = 0$, hence $\Delta W = 0$, so training
starts exactly at the pretrained model. The forward pass is $h = W_0 x + (\alpha/r)
B(Ax)$: the low-rank branch is computed separately and added, costing only
$O(r(d+k))$ extra multiply-adds. Because $\Delta W$ can be folded back into $W_0$
after training ($W \leftarrow W_0 + (\alpha/r)BA$), LoRA adds no inference latency,
and a single frozen $W_0$ can be served with different $(A, B)$ pairs swapped in
per task. The implicit assumption is that the task-specific adaptation has low
intrinsic rank, which holds empirically for fine-tuning but can fail when the
target task differs sharply from pretraining, in which case larger $r$ or full
fine-tuning is needed.


::: {.callout-tip title="When to use this"}
Reach for PEFT (LoRA in particular) when you want
fine-tuned quality but cannot afford to store or serve a full model copy per
task. You keep one frozen base model and a tiny set of swappable adapters, one
per task.
:::


### Prompting and Few-Shot Learning

The lightest-weight adaptation changes nothing in the model. You craft an input
(a prompt) that frames the task, optionally including a handful of worked
examples (few-shot) or none at all (zero-shot). This relies on the
in-context learning described above. It is fast to try and requires no training,
though it is sensitive to how the prompt is written.^[Because no weights change,
prompting is the cheapest thing to try first. Its main cost is that small changes
in wording, example order, or formatting can swing results, so prompts deserve
the same careful evaluation as any other modeling choice.]

### Instruction Tuning

A raw pretrained model predicts plausible continuations, which is not the same as
following a user's request. Instruction tuning fine-tunes the model on a
collection of tasks phrased as instructions paired with desired responses. The
result generalizes to following new instructions it was not explicitly trained
on, making the model far more usable as a general assistant.

### Reinforcement Learning from Human Feedback

Reinforcement learning from human feedback (RLHF) aligns model outputs with
human preferences in two conceptual stages. First, humans compare pairs of model
responses, and a reward model is trained to predict which response a human
would prefer. Second, the language model (the policy) is optimized to produce
responses that score highly under that reward model, typically with a policy
optimization algorithm, while a penalty keeps it from drifting too far from the
original model. Ouyang et al. (2022), the InstructGPT work, showed that this
procedure can make a smaller aligned model more helpful than a much larger
unaligned one. The takeaway is that alignment to preferences, not just scale,
drives perceived quality.

#### The reward model and the Bradley-Terry likelihood {.unnumbered}

The two stages have clean objectives. In stage one, a human ranks a pair of
responses $(y_w, y_l)$ to a prompt $x$, where $y_w$ is preferred to $y_l$. The
reward model $r_\phi(x, y)$ is fit under the Bradley-Terry model of pairwise
comparison, which posits

$$
P(y_w \succ y_l \mid x)
= \frac{\exp r_\phi(x, y_w)}{\exp r_\phi(x, y_w) + \exp r_\phi(x, y_l)}
= \sigma\!\big(r_\phi(x, y_w) - r_\phi(x, y_l)\big),
$$ {#eq-llms-bt}

with $\sigma$ the logistic function. Maximum likelihood over the preference
dataset minimizes $-\mathbb{E}\big[\log \sigma(r_\phi(x,y_w) - r_\phi(x,y_l))\big]$,
which is ordinary logistic regression on reward differences. Note that $r_\phi$
is identified only up to an additive constant per prompt, since the difference is
all that enters.

#### The KL-regularized policy objective {.unnumbered}

In stage two, the policy $\pi_\theta$ (the language model) is optimized to
maximize expected reward while staying near the supervised reference policy
$\pi_{\mathrm{ref}}$,

$$
\max_{\pi_\theta} \;
\mathbb{E}_{x,\, y \sim \pi_\theta(\cdot \mid x)}
\big[ r_\phi(x, y) \big]
- \beta \, D_{\mathrm{KL}}\!\big(\pi_\theta(\cdot \mid x) \,\|\,
\pi_{\mathrm{ref}}(\cdot \mid x)\big).
$$ {#eq-llms-rlhf}

The KL penalty with coefficient $\beta$ keeps the policy from collapsing onto
whatever degenerate output maximizes an imperfect reward model (reward hacking)
and from forgetting the fluency learned in pretraining. This objective has a
closed-form optimum. Treating each prompt separately and maximizing pointwise
over the distribution $\pi(\cdot \mid x)$ subject to normalization, the optimal
policy is the reference tilted by the exponentiated reward,

$$
\pi^\star(y \mid x) = \frac{1}{Z(x)}\,
\pi_{\mathrm{ref}}(y \mid x)\, \exp\!\Big(\tfrac{1}{\beta} r_\phi(x, y)\Big),
\qquad
Z(x) = \sum_{y} \pi_{\mathrm{ref}}(y \mid x)\, \exp\!\Big(\tfrac{1}{\beta} r_\phi(x, y)\Big).
$$ {#eq-llms-rlhf-opt}

In practice the partition function $Z(x)$ is intractable to sum over all
sequences, so proximal policy optimization (PPO) approximates this solution by
gradient ascent. Recognizing the closed form @eq-llms-rlhf-opt and inverting
it to express the reward in terms of the policy is precisely the insight behind
direct preference optimization (DPO), which substitutes
$r_\phi = \beta \log(\pi_\theta/\pi_{\mathrm{ref}}) + \beta \log Z(x)$ into the
Bradley-Terry likelihood @eq-llms-bt; the $\log Z(x)$ terms cancel in the
difference, yielding a supervised classification loss on preference pairs that
avoids reinforcement learning altogether.


::: {.callout-note}
Instruction tuning and RLHF are why a chat assistant feels so
different from a raw pretrained model. The base model only predicts plausible
text; these two steps teach it to predict text that a person actually asked for
and would rate highly.
:::


## Retrieval, Context, and Risks for Prediction

Adaptation makes a model useful; this section covers what still limits it once
deployed. A model's knowledge is fixed at training time, it can only read a
bounded amount of input, and it can state falsehoods fluently. The tools and
cautions below address each of those facts, and they matter most precisely when
you are using an LLM as a predictor whose outputs feed real decisions.

### Retrieval-Augmented Generation

A pretrained model's knowledge is frozen at training time and stored implicitly in
its weights. Retrieval-augmented generation (RAG) addresses this by fetching
relevant documents from an external store at query time and placing them in the
model's input, so the model can condition on up-to-date or proprietary
information it never memorized. RAG separates knowledge (in the retrievable store)
from reasoning (in the model), which makes facts easier to update and to cite. The
retrieval-augmented generation chapter (@sec-retrieval-augmented-generation)
develops this pattern in depth.


::: {.callout-tip title="When to use this"}
RAG is the standard answer when a model must work with
information it could not have memorized: documents written after training,
private internal data, or facts that change often. You update the document
store, not the model weights.
:::


### Context Windows

A model can only condition on a bounded number of tokens at once, its **context
window**. Anything outside the window is invisible to the model. Window sizes have
grown substantially, but they remain finite, and very long contexts cost more
compute and can dilute the model's attention to the most relevant parts. Context
length sets a hard limit on how much retrieved text, history, or document you can
feed in at once.

### Hallucination and Factuality

Because the model is trained to produce fluent, probable continuations, it can
generate statements that are confident and well-formed but false. This is called
hallucination. The model has no built-in mechanism to check claims against
ground truth, so fluency is not evidence of accuracy. For any prediction task
where correctness matters, outputs must be verified, and grounding the model with
retrieval reduces but does not eliminate the problem.


::: {.callout-warning}
Confidence and correctness are independent in an LLM. A smooth,
assured-sounding answer can be entirely wrong. Never treat fluency as a proxy
for truth, and build a verification step into any pipeline where mistakes have
consequences.
:::


### Calibration and Evaluation

For prediction use, two properties deserve scrutiny. Calibration
(@sec-probability-calibration) asks whether the model's stated or implied
confidence matches its actual accuracy; a well-calibrated model that says it is
70 percent sure is right about 70 percent of the time. Alignment procedures such as RLHF can improve helpfulness while
degrading calibration, so a model may sound more certain than it should.
Factuality asks whether claims are true. Evaluation therefore goes beyond
perplexity: held-out task accuracy, calibration curves, and task-specific
benchmarks are all needed, and human review remains important where stakes are
high.

### Limitations to Keep in Mind

LLMs are powerful but bounded. They can reflect biases present in their training
data, they are sensitive to prompt wording, they can be expensive to run, and
their behavior can change between versions. Treat them as one component in a
prediction pipeline whose outputs are measured and validated like any other
model, not as an oracle.

## A Runnable N-gram Language Model in Base R

The demonstration below builds everything from the chain-rule definition using
only base R. We define a small corpus inline, fit bigram and trigram models with
add-one smoothing, inspect a conditional next-token distribution, generate a
short sample, and compute perplexity on held-out text. This is the same machinery
as a large model in miniature: estimate $p(x_t \mid x_{<t})$, then use it to
score and to sample.


::: {.callout-important title="Key idea"}
Watch for one thread running through every chunk below. We first
estimate the conditional next-token distribution, then reuse that same estimate
two ways: sampling from it to generate text, and reading off its value on real
text to compute perplexity. A trillion-parameter model does nothing different in
kind; it only estimates that distribution far more accurately.
:::


```{r ngram-setup}
set.seed(123)

# A small training corpus defined inline. Each string is one "sentence".
train_corpus <- c(
  "the cat sat on the mat",
  "the dog sat on the log",
  "the cat ran to the dog",
  "the dog ran to the cat",
  "a cat and a dog sat on the mat",
  "the cat and the dog ran on the log",
  "the mat was on the floor",
  "the log was near the mat",
  "a dog sat near the cat",
  "the cat sat near the log"
)

# Held-out text the model never saw during training, for perplexity.
test_corpus <- c(
  "the cat sat on the log",
  "a dog ran to the mat"
)

# Tokenizer: lowercase, split on whitespace, and wrap each sentence with
# special start/end markers so the model can learn how sentences begin and end.
BOS <- "<s>"   # beginning of sentence
EOS <- "</s>"  # end of sentence

tokenize <- function(sentence) {
  words <- strsplit(tolower(trimws(sentence)), "\\s+")[[1]]
  c(BOS, words, EOS)
}

train_tokens <- lapply(train_corpus, tokenize)

# Vocabulary is every distinct token seen in training, including the markers.
vocab <- sort(unique(unlist(train_tokens)))
V <- length(vocab)
V
```

```{r ngram-fit}
# Count n-grams. For an n-gram model we need counts of the full n-gram and
# counts of the (n-1)-token context that precedes the final token.
ngram_counts <- function(token_lists, n) {
  context_counts <- list()  # count of each context
  ngram_counts   <- list()  # count of context + next token

  for (toks in token_lists) {
    if (length(toks) < n) next
    for (i in seq_len(length(toks) - n + 1)) {
      gram    <- toks[i:(i + n - 1)]
      context <- paste(gram[-n], collapse = " ")
      full    <- paste(gram, collapse = " ")
      context_counts[[context]] <-
        (if (is.null(context_counts[[context]])) 0 else context_counts[[context]]) + 1
      ngram_counts[[full]] <-
        (if (is.null(ngram_counts[[full]])) 0 else ngram_counts[[full]]) + 1
    }
  }
  list(context = context_counts, ngram = ngram_counts)
}

bigram_model  <- ngram_counts(train_tokens, 2)
trigram_model <- ngram_counts(train_tokens, 3)

# Add-one (Laplace) smoothed conditional probability p(word | context).
# Smoothing keeps unseen continuations from getting probability zero, which
# would make perplexity infinite the moment we hit a novel pair.
ngram_prob <- function(model, context, word, vocab_size) {
  full         <- paste(c(context, word), collapse = " ")
  ctx_key      <- paste(context, collapse = " ")
  count_full   <- if (is.null(model$ngram[[full]])) 0 else model$ngram[[full]]
  count_ctx    <- if (is.null(model$context[[ctx_key]])) 0 else model$context[[ctx_key]]
  (count_full + 1) / (count_ctx + vocab_size)
}

# Sanity check: probabilities over the whole vocabulary sum to 1 for any context.
probs_check <- sapply(vocab, function(w) ngram_prob(bigram_model, "the", w, V))
sum(probs_check)
```

@tbl-llms-ngram-table lists the eight most probable continuations of
the context word "the" under the smoothed bigram model.

```{r tbl-llms-ngram-table}
# Top next-token probabilities for the context "the" under the bigram model.
context_word <- "the"
next_probs <- sapply(vocab, function(w)
  ngram_prob(bigram_model, context_word, w, V))
next_probs <- sort(next_probs, decreasing = TRUE)

top_table <- data.frame(
  next_token  = names(next_probs),
  probability = round(as.numeric(next_probs), 4),
  row.names   = NULL
)
knitr::kable(
  head(top_table, 8),
  caption = "Top eight bigram continuations of the context word 'the' with their smoothed conditional probabilities."
)
```

@fig-llms-ngram-figure shows the smoothed bigram distribution over
the eight most likely continuations of the context word "the", the same
conditional next-token estimate that every step below reuses.

```{r fig-llms-ngram-figure, fig.cap = "Bigram conditional next-token distribution given the context word 'the' (top 8 continuations)."}
top8 <- head(top_table, 8)
op <- par(mar = c(7, 4, 3, 1))
barplot(
  height    = top8$probability,
  names.arg = top8$next_token,
  las       = 2,
  col       = "steelblue",
  ylab      = "p(next token | 'the')",
  main      = "Bigram distribution after 'the'"
)
par(op)
```

```{r ngram-generate}
# Generate text by sampling one token at a time from the bigram model,
# starting at the BOS marker and stopping at EOS or a length cap.
generate_text <- function(model, vocab, vocab_size, max_len = 15) {
  sentence <- c(BOS)
  for (i in seq_len(max_len)) {
    context <- sentence[length(sentence)]
    probs   <- sapply(vocab, function(w)
      ngram_prob(model, context, w, vocab_size))
    probs   <- probs / sum(probs)
    nxt     <- sample(vocab, size = 1, prob = probs)
    if (nxt == EOS) break
    sentence <- c(sentence, nxt)
  }
  paste(sentence[sentence != BOS & sentence != EOS], collapse = " ")
}

generated <- replicate(3, generate_text(bigram_model, vocab, V))
generated
```

@tbl-llms-ngram-perplexity reports the held-out perplexity of the
three models, and the same numbers are drawn as bars in the figure that follows.

```{r tbl-llms-ngram-perplexity}
# Perplexity on held-out text: exp of the average negative log-likelihood,
# exactly the definition from the start of the chapter.
perplexity <- function(model, n, test_lists, vocab_size) {
  total_logprob <- 0
  total_tokens  <- 0
  for (toks in test_lists) {
    if (length(toks) < n) next
    for (i in seq_len(length(toks) - n + 1)) {
      gram    <- toks[i:(i + n - 1)]
      context <- gram[-n]
      word    <- gram[n]
      p       <- ngram_prob(model, context, word, vocab_size)
      total_logprob <- total_logprob + log(p)
      total_tokens  <- total_tokens + 1
    }
  }
  exp(-total_logprob / total_tokens)
}

test_tokens <- lapply(test_corpus, tokenize)

ppl_bigram  <- perplexity(bigram_model,  2, test_tokens, V)
ppl_trigram <- perplexity(trigram_model, 3, test_tokens, V)

# A unigram baseline: ignore context entirely and use smoothed token frequencies.
unigram_counts <- table(unlist(train_tokens))
unigram_prob <- function(word, vocab_size) {
  c_w <- if (is.na(unigram_counts[word])) 0 else unigram_counts[word]
  (c_w + 1) / (sum(unigram_counts) + vocab_size)
}
ppl_unigram <- {
  lp <- 0; nt <- 0
  for (toks in test_tokens) {
    body <- toks[toks != BOS]
    for (w in body) { lp <- lp + log(unigram_prob(w, V)); nt <- nt + 1 }
  }
  exp(-lp / nt)
}

knitr::kable(
  data.frame(
    model      = c("unigram", "bigram", "trigram"),
    perplexity = round(c(ppl_unigram, ppl_bigram, ppl_trigram), 3)
  ),
  caption = "Held-out perplexity of the unigram, bigram, and trigram models on the test corpus."
)
```

@fig-llms-ngram-ppl-figure compares held-out perplexity across the
unigram, bigram, and trigram models, making the effect of conditioning on more
context visible at a glance.

```{r fig-llms-ngram-ppl-figure, fig.cap = "Held-out perplexity decreases as the n-gram order increases on this small corpus. Lower perplexity means the model is less surprised by the held-out text."}
ppl_vals <- c(unigram = ppl_unigram, bigram = ppl_bigram, trigram = ppl_trigram)
op <- par(mar = c(4, 4, 3, 1))
barplot(
  height    = ppl_vals,
  col       = c("grey70", "steelblue", "darkorange"),
  ylab      = "Held-out perplexity",
  xlab      = "Model order",
  main      = "Perplexity vs n-gram order"
)
par(op)
```

The pattern is the one to remember. Conditioning on more context (moving from
unigram to bigram to trigram) lowers perplexity because the model is less
surprised by real text, which is exactly the signal that drives the much larger
neural and Transformer models in the rest of this part of the book. The only
things that change at scale are how the conditional distribution
$p(x_t \mid x_{<t})$ is parameterized and how much data and compute go into
estimating it.

## Code You Would Run with Real Models

The base R demo above is for understanding. In practice you would call a
pretrained model through a library. The snippets below are not executed here,^[They
require the Python `transformers` and `peft` packages plus a downloaded model and
a deep-learning backend, none of which are part of this book's build, so they are
marked `eval = FALSE`.] but they show the shape of the workflow. Notice that the
quantities are the same ones we defined by hand: the library returns a
cross-entropy `loss`, and perplexity is just `exp(loss)`.

```{python, eval = FALSE}
# Scoring text and generating with a pretrained Transformer decoder (Python).
from transformers import AutoTokenizer, AutoModelForCausalLM
import torch

tokenizer = AutoTokenizer.from_pretrained("gpt2")
model = AutoModelForCausalLM.from_pretrained("gpt2")

text = "The cat sat on the"
inputs = tokenizer(text, return_tensors="pt")

with torch.no_grad():
    out = model(**inputs, labels=inputs["input_ids"])

# Cross-entropy loss, then perplexity = exp(loss), the same definition used above.
loss = out.loss.item()
perplexity = torch.exp(out.loss).item()

# Greedy continuation.
generated = model.generate(**inputs, max_new_tokens=10)
print(tokenizer.decode(generated[0]))
```

```{r, eval = FALSE}
# A parameter-efficient fine-tuning sketch (conceptual, Python via reticulate
# or run directly in Python). LoRA freezes the base weights and learns small
# low-rank updates, so only a tiny fraction of parameters are trained.
# from peft import LoraConfig, get_peft_model
# config = LoraConfig(r = 8, lora_alpha = 16, target_modules = c("q_proj", "v_proj"))
# model  = get_peft_model(base_model, config)
# Train `model` on your labeled task as usual; only the adapters update.
```

## Further Reading

- Vaswani, A., Shazeer, N., Parmar, N., Uszkoreit, J., Jones, L., Gomez, A. N.,
  Kaiser, L., and Polosukhin, I. (2017). Attention Is All You Need.
  *Advances in Neural Information Processing Systems*.
- Brown, T. B., et al. (2020). Language Models Are Few-Shot Learners (GPT-3).
  *Advances in Neural Information Processing Systems*.
- Kaplan, J., McCandlish, S., Henighan, T., Brown, T. B., Chess, B., Child, R.,
  Gray, S., Radford, A., Wu, J., and Amodei, D. (2020). Scaling Laws for Neural
  Language Models. *arXiv preprint*.
- Hoffmann, J., et al. (2022). Training Compute-Optimal Large Language Models
  (Chinchilla). *Advances in Neural Information Processing Systems*.
- Ouyang, L., et al. (2022). Training Language Models to Follow Instructions with
  Human Feedback (InstructGPT). *Advances in Neural Information Processing
  Systems*.
