# Large Language Models and Foundation Models

```{r}
#| include: false
suppressPackageStartupMessages(library(tidyverse))
```

A language model assigns probabilities to sequences of text. For a
prediction-focused course, the value is direct: once you can score how likely a
piece of text is, you can predict the next token, fill in blanks, classify
documents by comparing likelihoods, or generate text by sampling. **Large
language models** (LLMs) are the same idea scaled up by orders of magnitude in
data, parameters, and compute, trained once and then reused across many tasks.
That reuse pattern is what people mean by a foundation model.

This chapter builds the concept from the ground up. We start with what a language
model is mathematically, trace the path from simple counting models to
Transformer decoders, describe how these models are pretrained and then adapted,
and end with a runnable base R demonstration that builds a small n-gram language
model, generates text, and measures perplexity.

The reassuring news for anyone coming from the rest of this book is that nothing
here is conceptually new at its core. A language model is a conditional
probability estimator, the same object you have fit all along, just pointed at
text and scaled up dramatically. By the end you will be able to say precisely
what an LLM computes, why "next-token prediction" is enough to learn a great deal
about language, how perplexity measures a model's quality, and how a pretrained
model gets adapted to a specific task. The small n-gram model at the end makes
all of this concrete: it is a complete language model you can read line by line,
and it behaves, in miniature, exactly like the giants.

> **Intuition:** If you can estimate "given what came before, what comes next?"
> then prediction, classification, fill-in-the-blank, and generation are all the
> same machine used in different ways. Everything in this chapter is a variation
> on that single estimate.

## What a Language Model Is

Let a piece of text be a sequence of tokens $x_1, x_2, \dots, x_T$. A language
model defines a probability distribution over such sequences. Using the chain
rule of probability, any joint distribution factorizes exactly into a product of
conditional distributions:

$$
p(x_1, x_2, \dots, x_T) = \prod_{t=1}^{T} p(x_t \mid x_1, \dots, x_{t-1})
= \prod_{t=1}^{T} p(x_t \mid x_{<t}),
$$

where $x_{<t}$ denotes all tokens before position $t$. This factorization is not
an approximation. It is the definition of a joint distribution written one token
at a time. A model that estimates each factor $p(x_t \mid x_{<t})$ is called an
autoregressive language model, because it predicts the next token from the
tokens it has already seen.^[A token is the unit a model reads and predicts.
For now you can think of it as roughly a word; the section on tokenization below
explains why real models use subword pieces instead.]

> **Key idea:** The chain rule turns the impossible-looking task of modeling whole
> documents into a sequence of ordinary one-step prediction problems. Estimate
> $p(x_t \mid x_{<t})$ well at every position and you have, by construction, a
> model of the entire sequence.

### Next-Token Prediction

Training an autoregressive model is a supervised learning problem in disguise.
The input is the context $x_{<t}$, the target is the next token $x_t$, and the
labels come for free from the text itself. No human annotation is needed, which
is what makes it possible to train on enormous corpora. The model outputs a
probability distribution over the vocabulary $V$ at each position, and we want
that distribution to place high probability on the token that actually came next.

> **Note:** This is why LLMs train on text scraped from the open web, books, and
> code without paying for labels. The "label" for each position is simply the
> word that already follows it. Self-supervision of this kind is what makes
> training on trillions of tokens possible.

### Cross-Entropy Loss

The standard training objective is the cross-entropy loss, which is the
average negative log-likelihood of the observed tokens under the model. For a
single sequence,

$$
\mathcal{L} = -\frac{1}{T} \sum_{t=1}^{T} \log p(x_t \mid x_{<t}).
$$

Minimizing this loss is the same as maximizing the likelihood the model assigns
to real text. The logarithm is taken in natural base here, though base 2 is also
common (the two differ only by a constant factor of $\log 2$).^[Negative
log-likelihood is the same loss you have seen for logistic regression and softmax
classifiers elsewhere in the book. A language model is, at each position, a
classifier over the vocabulary, so it inherits exactly that objective.]

### Perplexity

The headline evaluation metric for a language model is perplexity (PPL),
defined as the exponential of the average negative log-likelihood:

$$
\mathrm{PPL} = \exp\!\left( -\frac{1}{T} \sum_{t=1}^{T} \log p(x_t \mid x_{<t}) \right)
= \exp(\mathcal{L}).
$$

Perplexity has a clean interpretation. If a model has perplexity $k$ on some
text, it is on average as uncertain as if it had to choose uniformly among $k$
equally likely tokens at each step. A perplexity of 1 means perfect prediction;
a perplexity equal to the vocabulary size means the model learned nothing beyond
a uniform guess. Lower is better. Because perplexity is a deterministic function
of cross-entropy, reducing the loss and reducing perplexity are the same goal.

> **Intuition:** Read perplexity as an "effective number of guesses." A perplexity
> of 5 means the model is about as confused as someone flipping fairly between 5
> options at each word. Cutting perplexity in half means the model has genuinely
> narrowed down its choices.

> **Warning:** Perplexity is only comparable across models that share the same
> tokenization and vocabulary. A model that splits text into more, smaller tokens
> faces an easier per-token prediction and can post a lower perplexity without
> being a better model. Never compare perplexity numbers across different
> tokenizers.

## From N-grams to Transformer Decoders

The definitions above tell us what to estimate; they say nothing about how. The
hard part of $p(x_t \mid x_{<t})$ is that the context $x_{<t}$ grows without
bound: by the thousandth word, the model must condition on the previous nine
hundred ninety-nine. No method can store a separate probability for every
possible history, so every model family makes a compromise in how it summarizes
that context. The history of language modeling is largely the history of better
and better summaries, and the three families below mark the main steps.

### N-gram Models

The oldest approach truncates the context to the previous $n-1$ tokens, an
$n$-th order Markov assumption:

$$
p(x_t \mid x_{<t}) \approx p(x_t \mid x_{t-n+1}, \dots, x_{t-1}).
$$

A bigram model ($n = 2$) conditions on one previous token, a trigram
($n = 3$) on two. The conditional probabilities are estimated by counting how
often each continuation follows each context in a training corpus and
normalizing. N-gram models are fast and transparent, but they suffer from data
sparsity (most long contexts are never seen) and they cannot capture
dependencies beyond the fixed window. Smoothing techniques such as add-one
(Laplace) smoothing are used to avoid assigning zero probability to unseen
continuations.^[Without smoothing, a single unseen continuation makes the
sequence probability zero and the perplexity infinite. Add-one smoothing pretends
every continuation was seen one extra time, which guarantees nonzero
probabilities everywhere.] The runnable demo at the end of the chapter builds
exactly this kind of model, so the summary here will become concrete shortly.

### Neural Language Models

Instead of counting discrete contexts, a neural language model maps each
token to a dense vector (an embedding) and uses a neural network to predict the
next token. Early versions used feed-forward networks over a fixed window;
recurrent networks (RNNs, LSTMs) then allowed the context to be summarized into a
hidden state that is carried forward step by step, in principle removing the
fixed-window limit. Recurrent models struggle with very long dependencies and are
hard to parallelize across positions, because each step must wait for the
previous one to finish.

> **Key idea:** Neural models replace counting with learning. Similar contexts
> map to similar vectors, so the model can generalize to histories it never saw,
> the exact weakness that sinks n-grams.

### Transformer Decoders

Modern LLMs are Transformer decoders. Rather than a recurrent hidden state,
they use self-attention so that each position can read directly from every
earlier position in a single step. To preserve the autoregressive property, the
attention is causally masked: position $t$ may attend to positions
$1, \dots, t$ but not to anything in the future. Stacking many masked
self-attention and feed-forward layers gives a model that predicts
$p(x_t \mid x_{<t})$ with a rich, learned summary of the entire prior context.
The attention mechanism, multi-head attention, and positional information are
covered in the Attention and Transformers material and in the chapter on BERT,
which uses the encoder half of the same architecture. The decoder used by LLMs is
that architecture with causal masking and a next-token objective.

> **Intuition:** Self-attention lets every word "look back" at every earlier word
> in one step and decide which ones matter, instead of squeezing all of history
> through a single running hidden state. Because the lookups happen in parallel,
> these models train efficiently on the huge corpora that scaling demands.

### Tokenization

Models do not operate on raw characters or whole words. They operate on
tokens produced by a subword tokenizer. Two common schemes are **Byte Pair
Encoding** (BPE), which starts from characters and greedily merges the most
frequent adjacent pairs into a fixed-size vocabulary, and WordPiece, which
makes similar merges guided by likelihood. Subword tokenization keeps the
vocabulary small while still representing rare and novel words by composing them
from known pieces. Every probability and perplexity figure is defined relative to
this token vocabulary, which is why comparisons require a shared tokenizer.

## Pretraining and the Foundation-Model Paradigm

The defining recipe is pretrain then adapt. A model is first pretrained once,
at great expense, on a broad corpus using the next-token objective. The result is
a general-purpose foundation model. It is then adapted, cheaply, to many
downstream tasks. The cost of learning language is paid once and amortized across
every later use.

> **When to use this:** The foundation-model pattern pays off whenever you have a
> task with limited labeled data but plenty of related unlabeled text. Instead of
> training from scratch, you start from a model that already understands language
> and nudge it toward your task. This is the text analogue of the transfer
> learning idea used with pretrained image networks.

### Scaling Laws

A central empirical finding is that the loss of a Transformer language model falls
smoothly and predictably as you increase three quantities: the number of model
parameters, the size of the training dataset, and the amount of compute. Kaplan
et al. (2020) reported that test loss follows approximate power-law relationships
in each of these factors over many orders of magnitude. This means you can
forecast how much better a larger model will be before training it, which turns
model design into something closer to an engineering calculation.^[A power law
means the loss falls by a roughly constant fraction each time you multiply a
resource by a constant factor. Plotted on log-log axes it looks like a straight
line, which is what makes extrapolation feasible.]

Hoffmann et al. (2022), the Chinchilla study, refined this picture for a
fixed compute budget. The earlier work tended to make models very large relative
to the data they were trained on. Chinchilla showed that, for a given amount of
compute, parameters and training tokens should be scaled roughly in proportion,
so many models of that era were undertrained: smaller models trained on more data
matched or beat larger models trained on less. The practical lesson is that data
and model size must be balanced, not just made big.

> **Tip:** If you ever have a fixed compute budget for training a model from
> scratch, the Chinchilla finding is the rule to remember: spend it on a smaller
> model fed more data rather than a giant model that never sees enough text.

### Emergent In-Context Learning

A striking property of large models is in-context learning. Without any
gradient updates, a sufficiently large pretrained model can perform a new task
when the task is described or demonstrated inside its input. You provide a few
input-output examples in the prompt (few-shot), and the model continues the
pattern. This behavior tends to appear or sharpen as scale increases and is not
explicitly trained for; it falls out of next-token prediction on diverse text.
It is the mechanism behind much of how LLMs are used in practice today.

## Adaptation Methods

Once you have a pretrained foundation model, several routes turn it into a useful
predictor for your problem. They differ in how many parameters they change and
how much labeled data and compute they require, and they form a ladder from
heaviest to lightest: full fine-tuning changes everything, parameter-efficient
methods change a sliver, and prompting changes nothing at all. The right rung
depends on how much labeled data, compute, and per-task isolation you need.

### Full Fine-Tuning

The most direct method updates all of the model's parameters by continuing
training on your labeled task data. It can reach high accuracy but is expensive in
memory and compute, and it produces a full separate copy of the model for each
task.

### Parameter-Efficient Fine-Tuning

Parameter-efficient fine-tuning (PEFT) freezes most of the pretrained weights
and trains only a small number of new ones. Adapters insert small trainable
layers between existing layers. LoRA (Low-Rank Adaptation) freezes the
original weight matrices and learns a low-rank update for each, so the number of
trainable parameters drops by orders of magnitude while accuracy stays close to
full fine-tuning. PEFT makes it feasible to maintain many task-specific
adaptations of one shared base model.

> **When to use this:** Reach for PEFT (LoRA in particular) when you want
> fine-tuned quality but cannot afford to store or serve a full model copy per
> task. You keep one frozen base model and a tiny set of swappable adapters, one
> per task.

### Prompting and Few-Shot Learning

The lightest-weight adaptation changes nothing in the model. You craft an input
(a prompt) that frames the task, optionally including a handful of worked
examples (few-shot) or none at all (zero-shot). This relies on the
in-context learning described above. It is fast to try and requires no training,
though it is sensitive to how the prompt is written.^[Because no weights change,
prompting is the cheapest thing to try first. Its main cost is that small changes
in wording, example order, or formatting can swing results, so prompts deserve
the same careful evaluation as any other modeling choice.]

### Instruction Tuning

A raw pretrained model predicts plausible continuations, which is not the same as
following a user's request. Instruction tuning fine-tunes the model on a
collection of tasks phrased as instructions paired with desired responses. The
result generalizes to following new instructions it was not explicitly trained
on, making the model far more usable as a general assistant.

### Reinforcement Learning from Human Feedback

Reinforcement learning from human feedback (RLHF) aligns model outputs with
human preferences in two conceptual stages. First, humans compare pairs of model
responses, and a reward model is trained to predict which response a human
would prefer. Second, the language model (the policy) is optimized to produce
responses that score highly under that reward model, typically with a policy
optimization algorithm, while a penalty keeps it from drifting too far from the
original model. Ouyang et al. (2022), the InstructGPT work, showed that this
procedure can make a smaller aligned model more helpful than a much larger
unaligned one. The takeaway is that alignment to preferences, not just scale,
drives perceived quality.

> **Note:** Instruction tuning and RLHF are why a chat assistant feels so
> different from a raw pretrained model. The base model only predicts plausible
> text; these two steps teach it to predict text that a person actually asked for
> and would rate highly.

## Retrieval, Context, and Risks for Prediction

Adaptation makes a model useful; this section covers what still limits it once
deployed. A model's knowledge is fixed at training time, it can only read a
bounded amount of input, and it can state falsehoods fluently. The tools and
cautions below address each of those facts, and they matter most precisely when
you are using an LLM as a predictor whose outputs feed real decisions.

### Retrieval-Augmented Generation

A pretrained model's knowledge is frozen at training time and stored implicitly in
its weights. Retrieval-augmented generation (RAG) addresses this by fetching
relevant documents from an external store at query time and placing them in the
model's input, so the model can condition on up-to-date or proprietary
information it never memorized. RAG separates knowledge (in the retrievable store)
from reasoning (in the model), which makes facts easier to update and to cite.

> **When to use this:** RAG is the standard answer when a model must work with
> information it could not have memorized: documents written after training,
> private internal data, or facts that change often. You update the document
> store, not the model weights.

### Context Windows

A model can only condition on a bounded number of tokens at once, its **context
window**. Anything outside the window is invisible to the model. Window sizes have
grown substantially, but they remain finite, and very long contexts cost more
compute and can dilute the model's attention to the most relevant parts. Context
length sets a hard limit on how much retrieved text, history, or document you can
feed in at once.

### Hallucination and Factuality

Because the model is trained to produce fluent, probable continuations, it can
generate statements that are confident and well-formed but false. This is called
hallucination. The model has no built-in mechanism to check claims against
ground truth, so fluency is not evidence of accuracy. For any prediction task
where correctness matters, outputs must be verified, and grounding the model with
retrieval reduces but does not eliminate the problem.

> **Warning:** Confidence and correctness are independent in an LLM. A smooth,
> assured-sounding answer can be entirely wrong. Never treat fluency as a proxy
> for truth, and build a verification step into any pipeline where mistakes have
> consequences.

### Calibration and Evaluation

For prediction use, two properties deserve scrutiny. Calibration asks whether
the model's stated or implied confidence matches its actual accuracy; a
well-calibrated model that says it is 70 percent sure is right about 70 percent of
the time. Alignment procedures such as RLHF can improve helpfulness while
degrading calibration, so a model may sound more certain than it should.
Factuality asks whether claims are true. Evaluation therefore goes beyond
perplexity: held-out task accuracy, calibration curves, and task-specific
benchmarks are all needed, and human review remains important where stakes are
high.

### Limitations to Keep in Mind

LLMs are powerful but bounded. They can reflect biases present in their training
data, they are sensitive to prompt wording, they can be expensive to run, and
their behavior can change between versions. Treat them as one component in a
prediction pipeline whose outputs are measured and validated like any other
model, not as an oracle.

## A Runnable N-gram Language Model in Base R

The demonstration below builds everything from the chain-rule definition using
only base R. We define a small corpus inline, fit bigram and trigram models with
add-one smoothing, inspect a conditional next-token distribution, generate a
short sample, and compute perplexity on held-out text. This is the same machinery
as a large model in miniature: estimate $p(x_t \mid x_{<t})$, then use it to
score and to sample.

> **Key idea:** Watch for one thread running through every chunk below. We first
> estimate the conditional next-token distribution, then reuse that same estimate
> two ways: sampling from it to generate text, and reading off its value on real
> text to compute perplexity. A trillion-parameter model does nothing different in
> kind; it only estimates that distribution far more accurately.

```{r ngram-setup}
set.seed(123)

# A small training corpus defined inline. Each string is one "sentence".
train_corpus <- c(
  "the cat sat on the mat",
  "the dog sat on the log",
  "the cat ran to the dog",
  "the dog ran to the cat",
  "a cat and a dog sat on the mat",
  "the cat and the dog ran on the log",
  "the mat was on the floor",
  "the log was near the mat",
  "a dog sat near the cat",
  "the cat sat near the log"
)

# Held-out text the model never saw during training, for perplexity.
test_corpus <- c(
  "the cat sat on the log",
  "a dog ran to the mat"
)

# Tokenizer: lowercase, split on whitespace, and wrap each sentence with
# special start/end markers so the model can learn how sentences begin and end.
BOS <- "<s>"   # beginning of sentence
EOS <- "</s>"  # end of sentence

tokenize <- function(sentence) {
  words <- strsplit(tolower(trimws(sentence)), "\\s+")[[1]]
  c(BOS, words, EOS)
}

train_tokens <- lapply(train_corpus, tokenize)

# Vocabulary is every distinct token seen in training, including the markers.
vocab <- sort(unique(unlist(train_tokens)))
V <- length(vocab)
V
```

```{r ngram-fit}
# Count n-grams. For an n-gram model we need counts of the full n-gram and
# counts of the (n-1)-token context that precedes the final token.
ngram_counts <- function(token_lists, n) {
  context_counts <- list()  # count of each context
  ngram_counts   <- list()  # count of context + next token

  for (toks in token_lists) {
    if (length(toks) < n) next
    for (i in seq_len(length(toks) - n + 1)) {
      gram    <- toks[i:(i + n - 1)]
      context <- paste(gram[-n], collapse = " ")
      full    <- paste(gram, collapse = " ")
      context_counts[[context]] <-
        (if (is.null(context_counts[[context]])) 0 else context_counts[[context]]) + 1
      ngram_counts[[full]] <-
        (if (is.null(ngram_counts[[full]])) 0 else ngram_counts[[full]]) + 1
    }
  }
  list(context = context_counts, ngram = ngram_counts)
}

bigram_model  <- ngram_counts(train_tokens, 2)
trigram_model <- ngram_counts(train_tokens, 3)

# Add-one (Laplace) smoothed conditional probability p(word | context).
# Smoothing keeps unseen continuations from getting probability zero, which
# would make perplexity infinite the moment we hit a novel pair.
ngram_prob <- function(model, context, word, vocab_size) {
  full         <- paste(c(context, word), collapse = " ")
  ctx_key      <- paste(context, collapse = " ")
  count_full   <- if (is.null(model$ngram[[full]])) 0 else model$ngram[[full]]
  count_ctx    <- if (is.null(model$context[[ctx_key]])) 0 else model$context[[ctx_key]]
  (count_full + 1) / (count_ctx + vocab_size)
}

# Sanity check: probabilities over the whole vocabulary sum to 1 for any context.
probs_check <- sapply(vocab, function(w) ngram_prob(bigram_model, "the", w, V))
sum(probs_check)
```

@tbl-llms-ngram-table lists the eight most probable continuations of
the context word "the" under the smoothed bigram model.

```{r tbl-llms-ngram-table}
# Top next-token probabilities for the context "the" under the bigram model.
context_word <- "the"
next_probs <- sapply(vocab, function(w)
  ngram_prob(bigram_model, context_word, w, V))
next_probs <- sort(next_probs, decreasing = TRUE)

top_table <- data.frame(
  next_token  = names(next_probs),
  probability = round(as.numeric(next_probs), 4),
  row.names   = NULL
)
knitr::kable(
  head(top_table, 8),
  caption = "Top eight bigram continuations of the context word 'the' with their smoothed conditional probabilities."
)
```

@fig-llms-ngram-figure shows the smoothed bigram distribution over
the eight most likely continuations of the context word "the", the same
conditional next-token estimate that every step below reuses.

```{r fig-llms-ngram-figure, fig.cap = "Bigram conditional next-token distribution given the context word 'the' (top 8 continuations)."}
top8 <- head(top_table, 8)
op <- par(mar = c(7, 4, 3, 1))
barplot(
  height    = top8$probability,
  names.arg = top8$next_token,
  las       = 2,
  col       = "steelblue",
  ylab      = "p(next token | 'the')",
  main      = "Bigram distribution after 'the'"
)
par(op)
```

```{r ngram-generate}
# Generate text by sampling one token at a time from the bigram model,
# starting at the BOS marker and stopping at EOS or a length cap.
generate_text <- function(model, vocab, vocab_size, max_len = 15) {
  sentence <- c(BOS)
  for (i in seq_len(max_len)) {
    context <- sentence[length(sentence)]
    probs   <- sapply(vocab, function(w)
      ngram_prob(model, context, w, vocab_size))
    probs   <- probs / sum(probs)
    nxt     <- sample(vocab, size = 1, prob = probs)
    if (nxt == EOS) break
    sentence <- c(sentence, nxt)
  }
  paste(sentence[sentence != BOS & sentence != EOS], collapse = " ")
}

generated <- replicate(3, generate_text(bigram_model, vocab, V))
generated
```

@tbl-llms-ngram-perplexity reports the held-out perplexity of the
three models, and the same numbers are drawn as bars in the figure that follows.

```{r tbl-llms-ngram-perplexity}
# Perplexity on held-out text: exp of the average negative log-likelihood,
# exactly the definition from the start of the chapter.
perplexity <- function(model, n, test_lists, vocab_size) {
  total_logprob <- 0
  total_tokens  <- 0
  for (toks in test_lists) {
    if (length(toks) < n) next
    for (i in seq_len(length(toks) - n + 1)) {
      gram    <- toks[i:(i + n - 1)]
      context <- gram[-n]
      word    <- gram[n]
      p       <- ngram_prob(model, context, word, vocab_size)
      total_logprob <- total_logprob + log(p)
      total_tokens  <- total_tokens + 1
    }
  }
  exp(-total_logprob / total_tokens)
}

test_tokens <- lapply(test_corpus, tokenize)

ppl_bigram  <- perplexity(bigram_model,  2, test_tokens, V)
ppl_trigram <- perplexity(trigram_model, 3, test_tokens, V)

# A unigram baseline: ignore context entirely and use smoothed token frequencies.
unigram_counts <- table(unlist(train_tokens))
unigram_prob <- function(word, vocab_size) {
  c_w <- if (is.na(unigram_counts[word])) 0 else unigram_counts[word]
  (c_w + 1) / (sum(unigram_counts) + vocab_size)
}
ppl_unigram <- {
  lp <- 0; nt <- 0
  for (toks in test_tokens) {
    body <- toks[toks != BOS]
    for (w in body) { lp <- lp + log(unigram_prob(w, V)); nt <- nt + 1 }
  }
  exp(-lp / nt)
}

knitr::kable(
  data.frame(
    model      = c("unigram", "bigram", "trigram"),
    perplexity = round(c(ppl_unigram, ppl_bigram, ppl_trigram), 3)
  ),
  caption = "Held-out perplexity of the unigram, bigram, and trigram models on the test corpus."
)
```

@fig-llms-ngram-ppl-figure compares held-out perplexity across the
unigram, bigram, and trigram models, making the effect of conditioning on more
context visible at a glance.

```{r fig-llms-ngram-ppl-figure, fig.cap = "Held-out perplexity decreases as the n-gram order increases on this small corpus. Lower perplexity means the model is less surprised by the held-out text."}
ppl_vals <- c(unigram = ppl_unigram, bigram = ppl_bigram, trigram = ppl_trigram)
op <- par(mar = c(4, 4, 3, 1))
barplot(
  height    = ppl_vals,
  col       = c("grey70", "steelblue", "darkorange"),
  ylab      = "Held-out perplexity",
  xlab      = "Model order",
  main      = "Perplexity vs n-gram order"
)
par(op)
```

The pattern is the one to remember. Conditioning on more context (moving from
unigram to bigram to trigram) lowers perplexity because the model is less
surprised by real text, which is exactly the signal that drives the much larger
neural and Transformer models in the rest of this part of the book. The only
things that change at scale are how the conditional distribution
$p(x_t \mid x_{<t})$ is parameterized and how much data and compute go into
estimating it.

## Code You Would Run with Real Models

The base R demo above is for understanding. In practice you would call a
pretrained model through a library. The snippets below are not executed here,^[They
require the Python `transformers` and `peft` packages plus a downloaded model and
a deep-learning backend, none of which are part of this book's build, so they are
marked `eval = FALSE`.] but they show the shape of the workflow. Notice that the
quantities are the same ones we defined by hand: the library returns a
cross-entropy `loss`, and perplexity is just `exp(loss)`.

```{python, eval = FALSE}
# Scoring text and generating with a pretrained Transformer decoder (Python).
from transformers import AutoTokenizer, AutoModelForCausalLM
import torch

tokenizer = AutoTokenizer.from_pretrained("gpt2")
model = AutoModelForCausalLM.from_pretrained("gpt2")

text = "The cat sat on the"
inputs = tokenizer(text, return_tensors="pt")

with torch.no_grad():
    out = model(**inputs, labels=inputs["input_ids"])

# Cross-entropy loss, then perplexity = exp(loss), the same definition used above.
loss = out.loss.item()
perplexity = torch.exp(out.loss).item()

# Greedy continuation.
generated = model.generate(**inputs, max_new_tokens=10)
print(tokenizer.decode(generated[0]))
```

```{r, eval = FALSE}
# A parameter-efficient fine-tuning sketch (conceptual, Python via reticulate
# or run directly in Python). LoRA freezes the base weights and learns small
# low-rank updates, so only a tiny fraction of parameters are trained.
# from peft import LoraConfig, get_peft_model
# config = LoraConfig(r = 8, lora_alpha = 16, target_modules = c("q_proj", "v_proj"))
# model  = get_peft_model(base_model, config)
# Train `model` on your labeled task as usual; only the adapters update.
```

## Further Reading

- Vaswani, A., Shazeer, N., Parmar, N., Uszkoreit, J., Jones, L., Gomez, A. N.,
  Kaiser, L., and Polosukhin, I. (2017). Attention Is All You Need.
  *Advances in Neural Information Processing Systems*.
- Brown, T. B., et al. (2020). Language Models Are Few-Shot Learners (GPT-3).
  *Advances in Neural Information Processing Systems*.
- Kaplan, J., McCandlish, S., Henighan, T., Brown, T. B., Chess, B., Child, R.,
  Gray, S., Radford, A., Wu, J., and Amodei, D. (2020). Scaling Laws for Neural
  Language Models. *arXiv preprint*.
- Hoffmann, J., et al. (2022). Training Compute-Optimal Large Language Models
  (Chinchilla). *Advances in Neural Information Processing Systems*.
- Ouyang, L., et al. (2022). Training Language Models to Follow Instructions with
  Human Feedback (InstructGPT). *Advances in Neural Information Processing
  Systems*.
