# Graph Neural Networks

```{r}
#| include: false
suppressPackageStartupMessages(library(tidyverse))
```

Most models in this book assume that observations are independent rows in a table. Each row carries its own features, and the model maps those features to a target. Many problems do not fit that picture. A user in a social network is described partly by who their friends are. A molecule's properties depend on how its atoms are bonded. A transaction looks normal in isolation but suspicious once you see the accounts it connects to. In these settings the relationships between observations carry as much signal as the observations themselves.

Graph Neural Networks (GNNs) are models that learn from data structured as a graph: a set of nodes (entities) connected by edges (relationships).^[A graph here is the mathematical object from graph theory, a collection of points and the links between them, not a chart or plot. The points are called nodes (or vertices) and the links are called edges.] The core idea is that a node's representation should be built from its own features plus information gathered from its neighbors. Stack a few rounds of this neighbor gathering and each node sees a widening region of the graph. This chapter develops the math of message passing, walks through the three architectures a practitioner meets first (GCN, GraphSAGE, GAT), implements one graph convolution layer in base R so the mechanics are concrete, and gives practical guidance including the over-smoothing failure mode.

> **Intuition:** Picture each node passing notes to its neighbors. In one round, you learn what your immediate friends know. In a second round, your friends have already absorbed what *their* friends know, so a piece of that reaches you too. A GNN is just this rumor-spreading process made differentiable, so the rules for what to pass and how to combine it can be learned from data.

> **What you will learn:** how to represent a graph as matrices, the single message-passing template that nearly every GNN layer follows, how GCN, GraphSAGE, and GAT each fill in that template, how to build a graph convolution by hand in base R, and the practical traps (over-smoothing, heterophily, scale) that separate a working GNN from a broken one.

## Where GNNs Fit in a Modern Workflow

In an applied ML or AI pipeline, GNNs occupy the slot you reach for when the relational structure is part of the data, not an afterthought. Typical entry points:

- Fraud and risk: accounts, devices, and transactions form a graph; fraud rings show up as suspicious subgraphs.
- Recommendation: users and items form a bipartite graph; predicting a missing edge is the recommendation.
- Molecules and materials: atoms are nodes, bonds are edges, and the task is to predict a property of the whole graph.
- Knowledge graphs and retrieval: entities and relations support link prediction and graph-aware retrieval that can feed a downstream LLM.
- Traffic, supply chains, and infrastructure: sensors and roads or warehouses and routes form a graph evolving over time.

A reasonable rule of thumb: if you can draw your data as nodes and edges and you believe a node's label depends on its neighborhood, a GNN is worth trying. If the relationships are weak or absent, a plain tabular model (gradient boosting, a feedforward network) is usually simpler and stronger. GNNs are not a replacement for those models; they are the relational extension of them.

> **When to use this:** reach for a GNN when the edges carry signal you would otherwise have to engineer by hand (counts of suspicious neighbors, average rating of co-purchased items, and so on). The GNN learns those neighborhood summaries for you. When the edges are noise or convenience, skip it.

## Graphs and Adjacency

A graph is $G = (V, E)$ with node set $V$ of size $n = |V|$ and edge set $E$. We store node features in a matrix $X \in \mathbb{R}^{n \times d}$, where row $i$ is the feature vector $x_i$ of node $i$ and $d$ is the number of input features.

The connectivity is encoded in the adjacency matrix $A \in \mathbb{R}^{n \times n}$:

$$
A_{ij} =
\begin{cases}
1 & \text{if there is an edge from } i \text{ to } j \\
0 & \text{otherwise}
\end{cases}
$$

For an undirected graph $A$ is symmetric, $A = A^\top$. Weighted graphs replace the $1$ with an edge weight. The degree of node $i$ is the number of its neighbors, $\deg(i) = \sum_j A_{ij}$, and the diagonal degree matrix $D$ has $D_{ii} = \deg(i)$ and zeros off the diagonal.

Two derived matrices appear constantly. The combinatorial graph Laplacian is

$$
L = D - A,
$$

and its symmetric normalized form is

$$
L_{\text{sym}} = I - D^{-1/2} A D^{-1/2}.
$$

The Laplacian measures how much a signal on the graph varies between connected nodes, and it is the bridge between GNNs and classical spectral graph theory.^[Concretely, the quadratic form $f^\top L f = \tfrac{1}{2}\sum_{i,j} A_{ij}(f_i - f_j)^2$ is small when the signal $f$ takes similar values on connected nodes. The normalization in $L_{\text{sym}}$ rescales this so that high-degree and low-degree nodes are compared on equal footing.]

> **Note:** You do not need spectral graph theory to use a GNN, but it explains *why* the normalization in the GCN layer below looks the way it does. We flag the connection and move on; the message-passing view in the next section is self-contained.

Before choosing an architecture, decide what you are predicting. Tasks on graphs come in three flavors, and the distinction drives everything downstream, from how you split the data to what the final layer outputs. Table @tbl-graph-neural-networks-task-levels summarizes the three task levels and what each one produces.

| Task level | What you predict | Example | Output shape |
|---|---|---|---|
| Node | a label per node | which users are bots | one prediction per node |
| Edge | whether or how two nodes link | will user $u$ buy item $v$ | one prediction per node pair |
| Graph | a label for the whole graph | is this molecule toxic | one prediction per graph |

: The three levels of prediction task on graphs, with an example and the output shape for each. {#tbl-graph-neural-networks-task-levels}

## Message Passing: The Common Framework

Almost every GNN layer is an instance of one pattern. Each node updates its representation by (1) collecting messages from its neighbors, (2) aggregating those messages into a single vector, and (3) combining the aggregate with its own current state. Let $h_i^{(k)}$ denote the representation of node $i$ after layer $k$, with $h_i^{(0)} = x_i$. Let $\mathcal{N}(i)$ be the neighbors of $i$. Then

$$
m_i^{(k)} = \underset{j \in \mathcal{N}(i)}{\text{AGGREGATE}} \left( \phi\!\left(h_i^{(k-1)}, h_j^{(k-1)}\right) \right),
$$

$$
h_i^{(k)} = \text{UPDATE}\!\left(h_i^{(k-1)}, m_i^{(k)}\right).
$$

Here $\phi$ builds a message from a node and one of its neighbors, AGGREGATE is a permutation-invariant pooling function (sum, mean, or max, because neighbors have no inherent order), and UPDATE merges the pooled message with the node's prior state, usually through a learned linear map and a nonlinearity.

> **Key idea:** AGGREGATE must be permutation invariant, meaning the answer cannot depend on the order you list the neighbors in. A node's neighbors form a set, not a sequence, so the only legal operations are ones like sum, mean, and max that ignore order. This single constraint is what distinguishes a GNN from feeding a node's neighbor list into an ordinary network.

The depth $k$ controls the receptive field. After $k$ layers, node $i$'s representation depends on every node within $k$ hops. This is the graph analogue of a convolutional network's growing receptive field over an image. The architectures below differ only in their choices of $\phi$, AGGREGATE, and UPDATE, so once you have this template the differences between GCN, GraphSAGE, and GAT are small and easy to keep straight.

## Graph Convolutional Network (GCN)

The GCN layer of Kipf and Welling (2017) is the most common starting point. Its propagation rule for a full layer is

$$
H^{(k)} = \sigma\!\left( \hat{A}\, H^{(k-1)} W^{(k)} \right),
$$

where $H^{(k-1)} \in \mathbb{R}^{n \times d_{k-1}}$ is the matrix of node representations entering the layer, $W^{(k)} \in \mathbb{R}^{d_{k-1} \times d_k}$ is a learnable weight matrix, $\sigma$ is an elementwise nonlinearity such as ReLU, and $\hat{A}$ is the symmetric normalized adjacency with self-loops.

The matrix $\hat{A}$ is built in two steps. First add self-loops so each node keeps its own information,

$$
\tilde{A} = A + I,
$$

with degree matrix $\tilde{D}$ where $\tilde{D}_{ii} = \sum_j \tilde{A}_{ij}$. Then normalize symmetrically,

$$
\hat{A} = \tilde{D}^{-1/2}\, \tilde{A}\, \tilde{D}^{-1/2}.
$$

Why this normalization? The product $\tilde{A} H$ alone is a sum over neighbors, so high-degree nodes would accumulate large magnitudes and dominate. Dividing by degrees rebalances this. The symmetric form $\tilde{D}^{-1/2} \tilde{A} \tilde{D}^{-1/2}$ scales entry $(i,j)$ by $1 / \sqrt{\deg(i)\,\deg(j)}$, which keeps the operator's spectrum bounded and stabilizes training. In message-passing terms, GCN uses a degree-weighted mean as AGGREGATE, a shared linear map as the message and update, and folds the node's own state in through the self-loop.

> **Warning:** If you forget the self-loop in $\tilde{A} = A + I$, each node throws away its own features and is described purely by its neighbors. This silently degrades accuracy and is one of the most common GNN implementation bugs. Always add self-loops before normalizing.

GCN is simple and fast, but the same weight matrix $W^{(k)}$ and the fixed normalization apply everywhere, so the model cannot weight some neighbors more than others, and it needs the whole adjacency matrix at training time (it is transductive in its basic form).^[Transductive means the model is trained on one fixed graph and can only score nodes that were present during training. The opposite, inductive, means the model can produce predictions for brand-new nodes or even entirely new graphs. GraphSAGE, below, is the inductive answer.]

## A GCN Layer in Base R

The propagation rule reduces to three matrix multiplications, so we can implement and inspect one layer with nothing but base R. No deep learning framework is needed to see the mechanism. We build a tiny 6-node graph with two communities loosely linked by a bridge, run one untrained GCN layer, and look at the resulting node embeddings.

> **Tip:** Reading code that uses random, untrained weights can feel pointless, but it is the cleanest way to see what the *architecture* does on its own, before any learning. Whatever structure shows up here (such as neighbors landing close together) comes purely from the graph operator, not from fitting.

```{r gcn-base-r, eval=TRUE, message=FALSE}
set.seed(1)

# 6 nodes: {1,2,3} form one cluster, {4,5,6} another, edge 3-4 is the bridge
A <- matrix(0, 6, 6)
edges <- rbind(
  c(1, 2), c(1, 3), c(2, 3),   # cluster 1 (a triangle)
  c(4, 5), c(4, 6), c(5, 6),   # cluster 2 (a triangle)
  c(3, 4)                      # bridge between clusters
)
for (e in seq_len(nrow(edges))) {
  i <- edges[e, 1]; j <- edges[e, 2]
  A[i, j] <- 1; A[j, i] <- 1   # undirected
}

# Symmetric normalized adjacency with self-loops: Ahat = Dt^{-1/2} (A+I) Dt^{-1/2}
gcn_norm <- function(A) {
  n  <- nrow(A)
  At <- A + diag(n)                 # add self-loops
  d  <- rowSums(At)                 # degrees including self-loop
  Dinv_sqrt <- diag(1 / sqrt(d))
  Dinv_sqrt %*% At %*% Dinv_sqrt
}

Ahat <- gcn_norm(A)

# One GCN layer: H_out = relu( Ahat %*% X %*% W )
relu <- function(z) pmax(z, 0)

gcn_layer <- function(X, Ahat, W) {
  relu(Ahat %*% X %*% W)
}

# Input features: 3 raw features per node, plus a fixed random projection W
d_in  <- 3
d_out <- 2
X <- matrix(rnorm(6 * d_in), nrow = 6, ncol = d_in)
W <- matrix(rnorm(d_in * d_out), nrow = d_in, ncol = d_out)

H1 <- gcn_layer(X, Ahat, W)
rownames(H1) <- paste0("node", 1:6)
colnames(H1) <- c("emb1", "emb2")
round(H1, 3)
```

The output is a 2-dimensional embedding per node. Because the layer mixes each node with its neighbors, nodes in the same cluster end up with more similar embeddings than nodes across the bridge, even with random untrained weights. We can confirm this by measuring how the layer pulls neighbors together: compute the average Euclidean distance between embeddings for within-cluster pairs versus across-cluster pairs.

```{r gcn-base-r-distance, eval=TRUE}
emb_dist <- function(H, i, j) sqrt(sum((H[i, ] - H[j, ])^2))

within  <- mean(c(
  emb_dist(H1, 1, 2), emb_dist(H1, 1, 3), emb_dist(H1, 2, 3),
  emb_dist(H1, 4, 5), emb_dist(H1, 4, 6), emb_dist(H1, 5, 6)
))
across  <- mean(c(
  emb_dist(H1, 1, 4), emb_dist(H1, 1, 5), emb_dist(H1, 2, 6), emb_dist(H1, 3, 6)
))

c(within_cluster = round(within, 3), across_cluster = round(across, 3))
```

The within-cluster distance comes out far smaller than the across-cluster distance. With no training at all, a single graph convolution has already pulled members of the same community together and kept the two communities apart, simply because each node was averaged with its neighbors. This is the entire promise of message passing in miniature: structure in the graph becomes structure in the embeddings.

## A Figure: Watching the Receptive Field Grow

A single GCN layer mixes 1-hop neighbors. Stacking layers mixes farther nodes. Figure @fig-graph-neural-networks-depth stacks the same untrained GCN layer up to six times on a small ring-of-clusters graph and plots how the mean pairwise distance between node embeddings shrinks as information spreads across a wider neighborhood. It also shows the over-smoothing tendency: with each added layer, all node embeddings move closer together.

```{r fig-graph-neural-networks-depth, eval=TRUE, message=FALSE, fig.cap="Average pairwise distance between node embeddings shrinks as GCN depth increases, illustrating over-smoothing.", fig.width=6, fig.height=4}
library(ggplot2)

# Build a larger graph: 4 clusters of 5 nodes arranged in a ring
make_ring_of_clusters <- function(n_clusters = 4, per = 5) {
  n <- n_clusters * per
  A <- matrix(0, n, n)
  idx <- split(seq_len(n), rep(seq_len(n_clusters), each = per))
  # dense within each cluster
  for (g in idx) for (i in g) for (j in g) if (i < j) { A[i, j] <- 1; A[j, i] <- 1 }
  # one bridge edge between consecutive clusters
  for (c in seq_len(n_clusters)) {
    nxt <- if (c == n_clusters) 1 else c + 1
    i <- idx[[c]][per]; j <- idx[[nxt]][1]
    A[i, j] <- 1; A[j, i] <- 1
  }
  A
}

A2   <- make_ring_of_clusters()
Ahat2 <- gcn_norm(A2)
n2 <- nrow(A2)

set.seed(7)
X2 <- matrix(rnorm(n2 * 4), nrow = n2)
W2 <- matrix(rnorm(4 * 4), nrow = 4)   # keep width fixed so we can iterate the layer

mean_pairwise_dist <- function(H) {
  D <- as.matrix(dist(H))
  mean(D[upper.tri(D)])
}

H  <- X2
records <- data.frame(depth = 0, dist = mean_pairwise_dist(H))
for (k in 1:6) {
  H <- gcn_layer(H, Ahat2, W2)
  records <- rbind(records, data.frame(depth = k, dist = mean_pairwise_dist(H)))
}

ggplot(records, aes(depth, dist)) +
  geom_line(color = "steelblue") +
  geom_point(size = 2, color = "steelblue") +
  labs(x = "Number of stacked GCN layers",
       y = "Mean pairwise embedding distance",
       title = "Over-smoothing: embeddings collapse with depth") +
  theme_minimal(base_size = 12)
```

The downward trend is the visual signature of over-smoothing: after only two or three layers the mean pairwise distance has nearly collapsed to zero, meaning the nodes have become almost indistinguishable. We return to why this happens, and what to do about it, in the pitfalls section below.

## GraphSAGE: Sampling and Inductive Learning

GraphSAGE (Hamilton, Ying, and Leskovec, 2017) addresses two GCN limitations: full-graph training and the inability to generalize to unseen nodes. Instead of using the entire neighborhood, it samples a fixed-size set of neighbors, and instead of folding the node's own state through a self-loop, it concatenates the aggregated neighborhood with the node's current representation:

$$
h_{\mathcal{N}(i)}^{(k)} = \text{AGGREGATE}_k\!\left( \{ h_j^{(k-1)} : j \in \mathcal{S}(i) \} \right),
$$

$$
h_i^{(k)} = \sigma\!\left( W^{(k)} \cdot \big[\, h_i^{(k-1)} \,\|\, h_{\mathcal{N}(i)}^{(k)} \,\big] \right),
$$

where $\mathcal{S}(i) \subseteq \mathcal{N}(i)$ is a sampled neighbor subset, $\|$ is concatenation, and AGGREGATE can be mean, max-pooling, or an LSTM over neighbors. A normalization step $h_i^{(k)} \leftarrow h_i^{(k)} / \lVert h_i^{(k)} \rVert_2$ usually follows.

The payoff is inductive learning: because the model learns aggregation functions rather than a fixed embedding per node, it can produce embeddings for nodes that did not exist at training time. The neighbor sampling also bounds per-batch computation, which is what makes GraphSAGE practical on graphs with millions of nodes.

> **When to use this:** choose GraphSAGE when new nodes keep arriving (new users, new products, new transactions) so retraining the whole model for every newcomer is not an option, or when the graph is simply too large to hold in memory at once. Sampling a fixed number of neighbors caps the cost per node no matter how large the graph grows.

## Graph Attention Network (GAT)

GCN and GraphSAGE weight neighbors by degree or treat them uniformly. GAT (Velickovic et al., 2018) lets the model learn how much attention to pay to each neighbor.

> **Intuition:** Not all neighbors are equally informative. In a citation network, a paper's most relevant citation may matter more than a passing reference; in a social graph, a close collaborator should outweigh a one-time contact. Attention lets the model decide, per edge, how much each neighbor contributes, instead of fixing those weights in advance by degree.

For a layer, first project all nodes with a shared weight $W$, then compute an unnormalized attention score for each edge $(i, j)$:

$$
e_{ij} = \text{LeakyReLU}\!\left( a^\top \big[\, W h_i \,\|\, W h_j \,\big] \right),
$$

where $a$ is a learnable attention vector. Normalize the scores over each node's neighborhood with a softmax:

$$
\alpha_{ij} = \frac{\exp(e_{ij})}{\sum_{l \in \mathcal{N}(i)} \exp(e_{il})}.
$$

The new representation is the attention-weighted sum of neighbor projections:

$$
h_i' = \sigma\!\left( \sum_{j \in \mathcal{N}(i)} \alpha_{ij}\, W h_j \right).
$$

In practice GAT runs several attention heads in parallel and concatenates or averages them, the same multi-head idea used in transformers. Attention helps when neighbors differ in relevance, for example a noisy graph where some edges are spurious, at the cost of more parameters and compute per edge.

### Comparing the Three Architectures

With all three layers in hand, it helps to see them side by side. Each makes a different choice for the message-passing template's three slots, and those choices map directly onto when you would pick one over another. Table @tbl-graph-neural-networks-architecture-comparison lays out how GCN, GraphSAGE, and GAT differ across the properties that matter in practice.

| Property | GCN | GraphSAGE | GAT |
|---|---|---|---|
| Neighbor weighting | fixed, by degree | uniform over sampled set | learned via attention |
| Aggregation | degree-weighted mean | mean / max / LSTM | attention-weighted sum |
| Own-state handling | self-loop | concatenation | self-loop (optional self-attention) |
| Uses neighbor sampling | no (full graph) | yes | no (basic form) |
| Inductive (unseen nodes) | no (basic form) | yes | yes |
| Relative compute | low | low to medium | medium to high |
| Best when | small, static, homophilous graph | large graph, new nodes arrive | neighbor relevance varies |

: Side-by-side comparison of GCN, GraphSAGE, and GAT across neighbor weighting, aggregation, own-state handling, sampling, inductive capability, compute cost, and the setting each suits best. {#tbl-graph-neural-networks-architecture-comparison}

## Over-smoothing and Other Pitfalls

No GNN survives contact with real data without running into at least one of the following. Knowing the failure modes in advance saves a great deal of debugging.

Figure @fig-graph-neural-networks-depth showed embeddings collapsing as depth grows. This is over-smoothing: repeated neighbor averaging acts like a low-pass filter on the graph, and in the limit every connected node converges to the same vector, erasing the distinctions a classifier needs. The practical consequence is that deep GNNs often perform worse than shallow ones, which is the opposite of the intuition from image CNNs. Most GNNs use only two to four layers.

> **Warning:** Adding layers to a GNN is not like adding layers to a CNN. Beyond a handful of layers, accuracy usually drops rather than improves. If your deep GNN underperforms a shallow one, suspect over-smoothing before anything else.

Several mitigations work in practice, and they can be combined:

- Residual or skip connections that add $h_i^{(k-1)}$ back into $h_i^{(k)}$, so the node retains its own signal.
- Jumping knowledge, which concatenates representations from all layers instead of using only the last.
- Normalization tailored to graphs (for example PairNorm) that keeps embeddings from collapsing.
- Simply keeping the network shallow and widening it instead of deepening it.

Over-smoothing is the most famous trap, but it is not the only one. A few others come up often enough to plan for:

- Over-squashing: information from an exponentially growing neighborhood is compressed into a fixed-size vector, so long-range dependencies get lost through narrow bottleneck edges.
- Heterophily: GCN assumes neighbors tend to share labels (homophily). On graphs where connected nodes usually differ, neighbor averaging hurts, and models designed for heterophily are preferable.
- Leakage in transductive splits: when train and test nodes live in one graph, message passing can pull test information into training. Use proper masked training and, where the use case is inductive, evaluate on held-out subgraphs.
- Scale: full-graph training does not fit in memory beyond modest sizes; use neighbor sampling (GraphSAGE), subgraph sampling (Cluster-GCN, GraphSAINT), or a framework that handles minibatching.

## torch geometric Reference Code

The base R layer above is for understanding, not production. For real work you would call a deep learning backend that gives you automatic differentiation, GPU support, and efficient sparse operations. The `torch` package and its `torchvision`/geometric extensions are not in this book's installed set, so the chunks below are marked `eval=FALSE`. They are written as correct, current code a reader can run after installing `torch`. The first defines a two-layer GCN as a `torch` module; the second is the equivalent in Python with PyTorch Geometric, which remains the most widely used GNN library.

```{r torch-gcn, eval=FALSE}
library(torch)

# A from-scratch 2-layer GCN in R torch.
# edge_index is a 2 x E integer tensor of (source, target) pairs (1-based here).
gcn_norm_torch <- function(edge_index, num_nodes) {
  # build dense adjacency with self-loops, then symmetric normalization
  A <- torch_zeros(num_nodes, num_nodes)
  src <- edge_index[1, ]; dst <- edge_index[2, ]
  A[cbind(as.integer(src), as.integer(dst))] <- 1
  A[cbind(as.integer(dst), as.integer(src))] <- 1   # undirected
  A <- A + torch_eye(num_nodes)
  deg <- A$sum(dim = 2)
  dinv <- torch_diag(deg$pow(-0.5))
  dinv$mm(A)$mm(dinv)
}

GCN <- nn_module(
  "GCN",
  initialize = function(d_in, d_hidden, d_out) {
    self$lin1 <- nn_linear(d_in, d_hidden, bias = FALSE)
    self$lin2 <- nn_linear(d_hidden, d_out, bias = FALSE)
  },
  forward = function(x, Ahat) {
    h <- nnf_relu(Ahat$mm(self$lin1(x)))
    h <- nnf_dropout(h, p = 0.5, training = self$training)
    Ahat$mm(self$lin2(h))   # logits per node
  }
)

# Training loop sketch for node classification with a train mask.
fit_gcn <- function(x, Ahat, y, train_mask, d_hidden = 16, epochs = 200) {
  model <- GCN(d_in = x$size(2), d_hidden = d_hidden, d_out = length(unique(as.integer(y))))
  opt <- optim_adam(model$parameters, lr = 0.01, weight_decay = 5e-4)
  for (epoch in seq_len(epochs)) {
    model$train()
    opt$zero_grad()
    logits <- model(x, Ahat)
    loss <- nnf_cross_entropy(logits[train_mask, ], y[train_mask])
    loss$backward()
    opt$step()
  }
  model
}
```

```{python pyg-gat, eval=FALSE}
# Python, PyTorch Geometric: a 2-layer GAT for node classification.
import torch
import torch.nn.functional as F
from torch_geometric.nn import GATConv
from torch_geometric.datasets import Planetoid

dataset = Planetoid(root="data/Cora", name="Cora")
data = dataset[0]

class GAT(torch.nn.Module):
    def __init__(self, in_dim, hidden, out_dim, heads=8):
        super().__init__()
        self.conv1 = GATConv(in_dim, hidden, heads=heads, dropout=0.6)
        self.conv2 = GATConv(hidden * heads, out_dim, heads=1, concat=False, dropout=0.6)

    def forward(self, x, edge_index):
        x = F.dropout(x, p=0.6, training=self.training)
        x = F.elu(self.conv1(x, edge_index))
        x = F.dropout(x, p=0.6, training=self.training)
        return self.conv2(x, edge_index)

model = GAT(dataset.num_features, 8, dataset.num_classes)
opt = torch.optim.Adam(model.parameters(), lr=0.005, weight_decay=5e-4)

for epoch in range(200):
    model.train()
    opt.zero_grad()
    out = model(data.x, data.edge_index)
    loss = F.cross_entropy(out[data.train_mask], data.y[data.train_mask])
    loss.backward()
    opt.step()

model.eval()
pred = model(data.x, data.edge_index).argmax(dim=1)
acc = (pred[data.test_mask] == data.y[data.test_mask]).float().mean().item()
print(f"test accuracy: {acc:.3f}")
```

## Practical Guidance

Pulling the chapter together, here is the short checklist worth keeping next to you when you build a GNN. Most of it is about resisting the urge to overcomplicate.

- Start simple. Fit a two-layer GCN before reaching for attention. If a plain tabular model on node features alone already does well, the graph may not be adding much.
- Keep it shallow. Two to four layers is typical. If you need a larger receptive field, prefer skip connections or jumping knowledge over raw depth.
- Always add self-loops and normalize the adjacency. Skipping normalization lets high-degree hubs swamp the signal.
- Match the split to the deployment. If new nodes will appear in production, train and evaluate inductively (GraphSAGE style) rather than transductively.
- Check homophily first. Compute the fraction of edges whose endpoints share a label. Low homophily means a vanilla GCN may underperform a label-agnostic baseline.
- Watch for over-smoothing and over-squashing as you scale depth or graph size, and validate with the embedding-distance diagnostic shown above.
- Mind compute. Use neighbor or subgraph sampling for large graphs, and reserve GAT's per-edge attention for cases where you have evidence that neighbors differ in relevance.

> **Key idea:** Every GNN layer is the same three-step recipe (gather messages from neighbors, aggregate them with an order-blind pooling, combine with the node's own state). GCN, GraphSAGE, and GAT differ only in how they fill those three slots. If you remember the recipe, you can read any new GNN paper as a variation on a theme rather than a fresh idea.

## Further Reading

- Kipf and Welling (2017), "Semi-Supervised Classification with Graph Convolutional Networks." The GCN propagation rule.
- Hamilton, Ying, and Leskovec (2017), "Inductive Representation Learning on Large Graphs." GraphSAGE and neighbor sampling.
- Velickovic, Cucurull, Casanova, Romero, Lio, and Bengio (2018), "Graph Attention Networks." Attention over neighbors.
- Gilmer, Schoenholz, Riley, Vinyals, and Dahl (2017), "Neural Message Passing for Quantum Chemistry." The unifying message-passing view.
- Xu, Hu, Leskovec, and Jegelka (2019), "How Powerful are Graph Neural Networks?" Expressive power and the Graph Isomorphism Network.
- Hamilton (2020), "Graph Representation Learning." A book-length treatment of the field.
- Bronstein, Bruna, Cohen, and Velickovic (2021), "Geometric Deep Learning." A broad framework that places GNNs among models built on symmetry and structure.
