# Mixture of Experts and Conditional Computation {#sec-mixture-of-experts}

```{r}
#| include: false
source("_common.R")
```

Imagine a help desk where, instead of one generalist trying to answer every question, the front desk listens to your problem and forwards you to the specialist who handles that kind of issue. The specialists never have to be good at everything; the front desk only has to be good at routing. A mixture of experts is exactly this arrangement turned into a statistical model.

More precisely, a mixture of experts (MoE) splits a prediction problem into pieces, trains a set of specialist sub-models (the experts), and learns a separate routing function (the gating network) that decides which experts handle which inputs. Instead of forcing one model to be good everywhere, you let several smaller models each be good in their own region of the input space, then combine them with input-dependent weights.

This chapter builds the idea from the ground up. We start with the intuition and the probabilistic model, derive the Expectation-Maximization (EM) algorithm that fits it, then show how the same machinery scales up to the *sparse* mixture of experts layers that power today's largest language models. Along the way you will fit a working two-expert mixture in base R and see it recover a regime change that a single line cannot. By the end you should understand both the classical model and why MoE has become a standard tool for scaling modern AI.

> **Key idea:** A mixture of experts is an ensemble whose combination weights depend on the input. The gate decides *who answers*, and that decision changes from one input to the next.

The idea goes back to @jacobs1991 and @jordan1994, who framed it as a probabilistic model fit with an EM-like procedure. The same idea now powers some of the largest neural networks in production, where it appears as the *sparse* mixture of experts layer inside transformers (@sec-transformers) [@shazeer2017; @fedus2022]. The modern motivation is *conditional computation*, the practice of activating only part of a model for each input: a model can hold a very large number of parameters but only switch on a small fraction of them per input, so the compute cost per token stays roughly fixed while capacity grows.

## Where this fits in a modern ML/AI workflow

For a data scientist or ML engineer, MoE shows up in three practical situations.

1.  Heterogeneous data with regime structure. When the relationship between predictors and target changes across regions (different customer segments, different operating regimes of a machine, piecewise dynamics), a single global model underfits. MoE lets the gate discover the regimes and assigns an expert to each.

2.  Scaling model capacity under a compute budget. In large language models, a dense feed-forward block is replaced by many parallel expert blocks plus a router that sends each token to only the top-$k$ experts (often $k = 1$ or $k = 2$). The model can have hundreds of billions of parameters while activating only a few billion per token. This is the dominant reason MoE is discussed in 2020s AI engineering.

3.  Ensembling with specialization. Unlike bagging (@sec-bagging) or stacking where every base model sees every input, MoE weights are input-dependent. The combination is local, not global, which is closer to a learned partition of the feature space than to a uniform average.

The core distinction from a standard ensemble is that the mixture weights depend on $x$. A stacked ensemble (see the model stacking chapter, @sec-model-stacking) learns fixed blend weights; a mixture of experts learns weights $g_k(x)$ that change with the input.

> **Intuition:** Stacking asks "what is the best fixed recipe for blending these models everywhere?" A mixture of experts asks "given *this* input, who should I trust right now?" The second question is harder, but it is the right one when no single model is good across the whole input space.

With that motivation in place, we now write the model down precisely.

## The model

The model has two parts, the experts and the gate, so we describe each in turn before combining them.

Let the input be $x \in \mathbb{R}^p$ and the target be $y$. We have $K$ experts. Expert $k$ produces a conditional distribution $p_k(y \mid x, \theta_k)$, for example a Gaussian whose mean is a linear function of $x$ for regression, or a softmax over classes for classification.^[Nothing forces the experts to be simple. In the classical statistical literature they are usually linear or generalized linear models because that keeps the M-step in closed form, but in deep learning each expert is itself a small neural network.]

The gating network produces a probability vector $g(x) = (g_1(x), \dots, g_K(x))$ with $g_k(x) \ge 0$ and $\sum_k g_k(x) = 1$. You can read $g_k(x)$ as "the share of the prediction we hand to expert $k$ for this particular input." The standard choice is a softmax over linear scores:

$$
g_k(x) = \frac{\exp(v_k^\top x)}{\sum_{j=1}^{K} \exp(v_j^\top x)},
$$

where $v_k \in \mathbb{R}^p$ are the gating parameters (include a 1 in $x$ for an intercept).

The mixture density is the gate-weighted sum of the expert densities:

$$
p(y \mid x, \Theta) = \sum_{k=1}^{K} g_k(x) \, p_k(y \mid x, \theta_k),
$$

with $\Theta = \{v_k, \theta_k\}_{k=1}^K$. For a regression expert with Gaussian noise,

$$
p_k(y \mid x, \theta_k) = \mathcal{N}\!\left(y \,;\, \beta_k^\top x, \, \sigma_k^2\right).
$$

The predicted mean is then the gate-weighted combination of expert means:

$$
\mathbb{E}[y \mid x] = \sum_{k=1}^{K} g_k(x) \, \beta_k^\top x .
$$

So to predict, you run every expert, ask the gate how much to weight each one for this input, and take the weighted average. The only thing left is to learn the parameters from data, which is what the next section does.

### Fitting with EM

Fitting an MoE looks circular at first: to know which expert should handle a point we need the experts to already be trained, but to train an expert we need to know which points belong to it. EM resolves the chicken-and-egg problem by alternating soft guesses for one while holding the other fixed.

Introduce a latent indicator $z \in \{1, \dots, K\}$ for which expert generated a given observation, with prior $P(z = k \mid x) = g_k(x)$.^[We never observe $z$; it is a modeling fiction that says "imagine each point really was produced by one specific expert." Treating $z$ as a hidden variable is the trick that turns a hard direct optimization into the easy two-step EM loop.] The data log-likelihood over $n$ observations is

$$
\ell(\Theta) = \sum_{i=1}^{n} \log \sum_{k=1}^{K} g_k(x_i) \, p_k(y_i \mid x_i, \theta_k).
$$

This is a log of a sum, which is awkward to maximize directly, so we use Expectation-Maximization.

E-step (who is responsible?). Given current parameters, compute the responsibility of expert $k$ for observation $i$, which is the posterior probability that $z_i = k$:

$$
r_{ik} = P(z_i = k \mid x_i, y_i) = \frac{g_k(x_i) \, p_k(y_i \mid x_i, \theta_k)}{\sum_{j=1}^{K} g_j(x_i) \, p_j(y_i \mid x_i, \theta_j)} .
$$

The responsibility $r_{ik}$ is a soft assignment: it blends what the gate expected (the prior $g_k$) with how well expert $k$ actually fit this point (the likelihood $p_k$). An expert that the gate favored but that fit poorly will see its responsibility shrink.

M-step (retrain given the soft assignments). Maximize the expected complete-data log-likelihood. Conveniently, this decouples into two independent pieces, one for the experts and one for the gate.

*Experts.* Each expert is fit by responsibility-weighted maximum likelihood. For Gaussian linear experts this is weighted least squares with weights $r_{ik}$:

$$
\beta_k = \left( X^\top W_k X \right)^{-1} X^\top W_k y, \qquad W_k = \mathrm{diag}(r_{1k}, \dots, r_{nk}),
$$

and the variance update

$$
\sigma_k^2 = \frac{\sum_{i} r_{ik} (y_i - \beta_k^\top x_i)^2}{\sum_{i} r_{ik}} .
$$

*Gate.* The gating parameters maximize $\sum_{i} \sum_{k} r_{ik} \log g_k(x_i)$, which is exactly the objective of a multinomial logistic regression with the responsibilities as soft labels. There is no closed form, so we take one or a few gradient steps per M-step. The gradient with respect to $v_k$ is

$$
\frac{\partial}{\partial v_k} \sum_i \sum_j r_{ij} \log g_j(x_i) = \sum_{i} \left( r_{ik} - g_k(x_i) \right) x_i .
$$

Notice the gate gradient has the same form you see in logistic regression: the prediction error $r_{ik} - g_k(x_i)$ times the input. The gate is simply being trained to match the responsibilities the E-step produced.

> **Note:** EM increases the log-likelihood at every iteration and converges to a *local* optimum, never guaranteed to be the global one. Because experts are interchangeable, different starts can land on different (but equally good) solutions, so running several random restarts and keeping the best log-likelihood is standard practice.

We will watch this monotone increase happen in the demo below.

## Sparse mixture of experts and top-k routing

So far we have a clean classical model. The reason MoE is everywhere in modern AI, though, is a single change that makes it scale, and that is the subject of this section.

The model above is *dense*: every expert is evaluated for every input, then combined. That is fine for $K = 2$ or $K = 10$, but it does not scale to thousands of experts because cost grows linearly in $K$.

The sparse variant keeps the same gating scores but evaluates only the top-$k$ experts per input.^[Here $k$ (lowercase) is the number of experts you keep per input, typically 1 or 2, while $K$ (uppercase) is the total number of experts, which can be in the thousands. Do not confuse the two.] Let $\mathrm{TopK}(s, k)$ keep the $k$ largest gating logits $s_j = v_j^\top x$ and set the rest to $-\infty$ before the softmax:

$$
\tilde{g}_j(x) = \frac{\exp(s_j) \cdot \mathbb{1}[j \in \mathrm{TopK}(s, k)]}{\sum_{m \in \mathrm{TopK}(s, k)} \exp(s_m)} .
$$

Only the selected experts run, so per-input compute depends on $k$, not on $K$. Adding more experts increases capacity (total parameters) without increasing per-input FLOPs. This is the conditional computation property that makes trillion-parameter language models feasible [@fedus2022].

> **Key idea:** Sparsity decouples capacity from cost. A dense model that doubles its parameters doubles its compute per input. A sparse MoE can multiply its parameters many times over while the work done per input barely moves, because each input still only touches $k$ experts.

### Load balancing

Sparse routing has a failure mode: the gate can collapse onto a few popular experts, leaving others untrained, which wastes capacity and unbalances hardware utilization. This is a real and common problem, not a theoretical worry, so production systems always guard against it. The standard fix is an auxiliary load-balancing loss added to the training objective. With $f_k$ the fraction of inputs in a batch routed to expert $k$ and $P_k$ the average gate probability assigned to expert $k$, a common penalty [@shazeer2017; @fedus2022] is

$$
L_{\text{balance}} = \alpha \cdot K \sum_{k=1}^{K} f_k \, P_k ,
$$

which is minimized when routing is uniform. Other practical devices include a per-expert capacity limit (drop or overflow tokens beyond a cap) and adding noise to the gating logits during training to encourage exploration.

> **Warning:** If you train a sparse MoE without a load-balancing term, expect a few experts to win all the traffic while the rest stay at their random initialization. The auxiliary loss is not optional in practice.

### Dense vs sparse mixture

@tbl-mixture-of-experts-dense-vs-sparse summarizes the trade-off between the two flavors so you can pick the right one for the job.

| Property | Dense MoE | Sparse MoE (top-$k$) |
|---|---|---|
| Experts evaluated per input | all $K$ | only $k \ll K$ |
| Per-input compute | grows with $K$ | roughly fixed in $K$ |
| Total parameters used per input | all | a fraction |
| Typical scale | small $K$ (2 to tens) | large $K$ (tens to thousands) |
| Training stability | straightforward EM or gradient | needs load balancing, capacity limits |
| Main use | regime/segment modeling | scaling LLM capacity |
| Routing | soft weights | hard top-$k$ selection |

: Comparison of dense and sparse top-$k$ mixtures of experts across compute cost, scale, training stability, and routing. {#tbl-mixture-of-experts-dense-vs-sparse}

## MoE compared to related methods

It helps to place MoE next to the other ensemble methods in this book. @tbl-mixture-of-experts-vs-ensembles lays out the comparison, and the single column that sets MoE apart is the last one: whether the combination weights are allowed to depend on the input.

| Method | Combination weights | Specialization | Weights depend on $x$? |
|---|---|---|---|
| Bagging / random forest | uniform average | none (same task) | no |
| Stacking | learned, fixed | mild | no |
| Boosting | sequential, additive | residual-focused | no (additive) |
| Mixture of experts | learned gate | strong, by region | yes |
| Sparse MoE layer | learned top-$k$ gate | strong, by token | yes (hard) |

: How a mixture of experts compares to bagging, stacking, and boosting, with input-dependent combination weights as the distinguishing feature. {#tbl-mixture-of-experts-vs-ensembles}

## Runnable demo: two-expert mixture on piecewise data

Enough theory. Let us watch the EM algorithm actually discover a hidden partition of the data with no labels telling it where the boundary is.

We simulate data where the slope of $y$ in $x$ flips sign at $x = 0$. A single linear regression is hopeless here, because the best single line through a V-shaped cloud is roughly flat, but two linear experts plus a softmax gate can recover the two regimes. We fit with the EM updates derived above, all in base R so nothing is hidden inside a package.

> **When to use this:** A worked-from-scratch fit like this one is exactly the case MoE was designed for, a relationship whose shape changes across regions of the input. If your relationship is globally smooth, reach for a GAM or boosted trees instead (see the guidance section at the end).

```{r moe-setup}
set.seed(1301)

# simulate piecewise-linear data with a regime change at x = 0
n  <- 400
x  <- runif(n, -3, 3)
# regime A (x < 0): negative slope; regime B (x >= 0): positive slope
mu <- ifelse(x < 0, 2 - 1.6 * x, 2 + 1.6 * x)
y  <- mu + rnorm(n, sd = 0.6)

X  <- cbind(1, x)          # design matrix for experts and gate
K  <- 2                    # number of experts
```

```{r moe-fit}
# Gaussian density helper
dnorm_safe <- function(y, mu, s2) {
  dnorm(y, mean = mu, sd = sqrt(pmax(s2, 1e-8)))
}

softmax_rows <- function(S) {
  S <- S - apply(S, 1, max)          # numerical stability
  E <- exp(S)
  E / rowSums(E)
}

fit_moe <- function(X, y, K = 2, iters = 100, gate_steps = 5,
                    gate_lr = 0.05, seed = 1) {
  set.seed(seed)
  n <- nrow(X); p <- ncol(X)

  # init experts by spreading their slopes apart so EM does not start in a
  # symmetric trap where every expert claims equal responsibility everywhere
  beta_ols <- as.numeric(solve(t(X) %*% X, t(X) %*% y))
  Beta <- matrix(beta_ols, nrow = p, ncol = K)
  Beta[p, ] <- beta_ols[p] + seq(-3, 3, length.out = K)   # vary the slope row
  Beta <- Beta + matrix(rnorm(p * K, sd = 0.3), p, K)
  s2   <- rep(var(y), K)
  V    <- matrix(rnorm(p * K, sd = 0.1), nrow = p, ncol = K)  # small random gate init

  ll_trace <- numeric(iters)

  for (it in 1:iters) {
    # gate probabilities
    G <- softmax_rows(X %*% V)            # n x K

    # expert means and densities
    M <- X %*% Beta                       # n x K
    D <- sapply(1:K, function(k) dnorm_safe(y, M[, k], s2[k]))  # n x K

    # E-step: responsibilities
    num  <- G * D
    den  <- rowSums(num) + 1e-12
    R    <- num / den
    ll_trace[it] <- sum(log(den))

    # M-step experts: weighted least squares per expert
    for (k in 1:K) {
      w  <- R[, k]
      XtWX <- t(X * w) %*% X
      XtWy <- t(X * w) %*% y
      Beta[, k] <- solve(XtWX + diag(1e-8, p), XtWy)
      resid <- y - X %*% Beta[, k]
      s2[k] <- sum(w * resid^2) / (sum(w) + 1e-12)
    }

    # M-step gate: a few gradient ascent steps on sum_i sum_k R_ik log G_ik
    for (g in 1:gate_steps) {
      G <- softmax_rows(X %*% V)
      grad <- t(X) %*% (R - G)            # p x K
      V <- V + gate_lr * grad
    }
  }

  list(Beta = Beta, s2 = s2, V = V, R = R, G = softmax_rows(X %*% V),
       ll = ll_trace)
}

fit <- fit_moe(X, y, K = 2, iters = 150, gate_steps = 8, gate_lr = 0.03, seed = 7)

# recovered expert slopes (one should be near -1.6, the other near +1.6)
round(fit$Beta, 3)
```

Read the output as two columns, one per expert. One expert recovered a slope near $-1.6$ and the other near $+1.6$, which are exactly the two slopes we built into the simulation, even though the fit was never told where the regime boundary was. The gate learned a soft boundary near $x = 0$ on its own. @fig-mixture-of-experts-fit plots the data colored by which expert claims responsibility, overlays each expert's fitted line, and shows the mixture prediction.

```{r fig-mixture-of-experts-fit, fig.cap="Two-expert mixture on piecewise data. Points are colored by the expert with higher responsibility, the dashed lines are the two linear experts, and the solid line is the gate-weighted mixture prediction.", fig.width=7, fig.height=5}
assign <- ifelse(fit$R[, 1] >= fit$R[, 2], 1L, 2L)

# gate-weighted mixture prediction on a grid
xg  <- seq(-3, 3, length.out = 200)
Xg  <- cbind(1, xg)
Gg  <- softmax_rows(Xg %*% fit$V)
Mg  <- Xg %*% fit$Beta
mix <- rowSums(Gg * Mg)

plot(x, y, col = ifelse(assign == 1, "#1b9e77", "#d95f02"),
     pch = 19, cex = 0.6,
     xlab = "x", ylab = "y",
     main = "Mixture of two linear experts")
abline(a = fit$Beta[1, 1], b = fit$Beta[2, 1], col = "#1b9e77", lwd = 2, lty = 2)
abline(a = fit$Beta[1, 2], b = fit$Beta[2, 2], col = "#d95f02", lwd = 2, lty = 2)
lines(xg, mix, col = "black", lwd = 3)
legend("top", bty = "n",
       legend = c("expert 1 points", "expert 2 points",
                  "expert 1 line", "expert 2 line", "mixture"),
       col = c("#1b9e77", "#d95f02", "#1b9e77", "#d95f02", "black"),
       pch = c(19, 19, NA, NA, NA),
       lty = c(NA, NA, 2, 2, 1), lwd = c(NA, NA, 2, 2, 3))
```

We can also check that the gate behaves like a learned soft router: the probability of expert 2 should rise as $x$ increases. @fig-mixture-of-experts-gate plots that learned gating probability across the input range.

```{r fig-mixture-of-experts-gate, fig.cap="Learned gating probability for expert 2 as a function of x. The softmax gate forms a soft routing boundary near the regime change.", fig.width=7, fig.height=4}
plot(xg, Gg[, 2], type = "l", lwd = 3, col = "#d95f02",
     ylim = c(0, 1), xlab = "x", ylab = "gate probability (expert 2)",
     main = "Softmax gating: soft routing boundary")
abline(v = 0, col = "grey60", lty = 3)
abline(h = 0.5, col = "grey60", lty = 3)
```

```{r moe-loglik}
# EM should increase the log-likelihood monotonically (up to gate sub-steps)
ll <- fit$ll
cat("first 3 log-likelihoods:", round(head(ll, 3), 1), "\n")
cat("last 3 log-likelihoods :", round(tail(ll, 3), 1), "\n")
cat("monotone non-decreasing:",
    all(diff(ll) > -1e-6), "\n")
```

The log-likelihood climbs from a poor start and flattens out, and the final check confirms it never went down (allowing a tiny tolerance for the gate's inner gradient steps). That monotone increase is the theoretical guarantee of EM made visible.

A single global linear fit for comparison makes the gain concrete:

```{r moe-baseline}
ols <- lm(y ~ x)
mse_ols <- mean(residuals(ols)^2)

# mixture training MSE using the gate-weighted mean at each point
G_full   <- fit$G
M_full   <- X %*% fit$Beta
yhat_moe <- rowSums(G_full * M_full)
mse_moe  <- mean((y - yhat_moe)^2)

c(ols_mse = round(mse_ols, 3), moe_mse = round(mse_moe, 3))
```

The mixture cuts training error substantially because it fits each regime with its own line rather than averaging the two slopes into a flat compromise.

## Sketch of top-k routing in base R

The dense demo above evaluated both experts everywhere. To make the sparse idea concrete without a deep learning stack, here is a tiny top-$k$ router over gating logits. With $K = 2$ and $k = 1$ it becomes hard assignment, where each input picks exactly one expert; the same code generalizes to many experts.

```{r moe-topk}
top_k_gate <- function(S, k) {
  # S: n x K matrix of gating logits; return sparse normalized weights
  n <- nrow(S); K <- ncol(S)
  W <- matrix(0, n, K)
  for (i in 1:n) {
    keep <- order(S[i, ], decreasing = TRUE)[1:k]
    e <- exp(S[i, keep] - max(S[i, keep]))
    W[i, keep] <- e / sum(e)
  }
  W
}

S <- X %*% fit$V
W1 <- top_k_gate(S, k = 1)     # hard routing: each row picks one expert
# fraction of inputs routed to each expert (load), should be away from 0 and 1
round(colMeans(W1 > 0), 3)
```

Here the load is close to a 50/50 split, which is what we want. In a real sparse MoE you would add the load-balancing penalty $L_{\text{balance}}$ from above so neither expert is starved, and a capacity cap so no expert is overloaded within a batch.

## A note on the deep learning version

In a transformer (@sec-transformers), the experts are feed-forward sub-networks and the gate is a small linear layer applied per token. The code below is the idiomatic Keras shape of a single sparse MoE feed-forward layer. It is kept `eval=FALSE` because it needs a configured deep learning backend (Keras and its TensorFlow backend), but it is correct, runnable code.

> **Tip:** Even in the deep learning version, the two moving parts are the same ones we derived classically: a gate that scores experts and a weighted combination of expert outputs. The novelty is engineering (sparsity, capacity, distributing experts across devices), not a new statistical idea.

```{r moe-keras, eval=FALSE}
library(keras)

# A simplified top-1 routed MoE feed-forward layer for one token embedding.
# d_model: embedding size, d_ff: expert hidden size, num_experts: K.
moe_ffn <- function(x, d_model = 256L, d_ff = 1024L, num_experts = 8L) {
  # gate logits per token, then softmax over experts
  gate_logits <- layer_dense(x, units = num_experts)          # (batch, K)
  gate        <- layer_activation_softmax(gate_logits)        # (batch, K)

  # build K independent expert FFNs and stack their outputs
  expert_outs <- lapply(seq_len(num_experts), function(k) {
    h <- layer_dense(x, units = d_ff, activation = "gelu")
    layer_dense(h, units = d_model)                           # (batch, d_model)
  })
  stacked <- k_stack(expert_outs, axis = 2L)                  # (batch, K, d_model)

  # weighted combination by gate (dense form shown; sparse would gather top-k)
  w <- k_expand_dims(gate, axis = -1L)                        # (batch, K, 1)
  k_sum(stacked * w, axis = 2L)                               # (batch, d_model)
}

# Auxiliary load-balancing loss to discourage expert collapse.
load_balance_loss <- function(gate, alpha = 0.01) {
  importance <- k_mean(gate, axis = 1L)                       # P_k
  load       <- k_mean(k_cast(gate > 0, "float32"), axis = 1L)# f_k
  K <- k_int_shape(gate)[[2]]
  alpha * K * k_sum(importance * load)
}
```

## Practical guidance, pitfalls, and when to use it

A mixture of experts is powerful but has sharp edges, so this section collects what tends to go right and wrong in practice.

When MoE helps.

-   The conditional mean or class boundary changes shape across regions of the input space, and you can plausibly describe those regions from the features.
-   You want to scale model capacity but keep per-input inference cost bounded (the sparse, deep learning case).
-   You have enough data per region. Each expert effectively trains on a fraction of the data, so MoE is data-hungry relative to a single model.

When to prefer something else.

-   If the relationship is globally smooth, a GAM (see the generalized additive models chapter, @sec-gam) or a gradient-boosted tree (see the boosting chapter, @sec-boosting) is simpler and usually stronger.
-   With little data, the latent assignment is unstable and the gate overfits; a single regularized model is safer.

Pitfalls.

-   *Local optima.* EM converges to a local maximum. Run several random restarts and keep the best log-likelihood. The demo uses a fixed seed for reproducibility, but in practice vary it.
-   *Degenerate experts.* An expert can grab almost no responsibility and have its variance collapse toward zero, inflating its density. Floor the variance (the `pmax(s2, 1e-8)` above) and consider pruning or reinitializing dead experts.
-   *Gate collapse in the sparse case.* Without a load-balancing term, the router sends everything to a few experts. Always add the auxiliary loss and a capacity limit when training sparse MoE.
-   *Identifiability.* Experts are exchangeable, so the labels (which expert is "expert 1") are arbitrary and can swap across restarts. Do not interpret expert indices as fixed.
-   *Choosing $K$.* Treat the number of experts as a hyperparameter. Use a validation metric or an information criterion; more experts is not automatically better and raises overfitting risk in the dense case.
-   *Routing instability.* Hard top-$k$ routing has discrete decisions that can flip during training; noisy gating and a small temperature on the gate logits help smooth optimization.

Operational notes for the sparse case. Expert parallelism places different experts on different devices, so the all-to-all communication that shuffles tokens to their chosen experts often dominates wall-clock time, not the matrix multiplies. Capacity factor, batch size, and the number of experts per device are the knobs that matter most for throughput.

To summarize the whole chapter in one line: a mixture of experts is an ensemble whose weights depend on the input, fit classically by EM and scaled in deep learning by routing each input to only its top few experts. The same two ideas, a gate and specialized experts, carry you from a two-line toy model to trillion-parameter language models.

## Further reading

The references below trace the arc from the original statistical formulation to the modern sparse layers.

-   Jacobs, Jordan, Nowlan, and Hinton (1991), "Adaptive mixtures of local experts." The original formulation.
-   Jordan and Jacobs (1994), "Hierarchical mixtures of experts and the EM algorithm." The EM treatment and the hierarchical (tree-structured) gate.
-   Shazeer et al. (2017), "Outrageously large neural networks: the sparsely-gated mixture-of-experts layer." Top-$k$ routing and load balancing at scale.
-   Fedus, Zoph, and Shazeer (2022), "Switch Transformers: scaling to trillion parameter models with simple and efficient sparsity." Top-1 routing, capacity factors, and large-scale training practice.
-   Bishop (2006), *Pattern Recognition and Machine Learning*, chapter on mixture models and EM, for the probabilistic background.
