# Bayesian Deep Learning and Uncertainty Quantification

Most predictive models return a point estimate $\hat{y}$ and stop there. In production, the more useful question is often not "what is the prediction?" but "how much should I trust this prediction?" A credit model that is confident on applicants similar to its training data and uncertain on applicants from a new region behaves very differently from one that reports the same confidence everywhere. Uncertainty quantification (UQ) is the set of tools that attaches a calibrated sense of doubt to model outputs.

This chapter covers the practical core of Bayesian deep learning and UQ for data scientists and ML engineers who work in R: the distinction between aleatoric and epistemic uncertainty, and the four methods that show up most often in practice, namely Monte Carlo (MC) dropout, deep ensembles, the Laplace approximation, and variational inference. The runnable demonstrations use base R and installed packages so the ideas are concrete. The deep learning code that needs Keras or Torch is shown as correct, idiomatic code but is not executed here.

By the end you will be able to name and motivate the two kinds of uncertainty, read the one formula that ties them together (the law of total variance), pick a UQ method that fits your constraints, and run a small ensemble in plain R that produces calibrated predictive intervals. The chapter builds on the Bayesian regression and Gaussian process material from earlier and connects forward to any model where a prediction alone is not enough.

> **Intuition:** A point prediction is a confident assertion with no escape hatch. UQ turns "the answer is 7" into "the answer is probably between 5 and 9, and I am especially unsure out here because I have never seen inputs like this." The second statement is far more useful for any decision that has a cost attached to being wrong.

## Where this fits in a modern ML workflow

A standard supervised pipeline produces a function $f_\theta(x)$ trained by minimizing a loss. UQ sits at two points in that pipeline.

- At training time, it changes how you fit the model (Bayesian priors, dropout kept on, an ensemble of fits) so that the model can express uncertainty at all.
- At serving time, it changes what you return: a predictive distribution $p(y \mid x)$ or an interval, instead of a single number.

Downstream consumers use this in concrete ways. Active learning queries the points where epistemic uncertainty is highest. Out-of-distribution detection flags inputs where the model is uncertain because they look unlike anything in training. Risk-sensitive decisions (medical triage, autonomous control, pricing) use the predictive variance directly in an expected-cost calculation. Selective prediction abstains when uncertainty exceeds a threshold and routes the case to a human.^[Selective prediction is sometimes called "reject option" classification. The model is allowed to say "I don't know" for the hardest cases, trading coverage (the fraction of inputs it answers) for higher accuracy on the cases it does answer.]

> **When to use this:** Reach for UQ whenever a wrong prediction has a cost that depends on how wrong it is, whenever you need to decide where to collect more data, or whenever the model will face inputs that may not resemble its training set. If every prediction is cheap and the data is stationary, a point estimate may be all you need.

## Two kinds of uncertainty

The central conceptual split is between aleatoric and epistemic uncertainty. They behave differently and call for different remedies, so keeping them separate is worth the effort.

Aleatoric uncertainty is the noise inherent in the data-generating process. If you measure the same $x$ many times and get different $y$ values, that spread is aleatoric. It does not shrink as you collect more data, because it is a property of the world (sensor noise, intrinsic randomness, unmeasured covariates). In a regression model with response

$$
y = f(x) + \epsilon, \qquad \epsilon \sim N(0, \sigma^2(x)),
$$

the term $\sigma^2(x)$ is the aleatoric variance. When it depends on $x$ it is called heteroscedastic^[Heteroscedastic simply means "different scatter": the noise level changes across the input space. The opposite, constant noise everywhere, is homoscedastic.]; a model can learn $\sigma^2(x)$ as a second output head.

Epistemic uncertainty is uncertainty about the model itself: which parameters $\theta$, or even which function $f$, generated the data. It is large where you have little data and shrinks toward zero as data accumulates. In the Bayesian view, epistemic uncertainty is encoded in the posterior over parameters $p(\theta \mid \mathcal{D})$, where $\mathcal{D} = \{(x_i, y_i)\}_{i=1}^n$ is the training set.

> **Key idea:** Aleatoric is noise in the world and will not go away with more data; epistemic is doubt in the model and shrinks as data accumulates. A useful one-word test: if collecting more samples at this input would tighten the prediction, the uncertainty is epistemic; if it would not, it is aleatoric.

The Bayesian predictive distribution makes the decomposition explicit. The posterior predictive for a new input $x_*$ is

$$
p(y_* \mid x_*, \mathcal{D}) = \int p(y_* \mid x_*, \theta)\, p(\theta \mid \mathcal{D})\, d\theta .
$$

The inner term $p(y_* \mid x_*, \theta)$ carries aleatoric noise; the spread of $\theta$ under the posterior carries epistemic uncertainty. Using the law of total variance, the total predictive variance splits as

$$
\underbrace{\operatorname{Var}(y_* \mid x_*)}_{\text{total}}
= \underbrace{E_\theta\!\big[\operatorname{Var}(y_* \mid x_*, \theta)\big]}_{\text{aleatoric}}
+ \underbrace{\operatorname{Var}_\theta\!\big[E(y_* \mid x_*, \theta)\big]}_{\text{epistemic}} .
$$

The first term averages the noise variance over the posterior; the second measures how much the mean prediction moves as $\theta$ varies. This formula is the practical workhorse of the chapter: every method below is, in effect, a recipe for sampling or approximating $p(\theta \mid \mathcal{D})$ so that you can estimate these two terms.

> **Note:** If you remember one equation from this chapter, make it the law-of-total-variance split above. Once you can sample plausible parameter settings $\theta$ from (an approximation of) the posterior, the rest is bookkeeping: average each member's noise for the aleatoric piece, and measure the scatter of the members' means for the epistemic piece.

## The four practical methods

The exact integral over $p(\theta \mid \mathcal{D})$ is intractable for any non-trivial network, so all methods are approximations. They trade off accuracy, compute, and how much they disturb your existing training code. The four below are ordered roughly from least to most intrusive: deep ensembles change only how many times you train, MC dropout changes nothing at all, Laplace bolts on after training, and variational inference rewrites the layers. Read each in three parts: the one-sentence idea, the formula that makes it precise, and the practical catch.

### Deep ensembles

Train $M$ networks independently from different random initializations (and, often, different data shuffles or bootstrap resamples). Each fit $\theta^{(m)}$ lands in a different mode of the loss surface. Treat the collection as samples that approximate the posterior. The predictive mean and epistemic variance are

$$
\bar{f}(x_*) = \frac{1}{M}\sum_{m=1}^M f_{\theta^{(m)}}(x_*),
\qquad
\widehat{\operatorname{Var}}_{\text{epi}}(x_*) = \frac{1}{M-1}\sum_{m=1}^M \big(f_{\theta^{(m)}}(x_*) - \bar{f}(x_*)\big)^2 .
$$

If each network also outputs a variance $\sigma^2_{\theta^{(m)}}(x_*)$, add its mean to get total variance. Ensembles are simple, embarrassingly parallel, and tend to give the best-calibrated uncertainty in benchmarks. The cost is $M$ times the training and storage.

> **Tip:** Because the $M$ fits are independent, you can train them on separate machines or GPUs at the same time. That makes ensembles the easiest UQ method to scale when you have hardware but not much patience for new code.

### MC dropout

Dropout randomly zeroes activations with probability $p$ during training as a regularizer. The MC dropout idea is to keep dropout on at prediction time and run $T$ stochastic forward passes. Each pass uses a different random mask, equivalent to sampling a different sub-network, so the spread across passes approximates epistemic uncertainty:

$$
\widehat{E}(y_* \mid x_*) \approx \frac{1}{T}\sum_{t=1}^T f_{\hat\theta, z_t}(x_*),
$$

where $z_t$ is the random dropout mask on pass $t$. This is almost free to add to an existing dropout network: no retraining, just turning off the usual "eval mode" deactivation of dropout. The catch is that the implied posterior is tied to the dropout rate and is often poorly calibrated unless the rate is tuned.

> **Warning:** MC dropout is cheap precisely because it reuses one set of weights, so its uncertainty does not necessarily shrink to zero even with unlimited data. Treat it as a useful rough signal, not a calibrated posterior, unless you have validated the dropout rate against held-out coverage.

### Laplace approximation

Fit the network normally to the maximum-a-posteriori (MAP) estimate $\hat\theta$, then approximate the posterior near that mode with a Gaussian. A second-order Taylor expansion of the log-posterior $\log p(\theta \mid \mathcal{D})$ around $\hat\theta$ gives

$$
p(\theta \mid \mathcal{D}) \approx N\big(\hat\theta,\, H^{-1}\big),
\qquad
H = -\nabla^2_\theta \log p(\theta \mid \mathcal{D})\big|_{\hat\theta},
$$

where $H$ is the Hessian of the negative log-posterior at the mode (often approximated by the Fisher information or a Kronecker-factored block-diagonal version, since the full Hessian is too large).^[The Hessian measures curvature: how sharply the log-posterior bends at its peak. A sharp peak (large curvature) means the data pins the parameters down tightly, so the posterior is narrow and epistemic uncertainty is low. A flat peak means many parameter settings fit nearly as well, so uncertainty is high. Inverting $H$ turns curvature into a covariance.] It is cheap because it reuses the trained weights and only adds a curvature estimate, and modern "last-layer" Laplace applies the Gaussian only to the final layer to keep $H$ small.

### Variational inference (brief)

Variational inference (VI) replaces the intractable posterior with a tractable family $q_\phi(\theta)$, typically a factorized Gaussian, and fits $\phi$ by maximizing the evidence lower bound (ELBO):

$$
\mathcal{L}(\phi) = E_{q_\phi(\theta)}\big[\log p(\mathcal{D} \mid \theta)\big] - \operatorname{KL}\!\big(q_\phi(\theta)\,\|\,p(\theta)\big).
$$

The first term rewards fitting the data; the KL term keeps $q_\phi$ close to the prior and acts as a complexity penalty. "Bayes by Backprop" makes this differentiable with the reparameterization trick so it trains with the usual stochastic gradient tooling. VI gives a genuine approximate posterior in one training run, but the factorized Gaussian usually underestimates variance and the implementation is more involved than the other three.

### Comparison

Table \@ref(tab:bayesian-deep-learning-method-comparison) lays the four methods side by side so the trade-offs are easy to scan.

| Method | Posterior approximation | Extra training cost | Extra inference cost | Captures | Calibration in practice | Code intrusiveness |
|---|---|---|---|---|---|---|
| Deep ensembles | $M$ modes (samples) | $\times M$ | $\times M$ forward passes | Epistemic (+ aleatoric if variance head) | Usually best | Low (train loop $M$ times) |
| MC dropout | Implicit, dropout-induced | None (reuse model) | $\times T$ forward passes | Epistemic, rate-dependent | Fair, needs tuning | Very low |
| Laplace | Gaussian at the mode | Small (curvature) | Small (last-layer) to moderate | Epistemic (local) | Good if curvature is reliable | Low (post-hoc) |
| Variational inference | Tractable family $q_\phi$ | Comparable to one fit | $\times T$ samples | Epistemic | Often underestimates | High (rewrite layers) |

Table: (\#tab:bayesian-deep-learning-method-comparison) Comparison of the four practical uncertainty quantification methods across posterior approximation, compute cost, what they capture, calibration, and how much they disturb existing training code.

Reading the table by column tells you what you are buying. The "code intrusiveness" column is the one that most often decides the choice in practice: a method you can add in an afternoon beats a theoretically nicer one that takes a week, especially for a first pass. A common default follows from this:

- Start with a small deep ensemble ($M = 5$); it is the strongest baseline and the easiest to reason about.
- Reach for MC dropout when retraining is impossible and the model already has dropout layers.
- Use last-layer Laplace when you need cheap post-hoc intervals on an already-trained model.
- Use VI when memory rules out an ensemble and you can afford the engineering.

With the methods surveyed, the next two sections make the ideas concrete in plain R, first by resampling (a bootstrap ensemble) and then in closed form (Bayesian linear regression).

## Runnable demo: bootstrap deep ensemble for predictive intervals

The cleanest way to see the aleatoric/epistemic split without a GPU is a bootstrap ensemble of flexible regressors. A bootstrap resample draws $n$ rows from the training set with replacement, so each member sees a slightly different dataset and lands on a slightly different fitted function. That collection of functions plays the role of samples from the posterior: a different $\theta^{(m)}$ per member. The spread of member means is then epistemic, and the residual noise around the truth is aleatoric.

To make both kinds of uncertainty visible at once, we rig the data deliberately. The noise grows with $x$ (heteroscedastic), so the aleatoric part should widen to the right regardless of how much data we have. And we make the data sparse beyond $x = 2.5$, so the epistemic part should widen on the right because the members disagree where they have few points to anchor them. A good UQ method should pick up both effects.

> **Intuition:** Bootstrapping is a poor person's posterior. We cannot enumerate every plausible model, so we perturb the data and refit many times; the wobble in the resulting predictions stands in for our uncertainty about which model is right.

```{r, message=FALSE, warning=FALSE}
set.seed(42)

# Heteroscedastic truth: noise grows with x. Data is dense on the left,
# sparse on the right, so epistemic uncertainty should grow to the right.
f_true   <- function(x) sin(2 * x) + 0.5 * x
sd_noise <- function(x) 0.10 + 0.18 * x          # aleatoric sd grows with x

n  <- 160
x  <- c(runif(120, 0, 2.5), runif(40, 2.5, 5))   # sparse beyond x = 2.5
y  <- f_true(x) + rnorm(n, 0, sd_noise(x))
dat <- data.frame(x = x, y = y)

grid <- data.frame(x = seq(0, 5, length.out = 200))

# Bootstrap ensemble of smoothing splines. Each member sees a resample,
# giving a different fitted function (a draw of theta).
M    <- 200
fits <- matrix(NA_real_, nrow = nrow(grid), ncol = M)
for (m in seq_len(M)) {
  idx <- sample.int(n, n, replace = TRUE)
  sm  <- smooth.spline(dat$x[idx], dat$y[idx], cv = FALSE)
  fits[, m] <- predict(sm, grid$x)$y
}

# Epistemic: spread of the member mean functions.
mean_pred <- rowMeans(fits)
epi_var   <- apply(fits, 1, var)

# Aleatoric: estimate residual variance, then smooth log of squared
# residuals against x so it can vary (heteroscedastic estimate).
res2      <- (dat$y - predict(smooth.spline(dat$x, dat$y, cv = FALSE), dat$x)$y)^2
ale_model <- smooth.spline(dat$x, log(res2 + 1e-6), cv = FALSE)
ale_var   <- pmax(exp(predict(ale_model, grid$x)$y), 1e-6)

# Total predictive sd from the law of total variance.
total_sd <- sqrt(epi_var + ale_var)

head(data.frame(x = grid$x, mean = mean_pred,
                epi_sd = sqrt(epi_var), ale_sd = sqrt(ale_var)))
```

Figure \@ref(fig:bayesian-deep-learning-bootstrap-bands) shows the decomposition. The dark band is epistemic only; the lighter band adds aleatoric noise to give a 95% predictive interval. Notice the epistemic band widens where data is sparse (right side), while the aleatoric band widens with $x$ regardless of data density.

```{r bayesian-deep-learning-bootstrap-bands, message=FALSE, warning=FALSE, fig.width=7, fig.height=4.5, fig.cap="Bootstrap deep ensemble. Epistemic uncertainty (dark band) grows where training data is sparse; aleatoric noise (light band) grows with x by construction."}
library(ggplot2)

band <- data.frame(
  x        = grid$x,
  mean     = mean_pred,
  epi_lo   = mean_pred - 1.96 * sqrt(epi_var),
  epi_hi   = mean_pred + 1.96 * sqrt(epi_var),
  tot_lo   = mean_pred - 1.96 * total_sd,
  tot_hi   = mean_pred + 1.96 * total_sd,
  truth    = f_true(grid$x)
)

ggplot() +
  geom_ribbon(data = band, aes(x, ymin = tot_lo, ymax = tot_hi),
              fill = "steelblue", alpha = 0.20) +
  geom_ribbon(data = band, aes(x, ymin = epi_lo, ymax = epi_hi),
              fill = "steelblue", alpha = 0.45) +
  geom_point(data = dat, aes(x, y), size = 0.9, alpha = 0.5) +
  geom_line(data = band, aes(x, mean), color = "navy", linewidth = 0.9) +
  geom_line(data = band, aes(x, truth), color = "firebrick",
            linetype = "dashed", linewidth = 0.7) +
  labs(x = "x", y = "y",
       title = "Predictive intervals from a bootstrap ensemble",
       subtitle = "Dark = epistemic 95%, light = epistemic + aleatoric 95%, red dashed = truth") +
  theme_minimal(base_size = 12)
```

The printed table confirms the intended behavior at a glance: the epistemic standard deviation (`epi_sd`) and aleatoric standard deviation (`ale_sd`) are reported separately for each grid point, so you can see the two pieces before they are combined into the total band.

A figure is persuasive but not proof. The honest test of any UQ method is calibration: a 95% predictive interval should contain close to 95% of fresh, held-out points. We check that directly on new data drawn from the same process.

```{r, message=FALSE, warning=FALSE}
# Coverage check on fresh test data drawn from the same process.
set.seed(7)
x_te  <- runif(2000, 0, 5)
y_te  <- f_true(x_te) + rnorm(2000, 0, sd_noise(x_te))

# Interpolate the fitted mean and total sd onto the test inputs.
mu_te  <- approx(grid$x, mean_pred, xout = x_te, rule = 2)$y
sd_te  <- approx(grid$x, total_sd,  xout = x_te, rule = 2)$y

lo <- mu_te - 1.96 * sd_te
hi <- mu_te + 1.96 * sd_te
coverage <- mean(y_te >= lo & y_te <= hi)
round(coverage, 3)   # target is about 0.95
```

The observed coverage lands somewhat below the nominal 95%, which is itself instructive. The aleatoric variance here is estimated by smoothing squared residuals, a deliberately rough plug-in that tends to undershoot the true noise, so the bands come out a little too narrow. This is exactly why the calibration check matters: a plausible-looking figure can still be overconfident, and only held-out coverage reveals it.

> **Warning:** Never judge an uncertainty method by how tight its bands look. Tight and wrong is worse than wide and honest. Always pair a figure with a numeric coverage or scoring-rule check on data the model did not see.

## Runnable demo: Bayesian linear regression posterior predictive

For a model where the posterior is available in closed form, we can show credible bands exactly rather than by resampling. Bayesian linear regression with a conjugate Gaussian prior is the canonical example, and it makes the aleatoric/epistemic split fully analytic.

Take the model $y = X\beta + \epsilon$ with $\epsilon \sim N(0, \sigma^2 I)$, a known $\sigma^2$ for simplicity, and a Gaussian prior $\beta \sim N(0, \tau^2 I)$. The posterior over coefficients is Gaussian,

$$
\beta \mid \mathcal{D} \sim N(\mu_n, \Sigma_n),
\qquad
\Sigma_n = \Big(\tfrac{1}{\sigma^2} X^\top X + \tfrac{1}{\tau^2} I\Big)^{-1},
\qquad
\mu_n = \tfrac{1}{\sigma^2}\, \Sigma_n X^\top y .
$$

For a new input row $x_*$, the posterior predictive is Gaussian with

$$
E(y_* \mid x_*, \mathcal{D}) = x_*^\top \mu_n,
\qquad
\operatorname{Var}(y_* \mid x_*, \mathcal{D}) = \underbrace{x_*^\top \Sigma_n x_*}_{\text{epistemic}} + \underbrace{\sigma^2}_{\text{aleatoric}} .
$$

The two variance terms are exactly the decomposition from the law of total variance, with the epistemic term shrinking as $X^\top X$ grows with more data. The aleatoric term $\sigma^2$ is constant here because we assumed homoscedastic noise. We use a polynomial basis so the mean function can curve.

> **Key idea:** This closed-form model is the reference standard the deep methods are imitating. The bootstrap ensemble above approximated $x_*^\top \Sigma_n x_*$ by resampling; here we compute it exactly. Seeing the exact answer once makes it clear what the approximations are reaching for.

```{r bayesian-deep-learning-bayes-poly, message=FALSE, warning=FALSE, fig.width=7, fig.height=4.5, fig.cap="Closed-form Bayesian polynomial regression. The narrow band is the credible interval on the mean function (epistemic); the wide band is the posterior predictive interval that also includes observation noise (aleatoric)."}
set.seed(123)

# Data on a limited x-range so epistemic uncertainty blows up on extrapolation.
n   <- 40
xb  <- runif(n, -2, 2)
yb  <- 0.5 * xb^3 - xb + rnorm(n, 0, 1.0)

sigma2 <- 1.0      # known observation noise (aleatoric)
tau2   <- 100      # weakly informative prior on coefficients
deg    <- 3

# Design matrices (intercept + polynomial terms).
make_X <- function(x) cbind(1, poly(x, deg, raw = TRUE))
X  <- make_X(xb)
p  <- ncol(X)

# Closed-form posterior over beta.
Sig_n <- solve(crossprod(X) / sigma2 + diag(p) / tau2)
mu_n  <- (Sig_n %*% t(X) %*% yb) / sigma2

# Predict on a grid that extends BEYOND the training range to show extrapolation.
xg <- seq(-3.5, 3.5, length.out = 200)
Xg <- make_X(xg)

mean_g    <- as.numeric(Xg %*% mu_n)
epi_var_g <- rowSums((Xg %*% Sig_n) * Xg)     # x*' Sigma_n x* per row
pred_var  <- epi_var_g + sigma2               # add aleatoric noise

bdf <- data.frame(
  x        = xg,
  mean     = mean_g,
  cred_lo  = mean_g - 1.96 * sqrt(epi_var_g),
  cred_hi  = mean_g + 1.96 * sqrt(epi_var_g),
  pred_lo  = mean_g - 1.96 * sqrt(pred_var),
  pred_hi  = mean_g + 1.96 * sqrt(pred_var)
)

library(ggplot2)
ggplot() +
  geom_ribbon(data = bdf, aes(x, ymin = pred_lo, ymax = pred_hi),
              fill = "darkorange", alpha = 0.18) +
  geom_ribbon(data = bdf, aes(x, ymin = cred_lo, ymax = cred_hi),
              fill = "darkorange", alpha = 0.45) +
  geom_line(data = bdf, aes(x, mean), color = "darkred", linewidth = 0.9) +
  geom_point(data = data.frame(x = xb, y = yb), aes(x, y),
             size = 1.3, alpha = 0.7) +
  geom_vline(xintercept = c(-2, 2), linetype = "dotted") +
  labs(x = "x", y = "y",
       title = "Bayesian polynomial regression: posterior predictive",
       subtitle = "Dark band = 95% credible (mean), light band = 95% predictive (mean + noise); dotted = training range") +
  theme_minimal(base_size = 12)
```

Figure \@ref(fig:bayesian-deep-learning-bayes-poly) shows the result. Outside the training range (the dotted lines), the epistemic band fans out fast: the model knows it is extrapolating. Inside the range the credible band is tight and the predictive band is dominated by the constant aleatoric floor $\sigma^2$. This is exactly the behavior we want from a UQ method and is the gold standard the approximate deep learning methods are trying to mimic.

## Deep learning implementations (not executed here)

The same ideas in a neural network need Keras or Torch. The code below is correct and current but is marked `eval=FALSE` because training is GPU-heavy and the goal here is the pattern, not the run. Read these as templates: the structure carries over directly, and the uncertainty bookkeeping at the end of each is the same law-of-total-variance split you have already seen.

MC dropout in Keras keeps dropout active at inference by calling the model in training mode:

```{r, eval=FALSE}
library(keras)

# A small regression network with dropout after each hidden layer.
inputs <- layer_input(shape = ncol(x_train))
out <- inputs |>
  layer_dense(64, activation = "relu") |>
  layer_dropout(rate = 0.1) |>
  layer_dense(64, activation = "relu") |>
  layer_dropout(rate = 0.1) |>
  layer_dense(1)
model <- keras_model(inputs, out)
model |> compile(optimizer = "adam", loss = "mse")
model |> fit(x_train, y_train, epochs = 200, batch_size = 32, verbose = 0)

# MC dropout: run T stochastic forward passes with training = TRUE so the
# dropout masks stay random at prediction time.
mc_predict <- function(model, x_new, T = 100) {
  draws <- replicate(T, as.numeric(model(x_new, training = TRUE)))
  list(mean = rowMeans(draws),
       epistemic_sd = apply(draws, 1, sd))
}
pred <- mc_predict(model, x_test, T = 200)
```

A deep ensemble with a variance head (so each member reports its own aleatoric noise) trained with the Gaussian negative log-likelihood:

```{r, eval=FALSE}
library(keras)
library(tensorflow)

# Each member outputs mean and log-variance; the loss is the heteroscedastic
# Gaussian NLL (Nix and Weigend, 1994; Lakshminarayanan et al., 2017).
gaussian_nll <- function(y_true, y_pred) {
  mu     <- y_pred[, 1, drop = FALSE]
  logvar <- y_pred[, 2, drop = FALSE]
  0.5 * tf$reduce_mean(tf$exp(-logvar) * (y_true - mu)^2 + logvar)
}

build_member <- function() {
  inp <- layer_input(shape = ncol(x_train))
  h <- inp |> layer_dense(64, activation = "relu") |>
              layer_dense(64, activation = "relu")
  out <- layer_concatenate(list(
    h |> layer_dense(1),                          # mean head
    h |> layer_dense(1)                           # log-variance head
  ))
  m <- keras_model(inp, out)
  m |> compile(optimizer = "adam", loss = gaussian_nll)
  m
}

M <- 5
members <- lapply(seq_len(M), function(i) {
  m <- build_member()
  m |> fit(x_train, y_train, epochs = 200, batch_size = 32, verbose = 0)
  m
})

# Combine members: epistemic from spread of means, aleatoric from mean of
# member variances, per the law of total variance.
preds   <- lapply(members, function(m) as.matrix(predict(m, x_test)))
mus     <- sapply(preds, function(p) p[, 1])
vars    <- sapply(preds, function(p) exp(p[, 2]))
mean_pred <- rowMeans(mus)
epistemic <- apply(mus, 1, var)
aleatoric <- rowMeans(vars)
total_sd  <- sqrt(epistemic + aleatoric)
```

The torch equivalent of a last-layer Laplace fit, which keeps the network's trained body fixed and fits a Gaussian only over the final linear layer, follows the same template using `nn_module` and the generalized Gauss-Newton approximation to the Hessian:

```{r, eval=FALSE}
library(torch)

# Sketch: after standard MAP training, freeze the feature extractor and
# compute the posterior precision of the last layer as
#   H = sum_i J_i' J_i / sigma^2 + I / tau^2
# where J_i is the Jacobian of the output wrt last-layer weights at x_i.
# Predictive variance at x* is then phi(x*)' H^{-1} phi(x*) + sigma^2,
# with phi(x*) the frozen features. Libraries such as laplace-torch
# automate this; the math matches the Laplace section above.
```

## Practical guidance, pitfalls, and when to use each

Always validate calibration, not just sharpness. A method can produce tight, confident-looking bands that are wrong. Check empirical coverage (does the 90% interval contain 90% of held-out points?) and use proper scoring rules such as the negative log predictive density or the continuous ranked probability score. A reliability diagram for classification is the analog. The coverage check in the bootstrap demo is the minimum you should run.

Do not confuse the two uncertainties when acting on them. High aleatoric uncertainty means "this region is inherently noisy; collecting more data will not help much." High epistemic uncertainty means "the model has not seen enough here; more data or abstention will help." Active learning targets epistemic; deciding whether a measurement is worth taking targets aleatoric. Reporting only total variance hides which lever to pull.

Method-specific traps.

- MC dropout uncertainty is sensitive to the dropout rate and does not generally go to zero with infinite data, so treat it as a rough, cheap signal rather than a calibrated posterior. Tune the rate on a validation set.
- Deep ensembles need genuine diversity. Identical initializations, identical data order, and over-regularization collapse the members onto the same solution and the epistemic band vanishes. Vary seeds and use bootstrap or different shuffles.
- The Laplace approximation is only as good as its Hessian. The full Hessian is intractable for large nets, and crude diagonal approximations can be badly miscalibrated; last-layer or Kronecker-factored versions are the practical compromise. It is also a local, single-mode approximation and will miss multimodality.
- Variational inference with a mean-field Gaussian systematically underestimates posterior variance because it cannot represent correlations between weights. Expect overconfidence and budget time for tuning the KL weight.

Distribution shift breaks everything. All these methods quantify uncertainty relative to the training distribution. Under genuine shift (new feature regimes, new label semantics) even well-calibrated models can be confidently wrong. Pair UQ with explicit out-of-distribution monitoring rather than assuming the predictive variance alone will catch it.

A pragmatic decision rule. If you can afford to train a handful of models, a deep ensemble of size 5 is the strong default and the hardest baseline to beat. If the model is already trained and frozen, add MC dropout (if it has dropout layers) or a last-layer Laplace fit for cheap post-hoc intervals. Reserve variational inference for cases where memory forbids an ensemble and you are prepared for the extra engineering. For small tabular problems, the closed-form Bayesian linear or Gaussian process models in the earlier chapters give exact uncertainty and should be tried before any deep method.

## Summary

Three ideas carry the whole chapter. First, uncertainty comes in two flavors that demand different responses: aleatoric noise that more data cannot remove, and epistemic doubt that more data can. Second, the law of total variance splits a predictive variance cleanly into those two pieces once you can sample plausible parameter settings, which reduces every method to the same recipe of averaging member noise and measuring member disagreement. Third, the four practical methods (deep ensembles, MC dropout, Laplace, variational inference) are just different ways to produce those parameter samples, trading off compute against how much they disturb your code, and a size-5 ensemble is the default that is hardest to beat. Whichever you choose, validate it by held-out coverage rather than by how convincing the bands look, and remember that all of these methods speak only about the training distribution, so genuine distribution shift still needs separate monitoring.

## Further reading

- Kendall and Gal (2017), "What Uncertainties Do We Need in Bayesian Deep Learning for Computer Vision?", on the aleatoric/epistemic decomposition.
- Gal and Ghahramani (2016), "Dropout as a Bayesian Approximation", the MC dropout paper.
- Lakshminarayanan, Pritzel, and Blundell (2017), "Simple and Scalable Predictive Uncertainty Estimation using Deep Ensembles".
- Blundell, Cornebise, Kavukcuoglu, and Wierstra (2015), "Weight Uncertainty in Neural Networks" (Bayes by Backprop), for variational inference.
- MacKay (1992), "A Practical Bayesian Framework for Backpropagation Networks", the original Laplace approximation for neural networks.
- Daxberger, Kristiadi, Immer, Eschenhagen, Bauer, and Hennig (2021), "Laplace Redux", on scalable post-hoc Laplace methods.
- Bishop (2006), *Pattern Recognition and Machine Learning*, chapters 3 and 5, for the conjugate Bayesian linear regression and Laplace foundations.
- Gneiting and Raftery (2007), "Strictly Proper Scoring Rules, Prediction, and Estimation", for evaluating probabilistic forecasts.
