# Meta-Learning and Few-Shot Learning

Most supervised methods in this book assume access to a large labeled training set for one fixed task. Meta-learning relaxes that assumption. The goal is to build a system that can adapt to a *new* task from only a handful of labeled examples, by reusing structure learned across many *related* tasks. The slogan is "learning to learn": rather than learning a single predictor, we learn a procedure that produces good predictors quickly.

This chapter covers the conceptual core of meta-learning (optimization-based methods such as MAML and metric-based methods such as prototypical and matching networks), the episodic training protocol, and the $N$-way $K$-shot evaluation setup. The runnable demonstration is a nearest-centroid (prototypical) few-shot classifier implemented in base R, evaluated on simulated episodes, where we measure how accuracy improves as the number of shots per class grows.

> **Intuition:** A standard model learns *what the answer is* for one task. A meta-learner learns *how to find the answer* for a whole family of tasks, so that when a brand new task arrives with only a few labels, it already knows where to look. Think of the difference between memorizing one map and learning to read maps in general.

## Where this fits in a modern ML workflow

In a typical applied setting you train once on a large dataset and serve that model. Few-shot learning targets a different operating regime:

- A new product category appears with only 5 labeled images per class.
- A fraud pattern emerges and you have 10 confirmed cases.
- A clinical phenotype is rare, so per-patient labels are scarce.

Collecting thousands of labels per new task is slow or impossible. Meta-learning amortizes that cost: you pay an expensive offline phase on many tasks that resemble the deployment tasks, and in return you get cheap, fast adaptation at deployment. In modern AI pipelines this idea sits next to two cousins that practitioners often confuse with it:

- Transfer learning / fine-tuning. Pretrain a large model, then fine-tune on the new task. This usually needs more than a handful of labels and a gradient-descent run per task.
- In-context learning with large language models. The model conditions on a few examples in its prompt and adapts with no weight update. This is, in effect, meta-learning whose "training across tasks" happened implicitly during pretraining.

Few-shot meta-learning is the explicit, supervised version of the same objective: optimize directly for fast adaptation from few examples.

> **Key idea:** The cost of learning a new task is not eliminated, it is moved. Meta-learning pays a heavy bill once, offline, by training across many tasks, and is repaid with cheap, fast adaptation every time a new task arrives.

### Notation and the task distribution

We assume a distribution over tasks $p(\mathcal{T})$. A task $\mathcal{T}_i$ is itself a small learning problem with its own data. For classification, each task draws $N$ classes and provides:

- a support set $\mathcal{S} = \{(x_j, y_j)\}_{j=1}^{N K}$ with $K$ labeled examples per class (so $N K$ examples total),
- a query set $\mathcal{Q} = \{(x_j^\*, y_j^\*)\}$ used to score how well the model adapted using $\mathcal{S}$.

This is the $N$-way $K$-shot setup: $N$ classes, $K$ labeled examples each.^[The vocabulary is borrowed from card games and learning theory: a "shot" is one labeled example you are allowed to see per class, and "$N$-way" counts how many classes the model must choose among. A 5-way 1-shot task is therefore a five-class problem with a single labeled example per class.] A "5-way 1-shot" task means five classes with a single labeled example per class. The important constraint is that the classes (and often the entire label space) differ between tasks. A model that simply memorizes a fixed set of categories cannot solve held-out tasks; it must learn a transferable adaptation rule.

> **Note:** The split into support and query within a single task is what makes meta-learning more than a renaming of supervised learning. The support set plays the role of "training data the model is allowed to adapt on" and the query set plays the role of "test data," but both live *inside* one task, and we repeat this train/test split thousands of times across tasks.

## Episodic training

The training signal mirrors the test condition. Instead of iterating over mini-batches of individual examples, we iterate over episodes, where each episode is a sampled task $(\mathcal{S}, \mathcal{Q})$. The meta-objective averages the query loss after adaptation on the support set:

$$
\min_{\theta}\; \mathbb{E}_{\mathcal{T}_i \sim p(\mathcal{T})}\;
\big[\, \mathcal{L}_{\mathcal{Q}_i}\big( A(\theta, \mathcal{S}_i) \big) \,\big],
$$

where $\theta$ are the meta-parameters, $A(\theta, \mathcal{S})$ is the adaptation procedure that turns $\theta$ and a support set into a task-specific predictor, and $\mathcal{L}_{\mathcal{Q}}$ is the loss on the query set. Different meta-learning families differ mainly in what $A$ is.

The reason episodic training works is that we never let the model see the query labels during adaptation; we only score on them. The gradient through that score teaches $\theta$ to make adaptation generalize, not to memorize.

> **Intuition:** Train the way you will be tested. If deployment means "adapt from a handful of labels, then predict on new points," then every training episode should rehearse exactly that, adapt on a support set, predict on a held-out query set. The meta-objective rewards adaptation rules that transfer.

With the training protocol in place, the families of methods differ only in the choice of the adaptation procedure $A$. We look at two: one that adapts by taking gradient steps, and one that adapts by comparing distances.

## Optimization-based meta-learning: MAML (conceptual)

Model-Agnostic Meta-Learning (Finn, Abbeel, and Levine, 2017) chooses $A$ to be a few steps of gradient descent. The meta-parameters $\theta$ are an *initialization*. For a task with support set $\mathcal{S}_i$, the adapted parameters after one inner step are

$$
\theta_i' = \theta - \alpha \nabla_{\theta}\, \mathcal{L}_{\mathcal{S}_i}(\theta),
$$

with inner step size $\alpha$. The outer (meta) objective evaluates these adapted parameters on the query set and updates the initialization:

$$
\theta \leftarrow \theta - \beta \,\nabla_{\theta} \sum_{i} \mathcal{L}_{\mathcal{Q}_i}(\theta_i').
$$

Because $\theta_i'$ already depends on $\theta$, the outer gradient differentiates through the inner update, producing a second-order term involving the Hessian $\nabla^2_\theta \mathcal{L}_{\mathcal{S}_i}(\theta)$. Expanding the chain rule for one inner step,

$$
\nabla_{\theta}\, \mathcal{L}_{\mathcal{Q}_i}(\theta_i')
= \big( I - \alpha\, \nabla^2_{\theta}\, \mathcal{L}_{\mathcal{S}_i}(\theta) \big)\,
\nabla_{\theta'} \mathcal{L}_{\mathcal{Q}_i}(\theta_i').
$$

The intuition: MAML does not look for parameters that are good on any single task. It looks for an initialization from which a *short* gradient run lands on a good task-specific solution. First-order variants (FOMAML, Reptile from Nichol, Achiam, and Schulman, 2018) drop the Hessian term and work well in practice at lower cost.

MAML is "model-agnostic" because $A$ is just gradient descent, so it applies to any differentiable model. The price is the inner loop: each meta-update requires adapting and backpropagating through adaptation, which is memory-hungry and sensitive to the inner step count and learning rate.

> **Warning:** Differentiating through the inner gradient step is what produces the Hessian term above, and it is the main source of MAML's fragility and cost. If MAML is unstable or too slow, the first thing to try is a first-order variant (FOMAML or Reptile), which drops that second-order term.

Optimization-based methods adapt by *moving the weights*. The next family adapts without touching the weights at all.

## Metric-based meta-learning

Metric-based methods skip the inner gradient loop. The adaptation procedure $A$ embeds inputs into a feature space with a learned encoder $f_\theta$, then classifies query points by comparison to the support set. There is no per-task weight update at deployment, only a distance computation, which makes these methods fast and stable.

### Prototypical networks

Prototypical networks (Snell, Swersky, and Zemel, 2017) represent each class by the mean of its support embeddings (its prototype). For class $c$ with support subset $\mathcal{S}_c$,

$$
\mathbf{p}_c = \frac{1}{|\mathcal{S}_c|} \sum_{(x,y)\in \mathcal{S}_c} f_\theta(x).
$$

A query point $x^\*$ is classified by its squared Euclidean distance to each prototype,^[The prototype $\mathbf{p}_c$ is just the centroid (average position) of class $c$'s support embeddings. Classifying a query by its nearest prototype is the centuries-old nearest-centroid rule, lifted into a learned feature space.] turned into a softmax over classes:

$$
P(y^\* = c \mid x^\*) =
\frac{\exp\!\big(-\, d(f_\theta(x^\*), \mathbf{p}_c)\big)}
{\sum_{c'} \exp\!\big(-\, d(f_\theta(x^\*), \mathbf{p}_{c'})\big)},
\qquad
d(u, v) = \lVert u - v \rVert_2^2 .
$$

Training minimizes the negative log probability of the true class on query points, averaged over episodes. A clean fact: with squared Euclidean distance, the prototype rule is a linear classifier in embedding space, since expanding $-\lVert u - \mathbf{p}_c \rVert^2$ gives a term linear in $u$ plus a per-class bias. This is why nearest-centroid classification is the natural base-R stand-in for a prototypical network with a fixed (or identity) encoder.

### Matching networks

Matching networks (Vinyals et al., 2016) predict by a weighted vote over support labels, where the weights are an attention kernel between the query and each support embedding:

$$
\hat{y}^\* = \sum_{(x_j, y_j) \in \mathcal{S}} a(x^\*, x_j)\, \mathbf{1}[y_j = c],
\qquad
a(x^\*, x_j) = \frac{\exp\!\big(\cos(f_\theta(x^\*), g_\theta(x_j))\big)}
{\sum_{k} \exp\!\big(\cos(f_\theta(x^\*), g_\theta(x_k))\big)}.
$$

When $K = 1$, prototypical and matching networks coincide up to the choice of distance (squared Euclidean versus cosine), because each prototype is just the single support point. For $K > 1$, prototypes average within a class while matching networks keep every support point as a separate voter.

### Comparison of the main families

Table \@ref(tab:meta-learning-family-comparison) contrasts the main meta-learning families along the axes that matter most when choosing one: how each adapts at test time, whether it updates weights per task, its deployment cost, and what it is sensitive to.

| Family | Adaptation $A$ at test time | Per-task weight update | Compute at deployment | Sensitive to | Representative work |
|---|---|---|---|---|---|
| Optimization-based (MAML, Reptile) | Few gradient steps from a learned init | Yes | High (inner loop, second-order) | Inner LR, step count | Finn et al. 2017; Nichol et al. 2018 |
| Prototypical networks | Compute class means, nearest prototype | No | Low (distance to $N$ prototypes) | Embedding quality, distance metric | Snell et al. 2017 |
| Matching networks | Attention-weighted vote over support | No | Low to medium ($NK$ comparisons) | Embedding quality, kernel | Vinyals et al. 2016 |
| Fine-tuning a pretrained model | Standard supervised training | Yes | Medium to high | Needs more labels than few-shot | Transfer learning, broadly |

Table: (\#tab:meta-learning-family-comparison) Comparison of the main meta-learning families by adaptation rule, whether weights update per task, deployment compute, key sensitivities, and representative work.

## Runnable demo: nearest-centroid few-shot classifier

We implement the prototypical idea with the simplest possible encoder, the identity map, so the "embedding" is the raw feature vector. We simulate a task distribution where each class is a Gaussian blob with a random center, draw episodes, and classify query points by nearest class centroid (the prototype). We then measure query accuracy as a function of the number of shots $K$.

This isolates the mechanism that prototypical networks rely on: averaging $K$ support points reduces the variance of the prototype estimate, which moves the decision boundary closer to the Bayes-optimal one and raises accuracy. With $K$ independent support points per class, the prototype is the sample mean, whose variance shrinks like $1/K$, so accuracy should rise and then saturate.

> **Key idea:** More shots help because each extra labeled example sharpens the estimate of the class center, not because the model "learns more." The $1/K$ variance reduction of an average is the whole story behind the rising accuracy curve we are about to plot.

```{r proto-setup, eval=TRUE}
set.seed(1301)

# Sample one N-way K-shot episode from a Gaussian task distribution.
# Each class is a blob with a random center in p dimensions and shared sd.
sample_episode <- function(n_way, k_shot, n_query, p = 2, sd = 1.0,
                           center_spread = 2.5) {
  centers <- matrix(rnorm(n_way * p, sd = center_spread), nrow = n_way)

  draw <- function(n_per_class) {
    X <- matrix(0, nrow = n_way * n_per_class, ncol = p)
    y <- integer(n_way * n_per_class)
    row <- 1L
    for (c in seq_len(n_way)) {
      for (j in seq_len(n_per_class)) {
        X[row, ] <- centers[c, ] + rnorm(p, sd = sd)
        y[row] <- c
        row <- row + 1L
      }
    }
    list(X = X, y = y)
  }

  list(support = draw(k_shot), query = draw(n_query))
}

# Prototypical classifier with identity encoder: class prototypes are the
# support means, queries go to the nearest prototype (squared Euclidean).
proto_predict <- function(support, query_X, n_way) {
  p <- ncol(support$X)
  protos <- matrix(0, nrow = n_way, ncol = p)
  for (c in seq_len(n_way)) {
    protos[c, ] <- colMeans(support$X[support$y == c, , drop = FALSE])
  }
  # squared Euclidean distance from each query point to each prototype
  d2 <- matrix(0, nrow = nrow(query_X), ncol = n_way)
  for (c in seq_len(n_way)) {
    diff <- sweep(query_X, 2, protos[c, ], "-")
    d2[, c] <- rowSums(diff^2)
  }
  max.col(-d2, ties.method = "first")  # argmin distance = argmax negative dist
}

# Average query accuracy over many episodes for a given K.
eval_shots <- function(k_shot, n_episodes = 400, n_way = 5,
                       n_query = 15, p = 2, sd = 1.0) {
  acc <- numeric(n_episodes)
  for (e in seq_len(n_episodes)) {
    ep <- sample_episode(n_way, k_shot, n_query, p = p, sd = sd)
    pred <- proto_predict(ep$support, ep$query$X, n_way)
    acc[e] <- mean(pred == ep$query$y)
  }
  mean(acc)
}
```

Now we sweep the number of shots and record mean accuracy. We repeat the sweep at two noise levels to show how task difficulty interacts with the number of shots. Table \@ref(tab:meta-learning-shot-accuracy) reports the mean query accuracy at each shot count under both noise levels.

```{r proto-sweep, eval=TRUE}
shots <- c(1, 2, 3, 5, 10, 20)

set.seed(7)
acc_easy <- sapply(shots, eval_shots, n_episodes = 400, n_way = 5, sd = 1.0)
set.seed(7)
acc_hard <- sapply(shots, eval_shots, n_episodes = 400, n_way = 5, sd = 1.6)

results <- data.frame(
  shots          = shots,
  acc_low_noise  = round(acc_easy, 3),
  acc_high_noise = round(acc_hard, 3)
)
```

```{r meta-learning-shot-accuracy, eval=TRUE}
knitr::kable(
  results,
  caption = "Mean query accuracy of the nearest-centroid few-shot classifier on simulated 5-way episodes, by shots per class and within-class noise level, averaged over 400 episodes.",
  col.names = c("shots (K)", "accuracy (low noise)", "accuracy (high noise)")
)
```

Figure \@ref(fig:meta-learning-proto-figure) plots accuracy against the number of shots. The chance baseline for a 5-way task is $1/5 = 0.2$.

```{r meta-learning-proto-figure, eval=TRUE, fig.cap="Few-shot accuracy of a nearest-centroid (prototypical) classifier on simulated 5-way episodes. Accuracy rises with the number of shots K and saturates; higher within-class noise lowers the whole curve.", fig.width=7, fig.height=4.5}
library(ggplot2)

plot_df <- rbind(
  data.frame(shots = shots, accuracy = acc_easy, noise = "low (sd = 1.0)"),
  data.frame(shots = shots, accuracy = acc_hard, noise = "high (sd = 1.6)")
)

ggplot(plot_df, aes(shots, accuracy, color = noise)) +
  geom_hline(yintercept = 0.2, linetype = "dashed", color = "grey50") +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  annotate("text", x = max(shots), y = 0.22,
           label = "chance (1/5)", hjust = 1, vjust = 0, size = 3,
           color = "grey40") +
  scale_x_continuous(breaks = shots) +
  scale_y_continuous(limits = c(0.15, 1)) +
  labs(x = "shots per class (K)", y = "mean query accuracy",
       color = "within-class noise",
       title = "Accuracy vs shots for a prototypical few-shot classifier") +
  theme_minimal(base_size = 12)
```

Two patterns are worth reading off the curve. First, even 1-shot accuracy is far above the chance line of $0.2$, because a single support point already locates each blob roughly. Second, the gain from $K=1$ to $K=5$ is large and then flattens, which is the $1/K$ variance reduction of the prototype estimate in action. Raising the noise shifts the whole curve down because the classes overlap more, so no amount of averaging fully separates them.

### A sketch of an encoder-based prototypical network (not run)

The demo above used the identity encoder. In practice the win comes from a *learned* encoder $f_\theta$ trained episodically so that distances in embedding space respect class structure. The following Keras sketch shows the shape of that training loop. It is left unevaluated because it is illustrative pseudocode for the episodic loss rather than a packaged model.

```{r proto-keras, eval=FALSE}
library(keras)

# f_theta: a small embedding network mapping inputs to R^d.
make_encoder <- function(input_dim, embed_dim = 16) {
  keras_model_sequential() |>
    layer_dense(64, activation = "relu", input_shape = input_dim) |>
    layer_dense(embed_dim)
}

# One episode's prototypical loss: embed support and query, form prototypes,
# score queries by negative squared distance, take softmax cross-entropy.
proto_loss <- function(encoder, support_x, support_y, query_x, query_y,
                       n_way) {
  z_s <- encoder(support_x)                     # (N*K, d)
  z_q <- encoder(query_x)                        # (Nq, d)
  protos <- tf$stack(lapply(seq_len(n_way) - 1L, function(c) {
    mask <- tf$equal(support_y, c)
    tf$reduce_mean(tf$boolean_mask(z_s, mask), axis = 0L)
  }))                                            # (N, d)
  # pairwise squared distances query-to-prototype
  d2 <- tf$reduce_sum((tf$expand_dims(z_q, 1L) -
                       tf$expand_dims(protos, 0L))^2, axis = 2L)
  logits <- -d2
  tf$reduce_mean(
    tf$nn$sparse_softmax_cross_entropy_with_logits(
      labels = query_y, logits = logits))
}

# Meta-training: sample episodes, accumulate gradients of proto_loss
# through the encoder, and step an optimizer. Each step is one episode.
```

The base-R demo and this sketch share the same prediction rule. The only difference is whether the feature map is fixed (identity) or trained end to end through the episodic loss.

> **Tip:** When you move from this toy to a real prototypical network, the identity encoder is replaced by `make_encoder`, but the loss, the prototypes, and the nearest-prototype decision rule stay exactly as written here. Read the base-R `proto_predict` and the Keras `proto_loss` side by side: they compute the same thing.

## Practical guidance, pitfalls, and when to use

The methods above are appealing, but the decision to use meta-learning at all is the one that matters most in practice. The single most common mistake is reaching for a meta-learner when a simpler tool would do better, so we start there.

> **When to use this:** Meta-learning earns its complexity only when two conditions hold at once: you face a *stream of related tasks*, and each new task gives you *very few labels*. If either condition fails, a standard model or a fine-tuned pretrained model is usually the better and simpler choice.

When to reach for meta-learning.

- You face a stream of new tasks that resemble each other, and each new task has only a few labels (single digits to low tens per class).
- You can assemble a meta-training set of many tasks drawn from a distribution close to deployment tasks. This last condition is the one people most often violate.

When not to.

- You have one fixed task with plenty of labels. Train a standard model.
- You have one task with few labels but a large related labeled corpus. Fine-tuning a pretrained model is usually simpler and stronger than building a meta-learner.
- Your deployment tasks are not exchangeable with your meta-training tasks. Meta-learning generalizes across tasks only to the extent that train and test tasks come from the same $p(\mathcal{T})$.

Pitfalls.

- Task leakage and overlapping classes. If the same classes appear in meta-training and meta-test, reported few-shot accuracy is inflated. Split by *class*, not just by example.
- Episode composition matters. Reported numbers depend on $N$, $K$, the query size, and how episodes are sampled. Always state the $N$-way $K$-shot configuration, and average over many episodes with confidence intervals, since per-episode accuracy is noisy.
- Choose the family by your deployment budget. Metric-based methods (prototypical, matching) need no per-task gradient step, so they are cheap and robust to serve. Optimization-based methods (MAML) are more flexible but require an inner adaptation loop and careful tuning of the inner learning rate and step count.
- Distance metric and feature scale. Prototypical networks assume distances are meaningful. Unnormalized or differently scaled features distort Euclidean distances; standardize features or learn an embedding that does so.
- Class imbalance within a task. Prototypes from very few points are high variance. With $K=1$ a single outlier support point can dislocate a whole class prototype.

Sanity checks. Compare against a nearest-centroid baseline on raw or pretrained features (exactly the demo above). A learned meta-learner that does not beat this simple baseline is not earning its complexity. Also report the chance level $1/N$ so improvements are interpretable.

> **Warning:** The most damaging error in few-shot work is silent: if a class appears in both meta-training and meta-test, the model has effectively already seen the answer, and reported accuracy is meaningless. Always split by class, not just by example, and state the exact $N$-way $K$-shot configuration alongside every number.

To summarize the chapter: meta-learning trades an expensive offline phase, training across many tasks under the episodic protocol, for cheap adaptation to new tasks from a few labels. Optimization-based methods like MAML adapt by taking gradient steps from a learned initialization, while metric-based methods like prototypical and matching networks adapt by comparing query points to a learned representation of the support set. The runnable demo showed the engine inside the metric-based family in its simplest form: averaging support points to form class prototypes, where accuracy climbs with the number of shots and then saturates as the $1/K$ variance reduction runs out. When you decide whether to use any of this, let the two questions in the callout above, "are the tasks related?" and "are labels truly scarce?", make the call.

## Further reading

- Vinyals, Blundell, Lillicrap, Kavukcuoglu, and Wierstra (2016). "Matching Networks for One Shot Learning."
- Finn, Abbeel, and Levine (2017). "Model-Agnostic Meta-Learning for Fast Adaptation of Deep Networks" (MAML).
- Snell, Swersky, and Zemel (2017). "Prototypical Networks for Few-shot Learning."
- Nichol, Achiam, and Schulman (2018). "On First-Order Meta-Learning Algorithms" (Reptile).
- Ravi and Larochelle (2017). "Optimization as a Model for Few-Shot Learning."
- Hospedales, Antoniou, Micaelli, and Storkey (2021). "Meta-Learning in Neural Networks: A Survey."
- Wang, Yao, Kwok, and Ni (2020). "Generalizing from a Few Examples: A Survey on Few-Shot Learning."
