# Transfer and Multi-Task Learning {#sec-transfer-multitask-learning}

```{r}
#| include: false
source("_common.R")
```

```{r setup-tml, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Most of the models in this book are trained on one dataset to solve one
prediction problem. Transfer learning and multi-task learning relax that
assumption. Transfer learning reuses knowledge learned on a source problem
to help a related target problem, usually one with less labeled data.
Multi-task learning trains a single model on several related tasks at once
so that the tasks share statistical strength. Both ideas rest on the same
bet: if two problems are related, then the data for one carries information
about the other, and a model that respects that relatedness can beat models
trained in isolation.

This chapter explains where these methods fit in a modern workflow, derives
the multi-task objective for the linear case, and gives a runnable base-R
demonstration that compares a joint multi-task estimator to independent
per-task models on simulated related tasks. By the end you should be able to
say, for a given problem, whether to train tasks separately, pool them into
one, or tie them together with a tunable penalty, and to recognize when
sharing backfires.


::: {.callout-tip title="Intuition"}
Think of two students studying for related exams. If they
compare notes, each learns faster than working alone, as long as the
subjects truly overlap. If the subjects are unrelated, swapping notes just
adds confusion. Transfer and multi-task learning automate that judgment:
share when sharing pays, and back off when it does not.
:::


## Intuition and workflow context

A data scientist rarely starts from zero. A churn model for a new product
line can borrow from a churn model on an established line. A sentiment
classifier for a niche domain can start from a general language model
(@sec-llms). A
demand forecast for a store that opened last month can borrow from stores
that have years of history. In each case the target task has little data but
a related source task has a lot, and we want to move usable structure from
source to target.

It helps to fix notation. Let a task be indexed by $t = 1, \dots, T$. Task
$t$ has data $\{(x_{ti}, y_{ti})\}_{i=1}^{n_t}$ drawn from a distribution
$P_t(x, y)$. In standard supervised learning we fit one model per task by
minimizing its own loss. Transfer and multi-task learning instead assume the
$P_t$ share structure, for example a common feature representation or
parameter vectors that are close to one another, and they exploit that
shared structure during fitting.

Two distinctions organize the field.

The first is the direction of reuse. *Transfer learning* is asymmetric: a
source model is trained first, then adapted to a target. *Multi-task
learning* is symmetric: all tasks are trained together and each is allowed
to help the others.

The second is what gets shared. In deep models the dominant transfer pattern
is to share a learned representation (the lower layers of a neural network, see
@sec-neural-networks) and keep
task-specific heads on top.^[A *head* is the small task-specific output layer
that sits on top of a shared *backbone* or *body*. The backbone turns raw
inputs into features; the head maps features to that task's predictions.] In
classical models the shared object is more often the coefficient vector or a
low-dimensional subspace that the coefficients live in.


::: {.callout-important title="Key idea"}
Transfer learning and multi-task learning are not separate
algorithms so much as a stance: assume related problems share structure,
then design the model so that structure can be reused. The rest of this
chapter is about *what* to share and *how strongly*.
:::


In a modern ML/AI workflow these methods show up at predictable points. When
you pull a pretrained image or text encoder and adapt it, that is transfer
learning. When you train one model to predict several related business
outcomes from the same features, that is multi-task learning. When your
training distribution differs from your serving distribution, the repair is
domain adaptation, a special case of transfer where the label space is fixed
but the input distribution shifts.

## Feature extraction versus fine-tuning

The previous section set up the two big choices, direction of reuse and what
gets shared. We now zoom in on the most common modern case: the shared object
is a pretrained representation, and you want to adapt it to your target task.
There are two ways to do that.

*Feature extraction* freezes the pretrained network and uses its output (or
an intermediate layer) as a fixed feature vector. You train only a small new
head, for example a logistic regression or a shallow dense layer, on those
features. This is cheap, needs little target data, and cannot overfit the
backbone because the backbone does not move.

*Fine-tuning* unfreezes some or all of the pretrained weights and continues
training them on the target task, usually with a smaller learning rate so the
pretrained structure is not destroyed. This is more expensive and needs more
target data, but it lets the representation itself adapt to the target.

The choice is governed by how much target data you have and how similar the
source and target are. @tbl-transfer-multitask-learning-adapt-guidance
summarizes the standard guidance, following @Yosinski_2014 and the practical
advice in @Howard_2018.

| Target data | Source vs target similarity | Recommended approach |
|---|---|---|
| Small | Similar | Feature extraction, train head only |
| Small | Different | Feature extraction from earlier layers |
| Large | Similar | Fine-tune the whole network |
| Large | Different | Fine-tune, or train from scratch |

: Recommended adaptation strategy as a function of target data size and source-target similarity. {#tbl-transfer-multitask-learning-adapt-guidance}


::: {.callout-tip title="When to use this"}
With only a few hundred labeled target examples,
default to feature extraction: it has few parameters to estimate and cannot
damage the backbone. Reach for fine-tuning only when target data are
plentiful enough to move the backbone safely without overfitting.
:::


A useful refinement is *gradual unfreezing*: start by training the head with
the backbone frozen, then unfreeze the top blocks, then deeper blocks, each
time with a lower learning rate. This protects the most general (lowest)
layers, which tend to transfer across tasks, while letting the most specific
(highest) layers adapt.

The following Keras sketch shows both patterns on an image backbone. It is
not run here because it needs a GPU and downloaded weights, but it is correct
idiomatic code.

```{r keras-finetune, eval=FALSE}
library(keras)

# Pretrained backbone without its classification head
base <- application_efficientnet_b0(
  include_top = FALSE,
  weights = "imagenet",
  input_shape = c(224, 224, 3),
  pooling = "avg"
)

# --- Pattern 1: feature extraction (backbone frozen) ---
freeze_weights(base)

model <- keras_model_sequential() %>%
  base() %>%
  layer_dropout(0.2) %>%
  layer_dense(units = 10, activation = "softmax")

model %>% compile(
  optimizer = optimizer_adam(learning_rate = 1e-3),
  loss = "categorical_crossentropy",
  metrics = "accuracy"
)
# model %>% fit(train_ds, epochs = 5, validation_data = val_ds)

# --- Pattern 2: fine-tuning (unfreeze top of backbone) ---
unfreeze_weights(base, from = "block6a_expand_conv")

model %>% compile(
  optimizer = optimizer_adam(learning_rate = 1e-5),  # small LR
  loss = "categorical_crossentropy",
  metrics = "accuracy"
)
# model %>% fit(train_ds, epochs = 10, validation_data = val_ds)
```

## Domain adaptation

Feature extraction and fine-tuning assume the source and target solve
different tasks. The next pattern is the opposite: the task stays the same but
the inputs come from a different distribution. Domain adaptation handles the
case where the task is fixed but the input distribution moves. Let the source domain be $P_S(x, y)$ and the target
domain $P_T(x, y)$. *Covariate shift* is the assumption that the conditional
$P(y \mid x)$ is the same in both domains while the marginal $P(x)$ differs,
so $P_S(x) \neq P_T(x)$ but $P_S(y \mid x) = P_T(y \mid x)$.

Under covariate shift the right correction is importance weighting. The
target risk can be written as an expectation over the source distribution
reweighted by the density ratio $w(x) = P_T(x) / P_S(x)$:

$$
\mathbb{E}_{P_T}[\ell(f(x), y)]
= \mathbb{E}_{P_S}\!\left[ \frac{P_T(x)}{P_S(x)}\, \ell(f(x), y) \right].
$$


::: {.callout-tip title="Intuition"}
Importance weighting says: trust source points that look
like target points, and discount source points that do not. A source
example sitting in a region the target rarely visits gets a small weight, so
the fitted model concentrates on the part of input space you actually care
about at serving time.
:::


So you fit the model on source data but weight each source point by how
likely it is under the target marginal. Estimating $w(x)$ directly by
density estimation is hard, so in practice you train a classifier to
distinguish source from target inputs and turn its probability into the
weight, as in @Sugiyama_2007.

### Why the identity holds, and what it costs

The reweighting identity is a one-line change of measure, valid whenever
$P_S(x) > 0$ wherever $P_T(x) > 0$ (the support condition, without which the
ratio is undefined and importance weighting cannot work):

$$
\mathbb{E}_{P_T}[\ell(f(x), y)]
= \int \ell(f(x), y)\, P_T(x)\, P_T(y \mid x)\, dx\, dy
= \int \ell(f(x), y)\, \frac{P_T(x)}{P_S(x)}\, P_S(x)\, P_S(y \mid x)\, dx\, dy ,
$$

where the second equality multiplies and divides by $P_S(x)$ and uses
$P_T(y\mid x) = P_S(y\mid x)$, the covariate-shift assumption. The right-hand
side is $\mathbb{E}_{P_S}[w(x)\,\ell(f(x),y)]$, so the weighted empirical risk
$\hat R_w(f) = \frac{1}{n_S}\sum_{i=1}^{n_S} w(x_i)\,\ell(f(x_i), y_i)$ is an
unbiased and (under standard conditions) consistent estimator of the target
risk. The crucial caveat is that the conditional $P(y\mid x)$ must genuinely be
invariant. If the label rule itself shifts (concept drift), reweighting the
inputs corrects the wrong thing and can make matters worse.

Unbiasedness is not free. The weighted estimator has variance inflated by the
spread of the weights. Its effective sample size is

$$
n_{\text{eff}}
= \frac{\left(\sum_i w(x_i)\right)^2}{\sum_i w(x_i)^2}
= \frac{n_S}{1 + \widehat{\operatorname{CV}}^2(w)} ,
$$ {#eq-transfer-multitask-learning-neff}

where $\widehat{\operatorname{CV}}^2(w)$ is the squared coefficient of variation
of the weights. When source and target overlap poorly, a few points carry
almost all the weight, $\operatorname{CV}^2(w)$ explodes, and $n_{\text{eff}}$
collapses far below $n_S$: the correction is unbiased but useless. This is why
practitioners clip or self-normalize the weights, trading a little bias for a
large reduction in variance, and why @eq-transfer-multitask-learning-neff is
the right diagnostic to monitor before trusting an importance-weighted fit. When representations are learned, an
alternative is to align the source and target feature distributions during
training, for example by matching their means and covariances or by an
adversarial domain classifier, as in @Ganin_2016.

## Parameter sharing: hard and soft

Domain adaptation reused a representation across input distributions. We now
turn to the symmetric, multi-task side: training several tasks together. When
tasks are trained together, the architecture decides how parameters are
shared. There are two canonical schemes, described in @Ruder_2017.

*Hard parameter sharing* uses one shared body for all tasks and a small
task-specific head per task. The shared body is forced to learn a
representation that serves every task. This is the most common multi-task
setup, it is parameter-efficient, and the shared body acts as a strong
regularizer because it must fit all tasks at once.

*Soft parameter sharing* gives each task its own full set of parameters but
adds a penalty that keeps the parameters of different tasks close. Nothing is
literally shared; the coupling is through the penalty. This is more flexible
when tasks are only loosely related, at the cost of more parameters.


::: {.callout-note}
Hard sharing forces tasks to use exactly the same body, so the
coupling is total and not tunable. Soft sharing keeps separate parameters
and tunes the coupling through a penalty, so you can dial sharing from none
to near-total. That tunability is what makes the soft case the right one to
study in detail.
:::


The base-R demonstration below is an instance of soft sharing in its purest
linear form. Each task has its own coefficient vector, and a penalty pulls
the vectors toward a common center.

## Multi-task linear model: derivation

Consider $T$ linear regression tasks. Task $t$ has design matrix
$X_t \in \mathbb{R}^{n_t \times p}$, response $y_t \in \mathbb{R}^{n_t}$, and
coefficient vector $\beta_t \in \mathbb{R}^p$. Independent ordinary least
squares solves $T$ separate problems:

$$
\hat\beta_t^{\,\text{ind}}
= \arg\min_{\beta_t} \; \lVert y_t - X_t \beta_t \rVert_2^2,
\qquad t = 1, \dots, T.
$$

When the tasks are related, their true coefficient vectors are close to a
shared center. We model this with a decomposition $\beta_t = \beta_0 + v_t$,
where $\beta_0$ is a common vector shared by all tasks and $v_t$ is a small
task-specific deviation. We then penalize the deviations.


::: {.callout-tip title="Intuition"}
The split $\beta_t = \beta_0 + v_t$ says each task is the
shared answer plus a small personal correction. Penalizing only $v_t$ keeps
the corrections small unless the data really demand them, which is exactly
the "borrow by default, deviate when justified" behavior we want.
:::


The joint objective is

$$
\min_{\beta_0,\, v_1, \dots, v_T}
\; \sum_{t=1}^{T} \lVert y_t - X_t (\beta_0 + v_t) \rVert_2^2
\;+\; \lambda \sum_{t=1}^{T} \lVert v_t \rVert_2^2 .
$$

The penalty $\lambda$ controls how much the tasks are tied together. As
$\lambda \to 0$ each task is free and we recover independent fits (each task
absorbs everything into its own $v_t$). As $\lambda \to \infty$ all
deviations are forced to zero, $\beta_t = \beta_0$ for every task, and we
recover a single pooled model fit on the stacked data. Intermediate
$\lambda$ interpolates between these two extremes, which is exactly the
multi-task regime: borrow strength across tasks without forcing them to be
identical. This is the regularized multi-task formulation of
@Evgeniou_2004.

The objective is a single quadratic in the stacked parameter vector
$\theta = (\beta_0, v_1, \dots, v_T)$, so it has a closed form. Build a block
design matrix $Z$ where the columns for $\beta_0$ repeat $X_t$ across all
task rows and the columns for $v_t$ contain $X_t$ only on task $t$'s rows.
Stack all responses into $y$. With a diagonal penalty matrix $P$ that
penalizes only the $v_t$ blocks (entries $\lambda$) and not $\beta_0$
(entries $0$), the solution is ridge-like:

$$
\hat\theta = (Z^\top Z + P)^{-1} Z^\top y .
$$


::: {.callout-note}
This is just ridge regression on a cleverly laid-out design
matrix. The only twist is that the penalty matrix $P$ leaves the shared
block $\beta_0$ unpenalized and applies $\lambda$ only to the deviation
blocks $v_t$. Once you see that, the multi-task estimator is no harder to
compute than any other ridge fit.
:::


Each task estimate is then $\hat\beta_t = \hat\beta_0 + \hat v_t$. The next
section builds exactly this $Z$ and $P$ in base R.

### Deriving the closed form from the normal equations

The block formulation above hides the structure. It is worth deriving the
solution directly, because the stationarity conditions expose exactly how the
shared center and the deviations are coupled. Write the objective as

$$
F(\beta_0, v_1, \dots, v_T)
= \sum_{t=1}^{T} \lVert y_t - X_t \beta_0 - X_t v_t \rVert_2^2
\;+\; \lambda \sum_{t=1}^{T} \lVert v_t \rVert_2^2 .
$$

Differentiate with respect to each block and set the gradient to zero. For a
fixed task $t$,

$$
\frac{\partial F}{\partial v_t}
= -2 X_t^\top (y_t - X_t \beta_0 - X_t v_t) + 2 \lambda v_t = 0 ,
$$

which rearranges to the per-task stationarity condition

$$
(X_t^\top X_t + \lambda I)\, v_t = X_t^\top (y_t - X_t \beta_0).
$$ {#eq-transfer-multitask-learning-vt-stationary}

So given the shared center, each deviation is a ridge fit to that task's
residual,

$$
\hat v_t(\beta_0) = (X_t^\top X_t + \lambda I)^{-1} X_t^\top (y_t - X_t \beta_0)
\equiv A_t (y_t - X_t \beta_0),
\qquad A_t := (X_t^\top X_t + \lambda I)^{-1} X_t^\top .
$$

Differentiating with respect to the shared block gives

$$
\frac{\partial F}{\partial \beta_0}
= -2 \sum_{t=1}^{T} X_t^\top (y_t - X_t \beta_0 - X_t v_t) = 0 ,
$$

so $\sum_t X_t^\top (y_t - X_t \beta_0 - X_t \hat v_t) = 0$. Substituting
@eq-transfer-multitask-learning-vt-stationary, namely
$X_t^\top X_t \hat v_t = X_t^\top(y_t - X_t\beta_0) - \lambda \hat v_t$, into
$X_t^\top(y_t - X_t\beta_0) - X_t^\top X_t \hat v_t = \lambda \hat v_t$ shows the
$\beta_0$ condition collapses to

$$
\sum_{t=1}^{T} \lambda \, \hat v_t = 0
\quad\Longleftrightarrow\quad
\sum_{t=1}^{T} \hat v_t = 0 .
$$ {#eq-transfer-multitask-learning-deviation-sum}

The deviations sum to zero at the optimum: the shared center is the point from
which the task-specific corrections balance out, which is the precise sense in
which $\beta_0$ is a "center." Substituting
$\hat v_t = A_t(y_t - X_t\beta_0)$ into
@eq-transfer-multitask-learning-deviation-sum yields a single linear system
for $\beta_0$,

$$
\left( \sum_{t=1}^{T} A_t X_t \right) \beta_0 = \sum_{t=1}^{T} A_t y_t ,
$$

after which each $\hat v_t$ follows in closed form. This is algebraically
identical to the block solution
$\hat\theta = (Z^\top Z + P)^{-1} Z^\top y$, but it makes the mechanism
transparent: profiling out the deviations leaves a $p \times p$ system in the
shared center, so the multi-task fit costs essentially one ridge solve plus $T$
small ridge solves, not one dense $(T+1)p$ solve, when implemented carefully.

### The ridge-on-deviations as a Gaussian prior

The penalty $\lambda \sum_t \lVert v_t \rVert_2^2$ is not arbitrary. It is the
negative log of a hierarchical Gaussian prior. Take

$$
\beta_t = \beta_0 + v_t, \qquad
v_t \sim \mathcal{N}(0, \tau^2 I), \qquad
y_t \mid X_t, \beta_t \sim \mathcal{N}(X_t \beta_t, \sigma^2 I),
$$

with a flat prior on $\beta_0$. The negative log posterior is, up to constants,

$$
\frac{1}{2\sigma^2} \sum_t \lVert y_t - X_t(\beta_0 + v_t) \rVert_2^2
+ \frac{1}{2\tau^2} \sum_t \lVert v_t \rVert_2^2 ,
$$

which is exactly the joint objective with $\lambda = \sigma^2 / \tau^2$. The
sharing strength is therefore an estimate of the noise-to-heterogeneity ratio.
Small $\tau^2$ (tasks tightly clustered around the center) sends
$\lambda \to \infty$ and recovers pooling; large $\tau^2$ (tasks free to roam)
sends $\lambda \to 0$ and recovers independent fits. This is the formal content
of the empirical observation that the best $\lambda$ is interior whenever the
tasks are related but not identical: $\tau^2$ is finite and positive.

::: {.callout-note}
The hierarchical-prior reading turns "tune $\lambda$ by cross-validation" into
"estimate the heterogeneity $\tau^2$." With many tasks one can estimate
$\sigma^2$ and $\tau^2$ directly by marginal (empirical Bayes) likelihood and
set $\lambda = \hat\sigma^2 / \hat\tau^2$, which is the multi-task analogue of
James-Stein shrinkage and avoids a cross-validation grid entirely.
:::

### Shrinkage, bias, and variance in the orthonormal case

To see the bias-variance trade-off in closed form, specialize to balanced
orthonormal designs: $n_t = n$ and $X_t^\top X_t = n I$ for every task. Then
$A_t = (n I + \lambda I)^{-1} X_t^\top$, and by symmetry the shared center is the
average of the per-task OLS solutions. Writing
$\hat\beta_t^{\text{ols}} = (X_t^\top X_t)^{-1} X_t^\top y_t$ and
$\bar\beta^{\text{ols}} = \tfrac{1}{T}\sum_t \hat\beta_t^{\text{ols}}$, the
multi-task estimate becomes a convex combination

$$
\hat\beta_t
= (1 - \alpha)\, \hat\beta_t^{\text{ols}} + \alpha\, \bar\beta^{\text{ols}},
\qquad
\alpha = \frac{\lambda}{\,n + \lambda\,} \in [0, 1].
$$ {#eq-transfer-multitask-learning-shrinkage}

The estimator shrinks each task's OLS solution toward the cross-task mean by a
fraction $\alpha$ that grows with $\lambda$ and shrinks with sample size $n$.
This is the exact analogue of ridge shrinkage and of James-Stein toward a
common mean. With $\beta_t = \beta_0 + \delta_t$ for true deviations $\delta_t$
and per-coordinate OLS variance $\sigma^2/n$, the per-task mean squared error of
@eq-transfer-multitask-learning-shrinkage decomposes (treating the mean as
approximately unbiased for $\beta_0$ when $T$ is large) as

$$
\mathbb{E}\,\lVert \hat\beta_t - \beta_t \rVert_2^2
\approx
\underbrace{\alpha^2 \lVert \delta_t \rVert_2^2}_{\text{bias}^2}
+ \underbrace{(1-\alpha)^2 \frac{p\,\sigma^2}{n}}_{\text{variance}} ,
$$

ignoring the $O(1/T)$ variance of the mean. Minimizing over $\alpha$ gives the
optimal shrinkage

$$
\alpha^\star
= \frac{p\,\sigma^2 / n}{\,p\,\sigma^2/n + \lVert \delta_t \rVert_2^2\,},
\qquad\text{equivalently}\qquad
\lambda^\star = \frac{p\,\sigma^2}{\lVert \delta_t \rVert_2^2}
\;\;\text{(per coordinate, } \tau^2 = \lVert\delta_t\rVert_2^2/p\text{)}.
$$

Three readings of this formula are worth stating. First, more noise or smaller
samples ($\sigma^2/n$ large) push $\alpha^\star$ toward 1: borrow more. Second,
more genuine heterogeneity ($\lVert\delta_t\rVert_2^2$ large) pushes
$\alpha^\star$ toward 0: borrow less. Third, $\alpha^\star$ is strictly interior
whenever both terms are positive and finite, which is the algebraic reason the
U-shaped curve in the demonstration has an interior minimum. The formula also
recovers the failure modes: when the tasks are unrelated,
$\lVert\delta_t\rVert_2^2 \to \infty$ forces $\alpha^\star \to 0$ and any
positive $\lambda$ is harmful, which is negative transfer in a single equation.

The convex-combination identity
@eq-transfer-multitask-learning-shrinkage is easy to confirm numerically.
The check below builds orthonormal designs so that $X_t^\top X_t = n I$ exactly,
solves the full block system, and compares the result to the shrinkage formula.

```{r mtl-shrinkage-check}
set.seed(7)
T_chk <- 4; p_chk <- 3; n_chk <- 20; lam <- 5

# Orthonormal designs: X_t^T X_t = n I via scaled orthonormal columns.
mk_orth <- function() {
  Q <- qr.Q(qr(matrix(rnorm(n_chk * p_chk), n_chk, p_chk)))
  Q * sqrt(n_chk)                      # columns now have squared norm n
}
Xs <- replicate(T_chk, mk_orth(), simplify = FALSE)
bt <- replicate(T_chk, rnorm(p_chk), simplify = FALSE)
ys <- Map(function(X, b) as.numeric(X %*% b) + rnorm(n_chk), Xs, bt)

# Block solve (same construction as fit_mtl).
N <- T_chk * n_chk
Z <- matrix(0, N, p_chk * (T_chk + 1)); yv <- numeric(N); r0 <- 0
for (t in 1:T_chk) {
  rows <- (r0 + 1):(r0 + n_chk)
  Z[rows, 1:p_chk] <- Xs[[t]]
  Z[rows, (t * p_chk + 1):((t + 1) * p_chk)] <- Xs[[t]]
  yv[rows] <- ys[[t]]; r0 <- r0 + n_chk
}
pen <- c(rep(0, p_chk), rep(lam, p_chk * T_chk))
theta <- solve(crossprod(Z) + diag(pen), crossprod(Z, yv))
b0 <- theta[1:p_chk]
beta_block <- lapply(1:T_chk, function(t)
  b0 + theta[(t * p_chk + 1):((t + 1) * p_chk)])

# Shrinkage formula: convex combo of per-task OLS and their mean.
ols   <- Map(function(X, y) solve(crossprod(X), crossprod(X, y)), Xs, ys)
obar  <- Reduce(`+`, ols) / T_chk
alpha <- lam / (n_chk + lam)
beta_form <- lapply(ols, function(b) (1 - alpha) * b + alpha * obar)

max_abs_diff <- max(abs(unlist(beta_block) - unlist(beta_form)))
cat("max |block - shrinkage formula| =", signif(max_abs_diff, 3), "\n")
```

The discrepancy is at the level of floating-point error, confirming that in the
orthonormal case the block estimator is exactly the shrinkage of each task's OLS
toward the cross-task mean with weight $\alpha = \lambda / (n + \lambda)$.

## Runnable demonstration

We simulate $T$ related regression tasks. Each task's true coefficients
equal a shared vector plus a small random deviation, so the tasks are related
but not identical. Each task gets only a modest sample, the situation where
borrowing strength should help. We compare three estimators: independent OLS
per task, a single pooled model, and the joint multi-task estimator across a
grid of $\lambda$. We evaluate by out-of-sample mean squared error against
the known truth.

```{r mtl-simulate}
set.seed(2026)

T_tasks <- 6      # number of tasks
p       <- 5      # number of predictors per task
n_train <- 25     # small training sample per task
n_test  <- 500    # large test sample per task

# Shared coefficient center plus small task-specific deviations.
beta0_true <- c(1.5, -2.0, 0.0, 1.0, -0.5)
sigma_dev  <- 0.4   # how far tasks drift from the shared center
sigma_eps  <- 1.0   # noise standard deviation

# True per-task coefficients: beta_t = beta0 + deviation_t
beta_true <- lapply(1:T_tasks, function(t) {
  beta0_true + rnorm(p, mean = 0, sd = sigma_dev)
})

make_task <- function(t, n) {
  X <- matrix(rnorm(n * p), nrow = n, ncol = p)
  y <- as.numeric(X %*% beta_true[[t]]) + rnorm(n, sd = sigma_eps)
  list(X = X, y = y)
}

train <- lapply(1:T_tasks, function(t) make_task(t, n_train))
test  <- lapply(1:T_tasks, function(t) make_task(t, n_test))
```

Independent OLS, fit one task at a time.

```{r mtl-independent}
fit_independent <- function(task) {
  # No intercept: data are centered around zero by construction.
  solve(crossprod(task$X), crossprod(task$X, task$y))
}
beta_ind <- lapply(train, fit_independent)
```

The pooled model stacks all tasks and fits a single coefficient vector,
the $\lambda \to \infty$ limit of the joint objective.

```{r mtl-pooled}
X_all <- do.call(rbind, lapply(train, `[[`, "X"))
y_all <- do.call(c,    lapply(train, `[[`, "y"))
beta_pool <- solve(crossprod(X_all), crossprod(X_all, y_all))
```

Now the joint multi-task estimator. We build the block matrix $Z$ and the
penalty $P$ described in the derivation, then solve the ridge-like system.

```{r mtl-joint}
fit_mtl <- function(train, lambda, p, T_tasks) {
  n_t <- sapply(train, function(z) length(z$y))
  N   <- sum(n_t)

  # Z has p shared columns followed by p columns per task.
  Z <- matrix(0, nrow = N, ncol = p * (T_tasks + 1))
  y <- numeric(N)

  row0 <- 0
  for (t in 1:T_tasks) {
    rows <- (row0 + 1):(row0 + n_t[t])
    Z[rows, 1:p] <- train[[t]]$X                       # shared block beta0
    cols <- (t * p + 1):((t + 1) * p)                  # deviation block v_t
    Z[rows, cols] <- train[[t]]$X
    y[rows] <- train[[t]]$y
    row0 <- row0 + n_t[t]
  }

  # Penalty: 0 on the shared block, lambda on every deviation block.
  pen <- c(rep(0, p), rep(lambda, p * T_tasks))
  P   <- diag(pen)

  theta <- solve(crossprod(Z) + P, crossprod(Z, y))

  beta0 <- theta[1:p]
  betas <- lapply(1:T_tasks, function(t) {
    beta0 + theta[(t * p + 1):((t + 1) * p)]
  })
  list(beta0 = beta0, betas = betas)
}
```

A common test-MSE helper, evaluated against held-out data per task.

```{r mtl-eval}
test_mse <- function(beta_list) {
  errs <- sapply(1:T_tasks, function(t) {
    pred <- as.numeric(test[[t]]$X %*% beta_list[[t]])
    mean((test[[t]]$y - pred)^2)
  })
  mean(errs)
}

mse_ind  <- test_mse(beta_ind)
mse_pool <- test_mse(rep(list(beta_pool), T_tasks))

lambda_grid <- 10^seq(-2, 3, length.out = 30)
mse_mtl <- sapply(lambda_grid, function(lam) {
  fit <- fit_mtl(train, lam, p, T_tasks)
  test_mse(fit$betas)
})

best_idx    <- which.min(mse_mtl)
best_lambda <- lambda_grid[best_idx]
best_mse    <- mse_mtl[best_idx]

results <- data.frame(
  method   = c("Independent OLS", "Pooled", "Multi-task (best lambda)"),
  test_mse = round(c(mse_ind, mse_pool, best_mse), 4)
)
print(results)
cat("Best lambda:", round(best_lambda, 3), "\n")
```

Reading the printed table: the multi-task estimator at its best $\lambda$ has
the lowest mean test MSE, beating both independent OLS (too noisy, because
each task fits only 25 points) and pooling (too biased, because the tasks are
not identical). The reported best $\lambda$ is an interior value, neither near
zero nor near infinity, which is the signature of genuine borrowing.

@fig-transfer-multitask-learning-mse-curve shows how multi-task
test error varies with $\lambda$, with the
independent and pooled baselines as horizontal references. The U shape is the
bias-variance trade-off across the sharing strength: too little sharing
behaves like the noisy independent fits, too much sharing behaves like the
biased pooled fit, and the minimum sits in between.

```{r fig-transfer-multitask-learning-mse-curve, fig.width=7, fig.height=4.5, fig.cap="Multi-task test MSE versus sharing strength, against independent and pooled baselines."}
library(ggplot2)

df <- data.frame(lambda = lambda_grid, mse = mse_mtl)

ggplot(df, aes(x = lambda, y = mse)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue") +
  geom_hline(yintercept = mse_ind, linetype = "dashed", color = "firebrick") +
  geom_hline(yintercept = mse_pool, linetype = "dotted", color = "darkgreen") +
  annotate("point", x = best_lambda, y = best_mse, color = "black", size = 3) +
  scale_x_log10() +
  annotate("text", x = min(lambda_grid), y = mse_ind,
           label = "Independent OLS", color = "firebrick",
           hjust = 0, vjust = -0.6, size = 3.3) +
  annotate("text", x = min(lambda_grid), y = mse_pool,
           label = "Pooled", color = "darkgreen",
           hjust = 0, vjust = -0.6, size = 3.3) +
  labs(x = expression(lambda~"(log scale)"),
       y = "Mean test MSE across tasks",
       title = "Borrowing strength across related tasks") +
  theme_minimal(base_size = 12)
```

With small per-task samples and genuinely related tasks, the multi-task
estimator at its best $\lambda$ should beat both baselines: it avoids the
high variance of independent OLS and the bias of pooling. If you increase
`n_train` substantially, the independent fits improve and the gap shrinks,
because each task no longer needs to borrow. If you increase `sigma_dev` so
the tasks drift far apart, pooling and small-$\lambda$ multi-task degrade,
which is negative transfer, discussed next.


::: {.callout-tip}
Try editing `n_train`, `sigma_dev`, and `sigma_eps` and rerunning.
Watching the U-shaped curve flatten, deepen, or invert builds far more
intuition than any single number in the table. The whole point of a runnable
demonstration is that the trade-off is yours to probe.
:::


## Negative transfer

Sharing helps only when tasks are actually related. When they are not, forcing
them to share hurts, and the shared model does worse than independent models.
This failure is called negative transfer, surveyed in @Pan_2010 and
@Zhang_2021.


::: {.callout-warning}
Negative transfer is not a rare edge case; it is the default
outcome when you assume relatedness without checking it. The discipline that
protects you is simple: always keep an independent-model baseline and refuse
to ship a shared model that loses to it on a clean target holdout.
:::


You can see it directly in the demonstration. The pooled model is the
$\lambda \to \infty$ limit, and when `sigma_dev` is large the pooled MSE rises
above the independent MSE: the tasks are too different to share a single
coefficient vector. The remedy is to let $\lambda$ be chosen by validation
rather than fixed, so the data decide how much sharing is appropriate, and to
let it go small when sharing does not pay.

Practical signals of negative transfer: a multi-task model whose per-task
metrics are worse than separately trained models; one dominant task whose
gradients swamp the others; or source and target distributions that look
unrelated under a two-sample test. The defenses are task weighting (next
section), grouping only tasks that are similar, and architectures that share
less (soft sharing, or task-specific layers). Closely related is ensemble
learning (@sec-ensemble-learning), where combining diverse models is preferred
to forcing one model to serve mismatched objectives.

## Task weighting

When tasks are trained jointly the total loss is a weighted sum,

$$
\mathcal{L} = \sum_{t=1}^{T} w_t \, \mathcal{L}_t ,
$$

and the weights $w_t$ matter.^[Setting every $w_t = 1$ looks neutral but is
not: it implicitly weights each task by its raw loss scale, so a task measured
in larger units silently dominates.] If one task has a much larger loss scale
or many more examples, it dominates the gradient and the others are under-fit.
The simplest fix is to normalize each task's loss to a comparable scale.
Beyond that, two learned schemes are common.

*Uncertainty weighting* (@Kendall_2018) treats each task's noise level as a
parameter $\sigma_t$ and derives weights from a Gaussian likelihood. For
regression tasks the objective becomes

$$
\mathcal{L}
= \sum_{t=1}^{T}
\left( \frac{1}{2 \sigma_t^2}\, \mathcal{L}_t + \log \sigma_t \right),
$$

so a task with high uncertainty gets a small weight $1 / (2\sigma_t^2)$, and
the $\log \sigma_t$ term prevents the weights from collapsing to zero. The
$\sigma_t$ are learned by gradient descent alongside the model.

To see where this objective comes from, model each regression task as Gaussian
with its own observation noise, $y_t \sim \mathcal{N}(f_t(x), \sigma_t^2)$. The
negative log likelihood of one task, dropping the constant $\frac12\log 2\pi$,
is

$$
-\log p(y_t \mid f_t(x))
= \frac{1}{2\sigma_t^2} \lVert y_t - f_t(x) \rVert_2^2 + \log \sigma_t ,
$$

and summing over tasks (which assumes conditional independence across tasks
given the shared representation) gives exactly the displayed objective with
$\mathcal{L}_t = \lVert y_t - f_t(x)\rVert_2^2$. The weighting is therefore not a
heuristic; it is maximum likelihood for a heteroscedastic multi-task Gaussian
model in which each task is allowed its own noise scale. Profiling out
$\sigma_t$ makes the mechanism explicit: setting
$\partial \mathcal{L} / \partial \sigma_t = 0$ gives

$$
-\frac{\mathcal{L}_t}{\sigma_t^3} + \frac{1}{\sigma_t} = 0
\quad\Longrightarrow\quad
\hat\sigma_t^2 = \mathcal{L}_t ,
$$

so at the optimum the learned noise variance equals the task's own residual
loss, and back-substituting yields a profiled objective
$\sum_t \big(\tfrac12 + \tfrac12\log \mathcal{L}_t\big)$, that is, a sum of log
losses. The practical effect is automatic, scale-free balancing: a task with a
large irreducible loss is downweighted in proportion to that loss, so no task
dominates merely because it is measured in larger units or is intrinsically
noisier. (In practice one optimizes $s_t = \log \sigma_t^2$ rather than
$\sigma_t$ for numerical stability and to keep the variance positive.)

*Gradient normalization* (@Chen_2018) instead adjusts the weights so that the
tasks train at similar rates, by balancing the magnitudes of their gradients
on the shared parameters. Both methods aim at the same goal: keep any single
task from dominating, which is one of the main causes of negative transfer.

## Practical guidance and pitfalls

When to reach for these methods:

- The target task has little labeled data but a related, data-rich source
  exists. Transfer learning is the default starting point.
- You must predict several related outcomes from the same inputs. A
  multi-task model can beat separate models and is cheaper to serve.
- Your training and serving input distributions differ. Domain adaptation,
  through importance weighting or representation alignment, is the repair.

Pitfalls to watch:

- *Assuming relatedness.* Verify it. Check whether sharing actually improves
  held-out per-task metrics versus independent baselines. If it does not, you
  have negative transfer.
- *Data leakage during transfer.* If you select the source model or tune
  $\lambda$ using the target test set, your reported gains are optimistic.
  Keep a clean target holdout.
- *Distribution shift in the source.* A pretrained backbone encodes its
  source distribution. If the target inputs differ sharply (different sensors,
  languages, time periods), early-layer features may not transfer.
- *Catastrophic forgetting during fine-tuning.* Too high a learning rate
  erases the pretrained structure. Use small learning rates and gradual
  unfreezing.
- *Imbalanced tasks.* Normalize loss scales and consider learned task
  weighting so that large or noisy tasks do not dominate.
- *Tuning the sharing strength on the wrong signal.* In the linear model,
  pick $\lambda$ by cross-validation on the target, not by training error,
  which always prefers small $\lambda$.

A reasonable default recipe: start with feature extraction or a moderately
penalized multi-task model, establish an independent-model baseline, tune the
sharing strength on validation data, and only move to full fine-tuning or
strong sharing if the data are abundant and the relatedness is confirmed.


::: {.callout-important title="Key idea"}
Every method in this chapter is a knob on a single dial, from
fully independent models to a single shared model. Transfer, multi-task,
domain adaptation, and task weighting are all ways of choosing, and
defending, where to set that dial. Let validation data, not optimism, make
the choice.
:::


## Further reading

- Caruana (1997), the foundational treatment of multi-task learning as
  inductive transfer.
- Evgeniou and Pontil (2004), regularized multi-task learning, the formulation
  used in this chapter's demonstration.
- Pan and Yang (2010), a broad survey of transfer learning.
- Yosinski, Clune, Bengio, and Lipson (2014), on how transferable features are
  across layers of deep networks.
- Ganin et al. (2016), domain-adversarial training for domain adaptation.
- Ruder (2017), an overview of multi-task learning in deep networks, including
  hard and soft parameter sharing.
- Kendall, Gal, and Cipolla (2018), uncertainty-based task weighting.
- Chen et al. (2018), GradNorm for gradient-based task balancing.
- Howard and Ruder (2018), discriminative fine-tuning and gradual unfreezing.
- Zhang and Yang (2021), a recent survey of multi-task learning methods.
