# Online and Streaming Learning {#sec-online-learning}

```{r}
#| include: false
source("_common.R")
```

Most of this book assumes a fixed training set: you collect $n$ observations, fit a model, and evaluate it. Online learning drops that assumption. Data arrive one observation (or one small batch) at a time, the model updates immediately, and the old data may never be stored again. This is the natural setting for systems that must keep learning in production: fraud scoring, recommendation, ad ranking, sensor monitoring, and any service where the data distribution moves over time.

This chapter develops the core ideas: the incremental update view of learning, the perceptron and online gradient descent, the stochastic approximation theory that explains why these updates converge, regret as the right notion of performance, Hoeffding (very fast) decision trees for unbounded streams, and concept drift. The runnable demos are implemented in base R so the mechanics are fully visible: a streaming mean and variance with Welford's algorithm, and an online logistic regression that updates one observation at a time.

> **Intuition:** Batch learning is like studying for an exam from a fixed textbook. Online learning is like a person who has to answer questions all day long, learning a little from each one as soon as they find out whether they were right, and never getting to reread the earlier questions. The skill is to improve continuously from a single forward pass through experience.

By the end you should be able to recognize when a problem calls for streaming methods rather than periodic retraining, write a simple incremental update by hand, reason about whether it will converge, and avoid the operational traps (leaky evaluation, runaway learning rates, silent drift) that make online systems fail in production.

## Where Online Learning Fits

In a batch workflow you retrain on a schedule (nightly, weekly) and ship a new model. That works well for many problems, but it breaks down in two situations:

- The data does not fit in memory, or the stream never ends, so there is no "full" training set to hold.
- The relationship between features and target changes faster than your retraining cadence, so a model trained last week is already stale.

Online learning addresses both. Each example is processed once, used to update the model, and discarded. Memory and per-step compute stay constant regardless of how many examples have been seen.^[This $O(1)$ per-step cost is the property that makes a stream of unbounded length feasible at all: whether you have seen a thousand examples or a billion, the work to absorb the next one is the same.] Because the update is incremental, the model tracks a moving target instead of freezing at a snapshot.

To place online learning next to its relatives, @tbl-online-learning-paradigm-comparison contrasts how each style accesses data, what it costs, and where it shines.

| Property | Batch learning | Mini-batch SGD | Online / streaming |
|---|---|---|---|
| Data access | Full set, many passes | Full set, shuffled batches | One pass, one example at a time |
| Memory | $O(n)$ | $O(\text{batch})$ | $O(1)$ in the number of examples |
| Update timing | After full fit | After each batch | After each example |
| Adapts to drift | Only on retrain | Only on retrain | Continuously |
| Typical use | Stable tabular problems | Deep nets, large but finite data | Production streams, huge or infinite data |

: Comparison of batch, mini-batch, and online learning paradigms by data access pattern, memory cost, update timing, and suitability for drifting distributions. {#tbl-online-learning-paradigm-comparison}

A common production pattern is a hybrid: train a strong batch model offline, then keep a lightweight online layer (a logistic regression or a small set of weights) updating on the live stream to absorb short-term drift, retraining the heavy model on a slower clock.

> **When to use this:** Reach for a streaming method when the data is too big to store, never stops arriving, or moves faster than you can retrain. If your dataset is finite, fits in memory, and comes from a stable distribution, batch learning is simpler to validate and usually more accurate per example. Section \@ref(practical-guidance-and-pitfalls) returns to this decision in detail.

## The Incremental Update View

Write the model as a parameter vector $\theta \in \mathbb{R}^d$. At step $t$ the learner holds $\theta_t$, sees an example $(x_t, y_t)$, suffers a loss $\ell_t(\theta_t) = \ell(\theta_t; x_t, y_t)$, and produces $\theta_{t+1}$. The generic online update is

$$
\theta_{t+1} = \theta_t - \eta_t \, g_t, \qquad g_t = \nabla_\theta \ell_t(\theta_t),
$$

where $\eta_t > 0$ is a step size (learning rate)^[The learning rate controls how far each example is allowed to move the parameters. It is the single most important knob in online learning, and most of the failure modes in Section \@ref(practical-guidance-and-pitfalls) trace back to choosing it badly.] and $g_t$ is the gradient of the current example's loss. Two features make this practical: the update touches only the current example, and the cost per step does not grow with $t$.

> **Key idea:** Almost every online method in this chapter is a special case of $\theta_{t+1} = \theta_t - \eta_t g_t$. Change the loss $\ell_t$ and you change the algorithm; the update template stays the same. Recognizing this one line under different disguises is most of what it takes to read the online-learning literature.

Even a running average is an instance of this template. If $\ell_t(\theta) = \tfrac12 (\theta - x_t)^2$ then $g_t = \theta_t - x_t$, and with $\eta_t = 1/t$ the update

$$
\theta_{t+1} = \theta_t - \tfrac{1}{t}(\theta_t - x_t)
$$

reproduces the exact sample mean of $x_1, \dots, x_t$. This is the bridge to stochastic approximation below.

## The Perceptron

The perceptron (Rosenblatt, 1958) is the earliest online classifier and the ancestor of modern neural networks (@sec-neural-networks); it is still the cleanest illustration of mistake-driven learning. Labels are $y_t \in \{-1, +1\}$ and the prediction is $\hat y_t = \operatorname{sign}(\theta_t^\top x_t)$. The update fires only on a mistake:

$$
\theta_{t+1} =
\begin{cases}
\theta_t + y_t x_t & \text{if } y_t (\theta_t^\top x_t) \le 0 \\
\theta_t & \text{otherwise.}
\end{cases}
$$

In words: if the prediction is correct, leave the weights alone; if it is wrong, nudge them toward getting that example right by adding $y_t x_t$. This is online (sub)gradient descent with step size 1 on the loss $\ell_t(\theta) = \max(0, -y_t \theta^\top x_t)$, the perceptron loss.^[A (sub)gradient generalizes the gradient to functions with kinks. The perceptron loss has a corner at $y_t\theta^\top x_t = 0$; one valid subgradient there is $0$, which is why a correct prediction triggers no update.] The classic guarantee bounds the number of mistakes rather than a probability of error, which is what makes it an online rather than a statistical result.

> **Note:** The mistake bound below is one of the oldest results in machine learning, and it is striking precisely because it makes no probabilistic assumption. It does not say "errors are unlikely"; it says the perceptron can make at most a fixed number of mistakes, ever, as long as the data is separable with a margin.

Perceptron mistake bound. Suppose there is a unit vector $u$ ($\lVert u \rVert = 1$) and a margin $\gamma > 0$ with $y_t (u^\top x_t) \ge \gamma$ for all $t$, and suppose $\lVert x_t \rVert \le R$ for all $t$. Then the total number of mistakes the perceptron makes is at most

$$
M \le \frac{R^2}{\gamma^2}.
$$

Sketch. Let the updates happen on the mistake rounds. On each mistake, the projection onto $u$ grows by at least $\gamma$, because $u^\top \theta_{t+1} = u^\top \theta_t + y_t (u^\top x_t) \ge u^\top \theta_t + \gamma$, so after $M$ mistakes $u^\top \theta \ge M\gamma$. At the same time the squared norm grows slowly: on a mistake $\lVert \theta_{t+1}\rVert^2 = \lVert\theta_t\rVert^2 + 2 y_t(\theta_t^\top x_t) + \lVert x_t\rVert^2 \le \lVert\theta_t\rVert^2 + R^2$, since the mistake condition makes the middle term nonpositive, so $\lVert\theta\rVert^2 \le M R^2$. Combining with $u^\top\theta \le \lVert\theta\rVert$ gives $M\gamma \le \sqrt{M} R$, hence $M \le R^2/\gamma^2$. The bound does not depend on the dimension $d$ or the number of examples, only on the geometry (margin and radius).

## Online Gradient Descent and Stochastic Approximation

Online gradient descent (OGD) is the perceptron's general form: apply $\theta_{t+1} = \theta_t - \eta_t g_t$ for any (sub)differentiable loss. Two questions follow. Does it converge to a good parameter, and how should $\eta_t$ behave?

Stochastic approximation answers the first. The Robbins-Monro method (Robbins and Monro, 1951) finds a root^[A root of $h$ is a value of $\theta$ where $h(\theta)=0$. When $h$ is the gradient of a smooth convex risk, its root is the risk minimizer, so root-finding and minimization coincide.] of $h(\theta) = \mathbb{E}[g(\theta; Z)] = 0$ using only noisy samples $g(\theta; z_t)$, with the update $\theta_{t+1} = \theta_t - \eta_t \, g(\theta_t; z_t)$. When $\theta$ minimizes a convex risk $R(\theta) = \mathbb{E}[\ell(\theta; Z)]$, the root of the gradient is exactly the minimizer, so OGD on a stream of i.i.d. examples is Robbins-Monro applied to $h = \nabla R$. Convergence holds under the step-size conditions

$$
\sum_{t=1}^\infty \eta_t = \infty, \qquad \sum_{t=1}^\infty \eta_t^2 < \infty.
$$

These two conditions encode a tug-of-war. The first condition keeps the steps large enough in aggregate to reach the solution from any start; the second forces the steps to shrink fast enough that gradient noise averages out. The canonical choice $\eta_t = c/t$ satisfies both. In practice $\eta_t = c/\sqrt{t}$ (which violates the second condition) is common because it adapts better when the target moves, trading exact convergence for responsiveness.

> **Intuition:** A step size that decays too slowly leaves the model permanently jittery, chasing noise. One that decays too fast freezes the model before it has reached a good answer. The $\sum\eta_t=\infty,\ \sum\eta_t^2<\infty$ pair is the precise statement of "fast enough to settle, slow enough to arrive."

### Regret: the right scorecard

When data is not i.i.d. (it may be adversarial, or drifting), "convergence to the risk minimizer" is not well defined, so online learning is judged by regret. After $T$ rounds, regret compares the learner's cumulative loss to the best single fixed parameter chosen in hindsight:

$$
\text{Regret}_T = \sum_{t=1}^T \ell_t(\theta_t) - \min_{\theta \in \Theta} \sum_{t=1}^T \ell_t(\theta).
$$

An algorithm is "no-regret" if $\text{Regret}_T / T \to 0$, meaning its average loss approaches that of the best fixed parameter, whatever the sequence. For convex losses with bounded gradients ($\lVert g_t \rVert \le G$) and a bounded domain ($\lVert \theta \rVert \le D$), OGD with $\eta_t = D/(G\sqrt{t})$ achieves (Zinkevich, 2003)

$$
\text{Regret}_T \le \tfrac{3}{2} G D \sqrt{T} = O(\sqrt{T}).
$$

Sublinear $O(\sqrt{T})$ regret means the per-round penalty for not knowing the future vanishes as $\sqrt{T}/T = 1/\sqrt{T}$. This is the central performance promise of online convex optimization and it requires no statistical assumption on how the data is generated.

> **Key idea:** Regret reframes the goal. Instead of asking "did we recover the true model?" (which assumes there is one), it asks "did we do nearly as well as the best fixed model we could have picked with hindsight?" That question stays meaningful even when an adversary chooses the data, which is why regret, not risk, is the native scorecard of online learning.

## Demo 1: Streaming Mean and Variance (Welford)

We start with the simplest possible streaming statistic, a running mean and variance, because it shows the incremental update template at its barest and introduces a numerical pitfall worth knowing.

A naive streaming variance that accumulates $\sum x$ and $\sum x^2$ and then computes $\frac{1}{n-1}(\sum x^2 - n\bar x^2)$ can lose precision badly: when the two large sums are close, subtracting them cancels most of the significant digits, and the result can even come out negative. Welford's algorithm (Welford, 1962) avoids this by updating the mean and the sum of squared deviations $M_2 = \sum_i (x_i - \bar x)^2$ in one pass with good numerical behavior. With $\bar x_t$ the mean after $t$ points,

$$
\bar x_t = \bar x_{t-1} + \frac{x_t - \bar x_{t-1}}{t}, \qquad
M_{2,t} = M_{2,t-1} + (x_t - \bar x_{t-1})(x_t - \bar x_t),
$$

and the sample variance is $s_t^2 = M_{2,t} / (t-1)$. Each step is $O(1)$ in time and memory. Notice that the mean update $\bar x_t = \bar x_{t-1} + (x_t - \bar x_{t-1})/t$ is exactly the running-average special case from Section \@ref(the-incremental-update-view), with the gradient $g_t = \theta_t - x_t$ and step size $\eta_t = 1/t$.

> **Warning:** The textbook one-pass formula $\operatorname{Var} = \frac{1}{n-1}\big(\sum x^2 - n\bar x^2\big)$ is numerically dangerous. For data with a large mean and small spread (say sensor readings around 1{,}000{,}000 that vary by a few units), catastrophic cancellation can return a variance that is wildly wrong or negative. Prefer Welford whenever you compute variance from a stream.

In the code below, `welford_update` carries the running state (count, mean, and $M_2$) and returns the new state after seeing one value. The `stopifnot` lines at the end are a self-check: they confirm the one-pass result equals what `mean()` and `var()` compute from the whole stored vector.

@fig-online-learning-welford-demo shows the running mean and standard deviation produced by the code below as the stream grows.

```{r fig-online-learning-welford-demo, message=FALSE, warning=FALSE, fig.cap="Welford's running mean and standard deviation converging to the true population values (dashed lines) as the stream grows. Estimates are noisy early and stabilize as more data arrives."}
set.seed(1)

welford_update <- function(state, x) {
  state$n <- state$n + 1L
  delta <- x - state$mean
  state$mean <- state$mean + delta / state$n
  delta2 <- x - state$mean
  state$M2 <- state$M2 + delta * delta2
  state
}

welford_variance <- function(state) {
  if (state$n < 2L) return(NA_real_)
  state$M2 / (state$n - 1L)
}

true_mu <- 5
true_sd <- 2
stream <- rnorm(2000, mean = true_mu, sd = true_sd)

state <- list(n = 0L, mean = 0, M2 = 0)
running_mean <- numeric(length(stream))
running_sd   <- numeric(length(stream))

for (i in seq_along(stream)) {
  state <- welford_update(state, stream[i])
  running_mean[i] <- state$mean
  running_sd[i]   <- sqrt(welford_variance(state))
}

# confirm the one-pass result matches base R on the full sample
stopifnot(abs(state$mean - mean(stream)) < 1e-10)
stopifnot(abs(welford_variance(state) - var(stream)) < 1e-8)

op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
plot(running_mean, type = "l", col = "steelblue",
     xlab = "samples seen", ylab = "running mean", main = "Welford mean")
abline(h = true_mu, lty = 2)
plot(running_sd, type = "l", col = "darkorange",
     xlab = "samples seen", ylab = "running sd", main = "Welford sd")
abline(h = true_sd, lty = 2)
par(op)
```

The checks pass and the two curves settle onto the dashed true values, which is the point: Welford reproduces the exact sample mean and variance while touching each value once and storing only three numbers, never the full vector at once.

## Demo 2: Online Logistic Regression

Now a supervised case. The model is logistic regression with parameter $\theta$ (including an intercept), predicting $p_t = \sigma(\theta^\top x_t)$ where $\sigma(z) = 1/(1+e^{-z})$. The per-example log loss and its gradient are

$$
\ell_t(\theta) = -\big[y_t \log p_t + (1-y_t)\log(1-p_t)\big], \qquad
\nabla_\theta \ell_t(\theta) = (p_t - y_t)\, x_t.
$$

The OGD update is therefore the compact $\theta \leftarrow \theta - \eta_t (p_t - y_t) x_t$. The gradient is just the prediction error $(p_t - y_t)$ scaled onto the input $x_t$: predict too high and the weights move down along $x_t$, predict too low and they move up. We process one observation at a time, never re-reading past data, and we track the running average log loss to see learning happen. For an honest read on generalization we also compute, before each update, the loss on a held-out test set; this is the prequential idea of test-then-train.^[Prequential (predictive-sequential) evaluation scores each example with the current model *before* that example is used to update it. Because the model never sees an example's label until after it has been graded on it, no future information leaks into the score, so prequential loss is an unbiased running estimate of how the deployed model is doing.]

> **Warning:** Never evaluate a temporal stream by shuffling it and taking a random train/test split. Doing so lets the model train on examples that came *after* the ones it is tested on, leaking the future and producing an optimistic, meaningless score. Prequential test-then-train respects the arrow of time.

@fig-online-learning-online-logit shows the resulting learning curves alongside the batch reference.

```{r fig-online-learning-online-logit, message=FALSE, warning=FALSE, fig.cap="Online logistic regression trained one observation at a time. The running-average training log loss (blue) falls as the weights improve, and the held-out test log loss (orange) tracks it, confirming the streaming fit generalizes. The dashed line is the log loss of the batch glm fit on all the data."}
set.seed(42)

sigmoid <- function(z) 1 / (1 + exp(-z))

# generate a linearly-separable-ish stream
n  <- 4000
d  <- 3
X  <- matrix(rnorm(n * d), n, d)
theta_true <- c(-0.5, 1.5, -2.0, 1.0)   # intercept + 3 weights
Xb <- cbind(1, X)
p_true <- sigmoid(Xb %*% theta_true)
y <- rbinom(n, 1, p_true)

# hold out a test set for prequential evaluation
test_idx  <- 1:500
train_idx <- 501:n
Xb_test <- Xb[test_idx, , drop = FALSE]
y_test  <- y[test_idx]

logloss <- function(p, y) {
  eps <- 1e-12
  p <- pmin(pmax(p, eps), 1 - eps)
  -mean(y * log(p) + (1 - y) * log(1 - p))
}

theta <- rep(0, d + 1)          # start at zero
c0 <- 0.5                       # base learning rate
running_train <- numeric(length(train_idx))
test_curve    <- numeric(length(train_idx))
avg_loss <- 0

for (k in seq_along(train_idx)) {
  i  <- train_idx[k]
  xi <- Xb[i, ]
  pi <- as.numeric(sigmoid(sum(theta * xi)))

  # online (instantaneous) loss before updating, for the running average
  inst <- logloss(pi, y[i])
  avg_loss <- avg_loss + (inst - avg_loss) / k
  running_train[k] <- avg_loss

  # held-out test loss with current weights (test-then-train)
  p_test <- sigmoid(Xb_test %*% theta)
  test_curve[k] <- logloss(as.numeric(p_test), y_test)

  # OGD step: decaying learning rate eta_t = c0 / sqrt(t)
  eta <- c0 / sqrt(k)
  theta <- theta - eta * (pi - y[i]) * xi
}

# batch reference fit on the same training rows
batch_fit <- glm(y ~ X, family = binomial,
                 subset = train_idx)
batch_test_loss <- logloss(
  predict(batch_fit,
          newdata = data.frame(X = I(X[test_idx, , drop = FALSE])),
          type = "response"),
  y_test
)

plot(running_train, type = "l", col = "steelblue", ylim = c(0, max(running_train)),
     xlab = "training examples seen", ylab = "log loss",
     main = "Online logistic regression")
lines(test_curve, col = "darkorange")
abline(h = batch_test_loss, lty = 2)
legend("topright", bty = "n",
       legend = c("running-avg train", "held-out test", "batch glm test"),
       col = c("steelblue", "darkorange", "black"),
       lty = c(1, 1, 2))

round(rbind(online = theta, batch = coef(batch_fit), truth = theta_true), 3)
```

The online weights end up close to both the batch `glm` estimate and the true coefficients, and the held-out loss settles near the batch model's loss, all from a single streaming pass that stored only the current example and a parameter vector.

## Trees for Streams: Hoeffding (Very Fast) Decision Trees

So far the parameters have been continuous weights that gradient descent can nudge. Trees are different: a split is a discrete choice, and the obvious way to pick the best split is to scan the whole dataset. That makes trees awkward online, because a split decision normally needs all the data at once to compare candidate features, exactly the assumption behind batch tree induction (@sec-regression-trees). The Hoeffding tree, also called the Very Fast Decision Tree or VFDT (Domingos and Hulten, 2000), solves this with a statistical stopping rule for how much data a node needs before it can commit to a split.

The idea: at a leaf, accumulate sufficient statistics (per-feature, per-class counts) as examples stream through. Let $G(\cdot)$ be a split-quality measure (information gain or Gini), and let $\bar G_a$ and $\bar G_b$ be the observed quality of the best and second-best features after $n$ examples have reached the leaf. The Hoeffding bound says that, with probability $1-\delta$, the true mean of a bounded quantity in range $[0,R]$ is within

$$
\epsilon = \sqrt{\frac{R^2 \ln(1/\delta)}{2n}}
$$

of its observed average. So if $\bar G_a - \bar G_b > \epsilon$, then with confidence $1-\delta$ feature $a$ really is the best split, and the node can split now without waiting for more data. As $n$ grows, $\epsilon$ shrinks, so even near-ties eventually resolve.

> **Intuition:** The Hoeffding bound answers "have I seen enough examples to be sure the leading feature really is leading, and is not ahead by luck?" When the gap between the top two features is larger than the statistical wiggle room $\epsilon$, the answer is yes and the node commits. The remarkable part is that this lets a single streaming pass build a tree whose splits match, with high probability, what a batch tree would have chosen on the same data, while reading each example only once.

@tbl-online-learning-hoeffding-vs-batch contrasts the streaming tree with its batch ancestors.

| Aspect | Batch CART / C4.5 | Hoeffding tree (VFDT) |
|---|---|---|
| Passes over data | Many | One |
| Memory | Grows with data | Bounded per node (sufficient statistics) |
| Split rule | Best split on full data | Split when $\bar G_a - \bar G_b > \epsilon$ |
| Suited to | Finite datasets | Unbounded streams |
| Drift handling | Refit from scratch | Extensions (CVFDT, Hoeffding Adaptive Tree) |

: Batch decision-tree induction versus the Hoeffding (Very Fast) Decision Tree, compared by number of passes, memory footprint, split criterion, data regime, and drift handling. {#tbl-online-learning-hoeffding-vs-batch}

Production stream-mining libraries (for example MOA in Java, and `river` in Python) implement VFDT and its drift-aware variants. R does not ship a maintained streaming-tree package, so the snippet below is illustrative pseudocode marked `eval=FALSE`: it shows the split-decision logic, not a full tree implementation.

```{r hoeffding-bound, eval=FALSE}
# Hoeffding split test at a leaf (illustrative).
# counts: accumulated sufficient statistics at this leaf.
hoeffding_epsilon <- function(R, n, delta) {
  sqrt((R^2 * log(1 / delta)) / (2 * n))
}

should_split <- function(gain_best, gain_second, n,
                         R = 1, delta = 1e-7, tau = 0.05) {
  eps <- hoeffding_epsilon(R, n, delta)
  diff <- gain_best - gain_second
  # split if the best feature clearly wins, OR if the contest is
  # effectively tied (diff < eps < tau) so waiting is pointless
  (diff > eps) || (eps < tau)
}
```

## Concept Drift

Everything so far assumed that the world a model learns is the world it is later tested on. Streams break that assumption: over time the joint distribution $P_t(x, y)$ drifts. Diagnosing the *kind* of drift matters because the fix differs, so it helps to keep three cases straight.

- Covariate shift (virtual drift): $P(x)$ changes but $P(y \mid x)$ does not. The decision boundary still holds, but inputs move into regions the model saw little of.
- Real concept drift: $P(y \mid x)$ changes, so the optimal boundary itself moves. This is what degrades a fixed model's accuracy.
- Drift can also differ in *speed*: abrupt (a sudden regime change), gradual (one concept fades as another grows), or recurring (seasonal patterns that return).

> **Note:** The distinction between covariate shift and real drift is practical, not academic. Covariate shift often just means the model needs more coverage of the newly common inputs; real drift means the model's current answer is now wrong and it has to forget. Misdiagnosing one as the other wastes effort.

No single mechanism handles every kind of drift, so production systems combine several. The main families are these.

- Forgetting through the learning rate. A non-decaying or slowly-decaying $\eta_t$ keeps the model plastic so recent examples dominate. The cost is more variance and never fully converging.
- Sliding windows. Train on the last $w$ examples and drop older ones. Window size sets the bias-variance and stability-plasticity trade-off: short windows adapt fast but are noisy.
- Explicit drift detectors. ADWIN (Adaptive Windowing, Bifet and Gavalda, 2007) maintains a variable-length window and shrinks it when a statistically significant change in the mean error is detected; DDM (Drift Detection Method, Gama et al., 2004) watches the online error rate and signals warning and drift thresholds. On a signal, the model is reset or retrained on recent data.
- Ensembles. Keep a pool of models trained on different recent segments and weight them by recent accuracy, retiring stale members; this is ensemble learning (@sec-ensemble-learning) adapted to a moving distribution.

A concrete decision rule: if accuracy is drifting, first check whether $P(x)$ moved (cheap to monitor) before assuming $P(y\mid x)$ did. Covariate shift often just needs more coverage or reweighting, while real drift needs the model to forget.

## Practical Guidance and Pitfalls

With the machinery in hand, the final question is when to actually use it. Online learning is the right call when any of the following holds.

- The data is too large for memory or never ends.
- Predictions and updates must happen in real time with bounded per-step cost.
- The target distribution moves and you cannot retrain fast enough to keep up.

Prefer batch learning in the opposite situation.

- The dataset is finite and fits in memory, and the distribution is stable. Batch optimization is more sample-efficient per example and easier to validate; online learning buys you nothing here and adds operational complexity.

Even when online learning is warranted, a handful of mistakes account for most failures in practice. The list below collects the ones worth memorizing.

- Learning rate. Too large diverges or oscillates; too small crawls and never catches drift. Start with a decaying schedule such as $\eta_t = c/\sqrt{t}$ and tune $c$ on a replayed stream. If you need permanent adaptation, cap the decay so $\eta_t$ never falls below a floor.
- Feature scaling on the fly. Standardize with running statistics (Welford), not a fixed scaler computed from an early batch, or the scale will be wrong once the stream drifts.
- Evaluation. Do not shuffle a temporal stream and do a random split; that leaks the future. Use prequential evaluation (test-then-train) so each example is scored before it is learned from.
- Single example variance. Per-example gradients are noisy. Small mini-batches (10 to 100) smooth the updates while keeping the streaming property.
- Catastrophic forgetting and label delay. If labels arrive late (a click is known only later), align the update to when the label actually lands, not when the features arrived.
- Stateful bugs. Online systems carry state across time, so a bad update can corrupt the model permanently. Keep a periodic checkpoint and a guardrail metric that can trigger a rollback.

> **Tip:** The single highest-leverage safeguard for a production online learner is a checkpoint plus a rollback trigger. Because the model's state evolves continuously, one pathological batch of inputs can poison it for good; being able to revert to last night's snapshot turns a permanent outage into a brief blip.

To summarize the chapter in one breath: online learning replaces "fit once, ship, retrain later" with "update on every example, forever." The same gradient step ($\theta_{t+1} = \theta_t - \eta_t g_t$) drives the perceptron, online logistic regression, and general OGD; stochastic approximation explains when it converges and regret measures it when it cannot; Hoeffding trees bring the streaming discipline to discrete split decisions; and concept drift, together with leak-free prequential evaluation and a sane learning rate, is where the real engineering lives.

## Further Reading

- Rosenblatt, F. (1958). The perceptron: a probabilistic model for information storage and organization in the brain. *Psychological Review*. Original perceptron.
- Robbins, H., and Monro, S. (1951). A stochastic approximation method. *Annals of Mathematical Statistics*. Foundation of stochastic approximation.
- Welford, B. P. (1962). Note on a method for calculating corrected sums of squares and products. *Technometrics*. The streaming variance update.
- Zinkevich, M. (2003). Online convex programming and generalized infinitesimal gradient ascent. *ICML*. Online gradient descent and $O(\sqrt{T})$ regret.
- Cesa-Bianchi, N., and Lugosi, G. (2006). *Prediction, Learning, and Games*. Cambridge University Press. Comprehensive treatment of regret and online learning.
- Shalev-Shwartz, S. (2012). Online learning and online convex optimization. *Foundations and Trends in Machine Learning*. Modern survey with proofs.
- Domingos, P., and Hulten, G. (2000). Mining high-speed data streams. *KDD*. The Hoeffding tree / VFDT.
- Gama, J., Medas, P., Castillo, G., and Rodrigues, P. (2004). Learning with drift detection. *SBIA*. The DDM drift detector.
- Bifet, A., and Gavalda, R. (2007). Learning from time-changing data with adaptive windowing. *SDM*. The ADWIN algorithm.
- Bifet, A., Gavalda, R., Holmes, G., and Pfahringer, B. (2018). *Machine Learning for Data Streams*. MIT Press. Book-length treatment, with the MOA framework.
- Bottou, L., Curtis, F. E., and Nocedal, J. (2018). Optimization methods for large-scale machine learning. *SIAM Review*. Stochastic gradient methods at scale.
