# Adversarial Learning

A well-trained classifier can reach high accuracy on a held-out test set and still
be brittle in a way that ordinary evaluation never reveals. Take an image that the
model labels "panda" with high confidence, add a perturbation so small that no human
notices any change, and the same model now says "gibbon" with even higher confidence.
The perturbation is not noise in the colloquial sense; it is a carefully chosen
direction in input space that the model's own gradient points to. This phenomenon,
documented by @szegedy2013intriguing and @goodfellow2014explaining, is the entry
point to *adversarial learning*.

The name covers two ideas that share a mathematical core but solve different problems.
The first is *adversarial robustness*: how do we attack a model with tiny input
perturbations, and how do we defend against such attacks? This matters most for
security-critical systems, spam filters, malware detectors, fraud scoring, and any
model whose inputs are supplied by a party who benefits from fooling it. The second
is *adversarial training as a learning objective*, where two models compete in a game.
The generator-versus-discriminator setup behind generative adversarial networks is the
famous case, and the same minimax structure reappears in domain-adversarial training
for domain adaptation.

What ties the two together is a single optimization pattern: one player tries to
minimize a loss while another tries to maximize it. In the robustness story the
maximizing player is an attacker perturbing inputs; in the GAN story it is a
discriminator trying to expose fakes. Once you see the minimax skeleton, attacks,
defenses, and generative training all read as instances of the same game.

> **Intuition:** A model that is accurate on average can still have a decision
> boundary that passes very close to many test points. "Close" is measured in input
> space, where a human's notion of similarity and the model's notion can diverge
> sharply. Adversarial examples live in that gap.

## Adversarial examples and the geometry of brittleness

Fix a classifier $f_\theta$ with parameters $\theta$ that maps an input
$x \in \mathbb{R}^d$ to class scores, and let $L(\theta, x, y)$ be the loss for the
true label $y$ (for instance cross-entropy). An *adversarial example* is a perturbed
input $x' = x + \delta$ that the model misclassifies, subject to a constraint that
keeps the perturbation small:

$$
\max_{\delta}\; L(\theta, x + \delta, y)
\quad\text{subject to}\quad \lVert \delta \rVert_p \le \epsilon .
$$

The norm $\lVert \cdot \rVert_p$ defines what "small" means. The $\ell_\infty$ norm,
$\lVert \delta \rVert_\infty = \max_i |\delta_i|$, caps how much any single feature may
move and is the most common choice for images. The $\ell_2$ norm caps the total
Euclidean size of the perturbation. The budget $\epsilon$ is the radius of the allowed
region around $x$.

Why does such a $\delta$ exist at all, and why is it often tiny? The short answer from
@goodfellow2014explaining is that high-dimensional models behave locally like linear
functions. Suppose $f$ scores a class as $w^\top x + b$. Then perturbing $x$ by
$\delta$ changes the score by $w^\top \delta$. To push the score as far as possible
under an $\ell_\infty$ budget $\epsilon$, set each coordinate of $\delta$ to
$\epsilon \cdot \operatorname{sign}(w_i)$, giving a change of
$\epsilon \sum_i |w_i| = \epsilon \lVert w \rVert_1$. In $d$ dimensions this grows with
$d$, so even a microscopic per-feature change $\epsilon$ accumulates into a large swing
in the score. Brittleness is partly a consequence of dimensionality, not just of
model pathology.

> **Key idea:** Many tiny coordinated nudges, each below the threshold of notice,
> sum to a large movement of the model's score. The attacker spends a small budget per
> feature and is repaid in proportion to the dimension.

### The fast gradient sign method

The fast gradient sign method (FGSM) turns the linear argument into a one-step attack.
Linearize the loss around $x$ using its gradient
$g = \nabla_x L(\theta, x, y)$. To first order, $L(\theta, x + \delta, y) \approx
L(\theta, x, y) + g^\top \delta$. Maximizing $g^\top \delta$ under
$\lVert \delta \rVert_\infty \le \epsilon$ has the closed-form solution

$$
\delta^{\mathrm{FGSM}} = \epsilon \cdot \operatorname{sign}\!\big(\nabla_x L(\theta, x, y)\big),
\qquad
x' = x + \delta^{\mathrm{FGSM}} .
$$

The sign function appears because, coordinate by coordinate, the largest contribution
to $g^\top \delta$ within $[-\epsilon, \epsilon]$ comes from moving each feature to the
edge of its allowed range in the direction that increases the loss. FGSM is a single
gradient evaluation, which is what makes it fast and a standard baseline.

### Projected gradient descent attacks

FGSM takes one big step along a linearized loss. When the loss is genuinely curved,
one step overshoots or underexploits. Projected gradient descent (PGD), the attack
studied by @madry2017towards, iterates small steps and projects back into the allowed
region after each:

$$
x^{(t+1)} = \Pi_{\mathcal{B}_\epsilon(x)}\!\Big( x^{(t)} + \alpha \cdot
\operatorname{sign}\!\big(\nabla_x L(\theta, x^{(t)}, y)\big) \Big),
$$

where $\alpha$ is the step size, $\mathcal{B}_\epsilon(x) = \{ x' : \lVert x' - x
\rVert_\infty \le \epsilon \}$ is the $\ell_\infty$ ball of radius $\epsilon$ around the
clean input, and $\Pi$ clips each coordinate back into that ball. Usually $x^{(0)}$ is
a random point inside the ball, which helps escape flat regions. PGD with enough steps
is treated as a strong first-order attack: if a defense survives PGD, it has cleared a
meaningful bar, though not a proof of security.

> **Warning:** A defense that lowers the gradient signal without changing the decision
> boundary can defeat FGSM and PGD by *gradient masking*, giving a false sense of
> safety. Such defenses fall to attacks that estimate gradients differently or transfer
> examples from a substitute model (@athalye2018obfuscated). Always report the attack
> used and its budget.

## Adversarial training as a defense

If the attacker solves an inner maximization, the natural defense is to train against
the worst case. @madry2017towards frame robust learning as a saddle-point problem:

$$
\min_{\theta}\; \mathbb{E}_{(x,y)}\!\left[\; \max_{\lVert \delta \rVert_p \le \epsilon}
L(\theta, x + \delta, y) \;\right].
$$

The inner $\max$ finds the strongest perturbation for the current model; the outer
$\min$ updates parameters to be robust to it. In practice you approximate the inner
problem with FGSM or PGD, generate adversarial inputs on each minibatch, and train on
them (often mixed with clean inputs). This is *adversarial training*, and it remains
one of the few defenses that holds up under careful evaluation.

Robustness has a price. Pushing the decision boundary away from the data along every
direction the attacker might exploit can lower clean accuracy, the *robustness-accuracy
tradeoff* analyzed by @tsipras2019robustness. Robust models also tend to need more data
and more capacity. The right operating point depends on the threat: a fraud model facing
adaptive adversaries may accept a clean-accuracy hit that a benign recommender would not.

> **When to use this:** Reach for adversarial training when inputs come from a party who
> gains by fooling the model, or when a wrong prediction on a perturbed input is costly
> (security, safety, finance). For ordinary i.i.d. deployment with no adversary, the
> clean-accuracy cost is usually not worth paying.

## A runnable demo: attacking and defending a linear classifier

The geometry is easiest to see with a model whose gradient is exact rather than
approximate. We use logistic regression, where the loss gradient with respect to the
input has a clean closed form, so FGSM is not an approximation but the genuine
worst-case linear perturbation. We simulate two Gaussian classes, fit the classifier,
build FGSM perturbations, and measure how accuracy collapses. Then we retrain on
adversarially perturbed data and measure recovery.

For logistic regression the predicted probability is
$\sigma(z) = 1 / (1 + e^{-z})$ with $z = \beta_0 + x^\top \beta$. The cross-entropy
loss for a label $y \in \{0,1\}$ has input gradient

$$
\nabla_x L = \big(\sigma(z) - y\big)\, \beta ,
$$

so the FGSM perturbation is
$\delta = \epsilon \cdot \operatorname{sign}\!\big((\sigma(z) - y)\,\beta\big)$. Because
$\operatorname{sign}((\sigma(z)-y)\beta) = \operatorname{sign}(\sigma(z)-y)\cdot
\operatorname{sign}(\beta)$, the attack simply moves every feature by $\pm\epsilon$
toward the wrong class. The direction is the same for every point of a given class,
which is exactly the linear picture from above.

```{r adversarial-learning-setup, eval=TRUE}
set.seed(1301)

# Two Gaussian classes in 2-D with overlap so the boundary is non-trivial.
n <- 400
mu0 <- c(-1.0, -1.0)
mu1 <- c( 1.0,  1.0)
Sigma <- matrix(c(1, 0.3, 0.3, 1), 2, 2)

rmvn <- function(n, mu, Sigma) {
  L <- chol(Sigma)
  matrix(rnorm(n * 2), n, 2) %*% L + matrix(mu, n, 2, byrow = TRUE)
}

X0 <- rmvn(n / 2, mu0, Sigma)
X1 <- rmvn(n / 2, mu1, Sigma)
X  <- rbind(X0, X1)
y  <- c(rep(0, n / 2), rep(1, n / 2))

# Train / test split.
idx   <- sample(seq_len(n), size = floor(0.6 * n))
Xtr   <- X[idx, ];  ytr <- y[idx]
Xte   <- X[-idx, ]; yte <- y[-idx]

sigmoid <- function(z) 1 / (1 + exp(-z))

# Fit logistic regression (clean model).
df_tr <- data.frame(x1 = Xtr[, 1], x2 = Xtr[, 2], y = ytr)
clean_fit <- glm(y ~ x1 + x2, data = df_tr, family = binomial())
beta  <- coef(clean_fit)[c("x1", "x2")]
beta0 <- coef(clean_fit)[["(Intercept)"]]
```

The attack function below builds the FGSM perturbation for a batch of points. Each row
of the input matrix is nudged by $\epsilon$ in the sign direction that raises its loss.

```{r adversarial-learning-attack, eval=TRUE}
# FGSM perturbation for logistic regression, exact input gradient.
fgsm <- function(Xmat, yvec, beta, beta0, epsilon) {
  z    <- as.numeric(beta0 + Xmat %*% beta)
  p    <- sigmoid(z)
  # grad_x L = (p - y) * beta, broadcast over rows.
  grad <- outer(p - yvec, beta)              # n x 2
  Xmat + epsilon * sign(grad)
}

predict_label <- function(Xmat, beta, beta0) {
  as.integer(sigmoid(as.numeric(beta0 + Xmat %*% beta)) > 0.5)
}

epsilon <- 1.0

# Clean and adversarial test accuracy for the clean model.
acc <- function(pred, truth) mean(pred == truth)

clean_acc_clean_model <- acc(predict_label(Xte, beta, beta0), yte)
Xte_adv <- fgsm(Xte, yte, beta, beta0, epsilon)
adv_acc_clean_model <- acc(predict_label(Xte_adv, beta, beta0), yte)
```

Now we apply the defense. We augment the training set with FGSM perturbations of the
training points (generated from the clean model's parameters), refit, and re-evaluate
against both clean and freshly attacked test data. The freshly attacked test set uses
the *defended* model's own gradient, so the attacker adapts to the new model rather than
reusing stale perturbations.

```{r adversarial-learning-defense, eval=TRUE}
# Adversarial training: augment with FGSM-perturbed training points.
Xtr_adv <- fgsm(Xtr, ytr, beta, beta0, epsilon)
df_aug  <- data.frame(
  x1 = c(Xtr[, 1], Xtr_adv[, 1]),
  x2 = c(Xtr[, 2], Xtr_adv[, 2]),
  y  = c(ytr, ytr)
)
robust_fit <- glm(y ~ x1 + x2, data = df_aug, family = binomial())
rbeta  <- coef(robust_fit)[c("x1", "x2")]
rbeta0 <- coef(robust_fit)[["(Intercept)"]]

clean_acc_robust_model <- acc(predict_label(Xte, rbeta, rbeta0), yte)
# Attacker adapts to the defended model's gradient.
Xte_adv_r <- fgsm(Xte, yte, rbeta, rbeta0, epsilon)
adv_acc_robust_model <- acc(predict_label(Xte_adv_r, rbeta, rbeta0), yte)

results <- data.frame(
  Model = c("Standard (clean training)", "Adversarial training"),
  `Clean accuracy` = round(c(clean_acc_clean_model, clean_acc_robust_model), 3),
  `Adversarial accuracy` = round(c(adv_acc_clean_model, adv_acc_robust_model), 3),
  check.names = FALSE
)
```

Table \@ref(tab:adversarial-learning-acc) reports the four numbers. The standard model
keeps high clean accuracy but its adversarial accuracy falls sharply: a per-feature
nudge of $\epsilon = `r epsilon`$ flips a large share of test predictions. Adversarial
training trades a little clean accuracy for a large gain in adversarial accuracy, the
robustness-accuracy tradeoff in miniature.

```{r adversarial-learning-acc, eval=TRUE}
knitr::kable(
  results,
  caption = "Clean versus adversarial test accuracy for a standard logistic-regression classifier and the same classifier after adversarial training (FGSM augmentation, epsilon = 1.0). Adversarial accuracy is measured against perturbations adapted to each model."
)
```

Figure \@ref(fig:adversarial-learning-boundary) shows the geometry. The left panel plots
the clean test points with the standard model's decision boundary. The right panel shows
the same points after the FGSM attack: every class-1 point slides down-left and every
class-0 point slides up-right, both toward the boundary, and many cross it. The arrows in
the right panel trace the perturbation applied to a sample of points.

```{r adversarial-learning-boundary, eval=TRUE, fig.cap="Left: clean test points and the standard model's linear decision boundary. Right: the same points after the FGSM attack (epsilon = 1.0); arrows show the perturbation on a sample of points, which pushes each class uniformly toward and across the boundary.", fig.width=9, fig.height=4.5}
op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))

# Boundary: beta0 + b1*x1 + b2*x2 = 0  =>  x2 = -(beta0 + b1*x1)/b2
abline_for <- function(b0, b) c(intercept = unname(-b0 / b[2]), slope = unname(-b[1] / b[2]))

cols <- ifelse(yte == 1, "#1b9e77", "#d95f02")

# Panel 1: clean
plot(Xte[, 1], Xte[, 2], col = cols, pch = 19, cex = 0.7,
     xlab = "x1", ylab = "x2", main = "Clean inputs")
ln <- abline_for(beta0, beta)
abline(a = ln["intercept"], b = ln["slope"], lwd = 2)

# Panel 2: adversarial
plot(Xte_adv[, 1], Xte_adv[, 2], col = cols, pch = 19, cex = 0.7,
     xlab = "x1", ylab = "x2", main = "FGSM-perturbed inputs")
abline(a = ln["intercept"], b = ln["slope"], lwd = 2)
samp <- sample(seq_len(nrow(Xte)), 25)
arrows(Xte[samp, 1], Xte[samp, 2],
       Xte_adv[samp, 1], Xte_adv[samp, 2],
       length = 0.05, col = "grey40")

par(op)
```

The lesson generalizes past logistic regression. Any model with a usable input gradient
admits FGSM and PGD attacks, and adversarial training is the corresponding defense. The
linear case just lets us see the closed form and the uniform per-class shift that more
complex models hide.

## The adversarial game behind GANs

The minimax pattern that defines robust training is the same one that drives generative
adversarial networks (see the [Generative Models] chapter for the full treatment). A
generator $G$ maps noise $z \sim p_z$ to samples $G(z)$, and a discriminator $D$ outputs
the probability that an input is real rather than generated. They play

$$
\min_{G}\; \max_{D}\;
\mathbb{E}_{x \sim p_{\mathrm{data}}}\!\big[\log D(x)\big]
+ \mathbb{E}_{z \sim p_z}\!\big[\log\big(1 - D(G(z))\big)\big].
$$

The discriminator maximizes its ability to separate real from fake; the generator
minimizes that ability by producing samples the discriminator cannot flag. Compare this
to the robustness saddle point: there the inner $\max$ is over input perturbations and
the outer $\min$ is over model parameters; here the inner $\max$ is over the
discriminator and the outer $\min$ is over the generator. Same skeleton, different
players. @goodfellow2014gan show that at the global optimum the generator reproduces the
data distribution and the discriminator is reduced to guessing, $D \equiv \tfrac{1}{2}$.

> **Note:** The shared minimax form is more than an analogy. Both settings inherit the
> same difficulties: the inner problem is rarely solved exactly, the outer updates chase
> a moving target, and naive alternation can oscillate instead of converging. Techniques
> that stabilize one game often inform the other.

### Domain-adversarial training

A third use of the same idea adapts a model trained on one data distribution (the
source) to perform on a different but related one (the target), the problem of domain
adaptation. @ganin2016domain attach a *domain classifier* that tries to tell source
features from target features, and train the feature extractor to fool it through a
gradient-reversal layer. If a feature representation makes source and target
indistinguishable, a label predictor trained on the source should transfer to the target.
The objective again has the competing-players form: the feature extractor minimizes label
loss while maximizing domain-classification loss, and the domain classifier does the
opposite.

```{r adversarial-learning-dann, eval=FALSE}
# Sketch of domain-adversarial training (requires a deep-learning backend, so eval=FALSE).
# feature_extractor: x -> h
# label_predictor:   h -> y_hat   (trained on labeled source only)
# domain_classifier: h -> d_hat   (real source vs target)
#
# Total loss = label_loss(label_predictor(h), y)            # minimized by extractor + predictor
#            - lambda * domain_loss(domain_classifier(h), d) # extractor MAXIMIZES this term
#
# The minus sign is implemented by a gradient-reversal layer: forward pass is identity,
# backward pass multiplies the gradient by -lambda before it reaches the feature extractor.
```

## Practical guidance and pitfalls

Several recurring mistakes undermine work on adversarial robustness, and a few habits
keep evaluations honest.

State the threat model before anything else. An attack is only meaningful relative to a
norm, a budget $\epsilon$, and an assumption about what the attacker knows. White-box
attacks see the model's gradients; black-box attacks do not and must transfer examples
from a substitute or estimate gradients by queries. Reporting "robust accuracy" without
the budget and access model is uninterpretable.

Do not trust a defense that only beats FGSM. A single gradient step is the weakest
standard attack. @athalye2018obfuscated showed that many published defenses worked by
*obfuscating gradients*, making the loss surface locally uninformative without moving the
boundary, and collapsed under PGD or transfer attacks. Evaluate against a strong
iterative attack with multiple random restarts, and be suspicious when robust accuracy is
high but the loss gradient is near zero.

Budget the inner optimization realistically. PGD adversarial training multiplies training
cost by the number of inner steps. Single-step adversarial training is cheaper but can
suffer *catastrophic overfitting*, where robustness to multi-step attacks suddenly
collapses mid-training. Monitor robust accuracy against a multi-step attack during
training, not just the single-step proxy you train on.

Expect to pay for robustness. The robustness-accuracy tradeoff (@tsipras2019robustness)
is real: clean accuracy usually drops, and robust models often need more data and
capacity. Decide up front how much clean accuracy you are willing to trade, and tie that
decision to the cost of an adversarial failure in your application.

Match the perturbation set to the real threat. The $\ell_\infty$ ball is a mathematical
convenience, not a model of every attacker. Real adversaries rotate images, paste
stickers, paraphrase text, or change the timing of network packets, none of which is a
small $\ell_p$ perturbation. Robustness inside one ball does not imply robustness to
transformations outside it.

For GAN-style training, watch for non-convergence and *mode collapse*, where the generator
maps many noise vectors to a few outputs that happen to fool the current discriminator.
These are the generative counterparts of the optimization instabilities that also afflict
adversarial robustness, and they trace back to the same fact: nobody is solving the inner
problem exactly.

## Further reading

- Szegedy, Zaremba, Sutskever, Bruna, Erhan, Goodfellow, and Fergus (2013), the paper
  that first reported adversarial examples for neural networks.
- Goodfellow, Shlens, and Szegedy (2015), introducing the fast gradient sign method and
  the linear explanation of adversarial examples.
- Madry, Makelov, Schmidt, Tsipras, and Vladu (2018), the projected gradient descent
  attack and the saddle-point view of robust training.
- Tsipras, Santurkar, Engstrom, Turner, and Madry (2019), on the robustness-accuracy
  tradeoff.
- Athalye, Carlini, and Wagner (2018), on obfuscated gradients and how to evaluate
  defenses honestly.
- Goodfellow, Pouget-Abadie, Mirza, Xu, Warde-Farley, Ozair, Courville, and Bengio
  (2014), the original generative adversarial networks paper.
- Ganin, Ustinova, Ajakan, Germain, Larochelle, Laviolette, Marchand, and Lempitsky
  (2016), on domain-adversarial training of neural networks.
