# Gaussian Process Regression

Suggested reading:

-   @rasmussen2005

-   @wikle2019 (section 4.2)

Under Gaussian Process (GP), we assume that $Y$ is a random (Gaussian) process that is controlled by a mean function and covariance function.

Response model:

$$
z_i = Y_i + \epsilon_i, i = 1, \dots, n
$$

Assumptions:

-   $Y$ = a latent process specified by a GP

-   $\epsilon_i$ independent and identically distributed mean-zero observation errors that are independent of $Y$ and have variance $\sigma_\epsilon^2$

-   the errors are assumed to have a normal (Gaussian) distribution $\epsilon_i \sim iid N (0, \sigma^2_\epsilon)$ these errors could be correlated

<br>

Consider a **linear predictor** of the form

$$
\hat{Y}(x_0) = k_0 + \sum_{i=1}^n w_i z_i
$$

where $w_i$ are weights and $k_0$ is some constant

To pick these weights, we choose "optimal" weight based on mean squared error notion, which requires us to consider the random process that controls $Y$

## Latent Process Model

Assume that we have the latent process follows the model

$$
Y(x) = f(x)
$$

where $f(x) \sim GP (m(x), c(x,x'))$ is a random process (specifically a Gaussian Process) and

-   $m(x)$ = mean function $E[f(x)]$

-   $c(x,x')$ = covariance function $c(x,x)'= E[(f(x) - m(x)) (f(x') - m(x'))^T]$

<br>

The difference between **Gaussian distribution** and a **Gaussian process** is

-   Gaussian Distribution: a distribution over vectors

    -   Fully specified by a **mean vector** and **covariance matrix**: (e.g., $\mathbf{Y} \sim N(\mathbf{\mu}, \mathbf{\Sigma})$ which is a multivariate normal distribution)

-   Gaussian Process: a distribution over functions

    -   Fully specified by a mean function ($m$) and covariance function ($c$) (e.g., $Y \sim GP(m,c)$ , where we specify a mean function and covariance function to define a GP)

## Gaussian Processes

-   A GP is an infinite-dimensional object

-   Allow us to make a prediction anywhere in our $x$ domain assuming that we know the mean and covariance functions (data generating process), we need the covariance between ANY two $x$ locations

-   We actually need to consider finite-dimensional distributions because

    -   We only have a finite set of data and also we are interested in a finite set of predictions

    -   Any finite collection of GP random variables has a joint Gaussian distributions (i.e., a multivariate normal distribution).

## Gaussian Distributions

$$
\left(
\begin{array}
{c}
\mathbf{v} \\
\mathbf{v}_*
\end{array}
\right)
\sim 
Gau 
\left(
\left(
\begin{array}
{c}
\mathbf{\mu} \\
\mathbf{\mu}_*
\end{array}
\right),
\left(
\begin{array}
{cc}
\mathbf{\Sigma} & \mathbf{\Sigma}_* \\
\mathbf{\Sigma}^T & \mathbf{\Sigma}_{**}
\end{array}
\right)
\right)
$$

where $E(\mathbf{v}) = \mathbf{\mu},E(\mathbf{v}_*) = \mathbf{\mu}_*, \mathbf{\Sigma}= cov(\mathbf{v}, \mathbf{v}), \mathbf{\Sigma}= cov(\mathbf{v}, \mathbf{v}_*),\mathbf{\Sigma}_{**} = cov(\mathbf{v}_*, \mathbf{v}_*)$

Then, partition

$$
\mathbf{v}_* | \mathbf{v} \sim Gau (\mathbf{\mu}_*+ \mathbf{\Sigma}^T_* \mathbf{\Sigma}^{-1} (\mathbf{v} - \mathbf{\mu}), \mathbf{\Sigma}_{**} - \mathbf{\Sigma}^T_* \mathbf{\Sigma}^{-1} \mathbf{\Sigma}_*)
$$

<br>

## Optimal Linear Prediction

We are interested in predicting $Y(x_0)$ (i.e., at some arbitrary $x$) given the data $\mathbf{z} \equiv (z_1, \dots, z_n)^T$

Assume that the mean function $\mu(x) \forall x$ (e.g., a constant)

We seek linear predictors that minimize the mean squared prediction error of the form

$$
\begin{aligned}
\hat{Y} (x_0) &=  k(x_0) + \sum_{i=1}^n w_i (x_0) z_i \\
&= k(x_0) + \mathbf{w}^T_0 \mathbf{z}
\end{aligned}
$$

The optimal linear predictor is also the conditional expectation $E(Y(x_0) |\mathbf{z})$.

Hence, we seek the conditional distribution $Y(x_0) |\mathbf{z}$

Rewriting the data model in vector form:

$$
\mathbf{z} = \mathbf{Y} + \epsilon, \epsilon \sim Gau(\mathbf{0}, \mathbf{C}_\epsilon)
$$

where $\mathbf{Y}$ and $\epsilon$ are ordered in the same way as the data vector $\mathbf{z}$

$$
cov(\mathbf{Y}, \mathbf{Y}) \equiv \mathbf{C}_y
$$

$$
cov(\epsilon,\epsilon) \equiv \mathbf{C}_\epsilon
$$

$$
cov(\mathbf{z}, \mathbf{z}) = \mathbf{C}_y + \mathbf{C}_\epsilon
$$

$$
\mathbf{c}^T_0 \equiv cov( Y(x_0), \mathbf{z})
$$

$$
c_{0,0} \equiv var(Y(x_0))
$$

$\mu_0$ is mean at $x_0$ and $\mathbf{\mu}$ is the mean corresponding to the observations.

Then, the joint Gaussian distribution:

$$
\left(
\begin{array}
{c}
Y (x_0) \\
\mathbf{v}_*
\end{array}
\right)
\sim 
Gau 
\left(
\left(
\begin{array}
{c}
\mathbf{\mu} \\
\mathbf{\mu}
\end{array}
\right),
\left(
\begin{array}
{cc}
c_{0,0} & \mathbf{c}^T_0 \\
\mathbf{c}_0 & \mathbf{C}_{z}
\end{array}
\right)
\right)
$$

Using the joint/conditional [Gaussian Distributions] formula

$$
Y(x_0)|\mathbf{z} \sim Gau (\mu_0+ \mathbf{c}^T_0 \mathbf{C}_z^{-1}(\mathbf{z}- \mathbf{\mu}), c_{0,0}- \mathbf{c}^T_0 \mathbf{C}_z^{-1} \mathbf{c}_0)
$$

The predictor is given by the **mean** of this distribution

$$
\hat{Y}(x_0) = \mu_0 + \mathbf{c}_0^T \mathbf{C}^{-1}_z(\mathbf{z} - \mathbf{\mu})
$$

and the prediction **variance** is

$$
\sigma^2_{Y_0} = c_{0,0} - \mathbf{c}^T_0 \mathbf{C}_z^{-1} \mathbf{c}_0
$$

This predictor is the **optimal predictor** because it minimizes the **mean squared prediction error**:

$$
MSPE = E[(Y(x_0) - \hat{Y}(x_0))^2]
$$

Hence, the **weights** are given by

$$
\mathbf{w}_0 = \mathbf{c}'_0 \mathbf{C}_z^{-1}
$$

and

$$
k(x_0) = \mu_0 - \mathbf{c}'_0 \mathbf{C}^{-1}_z \mathbf{\mu}
$$

which is an effective kernel (just like [Spline Regression] and [Local Regression]

These formulas can be extended to multiple locations (each location has its own vector of weights). However, if there are a lot of sample locations, there could be computational problem regarding calculating the matrix inverse $\mathbf{C}_z^{-1}$

In cases where you have no measurement errors (i.e., $z_i = y_i$) (equivalently, replacing $\mathbf{C}_z$ with $\mathbf{C}_y$)

## Covariance Functions

To predict at new $x$ locations, we must specify:

1.  a mean function (doable)
2.  a covariance function between ANY two locations (hard)

In the real world, we hardly ever know the variances and covariances that make up $\mathbf{C}_y, \mathbf{C}_\epsilon, \mathbf{c}_0, c_{0,0}$

<br>

A seemingly simple solution is to specify covariance functions in terms of parameters $\mathbf{\theta}$ that can give a covariance between any pairs of coordinates and then estimate the parameters, which is the same thing that we did in [Linear Mixed Models](https://bookdown.org/mike/data_analysis/linear-mixed-models.html) (e.g., AR models, exponential models)

-   **Covariance functions must be positive semi-definite** to ensure that prediction variances are non-negative

-   Covariance functions should be realistic (i.e., account for realistic dependencies), which can be difficult sometime. Typically, the assumption that "nearby" observations (similar in the case of [Kernel Smoothing]) are more dependent is sufficient.

<br>

In practice, we usually have to employ a **stationarity assumption**

Example:

If the random process is assumed to be second-order (weakly) stationary then it has constant expectation and a covariance function that can be expressed in terms of the distance between the $x$ locations ($h = x- x'$), we specify

$$
c(x,x') = c(h, \mathbf{\theta})
$$

In two or more dimensions, if this distance lag does not depend on direction (i.e., $h = ||\mathbf{h}||$), then we call the function **isotropic**

The benefits of stationarity assumption:

-   allows for more **parsimonious parameterization** of the covariance function

-   provides **pseudo-replication** of dependencies at given lags, which facilitate estimation

To construct covariance functions with this assumption, traditional machine learning would use **separable covariance functions** (e.g., for $x_1$ and $x_2$ - 2 dimension prediction)

$$
c(h_1;h_2) \equiv c(h_1) \times c_2(h_2)
$$

which is valid if both covariance functions are valid

This separability property helps with computation in prediction locations (e.g., matrix inverses we mention above), but it can also be unrealistic in cases like spatio-temporal data

Then to have valid spatial and temporal covariance functions, we can employ the following covariance functions:

-   The Matern

-   Power exponential: $c(\tau, \theta) = \sigma^2 \exp \{ - \frac{|\tau|}{\theta} \}$ where $\sigma^2$ is the variance parameter and $\theta$ is the dependence parameter

-   Gaussian classes

<br>

To estimate the parameters in the covariance functions, and in the mean function we can either

1.  Under the frequentist perspective, we can use REML, MLE (like [Linear Mixed Models](https://bookdown.org/mike/data_analysis/linear-mixed-models.html)) and BLUP for random process
2.  Under the Bayesian perspective, we use GP for $f(x)$ as prior distribution, and update this with the observations, $\mathbf{z}$

With a lot of data locations, REML, MLE, and Bayesian can be problematic because of the high-dimensional covariance matrix $\mathbf{C}_z, \mathbf{C}_y$ (which research are still being conducted to improve)

Notes:

-   If we model the mean function in terms of covariates ($\mu(x) = \mathbf{w}(x)'\mathbf{\beta}$m where $\mathbf{w}$ is a vector of covariates. Then, it's [Linear Mixed Models](https://bookdown.org/mike/data_analysis/linear-mixed-models.html) with difference that the covariates must be available for ANY $x$

-   In cases where you don't measurement uncertainty in the observation, ($\epsilon$) then the prediction at an observation location will be exactly the same as the observation. And the other way around will filter the uncertainty with the observation

-   we can also have non-Gaussian observations conditioned on the GP (like [GLMM](https://bookdown.org/mike/data_analysis/nonlinear-and-generalized-linear-mixed-models.html))

## Application

Other packages: `mlegp`

@macdonald2015

Simulation:

$$
y(x) = \log(x + .1) + sin(5 \pi x)
$$

A Latin Hypercube design is a space-filling design, used in the design of computer experiments.

The Maximin version does space-filling design by maximizing the minimum distance between design points.

```{r}
library(GPfit)
library(lhs)
    
set.seed(1)
n = 7
x = maximinLHS(n ,1) # draw a Maximin Latin Hypercube Sample
y = matrix(0, n, 1)
for (i in 1:n){
    y[i] = log(x[i] + .1 ) + sin(5*pi * x[i])
}

GPmodel = GP_fit(x,y)
print(GPmodel)
```

Predict at a grid of new x values

```{r}
xvec <- seq(from = 0, to = 1, length.out = 100)
yvec <- matrix(0,100,1)
for (i in 1:100){
    yvec[i] = log(xvec[i] + .1) + sin(5*pi*xvec[i])
}
GPpred <- predict(GPmodel,xvec)

# data points are circles
# prediction is in the bllue 
# 95% CIs (yhat + 2*s) are in red
plot(GPmodel, range= c(0,1), resolution = 100, colors = c("black","blue","red"), line_type = c(1,2),pch = 1)

# add the true function as a dotted black line
lines(xvec, yvec, col = "black", lty = 3)
```
