# Classification and Regression Trees

-   Segmenting the predictor space into a number of simple regions.

-   Develop a collection of splitting rules that make these segments.

-   The splitting rules can be thought of as "tree" (therefore, "decision trees")

-   The root of the tree (i.e., the first splitting variable) is typically chosen based on the the smallest sum of squared residuals (SSRs), then the process begins again to choose another branch.

-   Decision trees can be used for both classification and regression, ergo the name CART

    -   Under regression, the prediction is the average of the partitioned predictor space (i.e., segments)

    -   Under classification, we classify based on the most prevalent class in each segment

-   Compared to linear models

    -   If the underlying relationship is linear, the linear regression/ classification setting works better than CART

    -   IF the underlying relationship is nonlinear and complex between the features and the response, the CART model can be better.

-   Issues:

    -   Categorical predictors can prove to be difficult to implement CART when we have $q >2$ possible unordered values of the predictors as there are $2^{q-1} - 1$ possible partitions of the q values into 2 groups. Usually lead to over-fitting.

    -   Missing predictor values:

        -   For categorical variables, we can make a separate category for "missing". Or we can construct surrogate variables that mimic the split of the training data using only the non-missing variables

    -   Non-binary splits: not recommended because it splits the data too quickly and doesn't leave enough data for the next level of the tree

    -   Linear Combination Splits: we can make linear combination of the variables to choose the splits (akin to PC-regression), which can help prediction, but reduce interpretation

    -   Instability of trees: CART has high variance (i.e., a small change in the data can lead to different splits). To handle this problem, we can use "bagging" (i.e., bootstrapping procedure). This can happen even after you prune.

    -   Lack of Smoothness: CART can lead to non-smooth prediction surface, which can be problematic for regression applications. The MARS procedure can be used instead

    -   Additive structure: Since CART uses binary structure, it can't accommodate additive structure, but MARS can

    -   Trees do not have good predictive accuracy as other regression and classification approaches.

Advantages:

-   Easily explainable (to non-statisticians)

-   Decision trees may be closer to human decision making than traditional regression and classification

-   Trees are easily interpreted (with graphical depiction)

-   Trees can handle qualitative predictors without dummy variables

\
<br>

## Regression Trees

Say we have $p$ inputs and a response for each of $n$ observations, $(x_i, y_i), i = 1, \dots, n$ with $x_i = (x_{i1}, \dots , x_{ip})^T$

Suppose that we have a partition of the predictor space into $J$ regions $R_1, \dots, R_j$ . Then, our prediction in each region is some constant, $c_j$. Thus, we could write the response function as

$$
f(x) = \sum_{j=1}^J c_j I (x \in R_j)
$$

where $I(x \in R_j)$ is an indicator function (1 when $x$ is in the j-th region, 0 otherwise).

Given the partitions, the estimate $\hat{c}_j$ is the average of of $y_i$ in $R_j$

$$\hat{c}_j = \hat{y}_{R_j} \equiv ave(y_i |x_i \in R_j)$$ To find the partitions, we seek to divide the predictor space into high-dimensional rectangles (boxes) (i.e., we want to find the boxes $R_1, \dots, R_j$ that minimize the residual sum of squares)

$$
\sum_{j=1}^J \sum_{i \in R_j} (y_i - \hat{y}_{R_j})^2
$$

where $\hat{y}_{R_j}$ is the mean response for the training observations within the j-th box (e..g, $\hat{c}_j$ )

Since we can't consider every possible partition (computationally constraint), we seek binary partitions and specifically, a top down, greedy approach (i.e., at each step the best split is made at that step, once we make a split, we don't go back and change it) known as **recursive binary splitting**.

### Recursive Binary Splitting

Step 1: we select the predictors $X_j$ and the cutpoints such that splitting the predictor space into the regions $\{ X|X_j <s\}$ and $\{ X | X_j \ge s\}$ leads to the maximum reduction in RSS (i.e., for any j and s, we define the regions (half-planes):

$$
R_1 (j,s) = \{ X|X_j <s\} ; R_2(j,s) = \{ X|X_j \ge s\}
$$

and seek the value of j and s the minimizes

$$
\sum_{i: x_i \in R_1 (j,s)} (y_i - \hat{y}_{R_1})^2 + \sum_{i: x_i \in R_2 (j,s)} (y_i - \hat{y}_{R_2})^2
$$

where $\hat{y}_{R_1} = \hat{c}_1$ is the mean response in $R_1(j,s)$

Notes:

-   When $p$ is small, it's computationally efficient to find values of j and s.

-   After the partition is selected, we repeat the process, looking for the best predictor (j) and best cut point (s) to split one of the two new regions to minimize the RSS, which will give us three regions. We continue to split one of these three regions to minimize the RSS

-   The stopping criterion is usually when a region contains fewer than some specified number of observations (because we have to the averages of these regressions for our mean response)

-   The end regressions of the tree $(R_1, \dots, R_J)$ are known as **terminal nodes** or **leaves** of the tree

-   The points along the tree where the predictor space is spit are **internal nodes**

-   Regression trees can usually lead to over-fitting on the test set, to reduce this problem, we can prune (i.e., remove leaves) to obtain a **subtree** using cross-validation.

<br>

### Cost complexity pruning

-   also known as **weakest link pruning**

Intuitively,

1.  Calculate the total Sum of Squared Residual for each tree

2.  Calculate a Tree Score: $\text{Tree Score} = SSR + \alpha T$ where

    1.  SSR = Sum of Squared Residuals for the tree or sub-tree

    2.  T = number of leaves (or Terminal nodes) in the tree or sub-tree

    3.  $\alpha$ = tuning parameter (based on cross-validation)

3.  Pick tree with lowest tree score

The Tree Complexity Penalty ($\alpha T$) compensates for the difference in the number of leaves

<br>

Mathematically,

We first grow a very large tree $T_0$. Instead of considering every possible subtree, we consider a sequence of trees indeed a non-negative tuning parameters, $\alpha$ (i.e., for each value of $\alpha$ there is a subtree $T$ such that $T \in T_0$ that minimizes

$$
\sum_{j=1}^{|T|} \sum_{i:x_i \in R_j} (y_i - \hat{y}_{R_j})^2 + \alpha |T|
$$

where $|T|$ is the number of terminal nodes

The parameter $\alpha$ controls a tradoff between the subtree's complexity and how well it fits the training data ($\alpha = 0$ implies $T = T_0$). As $\alpha$ gets larger, the penalty for a larger tree is higher, so it penalizes similarly to the [Lasso regression]. Moreover, as $\alpha$ increases, the branches get pruned in a nested a predictable fashion. Hence, it's possible to get the whole sequence of subtrees as a function of $\alpha$ (based on cross-validation). After getting $\alpha$ using cross-validation, we can use that value with the full data set to obtain the fitted subtree

### Algorithm for regression tree

1.  [Recursive Binary Splitting] for growing a tree with your choice of stopping rule (e.g., observation-based)
2.  Use [Cost complexity pruning] to obtain a sequence of best subtrees as a function of $\alpha$
3.  Use [K-Fold Cross-Validation] to choose $\alpha$

```{r, warning=FALSE, message=FALSE}
library(ISLR2)
library(tree)

set.seed(1)
attach(Boston)

train <- sample(1:nrow(Boston), nrow(Boston)/2)
tree.boston <- tree(medv ~ ., Boston, subset = train)

# to fit a bigger tree
# tree.boston <-
#     tree(medv ~ .,
#          Boston,
#          subset = train,
#          control = tree.control(nobs = length(train), mindev = 0)) 

summary(tree.boston) # only 4 vars have been used (rm, lstat, crim, age)


```

`deviance` = sum of squared errors for the tree

```{r}
plot(tree.boston)
text(tree.boston, pretty = 0)

```

```{r}
cv.boston <- cv.tree(tree.boston) # tree complexity selected by cross-validation
plot(cv.boston$size, cv.boston$dev, type = "b")

```

If we want to prune complex tree

```{r}
prune.boston <- prune.tree(tree.boston, best = 5)
plot(prune.boston)
text(prune.boston, pretty = 0)

```

```{r}
yhat <- predict(tree.boston, newdata = Boston[-train, ])
boston.test <- Boston[-train, "medv"]
plot(yhat, boston.test)
abline(0, 1)
mean((yhat - boston.test) ^ 2) # Mean squared error 
```

The square root of the MSE is 5.9, meaning on average we predict within \$5.9k of the true median home value.

<br>

## Classification Trees

When we have a qualitative response we can use classification tree.

The main difference from [Regression Trees] is that we predict the response based on the most commonly occurring class in the region associated with that terminal node.

We may also be interested in the class proportion for each node when interpreting a classification tree

We grow a classification tree in the same manner as the [Regression Trees], with [Recursive Binary Splitting]. However, instead of using RSS we use other measure related to the classification summary.

First we define the proportion of class $k$ observation in the j-th node as

$$
\hat{p}_{jk} = \frac{1}{n_j} \sum_{x_i \in R_j} I(y_i = k)
$$we classify an observation in node j to class k with the maximum value of $\hat{p}_{jk}$

The three measures for growing tree:

1.  Classification error rate: $1 - \max_{{k}} (\hat{p}_{jk})$

    -   measures the fraction of the training observations in that region that do not belong to the most common class

2.  Gini index: $\sum_{k=1}^K \hat{p}_{jk}(1 - \hat{p}_{jk})$

    -   measures the total variance across the K classes

3.  Cross-Entropy or Deviance: $-\sum_{k=1}^K \hat{p}_{jk} \log (\hat{p}_{jk})$

    -   related to an information measures and the deviance

The Gini index and Cross-Entory both take small values when all $\hat{p}_{jk}$ are close to 1 or 0. Hence, we say they measure node purity, which are more sensitive to node probabilities than the classification error rate. Hence, we typically use the last two to grow the tree.

Rule of thumb: use Gini and Cross-Entropy to build and prune a tree and use classification error rate for pruning if prediction accuracy of the pruned tree is the goal

```{r}
library(tree)
library(ISLR2)
attach(Carseats)
High <- factor(ifelse(Sales <= 8, "No", "Yes")) # transform Sales (conitnuous) to binary variable

Carseats <- data.frame(Carseats, High)

tree.carseats <- tree(High ~ . - Sales, Carseats) # using all variables but Sales

summary(tree.carseats) # training error rate of 9%
# tree.carseats
```

Deviance formula:

$$
-2 \sum_m \sum_k n_{mk} \log \hat{p}_{mk}
$$

where

-   $n_{mk}$ = number of observations in the m-th terminal node in the k-th class

Smaller deviance is preferable (good fit to the training data).

Residual mean deviance formula

$$
\frac{-2 \sum_m \sum_k n_{mk} \log \hat{p}_{mk}}{n - |T_0|}
$$

where $|T_0|$ = number of terminal nodes

```{r}
plot(tree.carseats)
text(tree.carseats, pretty = 0) # shelving location is the most important indicator 
```

Evaluation on both training and test data

```{r}
set.seed(1)
train <- sample(1:nrow(Carseats), 200) # index 200 training datapoints

# get the test data
Carseats.test <- Carseats[-train,]

# get actual prediction
High.test <- High[-train]
tree.carseats <- tree(High ~ . - Sales, Carseats, subset = train)

tree.pred <- predict(tree.carseats, Carseats.test, type = "class") # return actual class predidction

table(tree.pred, High.test)

# correct prediction rate 
(table(tree.pred, High.test)[1, 1] + table(tree.pred, High.test)[2, 2])/nrow(Carseats.test) # 64% correct predictions

```

Tree with pruning

```{r}
set.seed(7)
cv.carseats <- cv.tree(tree.carseats, FUN = prune.misclass) # cross-validation to get the optimal level of tree complexity (depth); 
# prune.misclass = classification error rate guides the cross-validation and pruning process
# default = deviance guides the cross-validation and pruning process

cv.carseats

```

`size` = number of terminal nodes of each tree

`k` = $\alpha$ = cost--complexity parameter.

`dev` = number of cross-validation errors

```{r}
par(mfrow = c(1, 2))
plot(cv.carseats$size, cv.carseats$dev, type = "b")
plot(cv.carseats$k, cv.carseats$dev, type = "b")
```

It seems like 9 tree nodes have the lowest number of cv errors (`dev` in this case). Hence, we will pick this tree

```{r}
set.seed(7)
prune.carseats <- prune.misclass(tree.carseats, best = 9)
plot(prune.carseats)
text(prune.carseats, pretty = 0)

```

```{r}
tree.pred <- predict(prune.carseats, Carseats.test, type = "class")
table(tree.pred, High.test)

(table(tree.pred, High.test)[1, 1] + table(tree.pred, High.test)[2, 2]) /
    nrow(Carseats.test)
```

We can see slight improvement of prediction rate from 64 to over 70 percent

We can still increase the classification accuracy but with the price of tree complexity
