# Reproducible Pipelines with targets

```{r}
#| include: false
suppressPackageStartupMessages(library(tidyverse))
```

A data analysis is rarely a single script run once. It is a graph of steps: read raw data, clean it, build features, fit several models, score them, and render a report. When one step changes, only the steps that depend on it need to rerun. Rerunning everything wastes time and, worse, invites the silent error where some outputs reflect old code and others reflect new code. The `targets` package brings the discipline of build systems such as `make` to R, so that a pipeline reruns exactly the steps that are out of date and skips the rest, while guaranteeing that the final outputs are consistent with the current code and data.

By the end of this chapter you will be able to think of an analysis as a dependency graph, decide when a step is genuinely out of date, write a `targets` pipeline that rebuilds only what changed, scale it across many inputs with branching, and run it in parallel. We start with intuition, then make the rebuild rule precise with a little notation, then show both a real `targets` pipeline and a tiny base R engine that demonstrates the same mechanism on code you can run yourself.

> **Intuition:** Think of a recipe. If you change the frosting, you re-frost the cake; you do not re-bake it. A pipeline tool figures out the equivalent of "re-frost, do not re-bake" automatically, for every step, every time.

Because `targets` is probably not installed in your environment, the `targets` specific chunks are marked `eval=FALSE`. They are written to be correct and current so you can paste them into a session that has `targets` installed. To teach the core idea with code that actually runs, the chapter includes a base R pipeline engine that tracks dependencies with content hashes^[A *hash* is a function that maps any input to a short, fixed length string of characters (its *digest*), with the property that even a one character change in the input produces a completely different digest. We define a small one in the runnable demo below.] and skips up to date steps, which is the same mechanism `targets` uses under the hood.

## Where this fits in a modern workflow

In a small project you might keep one long script. As the project grows, the script becomes slow to rerun and fragile: you comment out the expensive model fit while you tweak a plot, then forget to uncomment it, and the report is built from stale objects. A pipeline tool removes this hazard by making the dependency structure explicit and by deciding, automatically, what must run.

The idea is borrowed from software build systems. `make` reads a set of rules of the form "target $T$ depends on prerequisites $P_1, \dots, P_k$ and is produced by command $C$." It rebuilds $T$ only when $T$ is missing or older than some prerequisite. `targets` adapts this to R by representing each step as a target object, tracking dependencies between R objects rather than files, and deciding staleness by comparing content hashes rather than file timestamps. The result is a workflow that is incremental (only stale steps run), reproducible (outputs always match current code and data), and inspectable (you can draw the dependency graph).

In an ML or AI engineering setting this matters because the expensive steps, fitting a gradient boosting model, computing embeddings, running a cross validated tuning grid, are exactly the ones you do not want to repeat unnecessarily. A pipeline lets you change a plot label and rebuild in seconds, while still guaranteeing that if you change the feature engineering, every downstream model and metric refits.

> **When to use this:** Reach for a pipeline tool once an end to end rerun is slow enough to break your concentration, or once you have lost confidence that the report on disk reflects the current code. Below that threshold, a single tidy script is fine.

## The dependency graph

To reason about "what depends on what," we need one piece of vocabulary. Model a pipeline as a directed acyclic graph (DAG) $G = (V, E)$.^[*Directed* means each edge has a direction (from a dependency to the thing that uses it). *Acyclic* means you can never follow the arrows in a loop back to where you started. *Graph* is just the collection of nodes and edges.] Each node $v \in V$ is a target: a named result computed by some R expression. A directed edge $(u, v) \in E$ means that computing $v$ uses the value of $u$, so $v$ depends on $u$. Acyclicity is required: a cycle would mean a target depends on itself, directly or transitively, which has no well defined build order.

Because $G$ is a DAG it admits a topological order, a sequence of the nodes in which every edge points forward. Build systems execute targets in a topological order so that each target's inputs are ready before it runs. Independent targets, those with no path between them, can run in any order relative to each other, which is what allows parallel execution.

### When does a target need to rebuild?

Assign to each target $v$ a signature $s(v)$ that summarizes everything the target's value depends on: the code of its command, the values (or signatures) of its upstream dependencies, and any external files it reads. A common choice for the signature is a hash, a function $h$ that maps arbitrary content to a fixed length digest such that different content almost surely yields different digests. Let $\mathrm{cmd}(v)$ be the source of the command, $\mathrm{deps}(v)$ the set of upstream targets, and $\mathrm{files}(v)$ the external files. Then

$$
s(v) = h\Big( \mathrm{cmd}(v),\; \{\, s(u) : u \in \mathrm{deps}(v) \,\},\; \{\, h(\text{contents of } f) : f \in \mathrm{files}(v) \,\} \Big).
$$

The pipeline stores $s(v)$ from the last successful run. On the next run it recomputes $s(v)$ and rebuilds $v$ if and only if the new signature differs from the stored one (or no stored value exists). Because each signature folds in the signatures of its dependencies, a change anywhere upstream propagates downstream: if you edit the code of target $u$, then $s(u)$ changes, which changes $s(v)$ for every $v$ that depends on $u$, and the whole affected subgraph rebuilds while everything else is skipped.

> **Key idea:** The signature folds in the signatures of its dependencies. That single recursive choice is what makes a change "ripple downstream" automatically: nothing has to track the ripple by hand, because the arithmetic of the hashes does it for you.

Two designs for the signature are worth distinguishing, and they differ in what question they ask about a target's inputs.

- Timestamps ask whether a target is older than any prerequisite. This is what `make` uses on files. It is cheap but fragile: touching a file without changing its contents forces a rebuild, and a clock skew across machines can give wrong answers.
- Content hashes ask whether the actual contents changed. This is what `targets` uses. It is robust to timestamp noise and to moving files between machines, at the cost of reading and hashing content.

> **Warning:** Timestamps lie in two directions. A file can be newer than its output yet have identical contents (a needless rebuild), or, after a clumsy file copy, be older than its output despite having changed (a missed rebuild). Content hashes avoid both failure modes, which is why modern tools prefer them.

The demo later in the chapter uses content hashing so the difference is concrete and runnable.

## The targets package and the `_targets.R` script

With the rebuild rule in hand, we can look at how a real pipeline is written. A `targets` pipeline is defined in a single file named `_targets.R` at the project root. It loads the package, sources your functions, sets options (such as which packages each target needs), and returns a list of target objects built with `tar_target()`. Each `tar_target(name, expression)` declares a node: `name` is the target's name and `expression` is the R code that produces it. The edges are not something you write down. `targets` discovers them automatically by static analysis of the expressions: if the expression for `model` mentions the symbol `features`, then `model` depends on `features`. You do not draw the graph by hand.

> **Note:** Inferring edges from code is what keeps the graph honest. A hand maintained list of dependencies drifts out of date the moment someone edits a function and forgets to update the list. Reading the dependencies straight from the code means the graph cannot disagree with what the code actually does.

```{r targets-script, eval=FALSE}
# _targets.R  (lives at the project root)
library(targets)

# Functions used by the pipeline, kept in a separate file
source("R/functions.R")

# Declare packages that targets should load when running each target
tar_option_set(packages = c("dplyr", "ggplot2", "glmnet"))

# The pipeline is a list of targets. Edges are inferred from the code:
# clean uses raw_file, features uses clean, model uses features, etc.
list(
  # An external file target: targets hashes the file's contents
  tar_target(raw_file, "data/raw.csv", format = "file"),

  tar_target(clean,    read_and_clean(raw_file)),
  tar_target(features, build_features(clean)),
  tar_target(model,    fit_model(features)),
  tar_target(metrics,  evaluate(model, features)),
  tar_target(plot,     plot_metrics(metrics))
)
```

The functions referenced above live in `R/functions.R`, which keeps the pipeline file a thin declaration of structure and puts the real work in testable functions.

```{r targets-functions, eval=FALSE}
# R/functions.R
read_and_clean <- function(path) {
  df <- read.csv(path)
  df[complete.cases(df), ]
}

build_features <- function(df) {
  df$logx <- log1p(df$x)
  df
}

fit_model <- function(df) {
  glmnet::cv.glmnet(as.matrix(df[, c("x", "logx")]), df$y, alpha = 1)
}

evaluate <- function(model, df) {
  pred <- as.numeric(predict(model, as.matrix(df[, c("x", "logx")]),
                             s = "lambda.min"))
  data.frame(rmse = sqrt(mean((df$y - pred)^2)))
}

plot_metrics <- function(metrics) {
  ggplot2::ggplot(metrics, ggplot2::aes(x = "model", y = rmse)) +
    ggplot2::geom_col()
}
```

Keeping the logic in `R/functions.R` and the structure in `_targets.R` is a deliberate split: the functions are ordinary R that you can test in isolation, and the pipeline file stays short enough to read at a glance.

Once the pipeline is declared, you drive it from the R console. The common verbs are these.

```{r targets-run, eval=FALSE}
library(targets)

tar_manifest()        # list the declared targets and their commands
tar_visnetwork()      # draw the dependency graph, colored by up-to-date status
tar_make()            # build everything that is out of date, skip the rest
tar_read(metrics)     # load a finished target's value into the session
tar_load(c(model, metrics))  # load several targets by name
tar_outdated()        # which targets would run on the next tar_make()
```

The first `tar_make()` runs every target and records each signature. If you then change only `plot_metrics()`, the next `tar_make()` reports that `clean`, `features`, `model`, and `metrics` are skipped and only `plot` runs. If instead you change `build_features()`, then `features` and everything downstream of it rebuilds, while `clean` is skipped. This is the staleness rule of the previous section in action.

> **Tip:** Run `tar_visnetwork()` before `tar_make()`. It draws the graph with each node colored by whether it is up to date, so you can confirm that the dependencies you intended are the ones `targets` inferred, and see exactly what the next build will touch, before spending any compute.

## A runnable base R pipeline that skips up-to-date steps

Reading about hashes and signatures is one thing; watching them skip steps is another. The following chunk runs. It implements a tiny pipeline engine in base R that does what `targets` does: it stores a signature for each step, and on rerun it executes a step only if its code or any upstream signature changed. The signature of a step folds in the hashes of its dependencies, so a change propagates downstream automatically.

> **Note:** This engine is a teaching model, not a replacement for `targets`. It keeps everything in memory, hashes deparsed code rather than using a cryptographic digest, and ignores parallelism. The point is to make the rebuild rule tangible in code you can step through, not to ship a build system.

```{r base-pipeline}
set.seed(1)

# A content hash. We use a simple, deterministic digest of the
# deparsed object so the demo needs no external package. A real
# pipeline would use a cryptographic hash; the design idea is the same.
hash <- function(x) {
  s <- paste(deparse(x), collapse = "\n")
  bytes <- as.integer(charToRaw(s))
  # Polynomial rolling hash kept inside 32-bit signed integer range.
  # Reducing modulo a prime below 2^31 after every step avoids overflow,
  # so different content reliably yields different digests.
  modulus <- 2147483647        # 2^31 - 1, a Mersenne prime
  base    <- 16777619
  h <- 2166136261 %% modulus
  for (b in bytes) {
    h <- (h * base + b) %% modulus
  }
  sprintf("%08x", as.integer(h))
}

# A "store" of signatures and cached values from the last build
store <- new.env()

# Define a target: a name, the names of its dependencies, and a
# function of the dependency values that produces the target value.
make_target <- function(name, deps, fun) {
  list(name = name, deps = deps, fun = fun)
}

# The build engine. Given targets in dependency order, build each
# only if its signature changed since the last run.
build <- function(targets, verbose = TRUE) {
  ran <- character(0)
  for (tg in targets) {
    dep_sigs <- vapply(tg$deps, function(d) store[[paste0("sig_", d)]],
                       character(1))
    # Signature folds in the code of the command and the signatures
    # of all upstream dependencies (see the staleness formula).
    sig <- hash(list(body = body(tg$fun), deps = dep_sigs))
    stored <- store[[paste0("sig_", tg$name)]]

    if (is.null(stored) || !identical(stored, sig)) {
      dep_vals <- lapply(tg$deps, function(d) store[[paste0("val_", d)]])
      val <- do.call(tg$fun, dep_vals)
      store[[paste0("val_", tg$name)]] <- val
      store[[paste0("sig_", tg$name)]] <- sig
      ran <- c(ran, tg$name)
      if (verbose) message("RUN  ", tg$name)
    } else {
      if (verbose) message("skip ", tg$name)
    }
  }
  invisible(ran)
}

# Define a four-step pipeline. The functions mimic real work.
pipeline <- list(
  make_target("clean", character(0), function() {
    Sys.sleep(0)                         # pretend this reads and cleans data
    data.frame(x = 1:200, y = 2 * (1:200) + rnorm(200, sd = 10))
  }),
  make_target("features", "clean", function(clean) {
    clean$logx <- log(clean$x)
    clean
  }),
  make_target("model", "features", function(features) {
    lm(y ~ x + logx, data = features)    # the "expensive" step
  }),
  make_target("metrics", "model", function(model) {
    data.frame(rmse = sqrt(mean(residuals(model)^2)))
  })
)

cat("First build (everything is new):\n")
first <- build(pipeline)

cat("\nSecond build with no changes (everything is up to date):\n")
second <- build(pipeline)

cat("\nTargets that ran on the first build: ", paste(first, collapse = ", "),
    "\nTargets that ran on the second build:",
    if (length(second)) paste(second, collapse = ", ") else "(none)", "\n")
```

The second build runs nothing because no signature changed. Now change one upstream step and rebuild. Editing the `features` command changes its signature, which changes the signature of `model` and then `metrics`, so those three rebuild while `clean` is skipped.

```{r base-pipeline-change}
# Edit the "features" step: add a squared term. Only this step and
# everything downstream of it should rerun; "clean" should be skipped.
pipeline[[2]] <- make_target("features", "clean", function(clean) {
  clean$logx <- log(clean$x)
  clean$x2   <- clean$x^2                 # the change
  clean
})

cat("Build after editing the 'features' step:\n")
changed <- build(pipeline)
cat("\nTargets that reran:", paste(changed, collapse = ", "), "\n")
```

This is the entire value proposition in miniature: a one line change to feature engineering reruns only the affected subgraph, and an unrelated change (say, to a downstream plot) would never trigger the model to refit. `targets` does the same thing, just with a robust hash, a persistent on disk store, automatic edge detection, and parallelism, all of which we get to next.

### A figure: cost of full reruns versus incremental builds

The payoff grows with pipeline length and with how expensive the unchanged steps are. Suppose a pipeline has $n$ steps in a chain, each costing $c$ seconds, and you edit only the last step. A naive rerun costs $n c$. An incremental build reruns only the edited step (and its descendants, here just one step), costing $c$ plus a small hashing overhead $\epsilon$ per skipped step. Figure @fig-reproducible-pipelines-targets-cost plots the two costs as the chain grows.

```{r fig-reproducible-pipelines-targets-cost, fig.cap="Wall-clock cost of rebuilding a chain pipeline when only the last step changes. Full rerun grows linearly with the number of steps; the incremental build stays nearly flat because skipped steps cost only a hash check.", fig.width=7, fig.height=4.2}
library(ggplot2)

c_step  <- 2.0      # seconds per step if it runs
epsilon <- 0.01     # seconds to hash and skip an up-to-date step
n_steps <- 1:30

full_rerun  <- n_steps * c_step
incremental <- c_step + (n_steps - 1) * epsilon  # only the last step runs

df <- rbind(
  data.frame(n = n_steps, seconds = full_rerun,  strategy = "Full rerun"),
  data.frame(n = n_steps, seconds = incremental, strategy = "Incremental build")
)

ggplot(df, aes(x = n, y = seconds, colour = strategy)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.6) +
  labs(x = "Number of steps in the pipeline",
       y = "Rebuild time (seconds)",
       colour = NULL,
       title = "Rerunning everything vs rebuilding only what changed") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")
```

The gap between the lines is the time `targets` saves you on every edit that does not touch the expensive upstream steps. In a real project where one step is a multi minute model fit, this is the difference between an interactive loop and a coffee break.

## Branching: one target, many values

So far each target has produced one value. Often, though, you want to apply the same computation across many inputs: fit the same model to each of fifty files, or sweep a hyperparameter over a grid, or bootstrap a statistic across many resamples. Writing one target per input by hand does not scale, and it loses the incremental benefit, since editing one input should not rerun the other forty nine. `targets` solves this with branching, which creates many sub targets from a pattern over upstream targets.

Two flavors exist, and the choice between them comes down to when you know the inputs.

- Static branching with `tar_map()` expands the pattern at the time you write the pipeline, producing named branches you can see in the manifest. Use it when the set of inputs is fixed and known in advance.
- Dynamic branching with the `pattern` argument creates branches at run time from the length of an upstream target. Use it when the number of inputs is not known until the pipeline runs, for example one branch per file found in a directory.

```{r targets-branching, eval=FALSE}
library(targets)
library(tarchetypes)   # provides tar_map() and helpers

list(
  # An upstream target that lists the inputs to branch over
  tar_target(files, list.files("data", full.names = TRUE)),

  # Dynamic branching: one branch per element of `files`.
  # map() runs the command once per file; targets tracks each branch.
  tar_target(
    cleaned,
    read_and_clean(files),
    pattern = map(files)
  ),

  # cross() forms the Cartesian product: every model type on every file
  tar_target(model_types, c("lasso", "ridge")),
  tar_target(
    fits,
    fit_one(cleaned, model_types),
    pattern = cross(cleaned, model_types)
  ),

  # A reduce-style target that combines all branch results into one table
  tar_target(summary, dplyr::bind_rows(fits))
)
```

The mathematics of branching is just an index set. If `files` has elements indexed by $i \in \{1, \dots, m\}$ and `model_types` by $j \in \{1, \dots, r\}$, then `map(files)` produces targets indexed by $i$, and `cross(cleaned, model_types)` produces targets indexed by the product set $\{(i, j)\}$, of size $m r$. Each branch has its own signature, so editing one input file rebuilds only the branches that read it, not the whole sweep.

> **Warning:** `cross()` multiplies. Two factors give $m r$ branches; a third gives $m r s$, and so on. A cross of three modest grids can quietly produce thousands of branches, each with its own metadata to track. Use `cross()` only over the dimensions you genuinely need to vary together.

## Scaling: parallel and distributed builds

Branching does more than save typing: by creating many independent targets, it gives the build something to parallelize. Independent targets, those with no path between them in $G$, can run at the same time. `targets` delegates this to the `crew` package, which manages a pool of worker processes (local or on a cluster or cloud). You change how the pipeline scales without changing what it computes: the targets and their edges stay the same, only the controller changes.

```{r targets-parallel, eval=FALSE}
# In _targets.R, add a controller for parallel local workers
library(targets)
library(crew)

tar_option_set(
  controller = crew_controller_local(workers = 4)
)

# Then build as usual; independent targets and branches run on the pool
# tar_make()  uses the controller automatically
```

The speedup is bounded by the structure of the graph, not just the number of workers. If the longest dependency chain (the critical path) has total cost $T_\infty$ and the total work across all targets is $T_1$, then with $p$ workers the build time obeys

$$
T_p \;\ge\; \max\!\left( T_\infty,\; \frac{T_1}{p} \right).
$$

No amount of parallelism beats the critical path $T_\infty$, because each step on it must wait for its predecessor. Branching helps precisely because it creates many independent targets, widening the graph so more work can proceed in parallel and $T_1 / p$ becomes the binding term rather than $T_\infty$.

> **Intuition:** A deep, narrow pipeline (one long chain) cannot be sped up much by adding workers, because every step waits on the one before it. A wide pipeline (many independent branches) can. If parallelism is not helping, the graph is probably too chain like, and the cure is to find work that can be split into independent branches.

## Comparison: targets, drake, make, and a plain script

`targets` is the current successor to `drake`, which pioneered hash based pipelines in R. `drake` is no longer actively developed and its author recommends `targets` for new projects. Table @tbl-reproducible-pipelines-targets-comparison contrasts the four approaches on the points that matter in practice.

| Property | Plain R script | GNU make | drake | targets |
|---|---|---|---|---|
| Unit of work | lines of code | file targets | R objects | R objects |
| Staleness test | none (always reruns all) | file timestamps | content hash | content hash |
| Dependency edges | implicit, in your head | written by hand in Makefile | inferred from code | inferred from code |
| Skips up-to-date steps | no | yes | yes | yes |
| Stores results | you manage `.RData` by hand | files on disk | hidden cache | `_targets/` object store |
| Branching over inputs | manual loops | manual rules | `map`/`cross` plans | `map`/`cross` + `tar_map()` |
| Parallel backend | manual (`future`, `parallel`) | `make -j` | `future`, `clustermq` | `crew` (local, cluster, cloud) |
| Visualize the graph | no | external tools | `vis_drake_graph()` | `tar_visnetwork()` |
| Status | works, error prone | language agnostic, R unaware | superseded, maintenance only | actively developed, recommended |

: Comparison of a plain R script, GNU make, drake, and targets across the workflow properties that matter in practice. {#tbl-reproducible-pipelines-targets-comparison}

The honest summary: a plain script is fine until reruns get slow or the object bookkeeping gets error prone. `make` is excellent and language agnostic but knows nothing about R objects, so you must marshal everything through files and hand write the rules. `drake` solved the R specific problem first; `targets` is its cleaner redesign with a smaller, more predictable internal model and the modern `crew` backend, and it is what you should reach for today.

## Practical guidance, pitfalls, and when to use it

Use `targets` when reruns have become slow enough to disrupt your iteration loop, when you have expensive steps you do not want to repeat, when a project has enough steps that you have lost track of what depends on what, or when you need a pipeline that a collaborator (or future you) can rebuild from scratch with one command and trust the result. It pairs naturally with R Markdown or Quarto reports as a final target, so the rendered document is always built from current upstream results.

Most of the trouble people hit comes from a handful of recurring mistakes. The common pitfalls are these.

- Hidden side effects. `targets` tracks values, not side effects. If a target writes to a database or downloads a file as a side effect and returns nothing meaningful, the dependency tracking cannot see it. Make external files explicit `format = "file"` targets so their contents are hashed, and return the path from the target.
- Global state and randomness. A target that depends on a global option, an environment variable, or an unset random seed can produce different results without its signature changing. Pass such inputs in as dependencies, and set seeds explicitly (`targets` manages per target seeds for you, but reproducibility still requires that your functions be deterministic given their inputs).
- Giant monolithic targets. If one target does everything, you lose the benefit of incremental rebuilds, because any change reruns the whole thing. Split work into the natural steps so that an edit reruns only the affected part.
- Functions defined in `_targets.R` instead of sourced files. Keep real logic in `R/` files that are sourced, and keep `_targets.R` a thin list of `tar_target()` calls. This keeps functions testable on their own and keeps the pipeline file readable.
- Branch explosion. `cross()` over several factors multiplies branch counts ($m r$ in the earlier notation, and more with additional factors). A cross of three modest grids can produce thousands of branches and a slow metadata bookkeeping step. Branch only over what you truly need to vary.
- Confusing the object store with version control. The `_targets/` directory is a cache, not source. Commit `_targets.R` and your `R/` functions to version control; the object store is regenerated by `tar_make()`.

A reasonable default workflow is short. Write your real work as functions in `R/`, declare a thin `_targets.R` listing the steps, run `tar_visnetwork()` to sanity check the graph, run `tar_make()`, and iterate. When the input sweep grows, add branching. When the build gets slow and the graph is wide, add a `crew` controller. Treat the rendered report as the terminal target so it is never out of date with respect to the data and code.

To recap the chapter in one breath: model an analysis as a DAG, give each target a content based signature that folds in its dependencies, and rebuild a target exactly when its signature changes. That one rule buys incremental rebuilds, reproducibility, and an inspectable graph. `targets` implements it for R objects with automatic edge detection, branching for many inputs, and a `crew` backend for parallel and distributed builds. The runnable base R engine above is the same idea stripped to its essentials, so if you understood why it skipped the right steps, you understand `targets`.

## Further reading

- Landau (2021). "The targets R package: a dynamic Make-like function-oriented pipeline toolkit for reproducibility and high-performance computing." Journal of Open Source Software.
- Landau (2018). "The drake R package: a pipeline toolkit for reproducibility and high-performance computing." Journal of Open Source Software. Background on the predecessor and the hash based design.
- The targets user manual and the targets R Markdown/Quarto guides, maintained by Landau, for current API and patterns (`_targets.R`, dynamic and static branching, `crew`).
- Stallman, McGrath, and Smith (2020). "GNU Make: A Program for Directing Recompilation." Free Software Foundation. The original timestamp based build system that motivates the design.
- Wilson, Bryan, Cranston, Kitzes, Nederbragt, and Teal (2017). "Good enough practices in scientific computing." PLoS Computational Biology. For the broader case for reproducible, automated analysis pipelines.
