# API

```{r}
#| include: false
suppressPackageStartupMessages(library(tidyverse))
```

A predictive model is only useful once other systems can *call* it. Training a
model in an R or Python session is one thing; making its predictions available to
a web app, a dashboard, a mobile client, or another team's service is another. You
can have the most accurate classifier in the world, but if it lives only inside your
laptop's R session, nobody else can use it. The standard solution is to wrap the
model behind an API (Application Programming Interface), most commonly an HTTP
REST API that accepts input data over the network and returns predictions.

This chapter is the bridge between *building* a model (everything in the earlier
chapters) and *operating* one. The goal is not to make you a backend engineer, but
to give you enough vocabulary and working examples that you can take a fitted model,
expose it as a service, and reason about the production concerns that follow. We
start with why serving through an API is the natural pattern, then walk through
minimal working examples in both R (`plumber`) and Python (FastAPI), and close with
the practical issues, validation, latency, scaling, security, and monitoring, that
separate a weekend demo from a service you can rely on.

> **Intuition:** Think of the API as a vending machine for predictions. The model is
> the stock inside; the API is the slot you put a request into and the tray the
> answer drops into. Callers never need to know what is inside or how it was built,
> only how to insert a request and read the result.

## Why Serve Models Through an API

An API is not the only way to use a model, but it is the one that scales across
teams, languages, and use cases. The reasons come down to four ideas that reinforce
one another.

The first is a clean separation of training and serving. Model training is
heavy, infrequent, and usually done by data scientists; serving is lightweight,
frequent, and consumed by applications. An API draws a clear boundary between the
two: you train offline, save the fitted model to disk, and load it into a serving
process whose only job is inference. The two halves can then evolve, scale, and fail
independently.

The second is language and platform independence. A model trained in R or Python
can be consumed by a JavaScript front end, a Java backend, or a one-line `curl`
command from a shell. The caller does not need R, Python, or any knowledge of how the
model works. It only needs to speak HTTP and JSON, two technologies that nearly every
language and platform already understands.^[HTTP (HyperText Transfer Protocol) is the
protocol your browser uses to talk to web servers. JSON (JavaScript Object Notation)
is a lightweight, human-readable text format for structured data, for example
`{"sepal_length": 5.1, "sepal_width": 3.5}`. Together they form the lingua franca of
web services.]

The third is the distinction between real-time and batch prediction. Some use
cases need real-time scoring: one request, one prediction, low latency, such as a
fraud check that must return before a checkout button finishes its click. Others are
batch: score millions of rows overnight and write the results to a table. APIs are
built for the real-time case. Batch scoring is usually a scheduled job rather than a
live service, though the same fitted model object can serve both.

The fourth, and the mechanism underneath all of this, is the **request/response
model. A client sends a request**, typically an HTTP `POST` whose body is a JSON
object containing the features. The server runs the model and returns a response,
JSON containing the prediction. Each request carries everything the server needs, so
the server keeps no memory of past calls.

> **Key idea:** Because each request is self-contained (the technical word is
> *stateless*), any server instance can answer any request. That is what lets you
> scale *horizontally*: when traffic grows, you add more identical server instances
> behind a load balancer instead of buying one bigger machine.

With the motivation in place, we now turn to how this looks in code, starting with R.

## Serving Models in R

The dominant tool in R is the [`plumber`](https://www.rplumber.io/) package, which
turns annotated R functions into a REST API. The idea is deliberately low-ceremony:
you write ordinary R functions and decorate them with special `#*` comments that
declare the HTTP method and path. `plumber` reads those comments, parses incoming
requests into function arguments, and serializes whatever the function returns back
into JSON. You never write parsing or networking code yourself.

> **When to use this:** Reach for `plumber` when your model is already in R and you
> want to expose it without porting anything to another language. It is the path of
> least resistance from a fitted R model to a live endpoint.

### A Minimal `plumber` Example

The typical workflow has two stages: train the model once and save it to disk, then
run a separate, lightweight serving script that loads the saved model and answers
requests. Suppose you trained a model and saved it with `saveRDS()`. The serving
script, conventionally named `plumber.R`, might look like this:

```{r, eval = FALSE}
# plumber.R
library(plumber)

# Load the fitted model once, when the API starts (not per request).
model <- readRDS("model.rds")

#* Health check
#* @get /health
function() {
  list(status = "ok")
}

#* Predict from input features
#* @param sepal_length:numeric
#* @param sepal_width:numeric
#* @param petal_length:numeric
#* @param petal_width:numeric
#* @post /predict
function(sepal_length, sepal_width, petal_length, petal_width) {
  newdata <- data.frame(
    Sepal.Length = as.numeric(sepal_length),
    Sepal.Width  = as.numeric(sepal_width),
    Petal.Length = as.numeric(petal_length),
    Petal.Width  = as.numeric(petal_width)
  )
  prediction <- predict(model, newdata, type = "response")
  list(prediction = unname(prediction))
}
```

Two details in that script are worth pausing on. The model is read with `readRDS()`
*outside* any function, so it loads a single time when the API process starts rather
than on every request (loading a model can be slow, and doing it per call would
cripple latency). And there are two endpoints: a trivial `/health` route that returns
`{"status": "ok"}`, and the real `/predict` route. The health route looks pointless
but earns its keep in production, where load balancers and monitoring systems
repeatedly ping it to confirm the service is alive.

> **Tip:** Always include a lightweight health-check endpoint. Orchestration tools
> (load balancers, Kubernetes, Posit Connect) use it to decide whether an instance
> is ready to receive traffic and when to restart one that has gone unhealthy.

With the script written, you launch the API by building a router from the file and
running it on a chosen port:

```{r, eval = FALSE}
library(plumber)
pr("plumber.R") |> pr_run(port = 8000)
```

The server is now listening on port 8000. A client can call it from any language; the
quickest way to test it is `curl` from a terminal, sending the features as a JSON
body:

```{bash, eval = FALSE}
curl -X POST "http://localhost:8000/predict" \
  -H "Content-Type: application/json" \
  -d '{"sepal_length":5.1,"sepal_width":3.5,"petal_length":1.4,"petal_width":0.2}'
```

The `-X POST` flag chooses the HTTP method, the `-H` flag declares that the body is
JSON, and `-d` carries the data. The server responds with JSON such as
`{"prediction": ["setosa"]}`. As a convenience, `plumber` also auto-generates
interactive API documentation (Swagger/OpenAPI), a web page where consumers can read
each endpoint's parameters and try requests in the browser, which makes the service
much easier to hand off to other developers.

### Versioning and Hosting

A single running endpoint is enough for a demo, but two production concerns deserve
dedicated tooling once a model matters to a business.

The first is model versioning. Models are not static: you retrain as new data
arrives, and a fresh model occasionally performs *worse* than the one it replaces. So
you need to track exactly which model is deployed and be able to roll back. The
[`vetiver`](https://rstudio.github.io/vetiver-r/) package standardizes the packaging,
versioning, and deployment of models. It stores model artifacts as versioned objects
through [`pins`](https://pins.rstudio.com/) (to a local folder, a cloud storage
bucket, or a Posit Connect server), and it can auto-generate both a `plumber` API and
a Dockerfile directly from a fitted model, so the path from model object to deployable
service is largely automated.

The second is hosting. Someone has to run the server, keep it up, secure it, and
scale it. Posit Connect (formerly RStudio Connect) is a managed platform that can
host `plumber` APIs and `vetiver` models with authentication, scaling, and monitoring
handled for you, so you *publish* a model rather than administer servers yourself.

> **Note:** Versioning and hosting are not R-specific ideas. Every serious serving
> setup, in any language, needs a way to track which model is live and a place to run
> it. `vetiver`, `pins`, and Posit Connect are simply R-friendly answers to those
> universal questions.

## Serving Models in Python

Much production model serving happens in Python, so it is worth seeing the same
pattern there. The two most common frameworks are FastAPI (modern, asynchronous,
with automatic input validation and OpenAPI documentation) and Flask (older and
more minimal). Both can serve models from scikit-learn, Keras/TensorFlow, PyTorch,
XGBoost, and other libraries. The example below uses FastAPI because its built-in
validation removes an entire class of bugs with almost no extra code.

A minimal FastAPI service for a scikit-learn (or Keras) model follows the same
load-once, define-endpoints structure as the `plumber` script:

```{python, eval = FALSE}
# app.py
from fastapi import FastAPI
from pydantic import BaseModel
import joblib

app = FastAPI()

# Load once at startup.
model = joblib.load("model.joblib")  # or keras.models.load_model("model.keras")

class Features(BaseModel):
    sepal_length: float
    sepal_width: float
    petal_length: float
    petal_width: float

@app.get("/health")
def health():
    return {"status": "ok"}

@app.post("/predict")
def predict(item: Features):
    X = [[item.sepal_length, item.sepal_width,
          item.petal_length, item.petal_width]]
    pred = model.predict(X)
    return {"prediction": pred.tolist()}
```

The structure mirrors the R version closely: the model loads once at startup, there is
a `/health` endpoint and a `/predict` endpoint, and the prediction is returned as
JSON. You run it with an ASGI server such as `uvicorn`, which is the process that
actually listens for network traffic and hands requests to your app:

```{bash, eval = FALSE}
uvicorn app:app --host 0.0.0.0 --port 8000
```

The one piece worth dwelling on is the `Features` class. By declaring the expected
fields and their types as a `pydantic` model, FastAPI gives you **input validation for
free**: a request with a missing field or a string where a number belongs is rejected
with a clear, automatic error before it ever reaches the model. Like `plumber`,
FastAPI also auto-generates interactive Swagger documentation.

> **Intuition:** The `pydantic` model is a bouncer at the door. It checks every
> request against the expected schema and turns away anything malformed, so your model
> code only ever sees clean, correctly typed inputs. We will see in a moment why that
> guarantee matters so much.

The R and Python examples differ in syntax but not in shape. That is the real lesson:
serving a model is the same handful of steps everywhere. Load the model once, define a
health check, define a prediction endpoint, validate inputs, and return JSON.

## Practical Concerns

Moving from a working demo to a reliable service raises several issues that apply
equally in R and Python. None of them is exotic, but each one is a common reason that
a model that "worked on my machine" misbehaves in production. We take them roughly in
the order a request flows through the system: validate it, decode it, run it quickly,
scale it, secure it, and watch it over time.

Input validation comes first because you can never trust incoming data. Requests
arrive from the open network, sometimes from buggy clients and occasionally from
malicious ones. Check types, ranges, required fields, and the allowed levels of any
categorical input, then reject anything malformed with an informative HTTP error code
(such as `400 Bad Request`) rather than letting it crash the model or, worse, produce
a confident but meaningless prediction.

> **Warning:** A model fed garbage rarely errors out. It usually returns a number, a
> wrong one, with no complaint. Silent garbage-in, garbage-out is far more dangerous
> than a loud crash, because nobody notices until decisions have been made on bad
> predictions. Validate aggressively.

JSON serialization is the next subtlety, because every input and output crosses
the wire as text. You have to be deliberate about how arrays, missing values
(`NA` in R, `null` in JSON), dates, and factor levels are encoded and decoded, so that
the model sees exactly the schema it was trained on. A factor level the model never
saw during training, or a date parsed in the wrong format, produces errors or quiet
nonsense that can be tedious to trace.

Batching addresses throughput. Scoring one row at a time wastes the fixed overhead
of each network round trip. If callers often need many predictions, design the
endpoint to accept an array of inputs so a whole batch arrives in a single request.
This amortizes the per-request cost and lets the model use its vectorized prediction
path, which is typically far faster than a loop over single rows.

Latency is what users actually feel, and the single biggest lever is the one the
examples already showed: load the model once at startup, never per request. Beyond
that, keep heavy preprocessing inside the serving process so it is not repeated
elsewhere, prefer vectorized operations, and, most importantly, *measure* end-to-end
response time under realistic load rather than guessing where the time goes.

> **Tip:** Profile before you optimize. The slow step is often something unglamorous,
> JSON parsing, a data-frame copy, a stray call that reloads the model, rather than the
> model's arithmetic itself. Measurement, not intuition, tells you where to spend
> effort.

Scaling and containerization handle growth in traffic. Package the service in a
Docker container so it runs identically on a laptop, a teammate's machine, and a
production server, eliminating the "works here, not there" class of problems.^[A
Docker container bundles your code together with its exact dependencies, the R or
Python runtime, system libraries, and your model, into one portable image. Running the
image anywhere reproduces the same environment, which is what makes deployments
predictable.] You can then run multiple replicas of that container behind a load
balancer, or on a platform such as Kubernetes, Posit Connect, or a cloud container
service, to absorb concurrent traffic. This is the horizontal scaling that the
stateless request/response model made possible at the start of the chapter.

Authentication keeps the endpoint from being open to the world. Protect it with
API keys, tokens (such as OAuth or JWT), or network-level controls so that only
authorized clients can request predictions. An unprotected prediction endpoint can be
abused, scraped, or simply overloaded by anyone who finds its address.

Monitoring is what turns a deployment into something you can trust over time. Log
requests, latencies, and errors, and, just as importantly, track the distribution of
incoming features and outgoing predictions. Watching those distributions is how you
detect data drift (the inputs gradually shifting away from the training
distribution) and model decay (accuracy eroding as the world changes). A model is
not a finished artifact; it ages as the data-generating process moves on.

> **Key idea:** A deployed model is a living system, not a delivered product.
> Monitoring is the feedback signal that tells you *when* to retrain and redeploy,
> which closes the loop back to the training chapters and connects naturally to the
> versioning tools (`vetiver`, `pins`) introduced earlier.

Taken together, these six concerns are the difference between a model that demos well
and one that survives contact with real traffic. You do not need to solve all of them
on day one, but you should know they exist and address each before the corresponding
failure finds you. With a fitted model, a serving framework, and these practices, you
can take any model from the earlier chapters and make it available to the rest of the
world.

## Further Reading

For going deeper, the following resources cover `plumber`, its surrounding ecosystem,
and example deployments:

- `plumber` documentation: <https://www.rplumber.io/>
- `plumber` community discussion: <https://community.rstudio.com/tag/plumber>
- Blair Drummond's plumber webinar:
  <https://github.com/blairj09-talks/plumber-webinar-2020>
- RStudio/Posit webinars: <https://github.com/rstudio/webinars>
- `rapidoc` (alternative API docs UI): <https://github.com/meztez/rapidoc>
- `plumberDeploy`: <https://github.com/meztez/plumberDeploy>
- `plumber` source: <https://github.com/rstudio/plumber>
- Curated plumber examples: <https://github.com/sol-eng/plumberExamples>
