# Boosting

-   Similar to [Bagging] but more powerful, where we can improve the predictions from decision trees

-   It was designed for classification problems (but now we can also apply to regression settings).

-   In the machine learning world, this method combines the output of many "weak" learners to produce a power "committee"

-   A classic boosting algorithm for classification is "AdaBoost.M1" [@freund1997b]

-   It can be considered as one of the best off-the-shelf classifiers

-   Under boosting, trees are *not grown independently*, but rather **sequentially**

-   There are no bootstrap samples with boosting because we build trees on the residuals in a step-by-step manner

-   In the literature, "fitting the data hard" means we fit a single large decision tree to the data, but we will **learn slowly** by fitting a decision tree to the residuals (not Y) from the model, which helps to improve the part of the tree that did not perform well in the existing tree structure.

-   We add this new decision tree into the fitted function with a shrinkage parameter ($\lambda$) to slow the process down to add more trees to areas that don't fit well (also helps prevent overfitting). Hence, each tree after will depend strongly on the trees that have already been grown.

## Regression Boosting

Algorithm for boosting regression trees

1.  Set $\hat{f}(x) = 0$ and $r_i = y_i \forall i$ (in the training set)

2.  For $b = 1, \dots, B$ repeat

    1.  Fit a tree $\hat{f}^b$ with $d$ splits ($d + 1$ terminal nodes) to the training data $(X,r)$

    2.  Update $\hat{f}$ by adding in a shrunken version of the new tree $\hat{f}(x) \leftarrow \hat{f}(x) + \lambda \hat{f}^b (x)$

    3.  Update the residuals $r_i \leftarrow r_i - \lambda \hat{f}^b (x_i)$

3.  Final boosted model $\hat{f}(x) = \sum_{b=1}^B \lambda \hat{f}^b (x)$

<br>

Boosting has 3 tuning parameters:

1.  $B$: the number of trees. boosting can overfit if $B$ is too large, we can use [Cross-validation Fitting] to select $B$
2.  $\lambda$: the shrinkage parameter, which controls the rate at which boosting learns. Typical values are 0.01 or 0.001 (best choice depends on the data at hand). The smaller the $\lambda$, the larger $B$ must be to achieve good performance
3.  $d$: the number of splits in each tree. $d = 1$ works well (surprisingly), which we call the tree a "stump." With $d =1$, the boosted ensemble is fitting an additive model since each term only involves one variable. More generally, $d$ is the interaction depth and controls the interaction order of the boosted model (since d splits can involve at most d variables).

<br>

## Classification Boosting

For classification problem, consider a two-class classification code such that $Y \in \{ -1, 1\}$ and a vector of predictor variables, $X$.

We have a classifier $G(X)$ that produces prediction that selects either -1 or 1. the error rate on the training sample is

$$
\bar{err} = \frac{1}{n} \sum_{i=1}^n I(y_i \neq G(x_i))
$$

A weak classifier is one whose error rate is only slightly better than random guessing. Now, boosting sequentially applies a weak classification algorithm to produce a sequence of weak classifiers, $G_m(x), m = 1, \dots, M$ ($M$ is above $B$)

The predictions are then combined through a **weighted majority vote** to produce the final prediction

$$
G(x) = sign (\sum_{m=1}^M \alpha_m G_m (x))
$$

where the weights $\alpha_m, m = 1, \dots, M$ are computed by the boosting algorithm and give highest influence to the best classifiers

At each boosting step, weights $w_1, \dots, w_n$ are applied to the training observations $(x_i, y_i), i = 1, \dots, n$.

To start, weights are set equal, $w_i = \frac{1}{n}$ so there is no preference for the first classification.

For each successive iteration, the observation weights at iteration $m$ are modified such that those observations that were misclassified by the classifier $G_{m-1}(x)$ have their weights increased, those that were correctly classified have their weights decreased

As iterations continues, observations that are difficult to classify correctly receive ever-increasing influence. Each successive classifier is forced to concentrate on those training observations that are missed by the previous classifiers in the sequence

<br>

Why boosting works so well?

From the perspective of a basis function expansion (\@ref(basis-function-representations):

$$
f(x) = \sum_{m=1}^M \beta_m b(x; \gamma_m)
$$

where

-   $\beta_m$ are the expansion coefficients

-   $b(x, \gamma)$ are simple functions of the multivariate inputs $x$ and parameters $\gamma$

then, boosting is a way of fitting such an additive expansion where the basis functions are the individual classifiers $G_m(x) \in \{ -1, 1\}$

To fit this model, we minimize some loss function averaged over the training data (e..g, squared error loss).

$$
\min_{\{\beta_m , \gamma_m \}^M_1} \sum_{i=1}^n L (y_i, \sum_{m=1}^M  \beta_m b(x_i; \gamma_m))
$$

For most loss functions $L(y, f(x))$ and basis functions $b(x; \gamma)$, the numerical optimization algorithms are usually computationally intensive.

In some cases, a solution is to solve the subproblem of fitting just a single basis function

$$
\min_{\{\beta, \gamma\}} \sum_{i=1}^n L(y_i, \beta b (x_i; \gamma))
$$

Then, we approximate the more general minimization problem by sequentially adding new basis functions to the expansion without adjusting the parameters and coefficients that have already been added. This process is called **forward stagewise additive modeling** (i.e., at each iteration $m$, we solve for the optimal basis function $b(x, \gamma_m)$ and coefficient $\beta_m$ to add to the current expansion $f_{m-1}(x)$, which gives $f_m(x)$ and we repeat the process while previously added terms remain constant)

Example: squared error loss

$$
L(y, f(x)) = (y - f(x))^2
$$

then

$$
\begin{aligned}
L(y_i, f_{m-1}(x_i) + \beta_m b(x_i; \gamma_m) &= (y_i - f_{m-1}(x_i) - \beta_mb(x_i; \gamma_m))^2 \\
&= (r_{im} - \beta_mb(x_i; \gamma_m))^2
\end{aligned}
$$

where $r_{im} = y_i - f_{m-1}(x)$ is the residual of the current model on the i-th observation.

The term $\beta_m b(x; \gamma_m)$ that best fits the current residuals is added to the expansion at each step

<br>

## AdaBoost

Ideas of AdaBoost

1.  It combines "weak learners" (e.g., stumps) to make classifications
2.  Weights of stumps are different
3.  Each stump is made sequentially based on previous stump's mistakes.

<br>

AdaBoost.M1

1.  Initialize the observation weights $w_i = \frac{1}{N}, i = 1, \dots, N$

2.  For $m = 1$ to M

    1.  Fit a classifier $G_m(x)$ to the training data with weights $w_i$

    2.  Calculate $err_m = \frac{\sum_{i=1}^{N} w_i I(y_i \neq G_m(x_i))}{\sum_{i=1}^N w_i}$

    3.  Calculate $\alpha_m = \log (\frac{1 - err_m}{err_m})$

    4.  Set $w_i \leftarrow w_i \times \exp(\alpha_m \times I (y_i \neq G_m(x_i))), i = 1, \dots, N$

3.  $G(x) = sign(\sum_{m=1}^M \alpha_m G_m(x))$

AdaBoost.M1 is equivalent to forward stagewise modeling using the loss function [@hastie2001a, 10.4]

$$
L(y, f(x)) = \exp(-y f(x))
$$

which is the "exponential loss" (closely related to the binomial negative log-likelihood (deviance), although they have different properties in finite data (real-world) settings )

This loss function is attractive because it facilitates computation

There are more loss functions. Typically we might choose a loss function because

-   it's robust to certain properties (e.g., outliers)

-   it facilitates computation

When a loss function is differentiable, we can make use of the gradient of the loss function to improve the optimization, which is the basis of gradient boosting ("gradient boosting models (GBM)"; formerly, Multiple Additive Regression Trees (MART)).

<br>

## Gradient Boosting

-   The algorithm XGBoost has become one of the standard implementations of gradient boosting, which is present in both R and Python.

    -   It sues Newton-Raphson for its optimization (i.e., a second order Taylor's approximation for the loss function to make the connection) instead of gradient descent

    -   It improves efficiency for more complicated settings than the original gradient boosting algorithms

-   Additional regularization and built-in cross-validation and handling of missing data are standard

-   Compare to [AdaBoost]: Gradient Boost makes a single leaf, representing an initial guess for the Weights of all of the samples. Similar to [AdaBoost], a new tree is made based on previous tree's errors and gradient boost scales the trees (but by the same amount). Unlike [AdaBoost], each tree can be larger than a stump.

-   Scaling the tree by the Learning Rate makes a small step in the right direction

<br>

### Gradient Boost Regression

**Input**: Data $\{(x_i, y_i)\}^n_{i=1}$ and a differentiable Loss Function $L(y_i, F(x))$ (e.g., Regression: squared residual: $0.5(\text{Observed} - \text{Predicted})^2$ )

Step 1: Initialize model with a constant value: $F_0(x) = \underset{\gamma}{\operatorname{argmin}} \sum_{i=1}^n L(y_i, \gamma)$

Step 2: for $m = 1 \to M$ (number of trees):

1.  Compute the (pseudo) residual: $r_{im} = - [ \frac{\partial L( y_i, F(x_i))}{\partial F(x_i)}]_{F(x) = F_{m-1}(x)}$ for $i = 1, \dots, n$ (where $\frac{\partial L( y_i, F(x_i))}{\partial F(x_i)}$ is the gradient, ergo the name gradient boost)
2.  Fit a regression tree to the $r_{im}$ values and create terminal regions $R_{jm}$ for $j = 1, \dots, J_m$
3.  For $j = 1, \dots, J_m$ compute $\gamma_{jm} = \underset{\gamma}{\operatorname{argmin}} \sum_{x_i \in R_{ij}} L(y_i, F_{m-1}(x_i) + \gamma)$
4.  Update $F_m (x) = F_{m-1}(x) + \nu \sum_{j=1}^{J_m} \gamma_{jm} I(x \in R_{jm})$ where $\nu$ is the learning rate (0 to 1) (smaller value means slower learning rate)

**Output**: $F_M(x)$

<br>

```{r, message=FALSE, warning=FALSE}
library(gbm)
library(ISLR2)
library(tree)

attach(Boston)


set.seed(1)
train <- sample(1:nrow(Boston), nrow(Boston)/2)
boston.test <- Boston[-train, "medv"]

boost.boston <-
    gbm(
        medv ~ .,
        data = Boston[train, ],
        distribution = "gaussian", # because this is regression setting
        # distribution = "bernoulli", # use this for classification
        n.trees = 5000, # number of trees
        interaction.depth = 4 # limits the depth of each tree 
    )

summary(boost.boston) #rm and lstate are most important
```

<br>

### Gradient Boost Classification

**Input**: Data $\{(x_i, y_i)\}^n_{i=1}$ and a differentiable Loss Function $L(y_i, F(x))$ (e.g., Classification: negative log likelihood: $- \sum_{i=1}^N \mathbf{y}_i \times\log (\mathbf{p}) + (1- \mathbf{y}_i) \times \log (1 - \mathbf{p})$ where $y_i$ is the observed value, equivalently, $- \text{Observed} \times \log(\text{odds}) + \log(1 + \exp(\log(\text{odds}))$ which is differentiable)

Step 1: Initialize model with a constant value: $F_0(x) = \underset{\gamma}{\operatorname{argmin}} \sum_{i=1}^n L(y_i, \gamma)$ where $\gamma$ is a log(odds) value

Step 2: for $m = 1 \to M$ (number of trees):

1.  Compute the (pseudo) residual: $r_{im} = - [ \frac{\partial L( y_i, F(x_i))}{\partial F(x_i)}]_{F(x) = F_{m-1}(x)}$ for $i = 1, \dots, n$ (where $\frac{\partial L( y_i, F(x_i))}{\partial F(x_i)}$ is the gradient, ergo the name gradient boost)
2.  Fit a regression tree to the $r_{im}$ values and create terminal regions $R_{jm}$ for $j = 1, \dots, J_m$
3.  For $j = 1, \dots, J_m$ compute $\gamma_{jm} = \underset{\gamma}{\operatorname{argmin}} \sum_{x_i \in R_{ij}} L(y_i, F_{m-1}(x_i) + \gamma)$
4.  Update $F_m (x) = F_{m-1}(x) + \nu \sum_{j=1}^{J_m} \gamma_{jm} I(x \in R_{jm})$ where $\nu$ is the learning rate (0 to 1) (smaller value means slower learning rate)

**Output**: $F_M(x)$

$$
\frac{\sum \text{Residual}_i}{\sum[\text{Previous Prob}_i \times (1 - \text{Previous Prob}_i)]}
$$

## Light Gradient Boosting

Reference: @machado2019 and [here](https://papers.nips.cc/paper/2017/hash/6449f44a102fde848669bdd9eb6b76fa-Abstract.html)

Based on new techniques:

1.  Gradient-based One Side Sampling (GOSS)
2.  Exclusive Feature Bundling (EFB)

Advantages:

-   Faster and higher efficiency

-   Lower Memory

-   Better accuracy

-   Can be used with parallel, distributed computing, GPU learning

-   Can handle big data.

-   An improvement on [XGBoost]

[Function Reference](https://lightgbm.readthedocs.io/en/latest/R/reference/)

Example by [lightgbm](https://lightgbm.readthedocs.io/en/latest/R/index.html)

```{r, message=FALSE, message=FALSE}
library(lightgbm)
data(agaricus.train, package='lightgbm')
train <- agaricus.train
dtrain <- lgb.Dataset(train$data, label = train$label)
model <- lgb.cv(
    params = list(
        objective = "regression"
        , metric = "l2"
    )
    , data = dtrain
)
```

Example by [Geeksforgeek](https://www.geeksforgeeks.org/lightgbm-light-gradient-boosting-machine/) (python)

[Data link](https://www.kaggle.com/lbronchal/breast-cancer-dataset-analysis?select=data.csv)

```{r}
# library(reticulate)
# py_install("lightgbm")
# py_install("pandas")
# py_install("xgboost")
```

```{python}
#importing standard libraries 
import numpy as np 
import pandas as pd 
from pandas import Series, DataFrame 

#import lightgbm and xgboost 
import lightgbm as lgb 
import xgboost as xgb 

#loading our training dataset 'adult.csv' with name 'data' using pandas 
data=pd.read_csv('images/adult.data',header=None) 

#Assigning names to the columns 
data.columns=['age','workclass','fnlwgt','education','education-num','marital_Status','occupation','relationship','race','sex','capital_gain','capital_loss','hours_per_week','native_country','Income'] 

#glimpse of the dataset 
data.head() 

# Label Encoding our target variable 
from sklearn.preprocessing import LabelEncoder,OneHotEncoder
l=LabelEncoder() 
l.fit(data.Income) 

l.classes_ 
data.Income=Series(l.transform(data.Income))  #label encoding our target variable 
data.Income.value_counts() 

 

#One Hot Encoding of the Categorical features 
one_hot_workclass=pd.get_dummies(data.workclass) 
one_hot_education=pd.get_dummies(data.education) 
one_hot_marital_Status=pd.get_dummies(data.marital_Status) 
one_hot_occupation=pd.get_dummies(data.occupation)
one_hot_relationship=pd.get_dummies(data.relationship) 
one_hot_race=pd.get_dummies(data.race) 
one_hot_sex=pd.get_dummies(data.sex) 
one_hot_native_country=pd.get_dummies(data.native_country) 

#removing categorical features 
data.drop(['workclass','education','marital_Status','occupation','relationship','race','sex','native_country'],axis=1,inplace=True) 

 

#Merging one hot encoded features with our dataset 'data' 
data=pd.concat([data,one_hot_workclass,one_hot_education,one_hot_marital_Status,one_hot_occupation,one_hot_relationship,one_hot_race,one_hot_sex,one_hot_native_country],axis=1) 

#removing duplicate columns 
_, i = np.unique(data.columns, return_index=True) 
data=data.iloc[:, i] 

#Here our target variable is 'Income' with values as 1 or 0.  
#Separating our data into features dataset x and our target dataset y 
x=data.drop('Income',axis=1) 
y=data.Income 

 

#Imputing missing values in our target variable 
y.fillna(y.mode()[0],inplace=True) 

#Now splitting our dataset into test and train 
from sklearn.model_selection import train_test_split 
x_train,x_test,y_train,y_test=train_test_split(x,y,test_size=.3)

#The data is stored in a DMatrix object 
#label is used to define our outcome variable
dtrain=xgb.DMatrix(x_train,label=y_train)
dtest=xgb.DMatrix(x_test)

#setting parameters for xgboost
parameters={'max_depth':7, 'eta':1, 'silent':1,'objective':'binary:logistic','eval_metric':'auc','learning_rate':.05}

#training our model 
num_round=50
from datetime import datetime 
start = datetime.now() 
xg=xgb.train(parameters,dtrain,num_round) 
stop = datetime.now()

#Execution time of the model 
execution_time_xgb = stop-start 
execution_time_xgb


#datetime.timedelta( , , ) representation => (days , seconds , microseconds) 
#now predicting our model on test set 
ypred=xg.predict(dtest) 
ypred


#Converting probabilities into 1 or 0  
for i in range(0,9769): 
    if ypred[i]>=.5:       # setting threshold to .5 
       ypred[i]=1 
    else: 
       ypred[i]=0  
       
#calculating accuracy of our model 
from sklearn.metrics import accuracy_score 
accuracy_xgb = accuracy_score(y_test,ypred) 
accuracy_xgb




# Light GBM


train_data=lgb.Dataset(x_train,label=y_train)
#setting parameters for lightgbm
param = {'num_leaves':150, 'objective':'binary','max_depth':7,'learning_rate':.05,'max_bin':200}
param['metric'] = ['auc', 'binary_logloss']

#Here we have set max_depth in xgb and LightGBM to 7 to have a fair comparison between the two.

#training our model using light gbm
num_round=50
start=datetime.now()
lgbm=lgb.train(param,train_data,num_round)
stop=datetime.now()

#Execution time of the model
execution_time_lgbm = stop-start
execution_time_lgbm


#predicting on test set
ypred2=lgbm.predict(x_test)
ypred2[0:5]  # showing first 5 predictions

#converting probabilities into 0 or 1
for i in range(0,9769):
    if ypred2[i]>=.5:       # setting threshold to .5
       ypred2[i]=1
    else:  
       ypred2[i]=0

#calculating accuracy
accuracy_lgbm = accuracy_score(ypred2,y_test)
accuracy_lgbm
y_test.value_counts()

from sklearn.metrics import roc_auc_score

#calculating roc_auc_score for xgboost
auc_xgb =  roc_auc_score(y_test,ypred)
auc_xgb



#calculating roc_auc_score for light gbm. 
auc_lgbm = roc_auc_score(y_test,ypred2)
auc_lgbm
comparison_dict = {
    "accuracy score":(accuracy_lgbm,accuracy_xgb),
    "auc score":(auc_lgbm,auc_xgb),
    "execution time":(execution_time_lgbm,execution_time_xgb)
    }

#Creating a dataframe ‘comparison_df’ for comparing the performance of Lightgbm and xgb. 
comparison_df = DataFrame(comparison_dict) 
comparison_df.index= ['LightGBM','xgboost'] 
comparison_df

```

<br>

## XGBoost

-   Extreme Gradient Boosting

-   

<br>

## Partial Dependence Plots

-   It can help interpret parameter effects from tree-based models (e..g, [Bagging] and [Boosting])

Consider a subvector $X_S$ of $l < p$ of the input predictor space $X$, indexed by $S \subset \{ 1, \dots, p \}$

Let $C$ be the complement such that $S \cup C = \{1, \dots, p \}$

We expect that $f(X) = f(X_S, X_C)$ and define the partial dependence of $f(X)$ on $X_S$ to be

$$
f_S(X_S) = E_{X_C} f(X_S, X_C)
$$

which is the marginal average of $f$

It can serve as a useful description of the effect of the chosen subset on $f(X)$ when elements of $X_S$ do not have strong interactions with elements of $X_C$

We can estimate this expectation by

$$
\bar{f}_S(X_S) = \frac{1}{n} \sum_{i=1}^n f(X_S, x_{iC})
$$

where $\{ x_{1C}, \dots, x_{nC}\}$ are the values of $X_C$ occurring in the training data

-   This can be computationally intensive to evaluate (but, for decision trees $\bar{f}_S(X_S)$ can be computed efficiently from the tree itself).

<!-- -->

-   The interpretation of the partial dependence plots is that they represent the effect of $X_S$ on $f(X)$ after accounting for the average effects of the other variables $X_C$ on $f(X)$

-   You can't assume that this present the effect of $X_S$ $f(X)$ *ignoring* the effects of $X_C$ (unless $X_S \perp X_C$)

Partial dependence plots for the 2 most important variables, showing marginal effect of the variables on the response after integrating out the other variables (i.e., conditioned on other variables)

```{r}
plot(boost.boston, i = "rm") # median house prices increase with rm
plot(boost.boston, i = "lstat") # median house prices decrease with lstate 
```

Examine test MSE

```{r}
attach(Boston)
set.seed(1)
train <- sample(1:nrow(Boston), nrow(Boston)/2)

yhat.boost <-
    predict(boost.boston, newdata = Boston[-train, ], n.trees = 5000)
mean((yhat.boost - boston.test) ^ 2)
```

The above analysis used $\lambda$ = 0.001 (shrinkage parameter)

But we can try other values

```{r}
boost.boston <-
    gbm(
        medv ~ .,
        data = Boston[train, ],
        distribution = "gaussian",
        n.trees = 5000,
        interaction.depth = 4,
        shrinkage = 0.2,
        verbose = F
    )
yhat.boost <-
    predict(boost.boston, newdata = Boston[-train, ], n.trees = 5000)
mean((yhat.boost - boston.test) ^ 2)
```

Lower shrinkage parameter (faster learning) might lead to lower test MSE
