# Prompt Engineering and Structured Output {#sec-prompt-engineering}

```{r}
#| include: false
suppressPackageStartupMessages(library(tidyverse))
```

Most of this book studies models whose parameters you fit. With a pretrained
large language model (LLM) you usually do not fit parameters at all. You shape
behavior by what you put in the context window: instructions, examples, and a
specification of the output you want back. Prompt engineering is the
discipline of designing that input so the model produces useful, parseable,
reliable results. Structured output is the part of that discipline that
forces the model to answer in a machine-readable format (typically JSON) so the
result can flow into the rest of a data pipeline.

Here is the intuition before any formalism. Think of a pretrained LLM as a fixed
function that takes a block of text in and produces a block of text out. You
cannot reach inside and retune it, but you can decide exactly what text goes in,
and you can decide how adventurous the model is allowed to be when it writes the
output. Those are the only two dials you have, and almost everything in this
chapter is a careful use of one of them. The reward for using them well is large:
a task that would normally need a labeled dataset and a training run can often be
solved by writing a few paragraphs of instructions and reading back a small,
well-shaped result.

This chapter treats prompting as a design problem with measurable trade-offs,
not folklore. We define the objects (system prompt, user prompt, few-shot
examples, tool schemas), give the probabilistic meaning of the decoding
parameters that control randomness (temperature and top-$p$), and show why
asking for JSON and then validating it is the difference between a demo and a
production component. By the end you will be able to read a prompt as a program,
explain in probabilistic terms what temperature does to the output, and turn a
chatty text generator into something a downstream job can call like a typed
function. The runnable demonstration is implemented in base R plus `glue`: a
prompt-template function that assembles a few-shot prompt from a table of
examples, plus a small JSON-schema validator that checks whether a model's
(simulated) reply conforms to the contract you asked for.^[We deliberately do
not call a paid API. A small deterministic stub stands in for the model, which
keeps the chapter fully reproducible and lets you see every moving part. The last
section shows the one-line change that swaps the stub for a real client.]

> **Key idea:** With a frozen LLM you never change the model. You change its
> input (the prompt) and you change how you sample from its output (the decoding
> parameters). Hold on to that distinction and the rest of the chapter is just
> two levers.

## Where this fits in a modern ML/AI workflow

In a classical supervised pipeline you collect labels, train a model, and serve
it. With a capable instruction-tuned LLM you often skip training entirely for a
first version of a task: classification, extraction, summarization, routing, and
data cleaning can be done by describing the task in a prompt and reading back
structured fields. The prompt is the program.

This matters for a data team in three concrete ways. The first is speed to a
baseline: a zero-shot or few-shot prompt is a working baseline in minutes, before
anyone labels a training set, and it tells you almost immediately whether the
task is even well posed. The second is that structured output acts as an API
contract. If the model returns free text, a human has to read it; if it returns
JSON that matches a fixed schema, a downstream job can consume it like any other
service. Validation is what turns a stochastic text generator into a typed
function. The third is that prompting sits at the bottom of a clean fallback
ladder. Prompting, few-shot prompting, retrieval (@sec-retrieval-augmented-generation),
tool use (@sec-llm-agents), and fine-tuning are
not competitors but rungs you climb only when the cheaper rung fails its
evaluation. Prompt engineering is the rung you should exhaust first.

> **When to use this:** Start with a prompt whenever the task can be described in
> words and the base model is already good at the underlying skill. Climb to
> retrieval or fine-tuning only after a held-out evaluation shows the prompt has
> plateaued below your target. Cheaper rungs first, always.

The mental model for the rest of the chapter: an LLM defines a conditional
distribution over output tokens given the input tokens, $p_\theta(y \mid x)$
with frozen parameters $\theta$. You do not change $\theta$. You change $x$ (the
prompt) and you change how you *sample* from $p_\theta(\cdot \mid x)$ (the
decoding parameters). Everything below is one of those two levers.

## The anatomy of a prompt

Before we can shape a prompt, we need names for its parts. Modern chat models
accept a list of messages, each tagged with a role, and you control three of
them. The system prompt holds persistent instructions that set the model's
task, tone, constraints, and output format; it is read once and conditions
everything after it, so this is where the contract belongs ("You are a
classifier. Return only JSON matching this schema."). The user prompt is the
specific request or input to act on, for example the one document you want
classified. Assistant messages are the model's replies, and you can also
fabricate them to supply worked examples (few-shot), where each example is a user
turn followed by the ideal assistant turn.

> **Tip:** The split between system and user prompts is the split between "rules
> that hold for every call" and "the one item this call is about." Putting the
> output schema in the system prompt and the data in the user prompt keeps the
> contract stable while the input varies.

Formally, the input $x$ is the concatenation of the rendered messages,
$x = \texttt{render}(m_1, \dots, m_k)$, where each $m_i = (\text{role}_i,
\text{content}_i)$. The model never sees roles as anything magical; they are
special tokens in $x$.^[Instruction-tuned models are trained so that text marked
as a system message is treated as higher-priority guidance than text in a user
message. The roles are not magic, but the training makes them behave as if they
were.] The practical consequence is that the *order* and *labeling* of content
changes the conditional distribution, which is why moving an instruction from the
user turn into the system turn can change behavior.

### Zero-shot, few-shot, and chain-of-thought

Three prompting patterns cover most applied use, and they differ only in what
extra content you place in $x$. Zero-shot prompting describes the task, gives
the input, and asks for the answer, with no examples at all; it works when the
task is common and the instruction is unambiguous. Few-shot prompting
prepends $n$ solved examples $(u_1, a_1), \dots, (u_n, a_n)$ before the real
input $u^\*$, so the examples teach the format and the decision boundary by
demonstration. This is in-context learning: the model adapts its behavior
from examples in the prompt with no weight update (Brown et al., 2020).
Chain-of-thought (CoT) prompting asks the model to produce intermediate
reasoning steps before the final answer (Wei et al., 2022); for multi-step
problems such as arithmetic, logic, or multi-hop questions this raises accuracy,
because the intermediate tokens give the autoregressive model a scratchpad, and
later tokens condition on the reasoning it already wrote.

> **Intuition:** A model generates one token at a time, each conditioned on
> everything before it. Letting it "think out loud" first means the final answer
> is generated in the presence of useful intermediate work, rather than being
> guessed in a single leap. The scratchpad is the point.

There is a real tension between CoT and structured output. Free-form reasoning
is text, not JSON. The standard resolution is to keep the reasoning *inside* a
field of the JSON object, for example a `"reasoning"` string emitted before the
`"answer"` field, so you get the accuracy benefit and a parseable result. The
ordering matters: because generation is left to right, the model must write the
reasoning field first for it to actually inform the answer field.

## Decoding parameters: the probabilistic view

We have talked about changing the input $x$. The second lever is how you draw
the output from $p_\theta(\cdot \mid x)$, and it is controlled by a handful of
decoding parameters. The intuition is simple: the model assigns a score to
every possible next word, and these parameters decide whether you always take the
top-scoring word or occasionally gamble on a lower-scoring one. Taking the top
word every time is repeatable and safe; gambling adds variety at the cost of
reliability. The math below just makes "gamble how much" precise.

At each step the model produces a vector of logits $z \in \mathbb{R}^{V}$
over a vocabulary of size $V$.^[A logit is an unnormalized score, the raw output
before it is turned into a probability. The vocabulary is the fixed set of tokens
the model can emit, often tens of thousands of word pieces.] The next-token
distribution is a softmax of those logits scaled by the temperature
$\tau > 0$:

$$
p_\tau(v) = \frac{\exp(z_v / \tau)}{\sum_{u=1}^{V} \exp(z_u / \tau)}, \qquad v = 1, \dots, V.
$$

Temperature reshapes the same logits without retraining anything.

- As $\tau \to 0^{+}$ the distribution concentrates on $\arg\max_v z_v$. Sampling
  becomes deterministic and equals greedy decoding.
- At $\tau = 1$ you sample from the model's native distribution.
- As $\tau$ grows the distribution flattens toward uniform, so rare tokens get
  more mass and output gets more diverse and less reliable.

A useful way to quantify "how flat" the distribution is is the entropy
$H(p_\tau) = -\sum_v p_\tau(v) \log p_\tau(v)$, which increases monotonically
with $\tau$ from $0$ (a point mass) toward $\log V$ (uniform). You can read
entropy here as a single number summarizing how undecided the model is at this
step: low entropy means it is confident in one token, high entropy means many
tokens look roughly equally good.

Top-$p$ (nucleus) sampling truncates the tail before sampling (Holtzman et
al., 2020). Sort tokens by probability and keep the smallest set
$\mathcal{N}_p$ whose cumulative mass first reaches $p$:

$$
\mathcal{N}_p = \min \Big\{ \mathcal{S} : \sum_{v \in \mathcal{S}} p_\tau(v) \ge p \Big\},
$$

then renormalize over $\mathcal{N}_p$ and sample. Setting $p = 1$ disables
truncation; small $p$ removes the long tail of implausible tokens while still
allowing variety among the plausible ones. Top-$k$ is the same idea but
keeps a fixed count $k$ of highest-probability tokens regardless of their mass.

The practical rule that follows from the math: for tasks with a single correct
answer (classification, extraction, anything you will parse), push $\tau$ toward
$0$ so the output is near-deterministic and reproducible. For open-ended
generation where you want variety (brainstorming, drafting), raise $\tau$ and
use top-$p$ around $0.9$ to cut the worst tail.

> **Warning:** Any temperature above $0$ makes the output irreproducible. The
> same prompt can give different answers on different runs, which turns bugs into
> intermittent ghosts and makes evaluation noisy. If you are going to parse the
> result, decode cold.

@tbl-prompt-engineering-decoding-knobs summarizes the levers. Read it
as a quick-reference card: each row is one knob, what it controls, and which
direction to turn it.

| Parameter | Symbol | Range | Effect | When to lower | When to raise |
|---|---|---|---|---|---|
| Temperature | $\tau$ | $(0, \infty)$ | scales logits before softmax | parsing, classification, reproducibility | brainstorming, creative drafts |
| Top-$p$ (nucleus) | $p$ | $(0, 1]$ | keep smallest set reaching mass $p$ | factual or structured output | diverse generation |
| Top-$k$ | $k$ | $\{1, 2, \dots\}$ | keep $k$ highest-probability tokens | tighten output | widen candidate set |
| Max tokens | | $\{1, 2, \dots\}$ | hard cap on output length | control cost and latency | long reasoning or documents |
| Stop sequences | | strings | end generation on a marker | enforce clean boundaries | rarely |

: Decoding parameters that control how output is sampled from the model, the quantity each one affects, and the direction to turn each knob for reliable parsing versus open-ended generation. {#tbl-prompt-engineering-decoding-knobs}

## Structured output and JSON schemas

So far we have shaped the input and tuned the sampling. The last piece is shaping
the output, because a free-text answer is hard to consume programmatically. The
fix is to specify the exact shape of the output and ask the model to return only
that. A JSON schema is a declarative description of an object: its fields,
their types, which are required, and any value constraints (for example an
enumeration of allowed labels).^[JSON, JavaScript Object Notation, is a simple
text format for nested key-value data. A "schema" is a description of what a
valid object looks like, much like a column specification for a data frame.] You
put the schema (or a compact description of it) in the system prompt, and you
validate the model's reply against it after generation.

Two failure modes make validation non-optional. The first is format drift:
the model wraps the JSON in prose ("Here is the result:") or in Markdown code
fences, so you must extract the JSON substring before parsing. The second is a
schema violation: the JSON parses cleanly but a required field is missing, a
type is wrong, or a label is outside the allowed set, and only schema validation
catches this.

> **Warning:** Valid JSON is not the same as correct JSON. A reply can parse
> perfectly and still carry a label your pipeline has never heard of. Parsing
> without validating is the single most common way an LLM component silently
> corrupts the data downstream of it.

The contract, then, is a short sequence: *parse, then validate, then act*. If
validation fails you retry with a corrective message, lower the temperature, or
fall back to a default record. Production LLM services also offer constrained
decoding that guarantees syntactically valid JSON, but valid JSON is not the same
as schema-conformant JSON, so you still validate the fields. The demo below
implements parse and validate from scratch so the moving parts are visible.

### Function and tool schemas

Tool use generalizes structured output. Instead of returning a final answer, the
model returns a structured call to a named function whose argument shape is
itself a JSON schema you provide. The model picks the function and fills the
arguments; your code executes it and feeds the result back.^[This is the
mechanism behind "agents" and tool-using assistants, developed in
@sec-llm-agents. The model never runs code
itself; it only proposes a structured call, and your program decides whether and
how to honor it.] A weather assistant
might expose a `get_weather(location: string, unit: enum["C","F"])` tool; the
model emits `{"name": "get_weather", "arguments": {"location": "Paris", "unit":
"C"}}`. From the model's side this is the same skill as structured output: emit
JSON that matches a declared schema. The only new piece is that you, not the
model, decide what to do with the validated arguments.

## Prompt templates: the runnable demo

A prompt template is a string with named slots that you fill from data. Keeping
templates as functions (rather than pasting strings) makes prompts versionable,
testable, and consistent across thousands of calls. The demo builds a few-shot
sentiment classifier prompt from a table of labeled examples, then validates a
(simulated) model reply against a JSON schema. We do not call a paid API here; a
small deterministic stub stands in for the model so the chapter is fully
runnable and reproducible. Swapping the stub for a real client is a one-line
change shown at the end.

```{r pe-setup, eval=TRUE, message=FALSE}
suppressPackageStartupMessages(library(glue))
suppressPackageStartupMessages(library(jsonlite))
set.seed(1301)

# Few-shot examples: an input text and its gold label.
examples <- data.frame(
  text  = c("This product exceeded my expectations.",
            "Worst purchase I have ever made.",
            "It works, nothing special either way."),
  label = c("positive", "negative", "neutral"),
  stringsAsFactors = FALSE
)

# The allowed labels define the enum constraint in our schema.
allowed_labels <- c("positive", "negative", "neutral")
```

The template function renders a system prompt that states the contract, a block
of few-shot examples, and the new input. Using `glue` keeps the slots explicit.

```{r pe-template, eval=TRUE}
render_examples <- function(ex) {
  # One example per shot: a user line then the ideal JSON assistant line.
  lines <- vapply(seq_len(nrow(ex)), function(i) {
    gold <- toJSON(list(reasoning = "stated explicitly in the text",
                        label = ex$label[i]),
                   auto_unbox = TRUE)
    glue("Input: {ex$text[i]}\nOutput: {gold}")
  }, character(1))
  paste(lines, collapse = "\n\n")
}

build_prompt <- function(new_text, ex, labels) {
  system_prompt <- glue(
    "You are a sentiment classifier. Read the input text and return ONLY a ",
    "JSON object with two fields: \"reasoning\" (a short string written first) ",
    "and \"label\" (one of: {paste(labels, collapse = ', ')}). ",
    "Do not add any text outside the JSON."
  )
  shots <- render_examples(ex)
  user_prompt <- glue("Input: {new_text}\nOutput:")
  list(system = as.character(system_prompt),
       user   = paste(shots, user_prompt, sep = "\n\n"))
}

prompt <- build_prompt("Absolutely fantastic, I love it.",
                       examples, allowed_labels)
cat(prompt$system, "\n\n---\n\n", prompt$user, sep = "")
```

Next, the JSON-schema validator. We define the schema as an R list and check a
parsed object field by field: required presence, type, and (for `label`) the
allowed enumeration. This is the same logic a production schema validator
applies, written out so each check is visible.

```{r pe-validate, eval=TRUE}
# Schema: required fields, their JSON types, and optional enum constraints.
schema <- list(
  reasoning = list(type = "string", required = TRUE),
  label     = list(type = "string", required = TRUE, enum = allowed_labels)
)

json_type <- function(x) {
  if (is.logical(x))   return("boolean")
  if (is.numeric(x))   return("number")
  if (is.character(x)) return("string")
  if (is.list(x))      return("object")
  "unknown"
}

extract_json <- function(txt) {
  # Strip code fences and surrounding prose: keep the first {...} block.
  m <- regmatches(txt, regexpr("\\{.*\\}", txt, perl = TRUE))
  if (length(m) == 0) NA_character_ else m
}

validate_against <- function(obj, schema) {
  errs <- character(0)
  for (field in names(schema)) {
    spec <- schema[[field]]
    if (is.null(obj[[field]])) {
      if (isTRUE(spec$required))
        errs <- c(errs, glue("missing required field '{field}'"))
      next
    }
    if (json_type(obj[[field]]) != spec$type)
      errs <- c(errs, glue("field '{field}' has type ",
                           "'{json_type(obj[[field]])}', expected '{spec$type}'"))
    if (!is.null(spec$enum) && !(obj[[field]] %in% spec$enum))
      errs <- c(errs, glue("field '{field}' value '{obj[[field]]}' ",
                           "not in allowed set"))
  }
  list(valid = length(errs) == 0, errors = errs)
}

# End to end: take raw model text, extract JSON, parse, validate.
process_reply <- function(raw_text, schema) {
  js <- extract_json(raw_text)
  if (is.na(js)) return(list(valid = FALSE, errors = "no JSON found", obj = NULL))
  obj <- tryCatch(fromJSON(js, simplifyVector = TRUE),
                  error = function(e) NULL)
  if (is.null(obj)) return(list(valid = FALSE,
                                errors = "JSON did not parse", obj = NULL))
  res <- validate_against(obj, schema)
  c(res, list(obj = obj))
}
```

Now a deterministic model stub. A real LLM is stochastic; here we mimic three
realistic reply styles so we can exercise the validator: a clean reply, a reply
wrapped in prose and a code fence, and a reply with an out-of-schema label.

```{r pe-stub, eval=TRUE}
model_stub <- function(prompt, style = c("clean", "fenced", "bad_label")) {
  style <- match.arg(style)
  switch(style,
    clean = '{"reasoning": "strong positive wording", "label": "positive"}',
    fenced = paste0("Here is the result:\n```json\n",
                    '{"reasoning": "clearly enthusiastic", "label": "positive"}',
                    "\n```"),
    bad_label = '{"reasoning": "seems good", "label": "happy"}'
  )
}

styles <- c("clean", "fenced", "bad_label")
results <- lapply(styles, function(s) {
  raw <- model_stub(prompt, s)
  process_reply(raw, schema)
})
names(results) <- styles

for (s in styles) {
  r <- results[[s]]
  cat(sprintf("[%s] valid = %s", s, r$valid))
  if (!r$valid) cat(sprintf("  (%s)", paste(r$errors, collapse = "; ")))
  cat("\n")
}
```

The clean and fenced replies validate, because the extractor strips the fence and
the surrounding prose. The `bad_label` reply, by contrast, parses as JSON but
fails the enum check, which is exactly the violation that silently corrupts a
pipeline if you parse without validating. This is the warning from earlier made
concrete: the validator, not the parser, is what protects you.

### A figure: temperature reshapes the next-token distribution

To make the decoding math concrete, fix a single set of logits and plot the
softmax distribution at several temperatures, with the entropy annotated. This is
the mechanism behind the reliability/diversity trade-off. Figure
@fig-prompt-engineering-temperature shows how the same logits give a peaked
distribution at low temperature and a flat one at high temperature.

```{r fig-prompt-engineering-temperature, eval=TRUE, fig.cap="Effect of temperature on the next-token distribution for a fixed set of logits. Low temperature concentrates mass on the top token (low entropy, near-deterministic); high temperature flattens the distribution (high entropy, diverse but less reliable).", fig.width=7, fig.height=4.5}
suppressPackageStartupMessages(library(ggplot2))

logits <- c(3.0, 2.0, 1.0, 0.5, 0.0, -1.0)
tokens <- factor(paste0("tok", seq_along(logits)),
                 levels = paste0("tok", seq_along(logits)))
temps  <- c(0.5, 1.0, 2.0)

softmax_t <- function(z, tau) {
  e <- exp(z / tau); e / sum(e)
}
entropy <- function(p) -sum(p * log(p))

df <- do.call(rbind, lapply(temps, function(tau) {
  p <- softmax_t(logits, tau)
  data.frame(token = tokens, prob = p,
             panel = sprintf("tau = %.1f  (H = %.2f)", tau, entropy(p)))
}))

ggplot(df, aes(token, prob)) +
  geom_col(fill = "steelblue") +
  facet_wrap(~ panel) +
  labs(x = "vocabulary token", y = "probability",
       title = "Temperature controls how peaked the sampling distribution is") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### A small simulation: temperature versus parse reliability

The argument that low temperature helps structured output can be simulated. Model
a generator that, with probability $q(\tau)$, emits schema-valid JSON and
otherwise emits malformed output, where $q$ decreases as $\tau$ rises (hotter
sampling wanders off-format more often). We estimate the valid-parse rate by
Monte Carlo across temperatures.

```{r pe-sim, eval=TRUE}
set.seed(7)

# Probability of a clean, schema-valid reply as a function of temperature.
# Monotone decreasing: a stylized but realistic relationship.
q_valid <- function(tau) plogis(3 - 1.6 * tau)

simulate_rate <- function(tau, n = 4000) {
  clean <- rbinom(n, 1, q_valid(tau))
  raw <- ifelse(clean == 1,
                '{"reasoning": "ok", "label": "positive"}',
                'Sure! label is positive')  # off-format, no JSON object
  valid <- vapply(raw, function(r) process_reply(r, schema)$valid,
                  logical(1))
  mean(valid)
}

taus <- c(0.0, 0.5, 1.0, 1.5, 2.0)
rates <- sapply(taus, simulate_rate)
data.frame(temperature = taus, valid_parse_rate = round(rates, 3))
```

The estimated valid-parse rate falls as temperature rises, which is the
empirical face of the entropy argument: hotter distributions put more mass on
tokens that break the format. For a task you intend to parse, this is a direct
reason to decode cold.

> **Note:** The relationship $q(\tau)$ here is stylized, chosen to illustrate the
> mechanism rather than measured from a specific model. The qualitative
> conclusion (cold decoding parses more reliably) is what holds in practice; the
> exact numbers will depend on the model and the task.

## Practical guidance, pitfalls, and when to use it

Knowing when to reach for prompting, and when to put it down, is half the skill.
Prompt engineering is the right first tool when the task is expressible in
instructions, the base model is already competent at the underlying skill, you
need a baseline fast, or labeled data is scarce. Classification, extraction,
routing, reformatting, and summarization all fit this description. You should
reach past it when prompting plateaus below your accuracy target on a held-out
set (consider retrieval or fine-tuning), when the task needs private knowledge
that is not in the prompt (retrieval), or when the latency and cost of long
few-shot prompts start to dominate (distill into a smaller fine-tuned model).

A handful of concrete practices pay off repeatedly, and they follow directly from
the ideas above:

- Put the contract in the system prompt and the data in the user prompt.
  State the output schema explicitly, including the allowed values.
- Order few-shot examples for coverage, including at least one example of
  each label and at least one hard or boundary case. Keep the example format
  byte-for-byte identical to what you ask for; the model copies the format it
  sees.
- Always validate, never trust. Parse, validate against a schema, and have a
  defined action on failure (retry with a corrective message, lower temperature,
  or emit a default record). Log validation failures; they are your evaluation
  signal, and feed naturally into a broader evaluation harness
  (@sec-evaluating-llms).
- Decode cold for anything you parse. Set temperature near $0$ and top-$p$
  modest. Save high temperature for genuinely open-ended generation.
- Pin versions. Prompts, schemas, model name, and decoding parameters are all
  part of the artifact. A prompt that worked on one model version can drift on
  the next, so evaluate on a fixed test set when any of these change.

The same ideas, read backward, name the most common ways teams get burned. Watch
for these pitfalls in particular:

- Parsing without validating. Valid JSON with a wrong field value passes
  `fromJSON` and poisons everything downstream.
- Putting chain-of-thought after the answer. Because generation is left to
  right, reasoning placed after the answer cannot inform it. Emit reasoning
  first, inside the JSON object.
- Overstuffed prompts. Past a point, more examples cost tokens and latency
  without raising accuracy, and can dilute the instruction. Measure, do not
  assume more is better.
- Hidden nondeterminism. Forgetting that temperature $> 0$ makes outputs
  irreproducible, which makes bugs intermittent and evaluation noisy.

> **Tip:** Treat the prompt, schema, model name, and decoding settings as one
> versioned artifact and keep a small fixed test set beside it. Then any change,
> including a silent model upgrade on the provider's side, can be caught by
> rerunning the evaluation rather than by a user noticing broken output.

### Swapping the stub for a real model

This is where the deterministic stub pays off. The demo isolates the model behind
one function, `model_stub(prompt, ...)`, so replacing it with a real client
changes only that call; the template builder and the validator are unchanged. The
chapter on calling LLM APIs from R (@sec-llm-apis-r) covers the client side in
depth. The snippet below sketches the shape of that call with a current R client. It is
marked `eval=FALSE` because it needs the `ellmer` package and a live API key,
neither of which is available in the book's build environment, so it is shown for
reference but not run.

```{r pe-real-client, eval=FALSE}
# Conceptual: route the assembled prompt to a real model and reuse the same
# validator. Requires a package not installed in this book's environment, so it
# is shown but not run. The 'ellmer' package provides a tidy R interface.
library(ellmer)

chat <- chat_anthropic(
  system_prompt = prompt$system,
  model = "claude-3-5-sonnet-latest",
  api_args = list(temperature = 0)   # decode cold for parseable output
)

raw_reply <- chat$chat(prompt$user)         # returns the model's text
result    <- process_reply(raw_reply, schema)  # same validator as above

if (!result$valid) {
  # Corrective retry: tell the model exactly what was wrong and ask again.
  fix_msg <- paste("Your reply was invalid:",
                   paste(result$errors, collapse = "; "),
                   "Return ONLY valid JSON matching the schema.")
  raw_reply <- chat$chat(fix_msg)
  result    <- process_reply(raw_reply, schema)
}
```

The discipline transfers directly to other interfaces. With OpenAI-style or
Anthropic tool schemas you would pass the JSON schema in the request so the
service constrains decoding, but you still run `validate_against` on the result,
because syntactic validity does not guarantee your field constraints.

## Further reading

- Brown, T. B., et al. (2020). Language Models Are Few-Shot Learners (GPT-3).
  *Advances in Neural Information Processing Systems*.
- Wei, J., Wang, X., Schuurmans, D., Bosma, M., Ichter, B., Xia, F., Chi, E.,
  Le, Q., and Zhou, D. (2022). Chain-of-Thought Prompting Elicits Reasoning in
  Large Language Models. *Advances in Neural Information Processing Systems*.
- Holtzman, A., Buys, J., Du, L., Forbes, M., and Choi, Y. (2020). The Curious
  Case of Neural Text Degeneration (nucleus sampling). *International Conference
  on Learning Representations*.
- Kojima, T., Gu, S. S., Reid, M., Matsuo, Y., and Iwasawa, Y. (2022). Large
  Language Models Are Zero-Shot Reasoners. *Advances in Neural Information
  Processing Systems*.
- Liu, P., Yuan, W., Fu, J., Jiang, Z., Hayashi, H., and Neubig, G. (2023).
  Pre-train, Prompt, and Predict: A Systematic Survey of Prompting Methods in
  Natural Language Processing. *ACM Computing Surveys*.
- Ouyang, L., et al. (2022). Training Language Models to Follow Instructions with
  Human Feedback (InstructGPT). *Advances in Neural Information Processing
  Systems*.
