# Evaluating LLMs and Guardrails {#sec-evaluating-llms}

```{r}
#| include: false
source("_common.R")
```

Once a large language model (LLM) is in a product, the hardest engineering
question is no longer "can it generate text" but "is the text good enough, safe
enough, and stable enough to ship." Evaluation answers that question. For a
prediction-focused practitioner the situation is familiar: an LLM is a function
that maps an input prompt to an output, and we want to estimate the quality of
that function on a distribution of inputs we care about. What changes is that
the output is free-form text rather than a class label or a number, so the loss
function is not handed to us.^[For a classifier the loss is obvious: did the
predicted label match the true label. For free-form text there is no single
"true" string, since many wordings can be equally good and many can be equally
bad, so we have to construct a scoring rule before we can measure anything.]
Most of the work in LLM evaluation is the work of defining a usable loss.


::: {.callout-important title="Key idea"}
An LLM is just a function from prompt to text. Evaluation is the
problem of inventing a sensible loss for that function, because, unlike
classification or regression, no loss comes built in.
:::


This chapter covers where evaluation fits in a modern ML/AI workflow, the
difference between generic benchmarks and task-specific evaluation, the common
scoring methods (exact match, token overlap, rubric scoring, LLM-as-judge), the
quality dimensions that guardrails target (hallucination and faithfulness,
toxicity and safety, calibration), and the regression-testing discipline that
keeps a prompt or model from silently degrading. The runnable demo is base R: it
computes exact-match and token-level $F_1$ for a question-answering (QA) task and
aggregates a scoring rubric over a set of mock model outputs, then prints a
scores table and a figure.

## Where Evaluation Fits

A typical LLM application pipeline has a few stages: a base or instruction-tuned
model, a prompt template (possibly with retrieved context), optional tool calls,
and a set of guardrails that filter or rewrite inputs and outputs. Evaluation
sits across all of these. During development you use it to choose a model and
tune a prompt. Before release you use it as an acceptance gate. In production you
use it as monitoring, sampling live traffic and re-scoring it to detect drift.

The key conceptual move is to treat the whole system, prompt and model and
guardrails together, as the object under test. A change to any component can
change behavior, so the evaluation set should exercise the system end to end, not
just the raw model. This is the same logic as testing a fitted pipeline rather
than a single estimator: the thing you deploy is the thing you score.


::: {.callout-tip title="Intuition"}
If you score the bare model but ship the model plus a retrieval
step plus three guardrails, your number describes something the user never
meets. Score the assembled system end to end.
:::


Two failure modes recur. The first is optimizing to a benchmark that does not
resemble your traffic, which gives a number that looks good and a product that
does not. The second is having no fixed evaluation set at all, so every prompt
edit is judged by eyeballing a handful of examples, which does not catch
regressions. The rest of the chapter is about avoiding both.

## Benchmarks Versus Task-Specific Evaluation

A benchmark is a fixed, public dataset with a fixed metric, built to compare
models on a general capability. Examples include broad knowledge and reasoning
suites and coding suites. Benchmarks are useful for model selection and for
tracking the field, but they have two limits. They saturate and leak: once a
dataset is widely published, later models may have seen it during pretraining, so
high scores can reflect memorization rather than capability.^[This is the text
analogue of testing a model on its own training data. If the answers were in the
pretraining corpus, a high benchmark score measures recall of seen examples, not
the generalization you actually care about.] And they measure a generic skill,
not your task. A model that ranks first on a reasoning benchmark can still be
wrong about your product's return policy.


::: {.callout-warning}
A leaderboard rank is a shortlisting signal, not a release
decision. The model that tops a public benchmark may know nothing about the
documents, jargon, and edge cases specific to your product.
:::


Task-specific evaluation is a dataset and metric you build from your own use
case: real or realistic prompts, reference answers or rubrics that encode your
quality bar, and a metric that reflects the cost of being wrong in your setting.
It is smaller and noisier than a benchmark but far more predictive of production
quality. The two are complements: use benchmarks to shortlist models, then use a
task-specific set to make the final call and to gate releases.

@tbl-evaluating-llms-benchcompare summarizes the trade-offs.

```{r tbl-evaluating-llms-benchcompare, echo = FALSE}
bench_tbl <- data.frame(
  Dimension = c("Goal", "Data source", "Coverage", "Leakage risk",
                "Predicts production quality", "Maintenance cost",
                "Best use"),
  `Public benchmark` = c("Compare general capability",
                         "Public, shared",
                         "Broad, generic",
                         "High (data may be in pretraining)",
                         "Weak", "Low (download once)",
                         "Shortlist models"),
  `Task-specific eval` = c("Measure quality on your task",
                          "Your prompts and references",
                          "Narrow, relevant",
                          "Low (private)",
                          "Strong", "Higher (curate, refresh)",
                          "Release gate and monitoring"),
  check.names = FALSE
)
knitr::kable(bench_tbl, caption = "Public benchmarks versus task-specific evaluation.")
```

## Scoring Methods and Their Math

The scoring method is the loss function for text. The right choice depends on
how constrained the output is: the more freedom the model has in how it phrases
an answer, the softer the metric needs to be. We move from the strictest metric
(exact match) to the most flexible (an LLM judge), and the trade is always the
same, since flexibility buys you tolerance for paraphrase at the cost of more
noise and more bias in the score.


::: {.callout-tip title="When to use this"}
Reach for the strictest metric the task allows. A
stricter metric is cheaper, more reproducible, and easier to audit. Only loosen
it when the output genuinely admits many correct surface forms.
:::


### Exact Match

When the correct answer is a short canonical string (a date, a name, a yes/no),
exact match (EM) is the simplest metric. Let the model produce prediction
$\hat{y}_i$ for example $i$ and let $y_i$ be the reference. After a normalization
function $g(\cdot)$ that lowercases, strips punctuation, and trims whitespace,

$$
\mathrm{EM} = \frac{1}{n} \sum_{i=1}^{n} \mathbb{1}\!\left[\, g(\hat{y}_i) = g(y_i) \,\right],
$$

where $\mathbb{1}[\cdot]$ is the indicator function and $n$ is the number of
examples. EM is strict: "Paris" and "Paris, France" both differ from "the city
of Paris," so EM understates quality when answers are phrased loosely. It is the
right metric only when there is a single accepted surface form.

### Token-Level F1

For short free-form answers, a softer metric compares the bag of tokens. Treat
the prediction and reference as multisets of tokens after normalization. Let
$\mathrm{TP}$ be the number of shared tokens (counted with multiplicity), so that

$$
\text{precision} = \frac{\mathrm{TP}}{|\hat{y}|}, \qquad
\text{recall} = \frac{\mathrm{TP}}{|y|},
$$

where $|\hat{y}|$ and $|y|$ are the token counts of the prediction and reference.
The $F_1$ score is their harmonic mean,

$$
F_1 = \frac{2 \cdot \text{precision} \cdot \text{recall}}{\text{precision} + \text{recall}},
$$

with $F_1 = 0$ when there are no shared tokens. This is the standard metric for
extractive QA datasets. It rewards partial overlap, so "Paris France" against the
reference "Paris" earns precision $1/2$ and recall $1$, giving $F_1 = 2/3$ rather
than zero. The number of shared tokens uses multiset intersection, so a token
that appears twice in the reference but once in the prediction contributes one to
$\mathrm{TP}$.^[The multiset (bag of tokens with counts) keeps the metric honest
about repetition: a prediction that repeats a correct word ten times does not earn
ten times the credit. It also means token $F_1$ ignores word order, so "dog bites
man" and "man bites dog" score identically, which is one of its known blind spots.]

### Rubric Scoring

When the output is a paragraph, a single string match is meaningless. A rubric
breaks quality into named criteria, each scored on an ordinal scale, and combines
them with weights. For example $i$ with criteria $j = 1, \dots, K$, let $s_{ij}$
be the score on criterion $j$ (say $0$ to $1$) and $w_j \ge 0$ a weight with
$\sum_j w_j = 1$. The aggregate score for example $i$ is

$$
S_i = \sum_{j=1}^{K} w_j \, s_{ij},
$$

and the dataset score is the mean $\bar S = \frac{1}{n}\sum_{i=1}^n S_i$. Rubrics
make the quality definition explicit and auditable, which matters when several
people disagree about what "good" means. The scores $s_{ij}$ can come from human
raters or from an LLM judge.


::: {.callout-tip}
Writing the rubric is itself the most valuable step. Forcing yourself
to name the criteria and assign weights surfaces hidden disagreements about
quality before they turn into arguments over a single output.
:::


### LLM-as-Judge

Asking a strong LLM to grade another model's output is LLM-as-judge. It scales
to free-form text where references are unavailable, and it can apply a rubric
written in natural language. It is also biased and noisy, so treat the judge as a
measurement instrument that itself needs validation against human labels. Known
biases include position bias (preferring the first option in a pairwise
comparison), length bias (preferring longer answers), and self-preference
(favoring outputs from the same model family). Mitigations include randomizing
option order, calibrating against a human-labeled subset, and reporting agreement
rates rather than treating the judge as ground truth.


::: {.callout-warning}
A judge that grades its own model's outputs can flatter them. Never
let a model be the sole arbiter of its own release without first checking how
well it agrees with human raters.
:::


@tbl-evaluating-llms-scorecompare compares the methods.

```{r tbl-evaluating-llms-scorecompare, echo = FALSE}
score_tbl <- data.frame(
  Method = c("Exact match", "Token F1", "Rubric (human)",
             "Rubric (LLM judge)", "Pairwise (LLM judge)"),
  `Output type` = c("Canonical string", "Short answer",
                   "Free-form text", "Free-form text", "Free-form text"),
  `Needs reference` = c("Yes", "Yes", "Optional", "Optional", "No"),
  Cost = c("Trivial", "Trivial", "High", "Low", "Low"),
  `Main weakness` = c("Too strict", "Ignores meaning/order",
                     "Slow, rater drift", "Judge bias", "Judge bias"),
  check.names = FALSE
)
knitr::kable(score_tbl, caption = "Scoring methods for LLM outputs.")
```

## Guardrail Dimensions

So far the metrics ask "is the answer right." Guardrails ask a different family
of questions: "is the answer supported, is it safe, and does it know what it does
not know." Guardrails are the checks and filters that keep a system within
acceptable behavior, and each one maps to a measurable evaluation dimension. We
take three in turn: faithfulness, safety, and calibration.

### Hallucination and Faithfulness

A hallucination is a confident statement not supported by the input or by
fact. In retrieval-augmented settings (@sec-retrieval-augmented-generation) the precise target is faithfulness:
every claim in the output should be entailed by the provided context. If a
response decomposes into claims $c_1, \dots, c_m$ and $e_k \in \{0,1\}$ marks
whether claim $k$ is supported by the context, a faithfulness score is

$$
\text{faithfulness} = \frac{1}{m}\sum_{k=1}^{m} e_k.
$$

The entailment labels $e_k$ come from a natural-language-inference model or an
LLM judge. Faithfulness is distinct from correctness: an answer can be faithful to
a wrong document, or correct but unsupported by the retrieved context. Report
both when you can.


::: {.callout-note}
Faithfulness measures the link between the output and the context you
gave the model, not the link between the output and the truth. A faithful
answer grounded in a wrong document is still wrong, which is why correctness
and faithfulness are reported as separate numbers.
:::


### Toxicity and Safety

Safety evaluation measures the rate at which a system produces disallowed content
(harassment, dangerous instructions, leaked secrets) and the rate at which it
incorrectly refuses benign requests. These are two error types. With a held-out
set of adversarial prompts and benign prompts, you estimate a violation rate on
the adversarial set and an over-refusal rate on the benign set.^[An over-refusal
is the safety analogue of a false positive: the system flags a harmless request as
dangerous and declines it. Track it explicitly, because a model that refuses
everything is perfectly "safe" and completely useless.] Pushing one to zero usually
raises the other, so safety tuning is an operating-point choice, the same trade-off
as a classification threshold.


::: {.callout-tip title="Intuition"}
Safety has two error types pulling against each other: letting
bad content through, and blocking good content. Driving the violation rate to
zero by refusing more aggressively just trades one failure for the other. The
goal is a deliberate operating point, not a corner.
:::


### Calibration

A model is calibrated if its stated confidence matches its accuracy, the same
property studied in the probability calibration chapter (@sec-probability-calibration). If the
model (or a wrapper) emits a probability $\hat p$ that its answer is correct,
calibration asks whether, among all answers with $\hat p \approx 0.8$, about 80
percent are in fact correct.


::: {.callout-tip title="Intuition"}
Calibration is about honesty, not accuracy. A weather forecaster
who says "70 percent chance of rain" is well calibrated if it rains on about 70
percent of such days, even if she is wrong half the time. A model can be
accurate but overconfident, or mediocre but well calibrated.
:::


Bin the predictions into $B$ confidence bins. For bin
$b$ with index set $I_b$, let

$$
\text{acc}(b) = \frac{1}{|I_b|}\sum_{i \in I_b} \mathbb{1}[\hat y_i = y_i],
\qquad
\text{conf}(b) = \frac{1}{|I_b|}\sum_{i \in I_b} \hat p_i .
$$

The expected calibration error aggregates the gap, weighting each bin by its
share of examples:

$$
\mathrm{ECE} = \sum_{b=1}^{B} \frac{|I_b|}{n}\,\bigl|\,\text{acc}(b) - \text{conf}(b)\,\bigr|.
$$

A low $\mathrm{ECE}$ means confidence is trustworthy, which lets you route
low-confidence answers to a human or a fallback. Calibration matters for
guardrails because an abstention policy ("answer only when confident") is only as
good as the confidence estimate behind it. When you need distribution-free
guarantees on such abstention decisions, conformal prediction
(@sec-conformal-prediction) supplies confidence sets with finite-sample coverage.

## Regression Testing for Prompts

A prompt is code.^[Treat it that way literally: keep prompts in version control,
review changes, and never edit the production prompt without running it past the
evaluation set first. A one-word change to a system message is a deploy.] A small
wording change, a new model version, or a tweaked system message can shift behavior
in ways that are invisible until a user hits them. Regression testing treats the
evaluation set as a test suite: pin a set
of inputs with expected properties, run the current system against it on every
change, and compare to the last accepted run. A change is accepted only if the
aggregate score does not drop and no protected example regresses.

Practical structure: keep golden examples (inputs with reference answers or
must-pass assertions) under version control, run them in continuous integration,
and fail the build when the score falls below a threshold or when any example
flagged as critical flips from pass to fail. Track scores over time so you can see
slow drift, not just single-commit breaks. Because LLM outputs can be
nondeterministic, fix the sampling temperature to zero for tests where you can,
and for the rest run several samples and compare distributions rather than single
outputs.


::: {.callout-tip}
A single passing example is weak evidence when the model samples
randomly. For nondeterministic tests, run each input several times and assert on
the distribution (for example, the answer must pass in at least 9 of 10 draws)
rather than on one lucky output.
:::


## Runnable Demo: EM, Token F1, and Rubric Aggregation

The following base R code implements the QA metrics and rubric aggregation from
the math above, runs them over a small set of mock model outputs, and reports a
scores table and a figure. It uses only base R so it runs anywhere.

```{r eval-setup}
# Normalization shared by exact match and token F1: lowercase, drop
# punctuation, collapse and trim whitespace. This mirrors standard QA
# evaluation so that "Paris." and "paris" compare equal.
normalize_text <- function(x) {
  x <- tolower(x)
  x <- gsub("[[:punct:]]", " ", x)
  x <- gsub("\\s+", " ", x)
  trimws(x)
}

tokenize <- function(x) {
  x <- normalize_text(x)
  if (nchar(x) == 0) return(character(0))
  strsplit(x, " ", fixed = TRUE)[[1]]
}

exact_match <- function(pred, ref) {
  as.integer(normalize_text(pred) == normalize_text(ref))
}

# Token-level F1 using multiset (bag-of-tokens) intersection.
token_f1 <- function(pred, ref) {
  p_tok <- tokenize(pred)
  r_tok <- tokenize(ref)
  if (length(p_tok) == 0 && length(r_tok) == 0) return(1)
  if (length(p_tok) == 0 || length(r_tok) == 0) return(0)

  p_counts <- table(p_tok)
  r_counts <- table(r_tok)
  shared   <- intersect(names(p_counts), names(r_counts))
  # multiset intersection: min count per shared token
  tp <- sum(pmin(p_counts[shared], r_counts[shared]))
  if (tp == 0) return(0)

  precision <- tp / length(p_tok)
  recall    <- tp / length(r_tok)
  2 * precision * recall / (precision + recall)
}
```

```{r eval-qa-data}
# Mock QA outputs: a model's predictions against gold references.
qa <- data.frame(
  question = c(
    "What is the capital of France?",
    "Who wrote Pride and Prejudice?",
    "What year did the first moon landing occur?",
    "What is the chemical symbol for sodium?",
    "How many continents are there?"
  ),
  reference = c("Paris", "Jane Austen", "1969", "Na", "seven"),
  prediction = c(
    "Paris, France",          # partial overlap, not exact
    "Jane Austen",            # exact
    "It happened in 1969.",   # contains the answer plus extra tokens
    "Sodium",                 # wrong surface form
    "7"                       # correct concept, different token
  ),
  stringsAsFactors = FALSE
)

qa$EM <- mapply(exact_match, qa$prediction, qa$reference)
qa$F1 <- round(mapply(token_f1, qa$prediction, qa$reference), 3)
qa
```

```{r tbl-evaluating-llms-qa-summary}
qa_summary <- data.frame(
  metric = c("Exact match", "Token F1"),
  score  = round(c(mean(qa$EM), mean(qa$F1)), 3)
)
knitr::kable(qa_summary, caption = "Aggregate QA scores over the mock set.")
```

@tbl-evaluating-llms-qa-summary reports the aggregate scores.
The gap between EM and $F_1$ is the point: EM credits only the second example,
while $F_1$ gives partial credit to "Paris, France" and to the answer embedded in
a longer sentence. Neither metric handles "7" versus "seven," which is a
reminder that surface-form metrics miss meaning and why semantic or judge-based
scoring is sometimes needed.

Now the rubric. We score four free-form summaries on three weighted criteria.

```{r eval-rubric}
# Rubric scores in [0,1] for three criteria, assigned here as mock
# judge or human ratings for four candidate summaries.
rubric_scores <- data.frame(
  output      = paste0("summary_", 1:4),
  relevance   = c(0.9, 0.6, 1.0, 0.4),
  faithfulness= c(1.0, 0.5, 0.8, 0.3),
  fluency     = c(0.8, 0.9, 0.7, 0.6),
  stringsAsFactors = FALSE
)

# Weights encode the quality bar: faithfulness matters most here.
weights <- c(relevance = 0.3, faithfulness = 0.5, fluency = 0.2)
stopifnot(abs(sum(weights) - 1) < 1e-8)

crit_mat <- as.matrix(rubric_scores[, names(weights)])
rubric_scores$aggregate <- round(as.numeric(crit_mat %*% weights), 3)
rubric_scores
```

```{r tbl-evaluating-llms-rubric-table}
rubric_summary <- data.frame(
  statistic = c("Mean aggregate", "Min aggregate", "Max aggregate"),
  value     = round(c(mean(rubric_scores$aggregate),
                      min(rubric_scores$aggregate),
                      max(rubric_scores$aggregate)), 3)
)
knitr::kable(rubric_summary,
             caption = "Weighted rubric aggregate across four mock outputs.")
```

@tbl-evaluating-llms-rubric-table summarizes the spread of the
weighted aggregate across the four mock outputs.

```{r fig-evaluating-llms-figure, fig.cap = "Weighted rubric aggregate score per candidate summary. The dashed line marks a release threshold of 0.7; only summaries above it would pass the gate."}
op <- par(mar = c(5, 4, 3, 1))
bp <- barplot(
  height    = rubric_scores$aggregate,
  names.arg = rubric_scores$output,
  ylim      = c(0, 1),
  col       = ifelse(rubric_scores$aggregate >= 0.7, "steelblue", "grey70"),
  ylab      = "weighted aggregate score",
  main      = "Rubric aggregate by candidate output"
)
abline(h = 0.7, lty = 2, lwd = 2, col = "firebrick")
text(bp, rubric_scores$aggregate + 0.04, labels = rubric_scores$aggregate)
par(op)
```

@fig-evaluating-llms-figure shows the release gate in action: with a
threshold of $0.7$ on the weighted aggregate, two of the four mock summaries pass. Lowering
faithfulness weight would let the unfaithful summary through, which is exactly
the kind of policy choice a rubric makes explicit.

## A Production-Style Judge Call (Not Run Here)

The metrics above need no external service. An LLM-as-judge does. The chunk below
shows current, idiomatic R for calling a judge through the `ellmer` package and
parsing a structured score. It is marked `eval = FALSE` because `ellmer` is not
installed in this build and a live call needs an API key.

```{r judge-call, eval = FALSE}
library(ellmer)

# A judge that returns a structured rubric score for one answer.
judge <- chat_anthropic(
  model  = "claude-sonnet-4-5",
  system_prompt = paste(
    "You are a strict grader. Score the ANSWER against the QUESTION and",
    "CONTEXT on three criteria, each from 0 to 1: relevance, faithfulness",
    "(every claim supported by CONTEXT), and fluency."
  )
)

score_type <- type_object(
  relevance    = type_number("Relevance to the question, 0 to 1."),
  faithfulness = type_number("Support by the context, 0 to 1."),
  fluency      = type_number("Readability, 0 to 1."),
  rationale    = type_string("One sentence explaining the scores.")
)

grade_one <- function(question, context, answer) {
  prompt <- sprintf(
    "QUESTION:\n%s\n\nCONTEXT:\n%s\n\nANSWER:\n%s",
    question, context, answer
  )
  judge$chat_structured(prompt, type = score_type)
}

# To reduce position and length bias, randomize option order in pairwise
# comparisons and validate the judge against a human-labeled subset before
# trusting its scores in a release gate.
```

The same structured output then flows into the weighted aggregation from the
runnable demo: collect `relevance`, `faithfulness`, and `fluency`, multiply by the
weight vector, and compare the mean to the release threshold.

## Practical Guidance and Pitfalls

Build the task-specific set first. A few hundred curated examples that look like
your traffic beat any public benchmark for deciding whether to ship. Stratify the
set by the slices you care about (topic, user type, difficulty) and report scores
per slice, because an aggregate can hide a regression in a minority slice.

Match the metric to the output. Use EM only for canonical short answers, token
$F_1$ for short free-form answers, rubrics for paragraphs, and a judge only after
you have measured its agreement with humans. Reporting a single headline number
across mixed output types hides more than it reveals.

Treat the judge as an instrument, not an oracle. Calibrate it against a
human-labeled subset, randomize option order, watch for length and
self-preference bias, and report inter-rater agreement (for example Cohen's
$\kappa$) between judge and humans. If the judge disagrees with humans more than
humans disagree with each other, the judge is not ready to gate releases.

Separate correctness from faithfulness from safety. They fail independently. A
RAG system can be faithful to a wrong document, correct from memory but
unsupported, or safe yet useless. Score each dimension on its own and decide the
operating point deliberately, since reducing over-refusal usually raises the
violation rate.

Pin the evaluation set and version it. Refresh it on a schedule, not on a whim,
and keep a frozen subset so historical scores stay comparable. Fix sampling
temperature to zero in tests where you can, and for nondeterministic cases compare
score distributions over several samples rather than single outputs.


::: {.callout-tip title="When to use this"}
Build a full evaluation harness once a change can reach
users, once several people must agree on what "good" means, or once you need to
detect drift over time. Throwaway prototypes and exploratory prompting do not
need one; a small golden set and a couple of assertions are enough.
:::


When not to do heavy evaluation: throwaway prototypes and exploratory prompting
do not need a full harness. The cost is justified once a change can reach users,
once several people must agree on quality, or once you need to detect drift over
time. Below that bar, a small golden set and a couple of assertions are enough.

To summarize the workflow: shortlist models with public benchmarks, then build a
private task-specific set that looks like your traffic; pick the strictest scoring
method the output allows, and validate any LLM judge against humans before
trusting it; score correctness, faithfulness, and safety as separate dimensions
and choose the operating point on purpose; and finally pin the set, run it on
every change, and watch the scores over time so that drift shows up as a trend
rather than a surprise.

## Further Reading

- Rajpurkar, Zhang, Lopyrev, and Liang (2016) introduced SQuAD and the token-level
  exact-match and $F_1$ metrics used for extractive QA.
- Lin (2004) defined ROUGE, the recall-oriented overlap metric for summarization.
- Papineni, Roukos, Ward, and Zhu (2002) defined BLEU, the precision-oriented
  n-gram overlap metric for generation.
- Zheng, Chiang, Sheng, and colleagues (2023) studied LLM-as-judge, documenting
  position, length, and self-preference biases and agreement with human raters.
- Guo, Pleiss, Sun, and Weinberger (2017) analyzed calibration of modern neural
  networks and the expected calibration error.
- Naeini, Cooper, and Hauskrecht (2015) introduced the binned estimator behind the
  expected calibration error.
- Es, James, Espinosa-Anke, and Schockaert (2024) proposed RAGAS, a framework for
  faithfulness and answer-relevance scoring in retrieval-augmented generation.
- Liang, Bommasani, and colleagues (2022) presented HELM, a holistic benchmarking
  framework emphasizing multiple metrics and scenarios over a single score.
