# Building AI Applications with Shiny

A trained model is not a product. The gap between a model object sitting in an R
session and something a colleague can actually use is filled by an application
layer: a way to take input, run inference, and return a result. Shiny is the
R framework for building that layer as an interactive web application, written
entirely in R. This chapter shows how Shiny works, how its reactive programming
model maps onto serving a model, and how to wrap a large language model (LLM)
backend in a chat interface that streams responses.

The focus is practical. We treat Shiny as the deployment surface for the models
built earlier in this book, with special attention to the patterns that an
LLM-powered application needs: managing conversation state, calling an external
inference API, streaming tokens back to the browser, and handling the failure
modes that come with network calls and paid endpoints. Because Shiny is not in
the set of packages we run live here, all Shiny code is shown with `eval=FALSE`.
The reactive idea itself, which is the part most worth understanding, is
demonstrated with a small runnable observer pattern in base R.

> **Key idea:** Shiny lets you declare *relationships* between values rather than
> writing a script that runs once from top to bottom. You say "this output
> depends on that input," and the framework figures out what to recompute when
> something changes. Holding that one idea in mind makes everything else in this
> chapter fall into place.

By the end you will be able to read and reason about a Shiny app, explain why its
reactive engine is a natural fit for serving a model interactively, wrap an LLM
behind a chat interface that streams its reply, and avoid the cost and
reliability traps that bite first-time builders of AI apps.

## Where This Fits in an ML/AI Workflow

A typical workflow has three stages: training, packaging, and serving. Training
produces a model object. Packaging turns it into something callable behind a
stable interface (a function, an HTTP endpoint, or a serialized artifact).
Serving exposes that interface to users or other systems. Shiny sits in the
serving stage, but it is a specific kind of serving: a human-facing, stateful,
interactive front end rather than a stateless machine-to-machine API.

That distinction matters for AI applications. A REST API built with `plumber`
answers one request at a time with no memory between calls. A chat application
needs the opposite: it must remember the conversation so far, react to each new
message, and update the display incrementally as a response arrives. Shiny's
reactive model is built for exactly this kind of incremental, state-dependent
update, which is why it has become a common choice for LLM demos, internal
tools, and data-science dashboards that embed a model.

Table \@ref(tab:ai-apps-shiny-serving-options) places Shiny among the serving
options an R user is likely to reach for.

| Tool | Interface | State | Best for | Limitation |
|------|-----------|-------|----------|------------|
| `shiny` | Interactive web UI | Per-session, reactive | Dashboards, chat apps, human-in-the-loop tools | One R process per session can limit scale |
| `plumber` | REST/HTTP API | Stateless | Model-as-a-service, microservices | No built-in UI |
| `vetiver` | REST API + versioning | Stateless | MLOps, model registry and monitoring | Not for interactive UIs |
| RMarkdown/Quarto | Static or parameterized report | None | Reproducible reporting | Not interactive at runtime |
| Saved `.rds`/`qs` | In-process function | In-memory | Embedding in another R program | Not exposed to non-R users |

Table: (\#tab:ai-apps-shiny-serving-options) Serving options available to an R user, comparing their interface, state model, ideal use case, and main limitation.

A common production pattern combines these: `plumber` or `vetiver` serves the
model behind a stable API, and Shiny calls that API as a client. This keeps the
heavy inference logic separate from the UI, lets the two scale independently, and
means the same model endpoint can serve a Shiny app, a scheduled job, and an
external consumer at once.

> **When to use this:** Choose Shiny when a human sits in the loop and the
> session needs memory, a chat assistant, a labeling tool, a what-if dashboard.
> Choose `plumber` or `vetiver` when the consumer is another program and each
> call is independent. Choose Quarto when the deliverable is a fixed report. The
> three are complementary, not competitors.

## Shiny Architecture

A Shiny app has two halves: a user interface (UI) object that describes the
layout and inputs, and a server function that contains the logic. A call to
`shinyApp(ui, server)` wires them together and starts the web server.

```{r shiny-skeleton, eval=FALSE}
library(shiny)

ui <- fluidPage(
  titlePanel("Minimal Shiny app"),
  textInput("name", "Your name", value = "world"),
  textOutput("greeting")
)

server <- function(input, output, session) {
  output$greeting <- renderText({
    paste0("Hello, ", input$name, "!")
  })
}

shinyApp(ui, server)
```

The UI is just an R object that Shiny renders to HTML.^[You never write HTML by
hand. Functions like `fluidPage` and `textInput` return nested R structures that
Shiny serializes into the page markup, so the layout is plain R you can build with
loops, conditionals, and helper functions like any other code.] Input widgets such as
`textInput("name", ...)` create entries in the `input` list keyed by their id, so
`input$name` holds whatever the user typed. Output placeholders such as
`textOutput("greeting")` are filled by matching entries in the `output` list,
which the server assigns with render functions. The `session` argument represents
one connected browser tab and is where per-user state lives.

### Reactivity

The piece that makes Shiny different from an ordinary script is reactivity.
You do not write code that runs top to bottom once. You declare relationships
between values, and Shiny re-runs the minimum amount of code needed when inputs
change.

> **Intuition:** Think of a spreadsheet. When you change a cell, every formula
> that refers to it updates automatically, and nothing else recalculates. Shiny
> works the same way: inputs are the cells you type into, outputs are the
> formula cells, and the framework tracks which depends on which so it only
> recomputes what is affected.

There are three kinds of reactive objects:

- Reactive sources are inputs. Reading `input$name` registers a dependency.
- Reactive conductors, created with `reactive(...)`, are cached intermediate
  values that depend on sources and feed other reactives. They recompute only
  when something they depend on changes.
- Reactive endpoints, created with `render*` functions or `observe(...)`,
  produce side effects such as updating the display.

Formally, the dependencies form a directed acyclic graph $G = (V, E)$ where each
node $v \in V$ is a reactive object and an edge $(u, v) \in E$ means $v$ read $u$
during its last evaluation. When a source $s$ changes, Shiny marks every node
reachable from $s$ as invalid and schedules the invalid endpoints for
re-execution. A conductor with value cache is recomputed only if it is read while
invalid, which gives lazy evaluation: work that nobody observes is never
done. If a node's recomputation produces the same value, propagation can stop
early, so an input change that does not actually alter a downstream value avoids
needless redraws.

The practical payoff: if a render block reads `input$a` and `input$b`, it re-runs
when either changes, and you never write the wiring by hand. The cost is that you
have to think in terms of dependencies rather than control flow, which is the
main conceptual hurdle for newcomers.

> **Note:** The dependency edges are discovered at runtime, not declared in
> advance. Shiny watches which reactive values a block actually reads while it
> executes, so a branch that is not taken creates no dependency. This is what
> makes the graph adapt as inputs change, but it also means a value you read
> only inside an `if` will not trigger a re-run until that branch runs.

### Reactive Expressions Versus Observers

The two workhorses of the server function look similar but behave oppositely, and
mixing them up is a frequent source of bugs. A `reactive()` returns a value and is
lazy: it runs only when read, and caches its result. An `observe()` (or
`observeEvent()`) returns nothing, runs for its side effect, and is eager: it runs
whenever its dependencies invalidate. Use a reactive to compute model inputs or
predictions you will display in several places; use an observer to write to a log,
call an API for its effect, or update a stored state value.

> **Warning:** Do not put a side effect, an API call, a file write, a database
> update, inside a `reactive()`. Because reactives are lazy and cached, the side
> effect would fire at unpredictable times and possibly not at all. Values belong
> in `reactive()`; actions belong in `observe()`.

```{r shiny-reactive, eval=FALSE}
server <- function(input, output, session) {
  # Conductor: cached, lazy, recomputed when input$x changes.
  scaled <- reactive({
    req(input$x)            # stop quietly until input$x exists
    (input$x - mean_x) / sd_x
  })

  # Endpoint: re-renders when scaled() changes.
  output$pred <- renderText({
    predict(model, newdata = data.frame(x = scaled()))
  })

  # Observer: side effect only, runs eagerly on button click.
  observeEvent(input$save, {
    saveRDS(scaled(), "last_input.rds")
  })
}
```

## A Runnable Reactive Demo in Base R

To make the dependency idea concrete without Shiny, we build a tiny reactive
system in base R. It has reactive values that notify dependents when they
change, and observers that re-run when a value they read becomes invalid.
This is the same source-and-endpoint pattern Shiny uses, stripped to its
essentials so it runs in a plain R session.

```{r reactive-sim}
# A minimal reactive system: values that notify, observers that react.

make_reactive_value <- function(initial) {
  state <- new.env(parent = emptyenv())
  state$value <- initial
  state$observers <- list()   # functions to call when value changes

  get <- function() state$value

  set <- function(new_value) {
    if (!identical(new_value, state$value)) {
      state$value <- new_value
      # Invalidate: re-run each dependent observer.
      for (obs in state$observers) obs()
    }
    invisible(new_value)
  }

  subscribe <- function(obs) {
    state$observers <- c(state$observers, obs)
    obs()                     # eager first run, like Shiny observers
    invisible(NULL)
  }

  list(get = get, set = set, subscribe = subscribe)
}

# Build a small dependency graph:
#   temperature_c  ->  observer that prints Fahrenheit and logs a count.
temperature_c <- make_reactive_value(20)

recompute_count <- 0L
temperature_c$subscribe(function() {
  recompute_count <<- recompute_count + 1L
  f <- temperature_c$get() * 9 / 5 + 32
  cat(sprintf("[recompute %d] %.1f C = %.1f F\n",
              recompute_count, temperature_c$get(), f))
})

# Changing the source triggers the observer automatically.
temperature_c$set(25)
temperature_c$set(100)

# Setting to the same value does NOT recompute (value-equality short-circuit).
temperature_c$set(100)

cat("Total recomputations:", recompute_count, "\n")
```

The observer ran once at subscription time, once for each genuine change, and was
skipped when the new value equaled the old one. That last behavior is the
base-R analogue of Shiny's early stopping: propagation halts when a value does
not actually change. The whole point is that you declare the relationship once in
`subscribe()` and never call the conversion by hand again, which is what reactive
programming buys you.

We can visualize how recomputations accumulate as a sequence of source updates
arrives, including the duplicate that is skipped. Figure
\@ref(fig:ai-apps-shiny-reactive-recompute) plots the running count of observer
runs against the stream of writes, showing that only genuine changes advance the
count.

```{r ai-apps-shiny-reactive-recompute, fig.cap="Cumulative observer recomputations across a stream of source updates. The flat step marks a duplicate write that the value-equality check skips, so no recomputation occurs.", fig.width=6, fig.height=3.6}
library(ggplot2)

# Replay a stream of writes and record whether each caused a recompute.
temp <- make_reactive_value(20)
events <- c(20, 21, 21, 23, 23, 23, 30, 18)  # some duplicates
runs <- integer(0)
temp$subscribe(function() runs[[length(runs) + 1L]] <<- temp$get())

cumulative <- integer(length(events))
for (i in seq_along(events)) {
  before <- length(runs)
  temp$set(events[i])
  cumulative[i] <- length(runs)
}

df <- data.frame(
  update = seq_along(events),
  written = events,
  recomputes = cumulative,
  changed = c(TRUE, events[-1] != events[-length(events)])
)

ggplot(df, aes(update, recomputes)) +
  geom_step(direction = "hv", color = "steelblue", linewidth = 1) +
  geom_point(aes(color = changed), size = 3) +
  scale_x_continuous(breaks = df$update) +
  scale_color_manual(values = c(`TRUE` = "steelblue", `FALSE` = "grey60"),
                     labels = c(`TRUE` = "value changed",
                                `FALSE` = "duplicate (skipped)"),
                     name = NULL) +
  labs(x = "Source update #", y = "Cumulative recomputations",
       title = "Reactive recomputation only on real change") +
  theme_minimal(base_size = 12)
```

## Integrating an LLM Backend

To turn a Shiny app into a chat application, the server needs to call a language
model. The model almost never lives in the R process. It is reached over HTTP,
either a hosted API (such as the Anthropic or OpenAI endpoints) or a model you
run yourself behind an HTTP server. The R side is a thin client: build a request,
send it, parse the response.

> **Intuition:** The LLM is just another remote service, like a weather API or a
> payment gateway. Strip away the hype and the R code is the same shape you would
> write for any web service: assemble some JSON, send it, read the JSON that comes
> back. Everything model-specific lives on the server you are calling.

The cleanest way to do this in current R is the `ellmer` package, which wraps the
major chat APIs behind one interface and handles message formatting,
authentication, and streaming. Below is the idiomatic shape. It is `eval=FALSE`
because `ellmer` is not installed in this build, but it is correct code a reader
can run after `install.packages("ellmer")` and setting an API key.

```{r ellmer-basic, eval=FALSE}
library(ellmer)

# The API key is read from an environment variable, never hard-coded.
# Set ANTHROPIC_API_KEY in .Renviron or the deployment secret store.
chat <- chat_anthropic(
  model = "claude-sonnet-4-5",
  system_prompt = "You are a concise data-science assistant."
)

# A single turn. chat$chat() sends the message and returns the reply text.
reply <- chat$chat("Explain reactivity in one sentence.")
cat(reply)
```

If you prefer to call an endpoint directly, the request is plain JSON over
`httr2`. The structure is the same for most providers: a list of messages, each
with a `role` (`"user"`, `"assistant"`, or `"system"`) and `content`, plus model
and sampling parameters.

```{r httr2-call, eval=FALSE}
library(httr2)

call_llm <- function(messages, model = "claude-sonnet-4-5",
                     max_tokens = 1024) {
  request("https://api.anthropic.com/v1/messages") |>
    req_headers(
      "x-api-key" = Sys.getenv("ANTHROPIC_API_KEY"),
      "anthropic-version" = "2023-06-01",
      "content-type" = "application/json"
    ) |>
    req_body_json(list(
      model = model,
      max_tokens = max_tokens,
      messages = messages
    )) |>
    req_perform() |>
    resp_body_json()
}

# messages is a list of role/content pairs:
msgs <- list(list(role = "user", content = "Hello"))
# resp <- call_llm(msgs)
# resp$content[[1]]$text
```

### Conversation State

A chat is not a single call. Each request must carry the whole conversation so
far, because the API itself is stateless: it has no memory between calls, so the
application supplies the history every time. Let the conversation be an ordered
list of messages $m_1, m_2, \dots, m_n$, where each $m_i = (r_i, c_i)$ pairs a
role $r_i$ with content $c_i$. To produce turn $n+1$ the app sends the full
prefix $(m_1, \dots, m_n)$ and appends the model's reply as $m_{n+1}$.

This has a direct cost consequence. If turn $t$ adds $\ell_t$ tokens, then the
number of input tokens billed across $T$ turns grows quadratically:

$$
\text{total input tokens} = \sum_{t=1}^{T} \sum_{i=1}^{t-1} \ell_i
= \sum_{i=1}^{T-1} (T - i)\,\ell_i,
$$

because every earlier message is resent on every later turn. A long session
resends its early messages many times.

> **Warning:** That quadratic growth is easy to miss because each individual call
> looks cheap. A fifty-turn conversation can bill the first message fifty times.
> On a public app this is exactly how a small demo runs up a surprising invoice.

Two standard mitigations are truncation (drop or summarize old turns once the
running token count exceeds a budget $B$) and prompt caching (have the
provider cache a stable prefix so repeated tokens are billed at a reduced
rate).^[Prompt caching only helps when the prefix is byte-for-byte identical
across calls, so put stable content, the system prompt and any fixed context,
first, and append the changing turns after it.] In Shiny, the conversation lives
in a `reactiveVal` so that appending a message invalidates the chat display and
triggers a re-render.

```{r shiny-state, eval=FALSE}
server <- function(input, output, session) {
  # Conversation history as a reactive value (a list of role/content pairs).
  history <- reactiveVal(list())

  append_msg <- function(role, content) {
    history(c(history(), list(list(role = role, content = content))))
  }

  observeEvent(input$send, {
    req(input$user_msg)
    append_msg("user", input$user_msg)
    updateTextInput(session, "user_msg", value = "")  # clear the box

    reply <- call_llm(history())          # send full history
    append_msg("assistant", reply$content[[1]]$text)
  })

  output$chat <- renderUI({
    lapply(history(), function(m) {
      div(class = paste0("msg ", m$role), strong(m$role), p(m$content))
    })
  })
}
```

### Streaming Responses

Waiting for a full reply before showing anything feels slow. LLM APIs support
streaming, where the response arrives as a sequence of small chunks (server-sent
events), and the UI appends each chunk as it lands. The user sees text appear
token by token instead of after a multi-second pause.

In Shiny, the clean approach is `ExtendedTask` (introduced in Shiny 1.8.1)
combined with a streaming-capable client. `ExtendedTask` runs the call without
blocking the rest of the app, and a `reactiveVal` holding the partial text is
updated as chunks arrive, which re-renders the bubble each time.

```{r shiny-stream, eval=FALSE}
library(shiny)
library(ellmer)
library(promises)

server <- function(input, output, session) {
  history <- reactiveVal(list())
  streaming_text <- reactiveVal("")

  chat <- chat_anthropic(model = "claude-sonnet-4-5")

  # ExtendedTask keeps the UI responsive while the model replies.
  reply_task <- ExtendedTask$new(function(prompt) {
    # stream_async yields chunks; we accumulate and push to a reactiveVal.
    generator <- chat$stream_async(prompt)
    promises::promise(function(resolve, reject) {
      acc <- ""
      coro::async(function() {
        for (chunk in coro::await_each(generator)) {
          acc <<- paste0(acc, chunk)
          streaming_text(acc)            # partial update -> re-render
        }
        resolve(acc)
      })()
    })
  })

  observeEvent(input$send, {
    req(input$user_msg)
    history(c(history(), list(list(role = "user", content = input$user_msg))))
    streaming_text("")
    reply_task$invoke(input$user_msg)
  })

  # When the task finishes, commit the full reply to history.
  observeEvent(reply_task$result(), {
    history(c(history(),
              list(list(role = "assistant", content = reply_task$result()))))
    streaming_text("")
  })

  output$live <- renderText(streaming_text())
  output$chat <- renderUI(lapply(history(), function(m) {
    div(class = m$role, p(m$content))
  }))
}
```

The key structural point is non-blocking execution. A naive `chat$chat()` call
inside an observer blocks the whole R process for that session, freezing every
other control until the model finishes. `ExtendedTask` plus `promises` moves the
wait off the reactive thread so the interface stays live, the stop button still
works, and partial text keeps flowing.

> **Tip:** If you only have one thing to add to a plain chat app, make it
> non-blocking. Streaming text is a nice touch, but a frozen interface during a
> multi-second model call is the difference users notice first.

### Chat UI

You can hand-build the chat layout with `div`s and CSS, but the `shinychat`
package provides a ready-made chat component (`chat_ui()` and `chat_append()`)
that handles message bubbles, the input box, auto-scroll, and streaming display.
It pairs directly with `ellmer`.

```{r shinychat, eval=FALSE}
library(shiny)
library(shinychat)
library(ellmer)

ui <- bslib::page_fluid(
  chat_ui("chat", placeholder = "Ask a data-science question...")
)

server <- function(input, output, session) {
  chat <- chat_anthropic(model = "claude-sonnet-4-5")

  observeEvent(input$chat_user_input, {
    stream <- chat$stream_async(input$chat_user_input)
    chat_append("chat", stream)   # appends streamed chunks to the UI
  })
}

shinyApp(ui, server)
```

## Deployment

Once the app runs locally, it has to go somewhere others can reach. Table
\@ref(tab:ai-apps-shiny-deployment-targets) lists the main options, roughly in
order of effort.

| Target | What it is | Scaling model | Notes |
|--------|------------|---------------|-------|
| `shinyapps.io` | Hosted service by Posit | Managed instances | Fastest path; secrets via dashboard |
| Posit Connect | Self-hosted/enterprise server | Managed, auth-aware | Adds access control, scheduling |
| Shiny Server (open source) | Self-hosted single server | Manual | Free; you manage the box |
| Docker + cloud | Container you build | Orchestrated (Kubernetes, ECS) | Most control; most work |
| `shinylive` | App compiled to WebAssembly | Runs in the browser | No R server; cannot hold API secrets |

Table: (\#tab:ai-apps-shiny-deployment-targets) Deployment targets for a Shiny application, ordered roughly by effort, with their scaling model and notable trade-offs.

For an LLM app, two deployment facts dominate. First, **never ship the API key in
the app bundle**. Read it from an environment variable or the platform's secret
store, and rotate it if it leaks. The `shinylive` option, which runs the app
entirely in the browser, cannot keep a secret at all, so an LLM key would be
exposed to every visitor; route those calls through a small server-side proxy
instead.^[Anything that reaches the browser is readable by the user: view-source,
the network tab, and the WebAssembly bundle all expose it. The rule "the client
cannot keep a secret" is not specific to Shiny; it applies to any browser-side
code.] Second, concurrency: each Shiny session ties up an R process, and an
LLM call can take seconds. Size the number of worker processes for the expected
number of simultaneous users, and prefer the non-blocking `ExtendedTask` pattern
so one slow call does not stall a worker that is serving several sessions.

> **Warning:** The single most common security mistake in AI apps is committing
> an API key to source control. Once a key is in git history it is compromised
> even after you delete the line, treat it as leaked and rotate it. Keep keys in
> `.Renviron` (git-ignored) locally and in the platform secret store in
> production.

A container definition makes the runtime reproducible and is the usual unit of
deployment on cloud platforms.

```{r dockerfile, eval=FALSE}
# Dockerfile (not R code; shown for completeness)
# FROM rocker/shiny:4.4.3
# RUN R -e "install.packages(c('shiny','ellmer','shinychat','bslib'))"
# COPY app.R /srv/shiny-server/app/
# EXPOSE 3838
# # API key injected at runtime, never baked into the image:
# #   docker run -e ANTHROPIC_API_KEY=... -p 3838:3838 my-shiny-llm
```

## Practical Guidance and Pitfalls

When to use Shiny. Reach for it when you need a human in the loop with a
stateful, interactive interface: a chat assistant, a dashboard that wraps a model,
a labeling or review tool, a what-if explorer. If the consumer is another program,
a stateless `plumber`/`vetiver` API is simpler and scales better. If the output
is a fixed report, Quarto is lighter.

Pitfalls specific to LLM apps:

- *Blocking the session.* A synchronous API call freezes the app for that user.
  Use `ExtendedTask` and `promises` for any call that can take more than a moment.
- *Unbounded history.* Conversation tokens grow quadratically (see the cost
  formula above). Cap or summarize history, and consider provider-side prompt
  caching for stable prefixes.
- *Leaking secrets.* Keep keys in environment variables or a secret manager, out
  of source control and out of any browser-side bundle.
- *No error handling.* Network calls fail, rate limits hit, the model returns
  malformed JSON. Wrap calls in `tryCatch`, show the user a clear message, and do
  not let one failed turn crash the session.
- *Reactivity misuse.* Putting a side effect (an API call, a file write) inside a
  `reactive()` makes it run at unpredictable times and possibly more than once.
  Side effects belong in `observe`/`observeEvent`; values belong in `reactive`.
- *Cost surprises.* Every turn is billed, retries included. Log token usage and
  set a per-session budget if the app is public.

General Shiny hygiene: use `req()` to short-circuit until inputs are ready,
prefer `reactive()` for shared computations so they are not recomputed per output,
and isolate values you want to read without creating a dependency with
`isolate()`.

To recap the arc of this chapter: a model becomes a product only when something
wraps it in an interface, Shiny is the R way to build a stateful, interactive one,
and its reactive engine, the same source-and-observer pattern we built by hand in
base R, is what makes incremental updates like streaming chat natural. The LLM
itself lives behind an HTTP call; the work on the R side is managing conversation
state, keeping calls non-blocking, guarding secrets, and watching cost. Get those
four right and the rest is layout.

## Further Reading

- Chang, W. and the Shiny authors. *shiny: Web Application Framework for R.*
  Package documentation and articles at shiny.posit.co.
- Wickham, H. (2021). *Mastering Shiny.* O'Reilly. The standard reference on
  reactivity, modules, and application structure.
- Sievert, C. (2020). *Interactive Web-Based Data Visualization with R, plotly,
  and shiny.* Chapman and Hall/CRC.
- Fay, C., Rochette, S., Guyader, V., and Girard, C. (2021). *Engineering
  Production-Grade Shiny Apps.* Chapman and Hall/CRC. Deployment, scaling, and
  testing.
- Posit. *ellmer* and *shinychat* package documentation, for current LLM client
  and chat-UI patterns in R.
- Vaswani, A. et al. (2017). "Attention Is All You Need." For the model
  architecture behind the LLM backends these apps call.
