# Bayesian Additive Regression Trees

-   also known as BART [@sparapani2021]

-   is related to [Bagging] and [Random Forests], and [Boosting] because

    -   Each tree is constructed in a random manner (similar to [Bagging] and [Random Forests])

    -   Each tree tries to capture signal not yet accounted for by the current model (similar to [Boosting])

Let $K$ be the number of regression trees, $B$ the number of (MCMC) iterations for which the BART algorithm will be run

Let $\hat{f}_k^b(x)$ represent the prediction at $x$ for the k-th regression tree used in the b-th iteration.

Then, the $K$ trees from that iteration will be summed as

$$
\hat{f}^b(x) = \sum_{k=1}^K \hat{f}_k^b(x), b = 1, \dots, B
$$

In the first iteration, all trees are initialized to have a single root node with $\hat{f}_k^1(x) = (1/nK)\sum_{i=1}^n y_i$, which is the mean of the response values divided by the total number of trees.

Thus,

$$
\hat{f}^1(x) = \sum_{k=1}^K \hat{f}_k^1 (x) = \frac{1}{n} \sum_{i=1}^n y_i \text{ (mean response)}
$$

In each later iteration, BART updates each of the $K$ trees, one at a time

In the b-th iteration, to update the k-th tree, we subtract from each response value the predictions from all but the k-th tree to obtain **a partial residual**:

$$
r_i = y_i - \sum_{k'<k} \hat{f}_{k'}^b (x_i) - \sum_{k'>k} f_{k'}^{b-1}(x_i), i = 1, \dots, n
$$

Instead of fitting a new tree to this partial residual, BART randomly chooses a perturbation to the tree in the previous iteration ($\hat{f}_k^{b-1}$) from a set of possible perturbations (this is where MCMC comes in), favoring ones that improve the fit to the partial residual. There are 2 components to this perturbation:

-   change the structure of the tree by adding or pruning branches

-   change the prediction in each terminal node of the tree

The output of BART is a collection of prediction models:

$$
\hat{f}^b(x) = \sum_{k =1}^K \hat{f}_k^b (x) , b = 1, \dots, B
$$

where we don't include the results from the iterations in the MCMC burn-in period (say $L$ burn-in samples):

$$
\hat{f}^b(x) = \frac{1}{B - L} \sum_{b = L +1}^B \hat{f}_k^b (x) 
$$

Since this method is under the Bayesian paradigm, we can use the posterior samples to get distributional estimates of the predictions

<br>

BART algorithm [@james2013] Alg 8.3

1.  Let $\hat{f}_1^1(x) = f^1_2(x) = \dots = \hat{f}_K^1(x) = \frac{1}{nK} \sum_{i=1}^n y_i$

2.  Compute $\hat{f}^1 (x) = \sum_{k=1}^K \hat{f}_k^1 (x) = \frac{1}{n}\sum_{i=1}^n y_i$

3.  For $b = 2, \dots, B$

    1.  For $k = 1, \dots, K$

        1.  For $i = 1, \dots , n$, compute the current partial residual ($r_i$

        2.  Fit a new tree $\hat{f}_k^b(x)$ to $r_i$ by randomly perturbing the k-th tree from the previous iteration, $\hat{f}_k^{b-1}(x)$ (favor perturbations that improve the fit)

    2.  Compute $\hat{f}^b (x) = \sum_{k=1}^K \hat{f}^b_k (x)$

4.  Compute the mean after $L$ burn-in samples $\hat{f}(x) = \frac{1}{B - L} \sum_{b = L +1}^B \hat{f}^b (x)$

<br>

Notes

-   Step 3.1.2 prevents overfitting since it limits how "hard we fit the data in each iteration

-   The individual trees are quite small. we limit the tree size to avoid overfitting the data (especially when we grew very large trees)

-   Each time we randomly perturb a tree to fit the residuals, we are drawing a new tree form a posterior distribution

-   When we use BART, we have to specify

    -   the number trees $K$

    -   the number of interactions $B$

    -   the number of burn-in interactions $L$

-   We typically choose large values for $B = 1000$ and $K =200$, and moderate value for $L=100$

-   BART is a good out-of-box tool (i.e., minimal tuning)

Example by [@james2013, Ch 8.3]

```{r, message=FALSE, warning=FALSE}
library(BART)
library(ISLR2)
library(tree)

set.seed(1)
attach(Boston)

train <- sample(1:nrow(Boston), nrow(Boston)/2)
# tree.boston <- tree(medv ~ ., Boston, subset = train)

x <- Boston[,1:12]
y <- Boston[,"medv"]
xtrain <- x[train,]
ytrain <- y[train]
xtest <- x[-train,]
ytest <- y[-train]

bartfit <- gbart(xtrain, ytrain, x.test = xtest)

# test error
yhat.bart <- bartfit$yhat.test.mean
mean((ytest - yhat.bart)^2)

# how many times each variable appeared in the colelction of trees
ord <- order(bartfit$varcount.mean, decreasing = T)
bartfit$varcount.mean[ord]
```

Example by the package's authors @sparapani2021

y = mdev = the median value of owner-occupied homes

x1 = rm = the average number of rooms

x2 = lstat = the percent of the population that is lower status

```{r}
library(MASS)
x <- Boston[,c(6,13)]
y <- Boston$medv
# head(cbind(x,y))
par(mfrow = c(2,2))
plot(x[, 1], y, xlab = "x1=rm", ylab = "y=mdev")
plot(x[, 2], y, xlab = "x2=lstat", ylab = "y=mdev")
plot(x[, 1], x[, 2], xlab = "x1=rm", ylab = "x2=lstat")
par(mfrow = c(1, 1))


```

```{r}
library("BART")
set.seed(99)
nd <- 200 # number of posterior draws
burn <- 50
# use wbart for continuous varaibles 
post <- wbart(x, y, nskip = burn, ndpost = nd)
```

Convergence assessment

```{r}
plot(post$sigma, type = "l")
abline(v = burn, lwd = 2, col = "red")
# good convergence 
```

Prediction comparison with lm

```{r}
lmf <- lm(y~., data.frame(x, y))
fitmat <- cbind(y, post$yhat.train.mean, lmf$fitted.values)
colnames(fitmat) <- c("y", "BART", "Linear")
cor(fitmat)
pairs(fitmat)
```

Prediction

```{r}
i <- order(post$yhat.train.mean)
boxplot(post$yhat.train[, i])

n <- length(y)
set.seed(14)
i <- sample(1:n, floor(0.75 * n))
x.train <- x[i, ]; y.train = y[i]
x.test <- x[-i, ]; y.test = y[-i]
cat("training sample size = ", length(y.train), "\n")
cat("testing sample size = ", length(y.test), "\n")

set.seed(99)
post1 <- wbart(x.train, y.train, x.test)
length(post1$yhat.test.mean)
```

Thinning (to speed things up)

```{r}
set.seed(4)
post3 <-
    wbart(
        x.train,
        y.train,
        nskip = 1000,
        ndpost = 10000,
        nkeeptrain = 0,
        nkeeptest = 0,
        nkeeptestmean = 0,
        nkeeptreedraws = 200
    )
yhatthin <- predict(post3, x.test)

fmat <- cbind(y.test, post1$yhat.test.mean, apply(yhatthin, 2, mean))
colnames(fmat) <- c("y", "yhat", "yhatThin")
pairs(fmat)

```

Friedman's partial dependence function to see variable importance

```{r}
x.train <- as.matrix(Boston[i,-14])
set.seed(12)
post4 <- wbart(x.train, y.train)
H <- length(y.train)
L <- 41
x <- seq(min(x.train[, 13]), max(x.train[, 13]), length.out = L)
x.test <- cbind(x.train[,-13], x[1])
for (j in 2:L)
    x.test <- rbind(x.test, cbind(x.train[,-13], x[j]))
pred <- predict(post4, x.test)
partial <- matrix(nrow = 1000, ncol = L)
for (j in 1:L) {
    h <- (j - 1) * H + 1:H
    partial[, j] <- apply(pred[, h], 1, mean)
}
plot(
    x,
    apply(partial, 2, mean),
    type = "l",
    ylim = c(10, 50),
    xlab = "lstat",
    ylab = "mdev"
)

lines(x, apply(partial, 2, quantile, probs = 0.025), lty = 2)
lines(x, apply(partial, 2, quantile, probs = 0.975), lty = 2)
```

For examples on categorical variables, see [@sparapani2021]
