# Model Tuning

```{r}
library("finetune")
```

[@Kuhn_2014] for detail

-   Grid Search
-   Annealing Search

# Model Stacking

Codes are from [stack vignettes](https://cran.r-project.org/web/packages/stacks/vignettes/basics.html) basic and

[Source Code](https://github.com/tidymodels/stacks)

Steps in doing model staking:

1.  Define your candidate (ensemble) models
2.  Initialize a `data_stack` object with `stacks()`
3.  Add candidate candidate models using `add_candidates()`
4.  Evaluate the combined predictions `blend_predictions()`
5.  'fit_members()\` allows you to fit models with non-zero staking coefficients
6.  `predict()` new data

```{r}
library(tidymodels)
library(stacks)
library(dplyr)
library(purrr)
library(tune)
```

```{r}
data("tree_frogs")
# subset the data
tree_frogs <- tree_frogs %>%
    filter(!is.na(latency)) %>%
    select(-c(clutch, hatched))
library(ggplot2)

ggplot(tree_frogs) +
    aes(x = age, y = latency, color = treatment) +
    geom_point() +
    labs(x = "Embryo Age (s)", y = "Time to Hatch (s)", col = "Treatment")
ggplot(tree_frogs) +
    aes(x = treatment, y = age, color = reflex) +
    geom_jitter() +
    labs(y = "Embryo Age (s)",
         x = "treatment",
         color = "Response")
```

```{r}
# some setup: resampling and a basic recipe
set.seed(1)

tree_frogs_split <- initial_split(tree_frogs)
tree_frogs_train <- training(tree_frogs_split)
tree_frogs_test  <- testing(tree_frogs_split)

folds <- rsample::vfold_cv(tree_frogs_train, v = 5)

tree_frogs_rec <-
    recipe(reflex ~ ., data = tree_frogs_train) %>%
    step_dummy(all_nominal(),-reflex) %>%
    step_zv(all_predictors())

tree_frogs_wflow <-
    workflow() %>%
    add_recipe(tree_frogs_rec)
ctrl_grid <- control_stack_grid() # same control settings
ctrl_res <- control_stack_resamples()
```

Add random forest model

```{r}
rand_forest_spec <-
    rand_forest(mtry = tune(),
                min_n = tune(),
                trees = 500) %>%
    set_mode("classification") %>%
    set_engine("ranger")

rand_forest_wflow <-
    tree_frogs_wflow %>%
    add_model(rand_forest_spec)

rand_forest_res <-
    tune_grid(
        object = rand_forest_wflow,
        resamples = folds,
        grid = 10,
        control = ctrl_grid
    )
```

Add neural network model

```{r}
nnet_spec <-
    mlp(hidden_units = tune(),
        penalty = tune(),
        epochs = tune()) %>%
    set_mode("classification") %>%
    set_engine("nnet")

nnet_rec <-
    tree_frogs_rec %>%
    step_normalize(all_predictors())

nnet_wflow <-
    tree_frogs_wflow %>%
    add_model(nnet_spec)

nnet_res <-
    tune_grid(
        object = nnet_wflow,
        resamples = folds,
        grid = 10,
        control = ctrl_grid
    )
```

Build stacks

```{r}
tree_frogs_model_st <-
    # initialize the stack
    stacks() %>%
    # add candidate members
    add_candidates(rand_forest_res) %>%
    add_candidates(nnet_res) %>%
    # determine how to combine their predictions
    blend_predictions() %>%
    # fit the candidates with nonzero stacking coefficients
    fit_members()

tree_frogs_model_st
```

Check performance

```{r}
theme_set(theme_bw())
autoplot(tree_frogs_model_st)
autoplot(tree_frogs_model_st, type = "members")
autoplot(tree_frogs_model_st, type = "weights") # to see penalty
```

See parameters

```{r}
collect_parameters(tree_frogs_model_st, "rand_forest_res")
```

Predict new data

```{r}
tree_frogs_pred <-
    tree_frogs_test %>%
    bind_cols(predict(tree_frogs_model_st, .,type = "prob"))
```

ROC AUC

```{r}
yardstick::roc_auc(tree_frogs_pred,
                   truth = reflex,
                   contains(".pred_"))
```

Compare predictions with actual observations

```{r}
# ggplot(tree_frogs_pred) +
#     aes(x = latency,
#         y = .pred) +
#     geom_point() +
#     coord_obs_pred()
```

Comparison between stacks and member model

```{r}
member_preds <-
    tree_frogs_test %>%
    select(latency) %>%
    bind_cols(predict(tree_frogs_model_st, tree_frogs_test, members = TRUE))
```

Compare different model predictions

```{r}
map_dfr(setNames(colnames(tree_frogs_pred), colnames(tree_frogs_pred)),
        ~ mean(tree_frogs_pred$reflex == pull(tree_frogs_pred, .x))) %>%
    pivot_longer(c(everything(),-reflex))
```
